---
title: "PCWD Anomalies and GINI calculation"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load relevant libraries:
```{r libraries, message=FALSE, echo=FALSE}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(ggplot2)
library(cwd)
library(ncdf4)
library(reshape2)
library(ggpubr)
library(maps)
library(raster)
library(rnaturalearth)
library(sf)
library(rgdal)
library(sp)
library(RColorBrewer)
library(rJava)
library(loadeR.java)
library(transformeR)
library(loadeR)
library(visualizeR)
library(geoprocessoR)
library(tidyverse)
library(abind)
library(patchwork)
library(gridExtra)
library(DescTools) #for GINI calculation
library(ggsci)
```

# Anomaly analysis of PCWD regional time series

- calculate timeseries of rolling anomalies wrt a reference: 1420-1450 as reference for epoch 1; 1851-1880 for epoch 2
- anomalies: value(t) - mean(1420-1450)


Load regional aggregated data:
```{r}
#read in aggregated regional data
# #Load with
regional_results_1420 <- readRDS("~/cwd_global/data/regionalResults_1420_1.RData")
regional_results_1850 <- readRDS("~/cwd_global/data/regionalResults_1850_1.RData")

### Remove m001_tidy for areas CAU, EAU, NAU, NZ, SAU, RFE, EAS, SEA
# List of regions where the first ensemble member should be removed
regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")

# Loop through each specified region and remove the first column (m001_tidy) - bias over Oceania
for (region in regions_to_modify) {
  regional_results_1420[[region]] <- regional_results_1420[[region]][, -1]
}


```

```{r}
# Define continent assignments
continent_mapping <- list(
  "Europe" = c("NEU", "WCE", "EEU", "MED"),
  "North America" = c("NWN", "NEN", "WNA", "CNA", "ENA", "GIC"),
  "Africa" = c("SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG"),
  "Asia" = c("WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "WSB", "ESB", "RFE", "RAR"),
  "Central America" =c("NCA", "SCA", "CAR"),
  "South America" = c("NWS", "NSA", "NES", "SAM", "SWS", "SES", "SSA"),
  "Australia" = c("NAU", "CAU", "EAU", "SAU", "NZ")
)
```

## Volcanic forcing impact on PCWD annual anomalies

Load volcanic forcing data for ModE-Sim set 1:
```{r}
#read in volcanic forcing data for set 1: 
#AOD years 1000 C.E. to 1900 C.E.:
input_file <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/ModESim_forcings/1420_1/eva_holo2.2_forcing_echam_T63_ir_1000-1900.nc"
nc_forcings <- nc_open(input_file)
aod = ncvar_get(nc_forcings, varid="aod")[6,,] # index 6 for 500nm wavelength
lat = ncvar_get(nc_forcings, varid="lat")
time_1420 = ncvar_get(nc_forcings, varid="time")
time_units <- ncatt_get(nc_forcings, "time", "units")  # Check the time units (e.g., "days since 0001-01-01")
nc_close(nc_forcings)

########### Calculate annual mean values for first epoch years:
# Define the years of interest
years_of_interest <- 1420:1849

# Find indices of columns corresponding to the desired years
selected_indices <- which(time_1420 %in% years_of_interest)

# Subset the matrix to include only those years
aod_selected <- aod[, selected_indices]
time_selected <- time_1420[selected_indices]  # Corresponding time values

### compute annual mean over latitudes:
# Compute the mean over latitudes (first dimension)
mean_latitude <- apply(aod_selected, 2, mean, na.rm = TRUE)
# Create a data frame from the two lists
df <- data.frame(
  AOD = mean_latitude,
  year = time_selected
)

# Calculate the annual mean AOD value (ignoring NA values)
annual_mean_AOD <- df %>%
  group_by(year) %>%
  summarise(mean_AOD = mean(AOD, na.rm = TRUE))

# Create the corresponding months (repeating each year 12 times)
months <- rep(1:12, length(time_selected) / 12)  # Assuming 12 months for each year

# Create the "yyyy-mm" format using the year and month
dates <- paste(time_selected, sprintf("%02d", months), sep = "-")  # sprintf adds leading zero for months < 10

# Now dates is in "yyyy-mm" format
head(dates)  # Check the first few entrie

# Convert to a data frame for easier manipulation
aod_1420_df <- data.frame(dates = dates, mean_value = mean_latitude)

# Compute the monthly !!! mean (since each year has 12 monthly values)
library(dplyr)
aod_annual_mean_1420 <- aod_1420_df %>%
  group_by(dates) %>%
  summarise(annual_mean_value = mean(mean_value, na.rm = TRUE))

#write.csv(aod_annual_mean_1420, "~/cwd_global/data/AOD_annual_mean_1420epoch.csv")


####read in data: 
##### volc forcing for 1420 - global annual mean: 
aod_annual_mean_1420 <- read.csv("~/cwd_global/data/AOD_annual_mean_1420epoch.csv")
##### volc forcing for 1850 - global annual mean: 
aod_annual_mean_1850 <- read.csv("~/cwd_global/data/AOD_annual_mean_1850epoch.csv")
aod_annual_mean_1850 <- aod_annual_mean_1850 %>% 
  rename(
    year = Year,
    annual_mean_value = mean_AOD
    )

```


```{r}
################################converting 1420 -1849 AOD into monthly format
# Convert year to Date if necessary (since it's currently in 'yyyy-mm' format)
# Convert the 'year' column to Date using lubridate's ym() function
aod_1420_df$dates <- lubridate::ym(aod_1420_df$dates)

# Check the result
head(aod_1420_df$dates)


# Line plot
ggplot(aod_1420_df, aes(x = dates, y = mean_value)) +
  geom_line(color = "blue", size = 1) +  # Line plot
  labs(title = "Line Plot of Mean Value Over Time", 
       x = "Year", 
       y = "Mean Value") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels if needed

##################Select all years where aod > 0.01
```


Calculate PCWD anomalies for each time step: 

```{r}
### calculate anomalies for both epochs based on reference time 1420-1450

### calculate baseline for anomalies: 1420-1450 mean for each region
# Initialize list to store baselines for each region
anom_baseline <- list()

# Loop over regions
for (region_name in names(regional_results_1420)) {
  
  # Extract all ensemble members for the region
  ensemble_members <- regional_results_1420[[region_name]]
  
  # Compute the mean over all values in the first 30 years, pooling everything
  anom_baseline[[region_name]] <- mean(ensemble_members[1:30, ], na.rm = TRUE)
}

# Initialize lists to store results
anoms_1420 <- list()      # Mean anomalies for each region
anoms_sd_1420 <- list()   # Standard deviation of anomalies for each region
ens_anoms_1420 <- list()

# Loop over regions
for (region_name in names(regional_results_1420)) {
  
  # Extract the ensemble members for the region
  ensemble_members <- regional_results_1420[[region_name]]  # Now using full ensemble data
  
  # Compute anomalies for each ensemble member
  ensemble_anoms <- sweep(ensemble_members, 2, anom_baseline[[region_name]], "-")  # Subtract baseline
  
  ens_anoms_1420[[region_name]] <- ensemble_anoms
  # Compute mean anomaly across ensemble members
  anoms_1420[[region_name]] <- rowMeans(ensemble_anoms, na.rm = TRUE)
  
  # Compute standard deviation (uncertainty) across ensemble members
  anoms_sd_1420[[region_name]] <- apply(ensemble_anoms, 1, sd, na.rm = TRUE)
}


### calculate baseline for anomalies: 1420-1450 mean for each region
# Initialize list to store baselines for each region
anom_baseline_1850 <- list()

# Loop over regions
for (region_name in names(regional_results_1850)) {
  
  # Extract all ensemble members for the region
  ensemble_members <- regional_results_1850[[region_name]]
  
  # Compute baseline (mean over first 30 years) for each ensemble member
  anom_baseline_1850[[region_name]] <- mean(ensemble_members[1:30,], na.rm = TRUE)
}

# Initialize lists to store results
anoms_1850 <- list()      # Mean anomalies for each region
anoms_sd_1850 <- list()   # Standard deviation of anomalies for each region

# Loop over regions
for (region_name in names(regional_results_1850)) {
  
  # Extract the ensemble members for the region
  ensemble_members <- regional_results_1850[[region_name]]  # Now using full ensemble data
  
  # Compute anomalies for each ensemble member
  ensemble_anoms <- sweep(ensemble_members, 2, anom_baseline_1850[[region_name]], "-")  # Subtract baseline
  
  # Compute mean anomaly across ensemble members
  anoms_1850[[region_name]] <- rowMeans(ensemble_anoms, na.rm = TRUE)
  
  # Compute standard deviation (uncertainty) across ensemble members
  anoms_sd_1850[[region_name]] <- apply(ensemble_anoms, 1, sd, na.rm = TRUE)
}

```

For simplicity: linear detrending of the data to remove warming trends etc:

```{r}
# Function to linearly detrend a time series
detrend_data <- function(anomaly_series) {
  time <- seq_along(anomaly_series)  # Create time index
  model <- lm(anomaly_series ~ time, na.action = na.exclude)  # Fit linear model
  trend <- predict(model)  # Extract trend
  detrended_series <- anomaly_series - trend  # Subtract trend
  return(detrended_series)
}

# Apply detrending to all regions
anoms_detrended_1420 <- lapply(anoms_1420, detrend_data)

# Apply detrending to all regions
anoms_detrended_1850 <- lapply(anoms_1850, detrend_data)

```


Create Dataframe region: 
```{r}
########### 1420 period
# Convert SNR list into a data frame
anom_df_1420 <- do.call(rbind, lapply(names(anoms_detrended_1420), function(region) {
  data.frame(
    time = seq(1420,1849),  # Extract time from names
    anomalies = anoms_detrended_1420[[region]],
    #std = anoms_sd_1420[[region]],
    region = region                                # Add region identifier
  )
}))

# Add a continent column to the SNR data frame
anom_df_1420$continent <- NA  # Initialize continent column

for (continent in names(continent_mapping)) {
  anom_df_1420$continent[anom_df_1420$region %in% continent_mapping[[continent]]] <- continent
}


###### 1850 period
# Convert SNR list into a data frame
anom_df_1850 <- do.call(rbind, lapply(names(anoms_detrended_1850), function(region) {
  data.frame(
    time = seq(1850,2009),  # Extract time from names
    anomalies = anoms_1850[[region]], 
    #std = anoms_sd_1850[[region]],
    region = region                                # Add region identifier
  )
}))

# Add a continent column to the SNR data frame
anom_df_1850$continent <- NA  # Initialize continent column

for (continent in names(continent_mapping)) {
  anom_df_1850$continent[anom_df_1850$region %in% continent_mapping[[continent]]] <- continent
}

```

Plot timeseries of anomalies for all regions: 

```{r}
#################all continents for 1420 epoch ###################################

#### Europe ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Europe"
selected_region <- "WCE"
anom_filtered <- anom_df_1420 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(annual_mean_AOD$mean_AOD, na.rm = TRUE)
aod_max <- max(annual_mean_AOD$mean_AOD, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
#scale_factor <- 3500

# Scale AOD values to match the anomalies scale
annual_mean <- annual_mean_AOD %>%
  mutate(scaled_aod = (mean_AOD - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.8, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t. 1420-1450"
    #,    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Global Mean AOD [μm]"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "#C355A7"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 14),  # Size for facet labels
    axis.title.y.right = element_text(color = "#C355A7", size = 13),
    axis.text.y.right  = element_text(color = "#C355A7", size = 12),
    strip.background = element_rect(fill = "#F5B0CB")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales

#### Asia ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Asia"
selected_region <- "EAS"
anom_filtered <- anom_df_1420 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(annual_mean_AOD$mean_AOD, na.rm = TRUE)
aod_max <- max(annual_mean_AOD$mean_AOD, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
#scale_factor <- 3500

# Scale AOD values to match the anomalies scale
annual_mean <- annual_mean_AOD %>%
  mutate(scaled_aod = (mean_AOD - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.8, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t. 1420-1450"
    #,    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Global Mean AOD [μm]"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "#C355A7"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 14),  # Size for facet labels
    axis.title.y.right = element_text(color = "#C355A7", size = 13),
    axis.text.y.right  = element_text(color = "#C355A7", size = 12),
    strip.background = element_rect(fill = "#E5973B")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales

#### Australia ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Australia"
selected_region <- "CAU"
anom_filtered <- anom_df_1420 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(annual_mean_AOD$mean_AOD, na.rm = TRUE)
aod_max <- max(annual_mean_AOD$mean_AOD, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
#scale_factor <- 3500

# Scale AOD values to match the anomalies scale
annual_mean <- annual_mean_AOD %>%
  mutate(scaled_aod = (mean_AOD - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.8, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t. 1420-1450"
    #,    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Global Mean AOD [μm]"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "#C355A7"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 14, color = "white"),  # Size for facet labels
    axis.title.y.right = element_text(color = "#C355A7", size = 13),
    axis.text.y.right  = element_text(color = "#C355A7", size = 12),
    strip.background = element_rect(fill = "#3C3980")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales

##### Africa ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Africa"
selected_region <- "WSAF"
anom_filtered <- anom_df_1420 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(annual_mean_AOD$mean_AOD, na.rm = TRUE)
aod_max <- max(annual_mean_AOD$mean_AOD, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
#scale_factor <- 3500

# Scale AOD values to match the anomalies scale
annual_mean <- annual_mean_AOD %>%
  mutate(scaled_aod = (mean_AOD - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.8, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t. 1420-1450"
    #,    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Global Mean AOD [μm]"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "#C355A7"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 14),  # Size for facet labels
    axis.title.y.right = element_text(color = "#C355A7", size = 13),
    axis.text.y.right  = element_text(color = "#C355A7", size = 12),
    strip.background = element_rect(fill = "#91C7B1")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales


#### North America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "North America"
selected_region <- "WNA"
anom_filtered <- anom_df_1420 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(annual_mean_AOD$mean_AOD, na.rm = TRUE)
aod_max <- max(annual_mean_AOD$mean_AOD, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
#scale_factor <- 3500

# Scale AOD values to match the anomalies scale
annual_mean <- annual_mean_AOD %>%
  mutate(scaled_aod = (mean_AOD - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.8, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t. 1420-1450"
    #,    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Global Mean AOD [μm]"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "#C355A7"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 14, color= "white"),  # Size for facet labels
    axis.title.y.right = element_text(color = "#C355A7", size = 13),
    axis.text.y.right  = element_text(color = "#C355A7", size = 12),
    strip.background = element_rect(fill = "#236A3F")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales

#### Central America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Central America"
selected_region <- "SCA"
anom_filtered <- anom_df_1420 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(annual_mean_AOD$mean_AOD, na.rm = TRUE)
aod_max <- max(annual_mean_AOD$mean_AOD, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
#scale_factor <- 3500

# Scale AOD values to match the anomalies scale
annual_mean <- annual_mean_AOD %>%
  mutate(scaled_aod = (mean_AOD - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.8, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t. 1420-1450"
    #,    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Global Mean AOD [μm]"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "#C355A7"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 14, color = "white"),  # Size for facet labels
    axis.title.y.right = element_text(color = "#C355A7", size = 13),
    axis.text.y.right  = element_text(color = "#C355A7", size = 12),
    strip.background = element_rect(fill = "#B33951")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales


#### South America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "South America"
selected_region <- "SES"
anom_filtered <- anom_df_1420 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(annual_mean_AOD$mean_AOD, na.rm = TRUE)
aod_max <- max(annual_mean_AOD$mean_AOD, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
#scale_factor <- 3500

# Scale AOD values to match the anomalies scale
annual_mean <- annual_mean_AOD %>%
  mutate(scaled_aod = (mean_AOD - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey40", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.8, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t. 1420-1450"
    #,    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Global Mean AOD [μm]"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "#C355A7"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 14),  # Size for facet labels
    axis.title.y.right = element_text(color = "#C355A7", size = 13),
    axis.text.y.right  = element_text(color = "#C355A7", size = 12),
    strip.background = element_rect(fill = "#E3D081")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales

```


```{r}
#################all continents for 1850 epoch ###################################
#### Europe ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Europe"
anom_filtered <- anom_df_1850 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
aod_max <- max(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
scale_factor <- 30000

# Scale AOD values to match the anomalies scale
annual_mean <- aod_annual_mean_1850 %>%
  mutate(scaled_aod = (annual_mean_value - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey2", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.5, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD Anomalies (w.r.t. 1850-2009)",
    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Mean AOD"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "red"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 10),  # Size for facet labels
    strip.background = element_rect(fill = "#F5B0CB")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales

#### Asia ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Asia"
anom_filtered <- anom_df_1850 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
aod_max <- max(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
scale_factor <- 30000

# Scale AOD values to match the anomalies scale
annual_mean <- aod_annual_mean_1850 %>%
  mutate(scaled_aod = (annual_mean_value - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey2", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.5, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD Anomalies (w.r.t. 1850-2009)",
    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Mean AOD"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "red"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 10, color = "white"),  # Size for facet labels
    strip.background = element_rect(fill = "#3C3980")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales


##### Africa ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Africa"
anom_filtered <- anom_df_1850 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
aod_max <- max(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
scale_factor <- 18000

# Scale AOD values to match the anomalies scale
annual_mean <- aod_annual_mean_1850 %>%
  mutate(scaled_aod = (annual_mean_value - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey2", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.5, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD Anomalies (w.r.t. 1850-2009)",
    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Mean AOD"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "red"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 10),  # Size for facet labels
    strip.background = element_rect(fill = "#91C7B1")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales


#### North America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "North America"
anom_filtered <- anom_df_1850 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
aod_max <- max(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
scale_factor <- 10000

# Scale AOD values to match the anomalies scale
annual_mean <- aod_annual_mean_1850 %>%
  mutate(scaled_aod = (annual_mean_value - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey2", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.5, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD Anomalies (w.r.t. 1850-2009)",
    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Mean AOD"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "red"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 10),  # Size for facet labels
    strip.background = element_rect(fill = "#236A3F")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales

#### Central America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Central America"
anom_filtered <- anom_df_1850 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
aod_max <- max(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
scale_factor <- 40000

# Scale AOD values to match the anomalies scale
annual_mean <- aod_annual_mean_1850 %>%
  mutate(scaled_aod = (annual_mean_value - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
  #geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey2", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.5, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD Anomalies (w.r.t. 1850-2009)",
    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Mean AOD"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "red"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 10),  # Size for facet labels
    strip.background = element_rect(fill = "#B33951")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales


#### South America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "South America"
anom_filtered <- anom_df_1850 %>% filter(continent == selected_continent)

# Normalize AOD data to fit the anomalies scale
aod_min <- min(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
aod_max <- max(aod_annual_mean_1850$annual_mean_value, na.rm = TRUE)
anom_min <- min(anom_filtered$anomalies, na.rm = TRUE)
anom_max <- max(anom_filtered$anomalies, na.rm = TRUE)

# Calculate scale factor to ensure zero alignment
aod_zero <- 0  # AOD zero point
anom_zero <- 0  # Anomalies zero point
scale_factor <- (anom_max - anom_min) / (aod_max - aod_min)  # Scale based on range
scale_factor <- 30000

# Scale AOD values to match the anomalies scale
annual_mean <- aod_annual_mean_1850 %>%
  mutate(scaled_aod = (annual_mean_value - aod_zero) * scale_factor + anom_zero)

# Plot anom time series
ggplot(anom_filtered, aes(x = time, y = anomalies, color = "PCWD anomalies")) +
  geom_line(show.legend = TRUE) +  # Plot lines for each region
 # geom_ribbon(aes(ymin = anomalies - std, ymax = anomalies + std, fill = "SD across EMs"), alpha = 0.3,color = NA, show.legend = TRUE) +
  geom_hline(yintercept = 0, color = "grey2", linetype = "dashed")+
  # Secondary Y-axis: Volcanic Forcing (AOD)
  geom_line(data = annual_mean, aes(x = year, y = scaled_aod, color = "Volcanic Forcing (AOD)"), linewidth = 0.5, linetype = "dotted", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD Anomalies [mm] (w.r.t. 1850-2009)",
    title = paste("Detrended annual PCWD anomalies for", selected_continent)
  ) +
  # Secondary axis transformation ensuring zero alignment
  scale_y_continuous(
    sec.axis = sec_axis(
      ~ (. - anom_zero) / scale_factor + aod_zero, 
      name = "Annual Mean AOD"
    )
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("PCWD anomalies" = "black", "Volcanic Forcing (AOD)" = "red"),
    labels = c("PCWD anomalies (Mean)", "Volcanic Forcing (AOD)")
  ) +
  scale_fill_manual(
    values = c("SD across EMs" = "honeydew4"), 
    labels = c("SD across EMs")
  ) +
    # Adjust legend appearance and keep the ribbon and line separate
  # Customize the legend to avoid the honeydew4 background around lines (using override.aes)
  guides(
    color = guide_legend(title = "Legend", 
                         order = 1, 
                         override.aes = list(fill = NA, shape = NA)),  # Remove background for color lines
    fill = guide_legend(title = "", order = 2)  # Fill scale: SD across Ensemble Members
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),  # Center the title
    strip.text = element_text(size = 10),  # Size for facet labels
    strip.background = element_rect(fill = "#E3D081")
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales

```

### Spatial PCWD anomalies after volcanic eruptions:

Maps showing annual anomalies
- showing Europe after 1783 eruption (Laki)
- showing North America 
- showing Central America

Read in lat, lon and time info: 
```{r}
#lat lon and time info from netcdf files
##read in data
input_file_1850 <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850/PCWD_ANNMAX.nc"
input_file_1420 <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1420/PCWD_ANNMAX.nc"

nc_pwcd_1850 <- nc_open(input_file_1850)
pcwd_annmax_1850 = ncvar_get(nc_pwcd_1850, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_1850, varid="lon")
lat = ncvar_get(nc_pwcd_1850, varid="lat")
time_1850 = ncvar_get(nc_pwcd_1850, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates_1850 <- reference_date + time_1850

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1850)

#1420 file for 1420 time
nc_pwcd_1420 <- nc_open(input_file_1420)
pcwd_annmax_1420 = ncvar_get(nc_pwcd_1420, varid="pcwd_annmax")
time_1420 = ncvar_get(nc_pwcd_1420, varid="time")
# Convert to actual dates (days since 2001-01-01)
time_dates_1420 <- reference_date + time_1420

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1420)
```

aggregate files: 
```{r}
#read in netcdf data and aggregate

#### 1420 set:
# Define the base path - have to seperate sets once there are more!
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)  #only want m001-m020 for now

# Extract ensemble member identifiers from folder names
ensemble_members <- basename(folders)

# Identify the target files
target_files <- file.path(folders, "04_result_1420/PCWD_ANNMAX.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included
target_files <- target_files[-1] #removes m001 due to spatial bias

# Filter ensemble members to match target files
ensemble_members <- ensemble_members[file.exists(target_files)]
ensemble_members <- ensemble_members[-1]

# Initialize a list to store data for each ensemble member
data_by_ensemble_1420 <- list()

# Extract dimensions from the first file to initialize the array size for consistency
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "pcwd_annmax"))[1]
dim_y <- dim(ncvar_get(nc, "pcwd_annmax"))[2]
dim_z <- dim(ncvar_get(nc, "pcwd_annmax"))[3]
nc_close(nc)

# Loop through all the target files and store the data in a list
for (i in seq_along(target_files)) {
  file <- target_files[i]
  member <- ensemble_members[i]
  
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "pcwd_annmax")  # Adjust "pcwd_annmax" to your variable name
  
  # Store the data in the list, keyed by the ensemble member
  data_by_ensemble_1420[[member]] <- data
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the data list
print(names(data_by_ensemble_1420))  # Check stored ensemble members
print(dim(data_by_ensemble_1420[[1]]))  # Check dimensions of the data for the first ensemble member


#### 1850 set: 
# Define the base path - have to think of way to seperate sets once there is data!!
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)

# Extract ensemble member identifiers from folder names
ensemble_members <- basename(folders)

# Identify the target files
target_files <- file.path(folders, "04_result_1850/PCWD_ANNMAX.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included

# Filter ensemble members to match target files
ensemble_members <- ensemble_members[file.exists(target_files)]

# Initialize a list to store data for each ensemble member
data_by_ensemble_1850 <- list()

# Extract dimensions from the first file to initialize the array size for consistency
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "pcwd_annmax"))[1]
dim_y <- dim(ncvar_get(nc, "pcwd_annmax"))[2]
dim_z <- dim(ncvar_get(nc, "pcwd_annmax"))[3]
nc_close(nc)

# Loop through all the target files and store the data in a list
for (i in seq_along(target_files)) {
  file <- target_files[i]
  member <- ensemble_members[i]
  
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "pcwd_annmax")  # Adjust "pcwd_annmax" to your variable name
  
  # Store the data in the list, keyed by the ensemble member
  data_by_ensemble_1850[[member]] <- data
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the data list
print(names(data_by_ensemble_1850))  # Check stored ensemble members
print(dim(data_by_ensemble_1850[[1]]))  # Check dimensions of the data for the first ensemble member

```

Compute anomalies for each time step:
```{r}
## Anomaly global grid cell values for 1420 epoch: 
# Convert list of 20 ensemble members into a single 4D array: [192, 96, 431, 19] - without m001
ensemble_array <- simplify2array(data_by_ensemble_1420)  # Shape: [192, 96, 431, 19]

# Compute the ensemble mean across the 4th dimension (ensemble members)
ensemble_mean_1420 <- apply(ensemble_array, c(1, 2, 3), mean, na.rm = TRUE)  # Shape: [192, 96, 431]

# `ensemble_mean_1420` now contains the ensemble-averaged values for each grid cell over time
# Define dimensions
num_years <- 431  # Total years in the dataset
window_size <- 30  # Reference period (first 30 years)

# Create an array of the same shape [192, 96, 431] to store means and standard deviations
mean_reference_1420 <- array(NA, dim = c(192, 96))  # Mean for first 30 years
anom_array_1420 <- array(NA, dim = c(192, 96, num_years))  # Store anomalies

# Compute reference mean & standard deviation over first 30 years (years 1:30)
for (i in 1:192) {
  for (j in 1:96) {
    reference_values <- ensemble_mean_1420[i, j, 1:window_size]  # First 30 years
    mean_reference_1420[i, j] <- mean(reference_values, na.rm = TRUE)
  }
}

# Compute anomalies for each time step
for (t in 1:num_years) {  # Loop through all years
  for (i in 1:192) {
    for (j in 1:96) {
        anom_array_1420[i, j, t] <- (ensemble_mean_1420[i, j, t] - mean_reference_1420[i, j])
    }
  }
}

# anom_array_1420 now contains anomalies for each grid cell at each time step


## Anomaly global grid cell values for 1850 epoch: 
# Convert list of 20 ensemble members into a single 4D array: [192, 96, 160, 20] 
ensemble_array <- simplify2array(data_by_ensemble_1850)  # Shape: [192, 96, 160, 20]

# Compute the ensemble mean across the 4th dimension (ensemble members)
ensemble_mean_1850 <- apply(ensemble_array, c(1, 2, 3), mean, na.rm = TRUE)  # Shape: [192, 96, 160]

# `ensemble_mean_1850` now contains the ensemble-averaged values for each grid cell over time
# Define dimensions
num_years <- 160  # Total years in the dataset
window_size <- 30  # Reference period (first 30 years)

# Create an array of the same shape [192, 96, 431] to store means and standard deviations
mean_reference_1850 <- array(NA, dim = c(192, 96))  # Mean for first 30 years
anom_array_1850 <- array(NA, dim = c(192, 96, num_years))  # Store anomalies

# Compute reference mean & standard deviation over first 30 years (years 1:30)
for (i in 1:192) {
  for (j in 1:96) {
    reference_values <- ensemble_mean_1850[i, j, 1:window_size]  # First 30 years
    mean_reference_1850[i, j] <- mean(reference_values, na.rm = TRUE)
  }
}

# Compute anomalies for each time step
for (t in 1:num_years) {  # Loop through all years
  for (i in 1:192) {
    for (j in 1:96) {
        anom_array_1850[i, j, t] <- (ensemble_mean_1850[i, j, t] - mean_reference_1850[i, j])
    }
  }
}

# anom_array_1850 now contains anomalies for each grid cell at each time step

```

plot maps: 
```{r}
# Load land data for filling continents
land <- ne_countries(scale = "medium", returnclass = "sf")

```

Region specifications: 
Continent	Longitude (xlim)	Latitude (ylim)
Europe	c(-15, 50)	c(30, 75)
North America	c(-170, -10)	c(25, 90)
Africa	c(-25, 60)	c(-35, 40)
Asia	c(30, 180)	c(5, 80)
Central America	c(-120, -60)	c(5, 35)
South America	c(-85, -30)	c(-60, 15)
Australia	c(110, 180)	c(-50, 0)

```{r}
#global anomaly map for 1783: 
# Find the index for the year specified
year_index <- 1783 - 1420 + 1  #-1420 or 1850 depending on dataset
plot_data <- anom_array_1420[,,year_index] #specify dataset, 1420 or 1850

# Convert the first year's data into a raster object
r <- raster(t(plot_data), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Load land polygons for masking
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Ensure that the layer values are numeric (continuous)
r_df$layer <- as.numeric(r_df$layer)

ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = layer), na.rm = TRUE) +
  scale_fill_distiller(name = "PCWD [mm]", 
                       palette = "RdBu",  # Red-Blue diverging palette
                       values = scales::rescale(c(-200, -10, 10, 200)),  # Rescales to emphasize 0-2 range
                       limits = c(-500, 500),  # Adjust scale to focus on values between -1 and 1
                       na.value = "grey50") +  # Color for missing values
  labs(
    #title = "PCWD Anomalies for 1783 w.r.t 1420-1450", 
       x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "none", legend.title = element_text(size=11), legend.text = element_text(size=11), 
        axis.text = element_text(size=14), axis.title = element_text(size=14))+
  #coord_sf(xlim =	c(-85, -30), ylim =	c(-60, 15), expand = FALSE) +  # Adjusted for Europe
   guides(fill = guide_legend(title = "PCWD anomalies [mm]", reverse = TRUE)) 


```
## Rolling 30 year anomaly timeseries for both epoch

### Check method for 30 year anomalies + anomalies for all regions

Calculate rolling 30-year means: 
```{r}
# Initialize list to store rolling 30-year means
rolling_means <- list()

# Define window size
window_size <- 30

# Compute rolling means for each ensemble member
for (region_name in names(regional_results_1420)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_results_1420[[region_name]]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means[[region_name]] <- rolling_mean_matrix
}

####################### 1850 epoch anomalies
# Initialize list to store rolling 30-year means
rolling_means_1850 <- list()

# Define window size
window_size <- 30

# Compute rolling means for each ensemble member
for (region_name in names(regional_results_1850)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_results_1850[[region_name]]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means_1850[[region_name]] <- rolling_mean_matrix
}
```


#### Calculate anomalies of 30yr running mean and plot

```{r}
# Initialize lists to store baselines
anom_baseline <- list()

# Compute baseline mean from the first 30 values of rolling mean
for (region_name in names(rolling_means_1850)) {
  rolling_mean_matrix <- rolling_means_1850[[region_name]]
  
  anom_baseline[[region_name]] <- mean(rolling_mean_matrix[30:1,], na.rm = TRUE)  ### for reference period of first 30 values: rolling_mean_matrix[1:30,]
}

# Initialize lists to store anomalies
ensemble_30anom_1850 <- list()

# Compute anomalies
for (region_name in names(rolling_means_1850)) {
  rolling_mean_matrix <- rolling_means_1850[[region_name]]
  num_years <- nrow(rolling_mean_matrix)
  
  # Initialize matrix to store anomalies
  anoms <- matrix(NA, nrow=num_years, ncol=ncol(rolling_mean_matrix))
  
  # Compute anomalies using rolling mean baseline
  for (t in 1:num_years) {
    anoms[t, ] <- (rolling_mean_matrix[t, ] - anom_baseline[[region_name]])
  }
  
  # Store results
  ensemble_30anom_1850[[region_name]] <- anoms
}

# Compute mean anomalies across ensemble members
mean_30anom_1850 <- list()
for (region_name in names(ensemble_30anom_1850)) {
  mean_30anom_1850[[region_name]] <- rowMeans(ensemble_30anom_1850[[region_name]], na.rm = TRUE)
}

# Compute region-specific pooled standard deviation using only the first 30 years
pooled_sd_by_region <- list()
for (region_name in names(ensemble_30anom_1850)) {
  # Extract only the first 30 years of anomalies
  first_30_years <- ensemble_30anom_1850[[region_name]][1:30, ]
  
  # Compute pooled standard deviation over these values
  pooled_sd_by_region[[region_name]] <- sd(unlist(first_30_years), na.rm = TRUE)
}

# Convert ensemble data into a long format for plotting individual members
anoms_long_1850 <- do.call(rbind, lapply(names(ensemble_30anom_1850), function(region) {
  df <- as.data.frame(ensemble_30anom_1850[[region]])
  df$time <- seq(1865, 1995)  # Adjust time range accordingly
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "anomalies")
  return(df_long)
}))

# Add a continent column
anoms_long_1850$continent <- NA  # Initialize continent column
for (continent in names(continent_mapping)) {
  anoms_long_1850$continent[anoms_long_1850$region %in% continent_mapping[[continent]]] <- continent
}

```


Plot for each region, showing ensemble mean anomalies and individual ensemble members:
```{r}
### Plot for each region
# Filter for selected continent
selected_continent <- "Europe"
selected_region <- "NEU"
anoms_filtered <- anoms_long_1850 %>% filter(region == selected_region)

#plot 2sigma from reference period as horizontal lines around reference mean 

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1850-1880"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14),
    strip.background = element_rect(fill = "#F5B0CB"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  
  facet_wrap(~region, scales = "free_y")

#### Asia ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Asia"
selected_region <- "EAS"
anoms_filtered <- anoms_long_1850 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )
ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1850-1880"
    #,  title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14),  # Size for facet labels
    strip.background = element_rect(fill = "#E5973B"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region,  scales = "free_y")  # Facet by region with independent y-axis scales

##### Australia ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Australia"
selected_region <- "EAU"
anoms_filtered <- anoms_long_1850 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1850-1880"
    #,  title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14, color = "white"),
    strip.background = element_rect(fill = "#3C3980"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region,  scales = "free_y")  # Facet by region with independent y-axis scales


##### Africa ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Africa"
selected_region <- "MDG"
anoms_filtered <- anoms_long_1850 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1850-1880"
    #,   title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14),  # Size for facet labels
    strip.background = element_rect(fill = "#91C7B1"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region,  scales = "free_y")  # Facet by region with independent y-axis scales


#### North America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "North America"
selected_region <- "ENA"
anoms_filtered <- anoms_long_1850 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1850-1880"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14, color= "white"),
    strip.background = element_rect(fill = "#236A3F"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region,  scales = "free_y")  # Facet by region with independent y-axis scales

#### Central America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Central America"
selected_region <- "SCA"
anoms_filtered <- anoms_long_1850 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1850-1880"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14, color = "white"),
    strip.background = element_rect(fill = "#B33951"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales


#### South America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "South America"
selected_region <- "NES"
anoms_filtered <- anoms_long_1850 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1850-1880"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14),
    strip.background = element_rect(fill = "#E3D081"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales


```

For 1420 period also: 
Calculate anomalies:
```{r}
# Initialize lists to store baselines
anom_baseline <- list()

# Compute baseline mean from the first 30 values of rolling mean
for (region_name in names(rolling_means)) {
  rolling_mean_matrix <- rolling_means[[region_name]]
  
  anom_baseline[[region_name]] <- mean(rolling_mean_matrix[1:30,], na.rm = TRUE)  ### for reference period of first 30 values: rolling_mean_matrix[1:30,]
}

# Initialize lists to store anomalies
ensemble_30anom_1420 <- list()

# Compute anomalies
for (region_name in names(rolling_means)) {
  rolling_mean_matrix <- rolling_means[[region_name]]
  num_years <- nrow(rolling_mean_matrix)
  
  # Initialize matrix to store anomalies
  anoms <- matrix(NA, nrow=num_years, ncol=ncol(rolling_mean_matrix))
  
  # Compute anomalies using rolling mean baseline
  for (t in 1:num_years) {
    anoms[t, ] <- (rolling_mean_matrix[t, ] - anom_baseline[[region_name]])
  }
  
  # Store results
  ensemble_30anom_1420[[region_name]] <- anoms
}

# Compute mean anomalies across ensemble members
mean_30anom_1420 <- list()
for (region_name in names(ensemble_30anom_1420)) {
  mean_30anom_1420[[region_name]] <- rowMeans(ensemble_30anom_1420[[region_name]], na.rm = TRUE)
}

# Compute region-specific pooled standard deviation using only the first 30 years
pooled_sd_by_region_1420 <- list()
for (region_name in names(ensemble_30anom_1420)) {
  # Extract only the first 30 years of anomalies
  first_30_years <- ensemble_30anom_1420[[region_name]][1:30, ]
  
  # Compute pooled standard deviation over these values
  pooled_sd_by_region_1420[[region_name]] <- sd(unlist(first_30_years), na.rm = TRUE)
}


# Convert ensemble data into a long format for plotting individual members
anoms_long_1420 <- do.call(rbind, lapply(names(ensemble_30anom_1420), function(region) {
  df <- as.data.frame(ensemble_30anom_1420[[region]])
  df$time <- seq(1435, 1835)  # Adjust time range accordingly
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "anomalies")
  return(df_long)
}))

# Add a continent column
anoms_long_1420$continent <- NA  # Initialize continent column
for (continent in names(continent_mapping)) {
  anoms_long_1420$continent[anoms_long_1420$region %in% continent_mapping[[continent]]] <- continent
}

```

Plot region wise:
```{r}
### plot 
### Plot for each region
# Filter for selected continent
selected_continent <- "Europe"
selected_region <- "WCE"
anoms_filtered <- anoms_long_1420 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1420-1450"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15),
    strip.background = element_rect(fill = "#F5B0CB"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  
  facet_wrap(~region, scales = "free_y")

#### Asia ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Asia"
selected_region <- "SAS"
anoms_filtered <- anoms_long_1420 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies (mm) w.r.t 1420-1450"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14),  # Size for facet labels
    strip.background = element_rect(fill = "#E5973B"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region,  scales = "free_y")  # Facet by region with independent y-axis scales

##### Australia ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Australia"
selected_region <- "EAU"
anoms_filtered <- anoms_long_1420 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1420-1450"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14, color = "white"),
    strip.background = element_rect(fill = "#3C3980"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  
  facet_wrap(~region, scales = "free_y")

##### Africa ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Africa"
selected_region <- "MDG"
anoms_filtered <- anoms_long_1420 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1420-1450"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14),  # Size for facet labels
    strip.background = element_rect(fill = "#91C7B1"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region,  scales = "free_y")  # Facet by region with independent y-axis scales


#### North America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "North America"
selected_region <- "ENA"
anoms_filtered <- anoms_long_1420 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1420-1450"
    #, title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14, color = "white"),
    strip.background = element_rect(fill = "#236A3F"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region,  scales = "free_y")  # Facet by region with independent y-axis scales

#### Central America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "Central America"
selected_region <- "CAR"
anoms_filtered <- anoms_long_1420 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1420-1450"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14, color = "white"),
    strip.background = element_rect(fill = "#B33951"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales


#### South America ####
# Filter data for the selected continent (e.g., Europe)
selected_continent <- "South America"
selected_region <- "SSA"
anoms_filtered <- anoms_long_1420 %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - 2 * pooled_sd,
    upper_bound = first_mean_anomaly + 2 * pooled_sd
  )

ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1420-1450"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold", size= 12),
    legend.text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
    strip.text = element_text(size = 14),
    strip.background = element_rect(fill = "#E3D081"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13)
  ) +
  facet_wrap(~region, scales = "free_y")  # Facet by region with independent y-axis scales


```


## Gini Index for inequality

### Gini Maps

- calculate Gini index for both each epoch for every gridcell
- Map variability/ measure of inequality or unequalness
- low number (close to 0 rather than 1) implies low inequality i.e. over time PCWD values stable for that gridcell
- Gini index is more sensitive to changes in the middle of income distribution and less sensitive to changes at the top and the bottom of income distribution (Atkinson, 1970)

Read in lat, lon and time info: 
```{r}
#lat lon and time info from netcdf files
##read in data
input_file_1850 <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850/PCWD_ANNMAX.nc"
input_file_1420 <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1420/PCWD_ANNMAX.nc"

nc_pwcd_1850 <- nc_open(input_file_1850)
pcwd_annmax_1850 = ncvar_get(nc_pwcd_1850, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_1850, varid="lon")
lat = ncvar_get(nc_pwcd_1850, varid="lat")
time_1850 = ncvar_get(nc_pwcd_1850, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates_1850 <- reference_date + time_1850

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1850)

#1420 file for 1420 time
nc_pwcd_1420 <- nc_open(input_file_1420)
pcwd_annmax_1420 = ncvar_get(nc_pwcd_1420, varid="pcwd_annmax")
time_1420 = ncvar_get(nc_pwcd_1420, varid="time")
# Convert to actual dates (days since 2001-01-01)
time_dates_1420 <- reference_date + time_1420

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1420)
```

aggregate files: 
```{r}
#read in netcdf data and aggregate

#### 1420 set:
# Define the base path - have to seperate sets once there are more!
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)  #only want m001-m020 for now

# Extract ensemble member identifiers from folder names
ensemble_members <- basename(folders)

# Identify the target files
target_files <- file.path(folders, "04_result_1420/PCWD_ANNMAX.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included

# Filter ensemble members to match target files
ensemble_members <- ensemble_members[file.exists(target_files)]

# Initialize a list to store data for each ensemble member
data_by_ensemble_1420 <- list()

# Extract dimensions from the first file to initialize the array size for consistency
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "pcwd_annmax"))[1]
dim_y <- dim(ncvar_get(nc, "pcwd_annmax"))[2]
dim_z <- dim(ncvar_get(nc, "pcwd_annmax"))[3]
nc_close(nc)

# Loop through all the target files and store the data in a list
for (i in seq_along(target_files)) {
  file <- target_files[i]
  member <- ensemble_members[i]
  
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "pcwd_annmax")  # Adjust "pcwd_annmax" to your variable name
  
  # Store the data in the list, keyed by the ensemble member
  data_by_ensemble_1420[[member]] <- data
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the data list
print(names(data_by_ensemble_1420))  # Check stored ensemble members
print(dim(data_by_ensemble_1420[[1]]))  # Check dimensions of the data for the first ensemble member


#### 1850 set: 
# Define the base path - have to think of way to seperate sets once there is data!!
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)

# Extract ensemble member identifiers from folder names
ensemble_members <- basename(folders)

# Identify the target files
target_files <- file.path(folders, "04_result_1850/PCWD_ANNMAX.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included

# Filter ensemble members to match target files
ensemble_members <- ensemble_members[file.exists(target_files)]

# Initialize a list to store data for each ensemble member
data_by_ensemble_1850 <- list()

# Extract dimensions from the first file to initialize the array size for consistency
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "pcwd_annmax"))[1]
dim_y <- dim(ncvar_get(nc, "pcwd_annmax"))[2]
dim_z <- dim(ncvar_get(nc, "pcwd_annmax"))[3]
nc_close(nc)

# Loop through all the target files and store the data in a list
for (i in seq_along(target_files)) {
  file <- target_files[i]
  member <- ensemble_members[i]
  
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "pcwd_annmax")  # Adjust "pcwd_annmax" to your variable name
  
  # Store the data in the list, keyed by the ensemble member
  data_by_ensemble_1850[[member]] <- data
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the data list
print(names(data_by_ensemble_1850))  # Check stored ensemble members
print(dim(data_by_ensemble_1850[[1]]))  # Check dimensions of the data for the first ensemble member

```
```{r}
##bind both into one for mean over time: 

# Create a new list to store the combined data
data_combined <- list()

# Loop over each ensemble member
for (i in names(data_by_ensemble_1420)) {
  # Remove last year from 1420 dataset (assumes 3rd dimension is time)
  data_1420_trimmed <- data_by_ensemble_1420[[i]][,,1:(dim(data_by_ensemble_1420[[i]])[3] - 1)]
  
  # Bind with 1850 data (all years)
  data_combined[[i]] <- abind::abind(
    data_1420_trimmed,
    data_by_ensemble_1850[[i]],
    along = 3
  )
}
```


```{r}
##### 1850 epoch
# Create an array of the same shape [192, 96, 160] to store means
mean_ensemble_1850 <- array(NA, dim = c(192, 96, 160))

# Compute the mean over all ensemble members
ensemble_array <- simplify2array(data_by_ensemble_1850)  # Convert list to array
mean_ensemble_1850 <- apply(ensemble_array, c(1,2,3), mean, na.rm = TRUE)

# Create an array to store Gini indices
gini_values_1850 <- matrix(NA, nrow = 192, ncol = 96)

# Loop through each grid cell and compute Gini index over time
for (i in 1:192) {
  for (j in 1:96) {
    gini_values_1850[i, j] <- Gini(mean_ensemble_1850[i, j, ], na.rm = TRUE)
  }
}

#### 1420 epoch
# Create an array of the same shape [192, 96, 160] to store means
mean_ensemble_1420 <- array(NA, dim = c(192, 96, 160))

# Compute the mean over all ensemble members
ensemble_array <- simplify2array(data_by_ensemble_1420)  # Convert list to array
mean_ensemble_1420 <- apply(ensemble_array, c(1,2,3), mean, na.rm = TRUE)

# Create an array to store Gini indices
gini_values_1420 <- matrix(NA, nrow = 192, ncol = 96)

# Loop through each grid cell and compute Gini index over time
for (i in 1:192) {
  for (j in 1:96) {
    gini_values_1420[i, j] <- Gini(mean_ensemble_1420[i, j, ], na.rm = TRUE)
  }
}

mean_ensemble_combined <- array(NA, dim = c(192, 96, 590))
# Compute the mean over all ensemble members
ensemble_array <- simplify2array(data_combined)  # Convert list to array
mean_ensemble_combined <- apply(ensemble_array, c(1,2,3), mean, na.rm = TRUE) #time mean

# Create an array to store Gini indices
gini_values_combined <- matrix(NA, nrow = 192, ncol = 96)

# Loop through each grid cell and compute Gini index over time
for (i in 1:192) {
  for (j in 1:96) {
    gini_values_combined[i, j] <- Gini(mean_ensemble_combined[i, j, ], na.rm = TRUE)
  }
}

```

plot epoch wise GINI index

```{r}
#plot map with 1850 GINI index
#Convert data into raster object

r <- raster(t(gini_values_1850), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84" # Set CRS for raster
r <- flip(r, direction='y') #Flip the raster to correct orientation
#Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs=st_crs(r)) #Ensure CRS alignment
#Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))
#Convert the masked raster to a dataframe for ggplot
r_df <- as.data.frame(r, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")

ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill = layer), na.rm = TRUE) +
  scale_fill_bs5("#E5973B", limits = c(0, 0.4)) +  
  labs(title = "GINI Index 1850 epoch (EM)", x= "Longitude", y="Latitude") +
  theme_classic() +
  geom_sf(data=land, fill=NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand=FALSE)+
  theme(legend.title = element_text(size=12), legend.text = element_text(size=10)) +
  guides(fill = guide_legend(title = "Gini index", reverse = TRUE))
```


```{r}
#plot map with 1420 GINI index
#Convert data into raster object

r <- raster(t(gini_values_1420), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84" # Set CRS for raster
r <- flip(r, direction='y') #Flip the raster to correct orientation
#Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs=st_crs(r)) #Ensure CRS alignment
#Mask the raster with land polygons to remove ocean values
#r_masked <- mask(r, as(land, "Spatial"))
#Convert the masked raster to a dataframe for ggplot
r_df <- as.data.frame(r, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")

ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill = layer), na.rm = TRUE) +
  scale_fill_bs5("#E5973B", limits = c(0, 0.4)) +  
  labs(title = "GINI Index 1420 epoch (EM)", x= "Longitude", y="Latitude") +
  theme_classic() +
  geom_sf(data=land, fill=NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand=FALSE)+
  theme(legend.title = element_text(size=12), legend.text = element_text(size=10)) +
  guides(fill = guide_legend(title = "Gini index", reverse = TRUE))
```

```{r}
library(ggsci)
#combined epochs: 

#plot map with 1420 GINI index
#Convert data into raster object

r <- raster(t(gini_values_combined), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84" # Set CRS for raster
r <- flip(r, direction='y') #Flip the raster to correct orientation
#Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs=st_crs(r)) #Ensure CRS alignment
#Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))
#Convert the masked raster to a dataframe for ggplot
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")

ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill = layer), na.rm = TRUE) +
  scale_fill_gradientn(
    colours = c("#EBD2CF", "#D49E98", "#BF6D63", "#7D3028", "#330B06"),
    limits = c(0.0, 0.4),
    name = "Gini Index"
  ) +
  labs(title = "Mean Gini over 600 years", x= "Longitude", y="Latitude") +
  theme_classic() +
  geom_sf(data=land, fill=NA, color = "black", lwd = 0.3) +
  #coord_sf(xlim = range(lon), ylim = range(lat), expand=FALSE)+
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.title = element_text(size=13), legend.text = element_text(size=13), 
        axis.text = element_text(size=13), axis.title = element_text(size=12)) +
  guides(fill = guide_legend(title = "Gini index", reverse = TRUE))

```

### Gini timeseries

Create and plot Gini timeseries with 30-year moving window for each region: 

```{r}
####################### Compute running GINI timeseries for each EM #########################
################## 1420 epoch #############
window_size <- 31  # Define window size

###################### Compute 30-year moving Gini coefficients
# Initialize lists to store Gini values
regional_gini_1420 <- list()

# Loop over regions
for (region_name in names(regional_results_1420)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_results_1420[[region_name]]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store Gini values (one per ensemble member per time step)
  gini_window <- matrix(NA, nrow = num_years - window_size + 1, ncol = ncol(ensemble_members))
  
  # Compute 30-year moving Gini values for each ensemble member
  for (t in 1:(num_years - window_size + 1)) {  # Prevent going out of bounds
    
    # Loop over ensemble members
    for (member in 1:ncol(ensemble_members)) {
      
      # Extract data for the current window
      window_data <- ensemble_members[t:(t + window_size - 1), member]
      
      # Compute Gini coefficient for this ensemble member
      gini_window[t, member] <- Gini(window_data, na.rm = TRUE)
    }
  }
  
  # Store results
  regional_gini_1420[[region_name]] <- gini_window
}

###################### Compute Mean Gini across Ensemble Members
# Initialize a list to store mean Gini values
mean_gini_1420 <- list()

# Loop over regions
for (region_name in names(regional_gini_1420)) {
  
  # Compute the mean Gini coefficient across ensemble members for each time step
  mean_gini_1420[[region_name]] <- rowMeans(regional_gini_1420[[region_name]], na.rm = TRUE)
}


################## 1850 epoch #############
window_size <- 31  # Define window size

###################### Compute 30-year moving Gini coefficients
# Initialize lists to store Gini values
regional_gini_1850 <- list()

# Loop over regions
for (region_name in names(regional_results_1850)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_results_1850[[region_name]]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store Gini values (one per ensemble member per time step)
  gini_window <- matrix(NA, nrow = num_years - window_size + 1, ncol = ncol(ensemble_members))
  
  # Compute 30-year moving Gini values for each ensemble member
  for (t in 1:(num_years - window_size + 1)) {  # Prevent going out of bounds
    
    # Loop over ensemble members
    for (member in 1:ncol(ensemble_members)) {
      
      # Extract data for the current window
      window_data <- ensemble_members[t:(t + window_size - 1), member]
      
      # Compute Gini coefficient for this ensemble member
      gini_window[t, member] <- Gini(window_data, na.rm = TRUE)
    }
  }
  
  # Store results
  regional_gini_1850[[region_name]] <- gini_window
}

###################### Compute Mean Gini across Ensemble Members
# Initialize a list to store mean Gini values
mean_gini_1850 <- list()

# Loop over regions
for (region_name in names(regional_gini_1850)) {
  
  # Compute the mean Gini coefficient across ensemble members for each time step
  mean_gini_1850[[region_name]] <- rowMeans(regional_gini_1850[[region_name]], na.rm = TRUE)
}

```

```{r}
##mean GINI 1420 timeseries with all regions in one plot
# Convert list to long-format data frame
df_gini <- bind_rows(
  lapply(names(mean_gini_1420), function(region) {
    tibble(
      year = 1435:1834,  # Assuming a running window of 400 years
      gini_value = mean_gini_1420[[region]],
      region = region
    )
  })
)

# Add continent column
df_gini$continent <- sapply(df_gini$region, function(r) {
  continent <- names(continent_mapping)[sapply(continent_mapping, function(regions) r %in% regions)]
  return(continent)
})

# Define continent colors
continent_colors <- c(
  "Europe" = "#F5B0CB",
  "North America" = "#236A3F",
  "Africa" = "#91C7B1",
  "Asia" = "#E5973B",
  "Central America" = "#B33951",
  "South America" = "#E3D081",
  "Australia" = "#3C3980",
  "Ocean" = "white"
)
# Convert continent to factor for proper ordering
df_gini$continent <- factor(df_gini$continent, levels = names(continent_colors))

# Compute global mean Gini index per year
df_global_mean <- df_gini %>%
  group_by(year) %>%
  summarise(global_mean_gini = mean(gini_value, na.rm = TRUE))

# Plot with both regional and global mean
ggplot(df_gini, aes(x = year, y = gini_value, color = continent, group = region)) +
  geom_line() +  # Regional lines
  geom_line(data = df_global_mean, aes(x = year, y = global_mean_gini), 
            color = "black", linewidth = 1.2) +  # Global mean in black
  scale_color_manual(values = continent_colors) +  # Apply continent colors
   scale_y_continuous(breaks = seq(0, 1, by = 0.05)) +  # y-axis ticks every 0.05
  theme_minimal() +
  labs(title = "30-Year Running Gini Index by IPCC Region",
       x = "Year",
       y = "Gini Index",
       color = "Region") +
  theme(legend.position = "none")

##############################################################
##mean GINI 1850 timeseries with all regions in one plot
# Convert list to long-format data frame
df_gini <- bind_rows(
  lapply(names(mean_gini_1850), function(region) {
    tibble(
      year = 1865:1994,  # Assuming a running window of 400 years
      gini_value = mean_gini_1850[[region]],
      region = region
    )
  })
)

# Add continent column
df_gini$continent <- sapply(df_gini$region, function(r) {
  continent <- names(continent_mapping)[sapply(continent_mapping, function(regions) r %in% regions)]
  return(continent)
})

# Define continent colors
continent_colors <- c(
  "Europe" = "#F5B0CB",
  "North America" = "#236A3F",
  "Africa" = "#91C7B1",
  "Asia" = "#E5973B",
  "Central America" = "#B33951",
  "South America" = "#E3D081",
  "Australia" = "#3C3980",
  "Ocean" = "white"
)
# Convert continent to factor for proper ordering
df_gini$continent <- factor(df_gini$continent, levels = names(continent_colors))

# Compute global mean Gini index per year
df_global_mean <- df_gini %>%
  group_by(year) %>%
  summarise(global_mean_gini = mean(gini_value, na.rm = TRUE))

# Plot with both regional and global mean
ggplot(df_gini, aes(x = year, y = gini_value, color = continent, group = region)) +
  geom_line() +  # Regional lines
  geom_line(data = df_global_mean, aes(x = year, y = global_mean_gini), 
            color = "black", linewidth = 1.2) +  # Global mean in black
  scale_color_manual(values = continent_colors) +  # Apply continent colors
   scale_y_continuous(breaks = seq(0, 1, by = 0.05)) +  # y-axis ticks every 0.05
  scale_x_continuous(breaks = seq(1880, 1980, by = 50)) +  # y-axis ticks every 0.05
  theme_minimal() +
  labs(title = "30-Year Running Gini Index by IPCC Region",
       x = "Year",
       y = "Gini Index",
       color = "Region") +
  theme(legend.position = "none", axis.text = element_text(size=12), axis.title = element_text(size = 12))

```



