---
title: "PCWD values from ModE-Sim vs ERA5Land output"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Define what to run:
flag_only_run_CHregion <- TRUE  # if TRUE, only runs CH, otherwise all regions (including CH)

dir.create(here::here("figures/"))
```

```{r libraries, message=FALSE, echo=FALSE}
library(here)
library(readr)
library(dplyr)
library(slider)
library(lubridate)
library(patchwork)
library(ggplot2)
library(ncdf4)
library(reshape2)
library(ggpubr)
library(ggtext)
library(maps)
library(terra)
library(rnaturalearth)
library(sf)
library(sp)
library(RColorBrewer)
library(tidyverse)
library(abind)
library(patchwork)
library(gridExtra)
library(ggsci)
library(purrr)
library(qmap)
library(cowplot)
library(extRemes)
library(ggpattern)
library(tidyr)
```

## Comparison of ModE-Sim and ERA5-Land PCWD data and bias correction - Regional comparisons 

## Compute regional dataset for ERA5Land data: 

```{r}
### read in IPCC Region information:
source(here::here("analysis","define_reference_regions.R"))
refregion_objects <- define_reference_regions()

regions <- names(refregion_objects)
# refregion_objects['WCE']
# refregion_objects['WCE_W']
# refregion_objects['WCE_W']
# refregion_objects['CHE']


# Define continent assignments
continent_mapping <- list(
  "Europe" = c("NEU", "WCE", "EEU", "MED"),
  "North America" = c("NWN", "NEN", "WNA", "CNA", "ENA", "GIC"),
  "Africa" = c("SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG"),
  "Asia" = c("WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "WSB", "ESB", "RFE", "RAR"),
  "Central America" =c("NCA", "SCA", "CAR"),
  "South America" = c("NWS", "NSA", "NES", "SAM", "SWS", "SES", "SSA"),
  "Australia" = c("NAU", "CAU", "EAU", "SAU", "NZ")
)
continent_colors <- c(
  "Europe" = "#F5B0CB",
  "North America" = "#236A3F",
  "Africa" = "#91C7B1",
  "Asia" = "#E5973B",
  "Central America" = "#B33951",
  "South America" = "#E3D081",
  "Australia" = "#3C3980",
  "Ocean" = "white"
)

# not mapping the extra regions: "WCE_E","WCE_W","CHE","BERN (25km)","Bern/Zollikofen"
```

### Load modern ERA5-Land reanalysis data

```{r}

if (flag_only_run_CHregion){
  # IF WE ONLY USED CHE AND WCE_W we could do:
  # regional_results_ERA5Land<- readRDS(here("data/regionalResults_CH_only_ERA5Land_cons.RData")) #1950 - 2024
  # incl26_regional_results_ERA5Land<- readRDS(here("data/regionalResults_CH_only_ERA5Land_cons_25_26.RData")) |> #1950 - 2024
  #   purrr::keep_at(c("CHE","WCE_W")) # TODO: need to hardcode this
  
  # NOW THAT WE HAVE RUN IT FOR ALL THE REGINOS WE CAN ALSO INCLUDE WCE:
  regional_results_ERA5Land<- readRDS(here("data/regionalResults_ERA5Land_cons.RData")) |> #1950 - 2024
    purrr::keep_at(c("CHE","WCE_W","WCE_E","WCE")) # TODO: need to hardcode this
  incl26_regional_results_ERA5Land<- readRDS(here("data/regionalResults_ERA5Land_cons_25_26.RData")) |> #1950 - 2024 #1950 - 2024
    purrr::keep_at(c("CHE","WCE_W","WCE_E","WCE")) # TODO: need to hardcode this
} else {
  regional_results_ERA5Land<- readRDS(here("data/regionalResults_ERA5Land_cons.RData")) #1950 - 2024
  incl26_regional_results_ERA5Land<- readRDS(here("data/regionalResults_ERA5Land_cons_25_26.RData")) #1950 - 2024
}
regional_results_ERA5Land        <- purrr::discard_at(regional_results_ERA5Land,        c("BERN (25km)","Bern/Zollikofen"))
incl26_regional_results_ERA5Land <- purrr::discard_at(incl26_regional_results_ERA5Land, c("BERN (25km)","Bern/Zollikofen"))

### to exclude potential precipitation problems in ERA5-Land only include years from 1970s onwards
regional_results_ERA5Land <- lapply(regional_results_ERA5Land, function(x) x[21:75])

ERA5_combined_26 <- mapply(c, regional_results_ERA5Land, incl26_regional_results_ERA5Land, SIMPLIFY = FALSE)
# save cropped ERA5Land data: 
saveRDS(ERA5_combined_26, file = here::here("data/regional_results_ERA5Land_cons_incl26.rds"))   # for 2025-2026
#saveRDS(regional_results_ERA5Land, file = here::here("data/regional_results_ERA5Land_cons.rds")) # for paper
```


### Bias correct regional ModE-Sim data

#### Apply Quantile Mapping to already regional average data in order to fit distribution better
- fit quantile mapping for an overlap period
- then apply to the rest of the dataset

Apply bias correction for chosen overlap period (21:50) - excludes 60s with potential precipitation problem in ModE-Sim and somewhat allows for stationary climate

```{r}
#---------- Read in raw ModE-Sim regional aggregate data and merge epochs
### read in ModE-Sim regional results
if (flag_only_run_CHregion){
  # IF WE ONLY USED CHE AND WCE_W we could do:
  #regional_results_1420 <- readRDS(here("data/regionalResults_CH_only_1420_1.RData"))
  #regional_results_1850 <- readRDS(here("data/regionalResults_CH_only_1850_1.RData"))
  # NOW THAT WE HAVE RUN IT FOR ALL THE REGINOS WE CAN ALSO INCLUDE WCE:
  regional_results_1420 <- readRDS(here("data/regionalResults_1420_1.RData"))|>
    purrr::keep_at(c("CHE","WCE_W","WCE_E","WCE")) # TODO: need to hardcode this
  regional_results_1850 <- readRDS(here("data/regionalResults_1850_1.RData"))|>
    purrr::keep_at(c("CHE","WCE_W","WCE_E","WCE")) # TODO: need to hardcode this
} else {
  regional_results_1420 <- readRDS(here("data/regionalResults_1420_1.RData"))
  regional_results_1850 <- readRDS(here("data/regionalResults_1850_1.RData"))
}
regional_results_1420        <- purrr::discard_at(regional_results_1420,        c("BERN (25km)","Bern/Zollikofen"))
regional_results_1850        <- purrr::discard_at(regional_results_1850,        c("BERN (25km)","Bern/Zollikofen"))

merged_results <- Map(rbind, regional_results_1420, regional_results_1850) # Stitch matrices together row-wise for each region

### Remove m001_tidy for areas CAU, EAU, NAU, NZ, SAU, RFE, EAS, SEA
# List of regions where the first ensemble member should be removed
regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")

# Loop through each specified region and remove the first column (m001_tidy) - bias over Australia
for (region in regions_to_modify) {
  merged_results[[region]] <- merged_results[[region]][, -1]
}

#install.packages("qmap") # uncomment if you haven't installed it
library(qmap)
library(Metrics)

corrected_full <- list()

for (region in names(regional_results_ERA5Land)) {

  obs <- regional_results_ERA5Land[[region]][1:30]  # [1:60] - 30 years (1970-1999)
  sim_overlap <- merged_results[[region]][551:580, ]  # 1970-1999 period for training as 60s have precip problems
  sim_full <- merged_results[[region]]  # 160 years

  corrected_matrix <- matrix(NA, nrow = nrow(sim_full), ncol = ncol(sim_full))
  colnames(corrected_matrix) <- colnames(sim_full)

  for (i in 1:ncol(sim_full)) {  #loop through ensemble members
    mod_train <- sim_overlap[, i]  # 60-year overlap
    mod_full <- sim_full[, i]      # 160-year full run

    # # Fit QM model on 60-year overlap on a parametric transfer function
    qm_model <- fitQmapPTF(obs = obs, mod = mod_train, wet.day = FALSE, qstep = 0.001)
    #
    # # Apply to full 160-year series
    corrected_matrix[, i] <- doQmapPTF(mod_full, qm_model)
  }

  corrected_full[[region]] <- corrected_matrix
}

###save bias corrected ModE-Sim data:
saveRDS(corrected_full, file = here::here("data/regionalResults_ModESim_bc.RData"))
# save cropped ERA5Land data: 
saveRDS(corrected_full, file = here::here("data/regional_results_ModESim_bc.rds"))

corrected_full <- readRDS(here("data/regionalResults_ModESim_bc.RData"))

```

```{r}
## Bias correction cross-validation: 
library(qmap)
library(Metrics)
library(dplyr)
library(tidyr)
library(ggplot2)

set.seed(123)
n_splits <- 10  # number of CV splits

# Initialize a list to store CV results in long format
cv_long <- list()

for (region in names(regional_results_ERA5Land)) {
  
  obs <- regional_results_ERA5Land[[region]][1:30]
  sim_overlap <- merged_results[[region]][551:580, ]
  
  # Store split results for this region
  region_cv <- data.frame()
  
  for (s in 1:n_splits) {
    # Randomly split 30 years into train (70%) and validation (30%)
    train_idx <- sample(1:30, size = 21)
    val_idx <- setdiff(1:30, train_idx)
    
    obs_train <- obs[train_idx]
    obs_val <- obs[val_idx]
    
    sim_train <- sim_overlap[train_idx, ]
    sim_val <- sim_overlap[val_idx, ]
    
    # Initialize matrices to store RMSE
    rmse_raw <- numeric(ncol(sim_overlap))
    rmse_corrected <- numeric(ncol(sim_overlap))
    
    for (i in 1:ncol(sim_overlap)) {
      # Fit QM on training
      qm_model <- fitQmapPTF(obs = obs_train, mod = sim_train[, i], wet.day = FALSE, qstep = 0.001)
      # Corrected validation
      sim_val_corrected <- doQmapPTF(sim_val[, i], qm_model)
      
      # Compute RMSE for raw and corrected
      rmse_raw[i] <- rmse(obs_val, sim_val[, i])
      rmse_corrected[i] <- rmse(obs_val, sim_val_corrected)
    }
    
    # Combine into long format for this split
    split_df <- data.frame(
      region = region,
      split = s,
      ensemble = 1:ncol(sim_overlap),
      raw = rmse_raw,
      corrected = rmse_corrected
    )
    
    # Pivot longer to have a 'type' column
    split_df_long <- pivot_longer(split_df, cols = c("raw", "corrected"),
                                  names_to = "type", values_to = "rmse")
    
    # Append to region CV
    region_cv <- bind_rows(region_cv, split_df_long)
  }
  
  # Store for all regions
  cv_long[[region]] <- region_cv
}

# Combine all regions into a single data frame
cv_long_df <- bind_rows(cv_long)

regions <- unique(cv_long_df$region)

shade_data <- data.frame(
  region = regions,
  xmin = seq_along(regions) - 0.5,
  xmax = seq_along(regions) + 0.5,
  shade = seq_along(regions) %% 2
)

shade_data <- subset(shade_data, shade == 1)

# Plot RMSE comparison raw vs corrected
RMSE_bc_boxplot <- ggplot(cv_long_df, aes(x = region, y = rmse, fill = type)) +
  # Background alternating bands
  geom_rect(
    data = shade_data,
    inherit.aes = FALSE,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    fill = "grey90"
  ) +
  geom_boxplot(position = position_dodge(width = 0.1), linewidth=0.2, outlier.alpha = 0.4) +
  scale_fill_manual(
    values = c("raw" = "#FFBD00", "corrected" = "#255C99"),
    labels = c("ModE-Sim (raw)", "ModESim* (bc)")  # Or customize as you like
  ) +
  labs(y = "RMSE", x = "Region", fill = "Simulation type",
       title = "") +
  theme_classic()  +
theme(axis.text.x = element_text(angle = 45, hjust = 1, color="black", size=11),
    axis.text.y = element_text(size = 12, color="black"),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    legend.position = "none") + # anchor point of the legend box) +  # Rotate x-axis labels
  labs(title = "",
       y = "RMSE",
       x = "",
       fill = "Data Set")
```

Validation plots: 
- violin plots for each region, showing the values from ModESim with ERA5 overlain to see where the match works well 
- scatterplot where data should lie on 1:1 line if it represents well 


```{r}
# Convert ModESim data to long format
ModESim_long <- bind_rows(lapply(names(merged_results), function(region) {
  mat <- merged_results[[region]]
  n_members <- ncol(mat)       # number of ensemble members (could be 19 or 20)
  n_years <- length(years)                # 550
  sub <- mat[seq_len(n_years), , drop = FALSE]  # take first 550 rows
  data.frame(
    Year = rep(1970:2009, times = n_members),  # Repeat years for each ensemble member
    Value = as.vector(merged_results[[region]][551:590,]),  # Flatten matrix
    Region = region,
    Source = "ModESim"
  )
}))

# Convert ModESim data to long format
ModESim_long_bc <- bind_rows(lapply(names(corrected_full), function(region) {
  mat <- corrected_full[[region]]
  n_members <- ncol(mat)       # number of ensemble members (could be 19 or 20)
  n_years <- length(years)                # 550
  sub <- mat[seq_len(n_years), , drop = FALSE]  # take first 550 rows
  
  data.frame(
    Year   = rep(1970:2009, times = n_members), # repeat years accordingly
    Value   = as.vector(mat[551:590, ]),           # flatten matrix
    Region = region,
    Source = "ModESim*"
  )
}))

# Convert ERA5Land data to long format (assuming it's stored in a similar list format)
ERA5Land_long <- bind_rows(lapply(names(regional_results_ERA5Land), function(region) {
  data.frame(
    Year = 1970:2009, #change to 2009 for overlap period
    Value = regional_results_ERA5Land[[region]][1:40],
    Region = region,
    Source = "ERA5Land"
  )
}))

# Combine both datasets
combined_data_raw <- bind_rows(ModESim_long, ModESim_long_bc, ERA5Land_long)
#combined_data_bc <- bind_rows(ModESim_long_bc, ERA5Land_long) #bias corrected ModE-Sim

#### colours x-axis

# Extract unique regions from your data
unique_regions <- unique(combined_data_raw$Region)

# Initialize a named character vector for storing formatted labels.
region_color_labels <- setNames(rep(NA, length(unique_regions)), unique_regions)


# For each continent, assign the corresponding region labels that match
for (continent in names(continent_mapping)) {
  regions <- continent_mapping[[continent]]
  for (region in regions) {
    if (region %in% unique_regions) {
      # Format the region label with the designated colour from continent_colors
      region_color_labels[region] <- sprintf('<span style="color:%s;">%s</span>', 
                                        continent_colors[continent], 
                                        region)
    }
  }
}

# In case any regions haven't been matched, assign a default colour (e.g. black)
for (region in unique_regions) {
  if (is.na(region_color_labels[region])) {
    region_color_labels[region] <- sprintf('<span style="color:%s;">%s</span>', "black", region)
  }
}


library(ggtext)  # <-- needed to render markdown/html in ggplot

# Create numeric index for Regions
regions <- unique(combined_data_raw$Region)

shade_data <- data.frame(
  Region = regions,
  xmin = seq_along(regions) - 0.5,
  xmax = seq_along(regions) + 0.5,
  shade = seq_along(regions) %% 2  # alternate
)

shade_data <- subset(shade_data, shade == 1)

# Plot the violin plot
bc_violin <- ggplot(combined_data_raw, aes(x = Region, y = Value, fill = Source)) +
    # Add alternating background bands
  geom_rect(
    data = shade_data,
    inherit.aes = FALSE,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    fill = "grey90"
  ) +
  geom_violin(scale = "width", linewidth=0.2, adjust = .6) +
  theme_classic() +
  scale_fill_manual(
    values = c( "ERA5Land" = "#B3001B", "ModESim" = "#FFBD00", "ModESim*" = "#255C99"),
    labels = c(  "ERA5-Land", "ModE-Sim (raw)", "ModESim* (bc)")  # Or customize as you like
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color="black", size=11),
    axis.text.y = element_text(size = 12, color="black"),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    legend.position = c(0.2, 0.95),  # top-right corner inside plot
    legend.justification = c("right", "top")) + # anchor point of the legend box) +  # Rotate x-axis labels
  labs(title = "",
       y = "PCWD [mm]",
       x = "Region",
       fill = "Data Set")

bc_violin

```

```{r}
# # Extract unique regions from your data
# unique_regions <- unique(merged_results$Region)
# 
# # Initialize a named character vector for storing formatted labels.
# region_color_labels <- setNames(rep(NA, length(unique_regions)), unique_regions)
# 
# 
# # For each continent, assign the corresponding region labels that match
# for (continent in names(continent_mapping)) {
#   regions <- continent_mapping[[continent]]
#   for (region in regions) {
#     if (region %in% unique_regions) {
#       # Format the region label with the designated colour from continent_colors
#       region_color_labels[region] <- sprintf('<span style="color:%s;">%s</span>', 
#                                         continent_colors[continent], 
#                                         region)
#     }
#   }
# }
# 
# # In case any regions haven't been matched, assign a default colour (e.g. black)
# for (region in unique_regions) {
#   if (is.na(region_color_labels[region])) {
#     region_color_labels[region] <- sprintf('<span style="color:%s;">%s</span>', "black", region)
#   }
# }
# 
# 
# library(ggtext)  # <-- needed to render markdown/html in ggplot
# 
# combined_data_raw <- bind_rows(ModESim_long, ERA5Land_long)
# 
# # Create numeric index for Regions
# regions <- unique(combined_data_raw$Region)
# 
# shade_data <- data.frame(
#   Region = regions,
#   xmin = seq_along(regions) - 0.5,
#   xmax = seq_along(regions) + 0.5,
#   shade = seq_along(regions) %% 2  # alternate
# )
# 
# shade_data <- subset(shade_data, shade == 1)
# 
# # Plot the violin plot
# bc_violin <- ggplot(combined_data_raw, aes(x = Region, y = Value, fill = Source)) +
#     # Add alternating background bands
#   geom_rect(
#     data = shade_data,
#     inherit.aes = FALSE,
#     aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
#     fill = "grey90"
#   ) +
#   geom_violin(scale = "width", linewidth=0.2, adjust = .6) +
#   theme_classic() +
#   scale_fill_manual(
#     values = c( "ERA5Land" = "#B3001B", "ModESim" = "#FFBD00"),
#     labels = c(  "ERA5-Land", "ModE-Sim (raw)")  # Or customize as you like
#   ) +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1, color="black", size=11),
#     axis.text.y = element_text(size = 12, color="black"),
#     axis.title.x = element_text(size = 13),
#     axis.title.y = element_text(size = 13),
#     legend.title = element_text(face = "bold"),
#     legend.text = element_text(size=12),
#     legend.position = c(0.2, 0.95),  # top-right corner inside plot
#     legend.justification = c("right", "top")) + # anchor point of the legend box) +  # Rotate x-axis labels
#   labs(title = "",
#        y = "PCWD [mm]",
#        x = "Region",
#        fill = "Data Set")
# 
# bc_violin
```


```{r}
## combine bc plots for SI: 
combined_bc_plots <- plot_grid(
  RMSE_bc_boxplot, bc_violin,
  ncol = 1,
  nrow = 2,
  rel_widths = c(1, 1),
  align = "h",
  axis = "tblr",
  labels = c(
    "a)",  "b) "
  )
  # label_size = 14,
  # label_x = 0.09,
  # hjust = 0,
  # label_y = 0.95,
  # vjust = 1
)

combined_bc_plots

ggsave(
  filename = here::here("figures/combined_bc_plots.pdf"),
  plot     = combined_bc_plots,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 7.5,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```


#### Internal variability vs forced signal in ModE-Sim
```{r}
library(tidyverse)

# # List of regions where the first ensemble member should be removed
# regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")
# 
# # Loop through each specified region and remove the first column (m001_tidy) - bias over Australia
# for (region in regions_to_modify) {
#   corrected_full[[region]] <- corrected_full[[region]][, -1]
# }

variability_summary <- map_dfr(names(corrected_full), function(region) {
  mat <- corrected_full[[region]]  # 590 x 20
  
  ensemble_mean <- rowMeans(mat)           # forced signal (590,)
  residuals     <- mat - ensemble_mean     # internal variability (590 x 20)
  
  tibble(
    region        = region,
    mean_pcwd     = mean(ensemble_mean),
    sd_internal   = mean(apply(residuals, 1, sd)),  # avg internal spread
    cv            = sd(rowMeans(mat)) / mean(rowMeans(mat)),
    snr           = mean(ensemble_mean) / mean(apply(residuals, 1, sd))
  )
})

# Convert list to a lookup data frame
continent_df <- utils::stack(continent_mapping) %>%
  rename(region = values, continent = ind)

# Join onto your variability_summary
variability_summary <- variability_summary %>%
  left_join(continent_df, by = "region")

# Assuming variability_summary has columns: region, snr, continent
variability_summary %>%
  arrange(snr) %>%
  mutate(region = factor(region, levels = region)) %>%   # fix order
  ggplot(aes(x = snr, y = region, color = continent)) +
  
  # Reference line at SNR = 1 (noise-dominated threshold)
  #geom_vline(xintercept = 5, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  
  # Lollipop stem
  geom_segment(aes(x = 0, xend = snr, y = region, yend = region),
               linewidth = 0.5, alpha = 0.8) +
  
  # Lollipop head
  geom_point(size = 3) +
  
  scale_color_manual(values = continent_colors) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  
  labs(
    x     = "SNR",
    y     = NULL,
    color = "Continent",
  ) +
  
  theme_classic() +
  theme(
    axis.text.y      = element_text(size = 10),
    legend.position  = c(0.8, 0.2),
    legend.title     = element_text(face = "bold"),
    text             = element_text(size = 12),
    axis.title.x     = element_text(margin = margin(t = 10)),
    plot.caption     = element_text(color = "grey50", size = 9)
  )

ggsave(
  filename = here::here("figures/SNR_regions.pdf"),
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 7.5,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```


```{r}
## Variations within the ensemble mean: 

# Define the timeframe matching your 590 rows
years <- 1420:2009

# Step 1: Compute the ensemble mean for each region across all years
# This reduces each region's matrix from [590, 20] or [590, 18] down to a single vector of 590
ensemble_means_list <- lapply(corrected_full, rowMeans)

# Step 2: Convert to a clean data frame and add the temporal axis
ensemble_means_df <- as.data.frame(ensemble_means_list)
ensemble_means_df$Year <- years

# Step 3: Reshape data and calculate temporal variability metrics per region
variability_summary_mean <- ensemble_means_df %>%
  pivot_longer(cols = -Year, names_to = "Region", values_to = "Ensemble_Mean") %>%
  group_by(Region) %>%
  summarize(
    Mean_Baseline = mean(Ensemble_Mean),
    SD_Absolute   = sd(Ensemble_Mean),
    CV_Relative   = sd(Ensemble_Mean) / mean(Ensemble_Mean),
    Min_Value     = min(Ensemble_Mean),
    Max_Value     = max(Ensemble_Mean),
    Range_Absolute= max(Ensemble_Mean) - min(Ensemble_Mean),
    .groups = "drop"
  )

# View summary table sorted by highest relative variability
print(variability_summary_mean %>% arrange(desc(CV_Relative)))

ggplot(variability_summary_mean, aes(x = reorder(Region, CV_Relative), y = CV_Relative)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Relative Temporal Variation (CV) of Ensemble Mean by IPCC Region",
    subtitle = "Years 1420-2009",
    x = "IPCC Region",
    y = "Coefficient of Variation (SD / Mean)"
  ) +
  theme_minimal()

# Extract the top 5 most variable regions based on CV
top_5_regions <- variability_summary_mean %>% 
  arrange(desc(CV_Relative)) %>% 
  slice_head(n = 5) %>% 
  pull(Region)

# Plot their historical time-series
ensemble_means_df %>%
  pivot_longer(cols = -Year, names_to = "Region", values_to = "Ensemble_Mean") %>%
  filter(Region %in% top_5_regions) %>%
  ggplot(aes(x = Year, y = Ensemble_Mean, color = Region)) +
  geom_line(alpha = 0.6) +
  facet_wrap(~Region, scales = "free_y", ncol = 1) +
  labs(
    title = "Historical Ensemble Mean Trajectories (Top 5 Most Volatile Regions)",
    x = "Year",
    y = "Drought Index Ensemble Mean"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```


#### Quantitative analysis 
How many 2000 - 2024 ERA5Land values lie outside the EM range of ModE-Sim (i.e. are unprecedented?)
Check how many regions lie outside of 2SD when considering whole ensemble member range? 

Calculate how many modern PCWD values are truly unprecedented (max(ERA5) > max(ref))
Calculate how many modern PCWD values are record shattering (max(ERA5) > (max(ref)+2sd*(ref))

```{r}
#------------reference values ERA5 - 2000-2024
#Convert ERA5Land data to long format 
ERA5Land_modern <- bind_rows(lapply(names(regional_results_ERA5Land), function(region) {
  data.frame(
    Year = 2000:2024, #change to 2009 for overlap period
    PCWD = regional_results_ERA5Land[[region]][31:55], #indexing year 2000-2024
    Region = region,
    Source = "ERA5Land"
  )
}))


#-----------reference values long context ModESim - 1420-1969
# Convert ModESim data to long format
years_ref <- 1420:1969
n_years <- length(years_ref)

ModESim_long <- bind_rows(
  lapply(names(corrected_full), function(region) {

    mat <- corrected_full[[region]]
    sub <- mat[1:n_years, , drop = FALSE]

    data.frame(
      Year = rep(years_ref, times = ncol(sub)),
      PCWD = as.vector(sub),
      Ensemble = rep(colnames(sub), each = n_years),
      Region = region,
      Source = "ModESim"
    )
  })
)


#reference values short context ModESim - 1970-1999
years_short <- 1970:1999
n_years <- length(years_short)

ModESim_short <- bind_rows(
  lapply(names(corrected_full), function(region) {

    mat <- corrected_full[[region]]
    sub <- mat[551:580, , drop = FALSE]

    data.frame(
      Year = rep(years_short, times = ncol(sub)),
      PCWD = as.vector(sub),
      Ensemble = rep(colnames(sub), each = n_years),
      Region = region,
      Source = "ModESim"
    )
  })
)

##------------------ Calculate whether PCWD is unprecedented
#Get reference ranges (1420–1969) for each region
ref_range_long <- ModESim_long %>%
  group_by(Region) %>%
  summarise(ref_max = max(PCWD, na.rm = TRUE),
            ref_maxsd = max(PCWD, na.rm = TRUE) + 2 * sd(PCWD, na.rm = TRUE),
            .groups = "drop")

#Get reference ranges (1970–1999) for each region
ref_range_short <- ModESim_short %>%
  group_by(Region) %>%
  summarise(ref_max = max(PCWD, na.rm = TRUE),
            ref_maxsd = max(PCWD, na.rm = TRUE) + 2 * sd(PCWD, na.rm = TRUE),
            .groups = "drop")

# 2. Compare modern period (2000–2024) to reference range
########Long reference
modern_outside_long <- ERA5Land_modern %>%
  left_join(ref_range_long, by = "Region") %>%
  filter(PCWD > ref_max) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_outside_long)
n_regions_outside_long <- nrow(modern_outside_long)
print(paste("In ", n_regions_outside_long, "out of 44 regions, PCWD is unprecendent in the long reference"))

##########Short reference
modern_outside_short <- ERA5Land_modern %>%
  left_join(ref_range_short, by = "Region") %>%
  filter(PCWD > ref_max) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_outside_short)
n_regions_outside_short <- nrow(modern_outside_short)
print(paste("In ", n_regions_outside_short, "out of 44 regions, PCWD is unprecendent in the short reference"))



##------------------ Calculate whether PCWD is "record-shattering"
# 2. Compare modern period (2000–2024) to reference range
########Long reference
modern_shattering_long <- ERA5Land_modern %>%
  left_join(ref_range_long, by = "Region") %>%
  filter(PCWD > ref_maxsd) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_shattering_long)
n_regions_shattering_long <- nrow(modern_shattering_long)
print(paste("In ", n_regions_shattering_long, "out of 44 regions, PCWD is record-shattering in the long reference"))

##########Short reference
modern_shattering_short <- ERA5Land_modern %>%
  left_join(ref_range_short, by = "Region") %>%
  filter(PCWD > ref_maxsd) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_shattering_short)
n_regions_shattering_short <- nrow(modern_shattering_short)
print(paste("In ", n_regions_shattering_short, "out of 44 regions, PCWD is record-shattering in the short reference"))



```
Compute the same for 25-year rolling means - general shift in PCWD distribution? 

```{r}
Modesim_25_long <- ModESim_long %>%
  group_by(Region, Ensemble) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(PCWD25 = slide_dbl(PCWD, mean, .before = 29, .complete = TRUE)) %>%
  ungroup()

Modesim_25_short <- ModESim_short %>%
  group_by(Region, Ensemble) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(PCWD25 = slide_dbl(PCWD, mean, .before = 29, .complete = TRUE)) %>%
  ungroup()


ERA5Land_means <- ERA5Land_modern %>%
  group_by(Region) %>%
  summarise(PCWD_mean = mean(PCWD, na.rm = TRUE), .groups = "drop")


#----------------Calculate whether ERA5Land means are above reference running means
#Get reference ranges (1420–1969) for each region
ref_range_long25 <- Modesim_25_long %>%
  group_by(Region) %>%
  summarise(ref_max = max(PCWD25, na.rm = TRUE),
            .groups = "drop")

#Get reference ranges (1970–1999) for each region
ref_range_short25 <- Modesim_25_short %>%
  group_by(Region) %>%
  summarise(ref_max = max(PCWD25, na.rm = TRUE),
            .groups = "drop")

# 2. Compare modern period (2000–2024) to reference range
########Long reference
modern_outside_long25 <- ERA5Land_means %>%
  left_join(ref_range_long25, by = "Region") %>%
  filter(PCWD_mean > ref_max) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_outside_long25)
n_regions_outside_long25 <- nrow(modern_outside_long25)
print(paste("In ", n_regions_outside_long25, "out of 44 regions, mean PCWD is unprecendent in the long reference"))

##########Short reference
modern_outside_short25 <- ERA5Land_means %>%
  left_join(ref_range_short25, by = "Region") %>%
  filter(PCWD_mean > ref_max) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_outside_short25)
n_regions_outside_short25 <- nrow(modern_outside_short25)
print(paste("In ", n_regions_outside_short25, "out of 44 regions, mean PCWD is unprecendent in the short reference"))

```

### Absolute values timeseries of bias corrected ModE-Sim and ERA5-Land - visualisation

Include rolling 25-year anomalies as thin black line with highlighted events

```{r}
# Initialize list to store rolling 30-year means
rolling_means <- list()

# Define window size
window_size <- 25
half_window <- (window_size - 1) / 2

start_year <- 1420
end_year   <- 2009

centers <- seq(start_year, end_year - window_size + 1) + half_window

# Compute rolling means for each ensemble member
for (region_name in names(corrected_full)) {
  
  # Extract ensemble members for the region
  ensemble_members <- corrected_full[[region_name]][2:590,]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means[[region_name]] <- rolling_mean_matrix
}

rolling_means_long <-  do.call(rbind, lapply(names(rolling_means), function(region) {
  df <- as.data.frame(rolling_means[[region]])
  df$time <- centers[1:nrow(df)]
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "rolling_means")
  return(df_long)
}))


```


Calculate 600 year rolling anomalies from bias corrected ModE-Sim data with complete data as reference period
```{r}

# Initialize lists to store baselines
anom_baseline <- list()

# Compute baseline mean from all values of rolling mean
for (region_name in names(rolling_means)) {
  rolling_mean_matrix <- rolling_means[[region_name]]
  
  anom_baseline[[region_name]] <- mean(rolling_mean_matrix, na.rm = TRUE)  ### for reference period of first 25 values: rolling_mean_matrix[1:25,]
}

# Initialize lists to store anomalies
ensemble_25anom <- list()

# Compute anomalies
for (region_name in names(rolling_means)) {
  rolling_mean_matrix <- rolling_means[[region_name]]
  num_years <- nrow(rolling_mean_matrix)
  
  # Initialize matrix to store anomalies
  anoms <- matrix(NA, nrow=num_years, ncol=ncol(rolling_mean_matrix))
  
  # Compute anomalies using rolling mean baseline
  for (t in 1:num_years) {
    anoms[t, ] <- (rolling_mean_matrix[t, ] - anom_baseline[[region_name]])
  }
  
  # Store results
  ensemble_25anom[[region_name]] <- anoms
}

# Compute mean anomalies across ensemble members
mean_25anom <- list()
for (region_name in names(ensemble_25anom)) {
  mean_25anom[[region_name]] <- rowMeans(ensemble_25anom[[region_name]], na.rm = TRUE)
}

# Compute region-specific pooled standard deviation using only the first 25 years
pooled_sd_by_region <- list()
for (region_name in names(ensemble_25anom)) {
  # Extract only the first 25 years of anomalies
  first_25_years <- ensemble_25anom[[region_name]]
  
  # Compute pooled standard deviation over these values
  pooled_sd_by_region[[region_name]] <- sd(unlist(first_25_years), na.rm = TRUE)
}

# Convert ensemble data into a long format for plotting individual members
anoms_long <- do.call(rbind, lapply(names(ensemble_25anom), function(region) {
  df <- as.data.frame(ensemble_25anom[[region]])
  df$time <- centers[1:nrow(df)]
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "anomalies")
  return(df_long)
}))

# Add a continent column
anoms_long$continent <- NA  # Initialize continent column
for (continent in names(continent_mapping)) {
  anoms_long$continent[anoms_long$region %in% continent_mapping[[continent]]] <- continent
}


```


Anomaly timeseries plot with bias corrected data for selected regions: 

```{r}
####-----------------------------------------------------------------------
if (!flag_only_run_CHregion){
  selected_region <- "MED"
  anoms_filtered <- anoms_long %>% filter(region == selected_region)
  
  # Compute mean anomalies per (time, region)
  mean_anomalies_filtered <- anoms_filtered %>%
    group_by(time, region) %>%
    summarise(
      mean_anomaly = mean(anomalies, na.rm = TRUE),
      pooled_sd = pooled_sd_by_region[[unique(region)]]
    ) %>%
    ungroup()
  
  # Extract the first year's mean anomaly per region
  first_year_mean <- mean_anomalies_filtered %>%
    group_by(region) %>%
    arrange(time) %>%
    slice_head(n = 1) %>%  # Take only the first year
    dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
    ungroup()
  
  # Join back and compute fixed bounds
  mean_anomalies_filtered <- mean_anomalies_filtered %>%
    left_join(first_year_mean, by = "region") %>%
    mutate(
      lower_bound = first_mean_anomaly - pooled_sd,
      upper_bound = first_mean_anomaly + pooled_sd
    )
  
  anom_MED <- ggplot() +
    geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+
  
    # Plot individual ensemble members with distinct colors
    geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.4, linewidth = 0.3, show.legend = TRUE) +
    # Plot mean anomaly as a bold black line
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
    # Dashed lines for ±2σ bounds
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
    labs(
      x = "",
      y = "PCWD anomalies [mm] w.r.t 1420-1445"
      #,title = paste("Rolling 25yr PCWD anomalies for", selected_continent)
    ) +
    # Manual color scale for legend
    scale_color_manual(
      values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
      labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
    ) +
  
    # Customize the legend
    guides(
      color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
    ) +
    theme_classic() +
    theme(
      legend.position = "none",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.text = element_text(size = 15),
      strip.background = element_rect(fill = "#F5B0CB"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
      axis.text.y = element_text(size = 11), 
      axis.title.x = element_text(size = 11),
      axis.title.y = element_text(size = 11)
    ) +
    
    facet_wrap(~region, scales = "free_y")
  
  #######---------------------------------------------------------------------------------
  selected_region <- "WNA"
  anoms_filtered <- anoms_long %>% filter(region == selected_region)
  
  # Compute mean anomalies per (time, region)
  mean_anomalies_filtered <- anoms_filtered %>%
    group_by(time, region) %>%
    summarise(
      mean_anomaly = mean(anomalies, na.rm = TRUE),
      pooled_sd = pooled_sd_by_region[[unique(region)]]
    ) %>%
    ungroup()
  
  # Extract the first year's mean anomaly per region
  first_year_mean <- mean_anomalies_filtered %>%
    group_by(region) %>%
    arrange(time) %>%
    slice_head(n = 1) %>%  # Take only the first year
    dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
    ungroup()
  
  # Join back and compute fixed bounds
  mean_anomalies_filtered <- mean_anomalies_filtered %>%
    left_join(first_year_mean, by = "region") %>%
    mutate(
      lower_bound = first_mean_anomaly - pooled_sd,
      upper_bound = first_mean_anomaly + pooled_sd
    )
  
  anom_WNA <- ggplot() +
    geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+
  
    # Plot individual ensemble members with distinct colors
    geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
    # Plot mean anomaly as a bold black line
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
    # Dashed lines for ±2σ bounds
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
    labs(
      x = "",
      y = ""
      #,title = paste("Rolling 25yr PCWD anomalies for", selected_continent)
    ) +
    # Manual color scale for legend
    scale_color_manual(
      values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
      labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
    ) +
  
    # Customize the legend
    guides(
      color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
    ) +
    theme_classic() +
    theme(
      legend.position = "none",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.text = element_text(size = 15, color = "white"),
      strip.background = element_rect(fill = "#236A3F"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
      axis.text.y = element_text(size = 11), 
      axis.title.x = element_text(size = 11),
      axis.title.y = element_text(size = 11)
    ) +
    
    facet_wrap(~region, scales = "free_y")
  
  ###########------------------------------------------------------------
  selected_region <- "ARP"
  anoms_filtered <- anoms_long %>% filter(region == selected_region)
  
  # Compute mean anomalies per (time, region)
  mean_anomalies_filtered <- anoms_filtered %>%
    group_by(time, region) %>%
    summarise(
      mean_anomaly = mean(anomalies, na.rm = TRUE),
      pooled_sd = pooled_sd_by_region[[unique(region)]]
    ) %>%
    ungroup()
  
  # Extract the first year's mean anomaly per region
  first_year_mean <- mean_anomalies_filtered %>%
    group_by(region) %>%
    arrange(time) %>%
    slice_head(n = 1) %>%  # Take only the first year
    dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
    ungroup()
  
  # Join back and compute fixed bounds
  mean_anomalies_filtered <- mean_anomalies_filtered %>%
    left_join(first_year_mean, by = "region") %>%
    mutate(
      lower_bound = first_mean_anomaly - pooled_sd,
      upper_bound = first_mean_anomaly + pooled_sd
    )
  
  anom_ARP <- ggplot() +
    geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+
  
    # Plot individual ensemble members with distinct colors
    geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
    # Plot mean anomaly as a bold black line
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
    # Dashed lines for ±2σ bounds
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
    labs(
      x = "Year",
      y = "PCWD anomalies [mm] w.r.t 1420-1445"
      #,title = paste("Rolling 25yr PCWD anomalies for", selected_continent)
    ) +
    # Manual color scale for legend
    scale_color_manual(
      values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
      labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
    ) +
  
    # Customize the legend
    guides(
      color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
    ) +
    theme_classic() +
    theme(
      legend.position = "none",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.text = element_text(size = 15),
      strip.background = element_rect(fill = "#E5973B"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
      axis.text.y = element_text(size = 11), 
      axis.title.x = element_text(size = 12),
      axis.title.y = element_text(size = 11)
    ) +
    
    facet_wrap(~region, scales = "free_y")
  
  ###----------------------------------------------------------------------------
  selected_region <- "CAU"
  anoms_filtered <- anoms_long %>% filter(region == selected_region)
  
  # Compute mean anomalies per (time, region)
  mean_anomalies_filtered <- anoms_filtered %>%
    group_by(time, region) %>%
    summarise(
      mean_anomaly = mean(anomalies, na.rm = TRUE),
      pooled_sd = pooled_sd_by_region[[unique(region)]]
    ) %>%
    ungroup()
  
  # Extract the first year's mean anomaly per region
  first_year_mean <- mean_anomalies_filtered %>%
    group_by(region) %>%
    arrange(time) %>%
    slice_head(n = 1) %>%  # Take only the first year
    dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
    ungroup()
  
  # Join back and compute fixed bounds
  mean_anomalies_filtered <- mean_anomalies_filtered %>%
    left_join(first_year_mean, by = "region") %>%
    mutate(
      lower_bound = first_mean_anomaly - pooled_sd,
      upper_bound = first_mean_anomaly + pooled_sd
    )
  
  anom_CAU <- ggplot() +
    geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+
  
    # Plot individual ensemble members with distinct colors
    geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
    # Plot mean anomaly as a bold black line
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
    # Dashed lines for ±2σ bounds
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
    geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
    labs(
      x = "Year",
      y = ""
      #,title = paste("Rolling 25yr PCWD anomalies for", selected_continent)
    ) +
    # Manual color scale for legend
    scale_color_manual(
      values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
      labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
    ) +
  
    # Customize the legend
    guides(
      color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
    ) +
    theme_classic() +
    theme(
      legend.position = "none",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      strip.text = element_text(size = 15,color = "white"),
      strip.background = element_rect(fill = "#3C3980"),
      axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
      axis.text.y = element_text(size = 11), 
      axis.title.x = element_text(size = 12),
      axis.title.y = element_text(size = 11)
    ) +
    
    facet_wrap(~region, scales = "free_y")
}
```

Test - ModESim MED annmax PCWD distribution as context for ERA5Land extremes

```{r}
## select MED region data
#MED_ERA5 <- ERA5_combined_26[["WCE"]] #regional_results_ERA5Land for only until 2024 as in paper
MED_ERA5 <- ERA5_combined_26[["WCE_W"]] #regional_results_ERA5Land for 2025/2026
# TODO: change here??? to WCE_W

#ModESim data
#MED_MODE <- corrected_full[["WCE"]]
MED_MODE <- corrected_full[["WCE_W"]]

#attach time dimension
time_info_mode <- seq(from = as.numeric(1420), to = as.numeric(2009))
time_info_era <- seq(from = as.numeric(1970), to = as.numeric(2026))


# Convert ModE dataset to long format
MED_MODE_long <- MED_MODE %>%
  as.data.frame() %>%
  mutate(time = time_info_mode) %>%
  pivot_longer(cols = -time, names_to = "ensemble", values_to = "PCWD")

# Compute means
MED_MODE_mean <- MED_MODE_long %>%
  group_by(time) %>%
  summarise(PCWD = mean(PCWD, na.rm = TRUE), .groups = "drop")

# Convert ERA5 dataset into long format
MED_ERA5_df <- data.frame(time = time_info_era, PCWD = MED_ERA5)

## Calculation return period values 

baseline <- MED_ERA5[5:30] ##1975-1999
fit_ref <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_era_ref_5yr <- round(as.numeric(extRemes::return.level(fit_ref, return.period = 5)))

baseline <- MED_ERA5[31:55] ##1970-1999
fit_mod <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_era_mod_5yr <- round(as.numeric(extRemes::return.level(fit_mod, return.period = 5)))

baseline <- as.vector(MED_MODE[535:560,])  #1975-1999
fit_ref <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_ModESim_ref_5yr <- round(as.numeric(extRemes::return.level(fit_ref, return.period = 5)))

baseline <- as.vector(MED_MODE[1:534,])  #1420-1969
fit_full <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_ModESim_full_5yr <- round(as.numeric(extRemes::return.level(fit_full, return.period = 5)))


# 1) extract 1970-1999 of the wide matrix, then pivot
MED_last60_long <- MED_MODE[(nrow(MED_MODE)-35):(nrow(MED_MODE)-10), ] %>%
  as.data.frame() %>%
  pivot_longer(cols = everything(),
               names_to  = "ensemble",
               values_to = "PCWD")

# Convert ModE dataset to long format 1420 - 1969
MED_MODE_long_sel <- MED_MODE %>%
  as.data.frame() %>%
  mutate(time = time_info_mode) %>%
  filter(time < 1975) %>%
  pivot_longer(cols = -time, names_to = "ensemble", values_to = "PCWD")

# 1) extract last 60 rows of the wide matrix, then pivot
MED_ERA5_long <- MED_ERA5[5:30] %>%
  as.data.frame() %>%
  pivot_longer(cols = everything(),
              names_to  = "year",
               values_to = "PCWD")

MED_ERA5_last30 <- MED_ERA5[31:55] %>%
  as.data.frame() %>%
  pivot_longer(cols = everything(),
              names_to  = "year",
               values_to = "PCWD")

ggplot() +
  # full‑data density
  geom_density(
    data    = MED_MODE_long_sel,
    aes(x   = PCWD,
    color   = factor("ModESim_fill_full"),
    fill    = factor("ModESim_fill_full")),
    alpha   = 0.8,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
  # last‑60 density
  geom_density(
    data    = MED_last60_long,
    aes(x   = PCWD,
    color   = factor("ModESim_fill"),
    fill    = factor("ModESim_fill")),
    alpha   = 0.8,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
    # first‑60 density
  geom_density(
    data    = MED_ERA5_long,
    aes(x   = PCWD,
    color   = factor("ERA5_fill"),
    fill    = factor("ERA5_fill")),
    alpha   = 0.6,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
  geom_density(
    data    = MED_ERA5_last30,
    aes(x   = PCWD,
    color   = factor("ERA5_fill_last30"),
    fill    = factor("ERA5_fill_last30")),
    alpha   = 0.6,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
  geom_vline(aes(xintercept = MED_ModESim_full_5yr, color = factor("ModESim_full")), linetype = "dashed", linewidth=0.9) +
  geom_vline(aes(xintercept = MED_ModESim_ref_5yr, color = factor("ModESim_ref")), linetype = "dashed", linewidth=0.9) +
  geom_vline(aes(xintercept = MED_era_ref_5yr, color = factor("ERA_ref")), linetype = "dashed", linewidth=0.9) +
  geom_vline(aes(xintercept = MED_era_mod_5yr, color = factor("ERA_mod")), linetype = "dashed", linewidth=0.9) +
    labs(
    x = "PCWD [mm]",
    y = "Density",
    title = "WCE_W Extreme Value Distribution"
  ) +
  theme_minimal() +
 scale_fill_manual(
    name = "Data Source",
    values = c("ModESim_fill_full" = "#255C99","ModESim_fill" = "#91C499", "ERA5_fill" = "#A25E8E", "ERA5_fill_last30" = "#B3001B") ,
    labels = c("ModE-Sim* (1420-1974)", "ModE-Sim* (1975-1999)", "ERA5-Land (1975-1999)", "ERA5-Land (2000-2024)")
  )+
  scale_color_manual(
    name = "5-year Event",
    values = c("ModESim_full" = "#17385E", "ModESim_ref" = "#628567", "ERA_ref" = "#E8B5CC", "ERA_mod" = "#870014"),
    labels = c("ModE-Sim* (1420 - 1974)", "ModE-Sim* (1975 - 1999)", "ERA5Land (1975 - 1999)", "ERA5Land (2000 - 2024)")
    #guide = guide_legend(override.aes = list(fill= NA,shape = NA, color = "white"))
  )+

    theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.key = element_rect(fill   = NA,    # transparent fill
                              colour = NA),# no border
    legend.title = element_text(face = "bold")
  )


# ggsave(
#   filename = here::here("manuscript/Annmax_PCWD_Timeseries_Density_legend.pdf"),
#   device   = cairo_pdf,    # better text rendering
#   width    = 12,           # in inches
#   height   = 6,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )

```


Function that does the plotting for all regions: 

```{r}
# ------------------------------------------------------------------------------
# Helper: prepare the four slices + FEVD-based 5‑yr return levels
# ------------------------------------------------------------------------------
prepare_region_data <- function(region,
                                corrected_full,
                                ERA5_combined_26,
                                mode_years = 1420:2009,
                                era5_years = 1970:2026,
                                drop_initial_years = 2) {
  MODE_mat <- corrected_full[[region]]
  ERA5_vec <- ERA5_combined_26[[region]]
  
  # Time‐series data
  MODE_long <- MODE_mat %>%
    as.data.frame() %>%
    mutate(time = mode_years) %>%
    pivot_longer(-time, names_to = "ensemble", values_to = "PCWD")
  
  MODE_mean <- MODE_long %>% group_by(time) %>% summarise(PCWD = mean(PCWD, na.rm=TRUE), .groups="drop")
  
  ERA5_df   <- data.frame(time = era5_years, PCWD = ERA5_vec)
  
  # ———————— drop the first `drop_initial_years` rows ————————
  MODE_long <- MODE_long %>% 
    filter(time >= mode_years[drop_initial_years])
  
  MODE_mean <- MODE_mean %>% 
    filter(time >= mode_years[drop_initial_years])
  # ——————————————————————————————————————————————————————————————
  
  # Four density slices
  MODE_full <- MODE_long %>% filter(time < 1975)
  MODE_ref  <- MODE_long %>% filter(time >= 1975 & time <= 1999)
  ERA5_ref  <- ERA5_df   %>% filter(time >= 1975 & time <= 1999)
  ERA5_mod  <- ERA5_df   %>% filter(time >= 2000 & time <= 2025)
  ERA5_2026  <- ERA5_df   %>% filter(time == 2026)
  
  # Extract numeric scalar value for 2026
  val_2026   <- ERA5_df %>% filter(time == 2026) %>% pull(PCWD)
  
  # FEVD + 5‑yr return level
  fit_rl <- function(x) {
    fit <- fevd(x, type="Gumbel", method="MLE", units="years")
    round(as.numeric(return.level(fit, return.period = 5)))
  }
  rl <- list(
    MODE_full    = fit_rl(MODE_full$PCWD),
    MODE_ref     = fit_rl(MODE_ref$PCWD),
    ERA5_ref     = fit_rl(ERA5_ref$PCWD),
    ERA5_mod     = fit_rl(ERA5_mod$PCWD)
  )
  
  list(
    MODE_long   = MODE_long,
    MODE_mean   = MODE_mean,
    ERA5_df     = ERA5_df,
    MODE_full   = MODE_full,
    MODE_ref    = MODE_ref,
    ERA5_ref    = ERA5_ref,
    ERA5_mod    = ERA5_mod,
    ERA5_2026   = ERA5_2026,
    val_2026   = val_2026, # Store the numeric value here
    rl          = rl
  )
}

```

```{r}
#Timeseries function alone: 
plot_region_PCWD_ts <- function(region_code) {
  
  dat <- prepare_region_data(region_code,
                             corrected_full,
                             ERA5_combined_26)
  
  # Compute anomalies + bounds + exceedance runs
  anoms_filtered <- anoms_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      mean_anomaly = mean(anomalies, na.rm=TRUE),
      pooled_sd    = pooled_sd_by_region[[region_code]],
      .groups = "drop"
    )
  
  first_mean <- anoms_filtered %>%
    arrange(time) %>%
    slice(1) %>%
    pull(mean_anomaly)
  
  anoms_filtered <- anoms_filtered %>%
    mutate(
      lower_bound = first_mean - pooled_sd,
      upper_bound = first_mean + pooled_sd,
      exceeds     = mean_anomaly < lower_bound | mean_anomaly > upper_bound,
      direction   = case_when(
        mean_anomaly > upper_bound ~ "drier",
        mean_anomaly < lower_bound ~ "wetter",
        TRUE ~ NA_character_
      ),
      run_id      = data.table::rleid(exceeds)
    )
  
  exceed_periods <- anoms_filtered %>%
    filter(exceeds) %>%
    group_by(run_id, direction) %>%
    summarise(
      start = min(time),
      end   = max(time),
      .groups="drop"
    )
  
  # Filter rolling means
  rolling_filtered <- rolling_means_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      mean_rolling = mean(rolling_means, na.rm=TRUE),
      .groups = "drop"
    )
  
  # y-limits
  y_limits     <- range(c(dat$MODE_long$PCWD, 200), na.rm=TRUE)
   #y_limits     <- range(c(dat$MODE_long$PCWD, dat$ERA5_df$PCWD), na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  # Compute max of each ensemble for dotted lines
  ensemble_max <- dat$MODE_long %>%
    group_by(ensemble) %>%
    summarise(max_PCWD = max(PCWD, na.rm=TRUE), .groups = "drop") %>%
    summarise(max_max = max(max_PCWD, na.rm = TRUE), .groups="drop") %>%
    mutate(label = "Ens_maximum")
  
  # Compute max of each ensemble for dotted lines
  ERA5_2026_max <- dat$ERA5_2026 %>%
  summarise(max_PCWD = max(PCWD, na.rm=TRUE), .groups = "drop") %>%  
  summarise(max_max = max(max_PCWD, na.rm = TRUE), .groups="drop") %>%
    mutate(label = "2026")
  
  # Build ggplot
  p_ts <- ggplot()
  
  # Exceedance rectangles colored by direction
  if (nrow(exceed_periods) > 0) {
    p_ts <- p_ts + 
      geom_rect(
        data = exceed_periods,
        aes(xmin = start, xmax = end, 
            ymin = y_limits[1], ymax = y_limits[2],
            fill = direction),
        inherit.aes = FALSE,
        alpha = 0.4
      ) 
  }
  p_ts <- p_ts +
    geom_line(data = dat$MODE_long, aes(time, PCWD, group=ensemble, color = factor("EMs_ModE")), linewidth=0.3, alpha=0.4) +
    geom_line(data = dat$MODE_mean, aes(time, PCWD, color = factor("Mean_ModE")), linewidth = 0.8) +
    geom_line(data = rolling_filtered, aes(time, mean_rolling, color="rolling_mean"), linewidth=0.7) +
    geom_line(data = dat$ERA5_df, aes(time, PCWD, color = factor("ERA5")), linewidth = 0.8) +
    geom_hline(data = ensemble_max, aes(yintercept = max_max, color= factor("Ens_maximum")),linetype="dotted", linewidth=0.8)+
    geom_hline(data = ERA5_2026_max, aes(yintercept = max_max, color= factor("2026")),linetype="dashed", linewidth=0.8)+
    scale_y_continuous(limits=y_limits, breaks=common_breaks) +
    scale_color_manual(name = "Data",
      values = c(
        "EMs_ModE"  = "#7EA3CC",
        "Mean_ModE" = "#255C99",
        "rolling_mean"="black", 
        "ERA5"      = "#B3001B", 
        "Ens_maximum"= "grey30",
        "2026" = "#ffb800"
      ),
      labels = c(
        "EMs - ModE-Sim*", "Mean - ModE-Sim*", "25-yr rolling\nmean - ModE-Sim*", "ERA5-Land", "ModE-Sim* - max PCWD", "ERA5-Land 2026")
      ) +
      scale_fill_manual(name ="Exceedances", values = c("drier" = "#F4DE70", "wetter" = "#136E40"),
      labels = c("drier", "wetter")
    ) +
    labs(x = "Year", y = "PCWD [mm]", color = "Data") +
    theme_classic(base_size = 14) +
    theme(axis.title.y = element_text(margin=margin(r=10)),
          legend.position="none", legend.title = element_text(face="bold"))
  
  return(p_ts)
}

# ts_MED <- plot_region_PCWD_ts("WCE")
ts_MED <- plot_region_PCWD_ts("CHE")
# ts_MED <- plot_region_PCWD_ts("WCE_W")
ts_MED
```

```{r}
#density plot function alone: 
plot_region_PCWD_density <- function(region_code) {
  
  dat <- prepare_region_data(region_code,
                             corrected_full,
                             ERA5_combined_26)
  
  # y-limits
  y_limits     <- range(c(dat$MODE_long$PCWD, 200), na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  p_den <- ggplot() +
    geom_density(data = dat$MODE_full,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ModESim_full"),
                 alpha=0.8, size=0.6, trim = FALSE) +
    geom_density(data = dat$MODE_ref,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ModESim_ref"),
                 alpha=0.8, size=0.6, trim = FALSE) +
    geom_density(data = dat$ERA5_ref,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ERA5_70_99"),
                 alpha=0.6, size=0.6, trim = FALSE) +
    geom_density(data = dat$ERA5_mod,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ERA5_00_24"),
                 alpha=0.6, size=0.6, trim = FALSE) +

    # geom_hline(aes(yintercept=dat$rl$MODE_full,  color="ModESim_full_lin"),
    #            linetype="dashed", size=0.8) +
    # geom_hline(aes(yintercept=dat$rl$MODE_ref,   color="ModESim_ref_lin"),
    #            linetype="dashed", size=0.8) +
    # geom_hline(aes(yintercept=dat$rl$ERA5_ref,   color="ERA5_70_99_lin"),
    #            linetype="dashed", size=0.8) +
    # geom_hline(aes(yintercept=dat$rl$ERA5_mod,   color="ERA5_00_24_lin"),
    #            linetype="dashed", size=0.8) +
# Horizontal line for 2026 ERA5 value
    geom_hline(aes(yintercept = dat$val_2026, color = "ERA5_26_lin"),
               linetype = "dashed", linewidth = 0.8) +    
    scale_fill_manual(
      name   = "Period",
      breaks = c("ModESim_full", "ModESim_ref", "ERA5_70_99", "ERA5_00_24"),
      values = c(
        ModESim_full  = "#255C99",
        ModESim_ref   = "#91C499",
        ERA5_70_99    = "#A25E8E",
        ERA5_00_24    = "#B3001B"
      ),
      labels = c(
        "ModE-Sim* (1420-1974)",
        "ModE-Sim* (1975-1999)",
        "ERA5-Land (1975-1999)",
        "ERA5-Land (2000-2026)"
      )
    ) +
    scale_color_manual(
      name   = "Single Year Event",
      breaks = c("ModESim_full_lin", "ModESim_ref_lin", "ERA5_70_99_lin", "ERA5_00_24_lin", "ERA5_26_lin"),
      values = c(
        ModESim_full_lin  = "#17385E",
        ModESim_ref_lin   = "#628567",
        ERA5_70_99_lin    = "#E8B5CC",
        ERA5_00_24_lin    = "#870014",
        #ERA5_26_lin       = "#f4a259"
        ERA5_26_lin       = "#ffb800"
      ),
      labels = c(
        "ModE-Sim* (1420-1969)",
        "ModE-Sim* (1970-1999)",
        "ERA5-Land (1970-1999)",
        "ERA5-Land (2000-2024)",
        "ERA5-Land (2026)"
      ))+
    labs(x="Density", y=NULL) +
    theme_classic(base_size=14) +
    theme(
      plot.title     = element_text(hjust=0.5, face="bold"),
      legend.key     = element_rect(fill=NA, colour=NA),
      legend.title   = element_text(face="bold"),
      text           = element_text(size=14),
      axis.title.y   = element_text(margin=margin(r=10)),
      legend.position = "none"
    )+
    scale_y_continuous(limits = y_limits, breaks = common_breaks)
  
  return(p_den)
}

# TEST CASES:
# den_MED <- plot_region_PCWD_density("WCE")
# den_MED <- plot_region_PCWD_density("CHE")
# den_MED <- plot_region_PCWD_density("WCE_W")
# den_MED
```

Combine multiple panels into one and save: 

```{r}
##for Fig. A in press release:
plot_WCE_ts  <- plot_region_PCWD_ts("WCE")
plot_WCE_den <- plot_region_PCWD_density("WCE")
plot_WCE_den <- plot_WCE_den + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

plot_WCE_E_ts <- plot_region_PCWD_ts("WCE_E")
plot_WCE_E_den <- plot_region_PCWD_density("WCE_E")
plot_WCE_E_den <- plot_WCE_E_den + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

plot_WCE_W_ts <- plot_region_PCWD_ts("WCE_W")
plot_WCE_W_den <- plot_region_PCWD_density("WCE_W")
plot_WCE_W_den <- plot_WCE_W_den + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

plot_CHE_ts <- plot_region_PCWD_ts("CHE")
plot_CHE_den <- plot_region_PCWD_density("CHE")
plot_CHE_den <- plot_CHE_den + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

# plot_WCE_ts + theme(legend.position = "bottom")
# plot_WCE_E_ts + theme(legend.position = "bottom")
# plot_WCE_W_ts + theme(legend.position = "bottom")
# plot_CHE_ts + theme(legend.position = "bottom")
# plot_WCE_den + theme(legend.position = "bottom")
# plot_WCE_E_den + theme(legend.position = "bottom")
# plot_WCE_W_den + theme(legend.position = "bottom")
# plot_CHE_den + theme(legend.position = "bottom")

#legend_a <- get_legend(plot_WCE_W_ts  + theme(legend.position = "bottom", legend.box.margin = margin(0, 0, 0, 12)))
#legend_b <- get_legend(plot_WCE_W_den + theme(legend.position = "bottom", legend.box.margin = margin(0, 0, 0, 12)))

combined2all <- plot_grid(
  plot_WCE_ts,  plot_WCE_den, 
  plot_WCE_E_ts,plot_WCE_E_den, 
  plot_WCE_W_ts,plot_WCE_W_den, 
  plot_CHE_ts,  plot_CHE_den,
  # legend_a, legend_b, 
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) WCE – West Central Europe",  "e)", 
    "b) WCE_E – Eastern half of WCE",  "f)", 
    "c) WCE_W – Western half of WCE",  "g)", 
    "d) CHE – Switzerland", "h)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(2, 1),
  expand      = TRUE  
)


print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_WCEW_CHE_Fig.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,       # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```

```{r}
# redo above plot with zoom into 2000's
combined2all <- plot_grid(
  plot_WCE_ts   + coord_cartesian(xlim = c(2000,2028)) + scale_x_continuous("Year",minor_breaks = c(2000:2028)) + theme_bw(), plot_WCE_den, 
  plot_WCE_E_ts + coord_cartesian(xlim = c(2000,2028)) + scale_x_continuous("Year",minor_breaks = c(2000:2028)) + theme_bw(), plot_WCE_E_den, 
  plot_WCE_W_ts + coord_cartesian(xlim = c(2000,2028)) + scale_x_continuous("Year",minor_breaks = c(2000:2028)) + theme_bw(), plot_WCE_W_den, 
  plot_CHE_ts   + coord_cartesian(xlim = c(2000,2028)) + scale_x_continuous("Year",minor_breaks = c(2000:2028)) + theme_bw(), plot_CHE_den,
  # legend_a, legend_b, 
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) WCE – West Central Europe",  "e)", 
    "b) WCE_E – Eastern half of WCE",  "f)", 
    "c) WCE_W – Western half of WCE",  "g)", 
    "d) CHE – Switzerland", "h)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(2, 1),
  expand      = TRUE  
)


print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_WCEW_CHE_Fig_zoom2026.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,       # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```
