---
title: "PCWD values from ModE-Sim vs ERA5Land output"
author: "Patricia Helpap, Fabian Bernhard, Benjamin Stocker"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

dir.create(here::here("figures"), showWarnings = FALSE)
```

```{r libraries, message=FALSE, echo=FALSE}
library(cowplot)
library(dplyr)
library(ggplot2)
library(here)
library(purrr)
library(tidyr)
```

## Load data

Run `ERA5Land_vs_ModESim_forPressRelease.Rmd` first to generate the input files.

```{r load-data}
era5_combined_26 <- readRDS(
  here("data/regional_results_ERA5Land_cons_incl26.rds")
)
corrected_full <- readRDS(
  here("data/regionalResults_ModESim_bc.RData")
)
```

## Prepare data

```{r prepare-rolling-means}
mode_years <- 1420:2009
era5_years <- 1970:2026
window_size <- 25L
half_window <- (window_size - 1) / 2

calculate_rolling_means <- function(mode_matrix, region) {
  ensemble_members <- mode_matrix[2:590, , drop = FALSE]
  number_of_windows <- nrow(ensemble_members) - window_size + 1

  rolling_matrix <- t(vapply(
    seq_len(number_of_windows),
    function(start_index) {
      end_index <- start_index + window_size - 1
      colMeans(
        ensemble_members[start_index:end_index, , drop = FALSE],
        na.rm = TRUE
      )
    },
    numeric(ncol(ensemble_members))
  ))

  rolling_centers <-
    seq(min(mode_years), max(mode_years) - window_size + 1) + half_window

  rolling_matrix |>
    as.data.frame() |>
    mutate(
      time = rolling_centers[seq_len(nrow(rolling_matrix))],
      region = region
    ) |>
    pivot_longer(
      cols = -c(time, region),
      names_to = "ensemble_member",
      values_to = "rolling_mean"
    )
}

rolling_means_long <- corrected_full |>
  imap(calculate_rolling_means) |>
  list_rbind()
```

```{r prepare-region-data}
prepare_region_data <- function(
  region,
  corrected_full,
  era5_combined_26,
  mode_years = 1420:2009,
  era5_years = 1970:2026,
  drop_initial_years = 2L
) {
  mode_matrix <- corrected_full[[region]]
  era5_values <- era5_combined_26[[region]]

  stopifnot(
    nrow(mode_matrix) == length(mode_years),
    length(era5_values) == length(era5_years)
  )

  mode_long <- mode_matrix |>
    as.data.frame() |>
    mutate(time = mode_years) |>
    pivot_longer(
      cols = -time,
      names_to = "ensemble",
      values_to = "PCWD"
    ) |>
    filter(time >= mode_years[[drop_initial_years]])

  mode_mean <- mode_long |>
    group_by(time) |>
    summarise(PCWD = mean(PCWD, na.rm = TRUE), .groups = "drop")

  era5_data <- tibble(time = era5_years, PCWD = era5_values)
  era5_modern <- era5_data |>
    filter(between(time, 2000, 2026))

  pcwd_2026 <- era5_data |>
    filter(time == 2026) |>
    pull(PCWD)

  list(
    mode_long = mode_long,
    mode_mean = mode_mean,
    era5_data = era5_data,
    era5_modern = era5_modern,
    pcwd_2026 = pcwd_2026
  )
}
```

## Plot functions

```{r plot-helpers}
get_pcwd_y_limits <- function(region_code, pcwd_values) {
  upper_limit <- if (region_code == "CHE") 150 else 200
  c(min(pcwd_values, na.rm = TRUE), upper_limit)
}

get_pcwd_y_breaks <- function(region_code, y_limits) {
  if (region_code == "CHE") {
    seq(0, 150, by = 50)
  } else {
    pretty(y_limits, n = 6)
  }
}
```

```{r plot-time-series}
plot_region_pcwd_ts <- function(region_code) {
  data <- prepare_region_data(
    region_code,
    corrected_full,
    era5_combined_26
  )

  rolling_mean <- rolling_means_long |>
    filter(region == region_code) |>
    group_by(time) |>
    summarise(
      mean_rolling = mean(rolling_mean, na.rm = TRUE),
      .groups = "drop"
    )

  y_limits <- get_pcwd_y_limits(region_code, data$mode_long$PCWD)
  common_breaks <- get_pcwd_y_breaks(region_code, y_limits)
  record_pcwd <- max(data$mode_long$PCWD, na.rm = TRUE)

  ggplot() +
    geom_line(
      data = data$mode_long,
      aes(
        x = time,
        y = PCWD,
        group = ensemble,
        color = "ensemble_members"
      ),
      linewidth = 0.3,
      alpha = 0.4
    ) +
    geom_line(
      data = data$mode_mean,
      aes(x = time, y = PCWD, color = "ensemble_mean"),
      linewidth = 0.6
    ) +
    geom_line(
      data = rolling_mean,
      aes(x = time, y = mean_rolling, color = "rolling_mean"),
      linewidth = 0.6
    ) +
    geom_line(
      data = data$era5_data,
      aes(x = time, y = PCWD, color = "era5_land"),
      linewidth = 0.6
    ) +
    geom_hline(
      yintercept = record_pcwd,
      color = "#255C99",
      linewidth = 0.4
    ) +
    geom_hline(
      yintercept = data$pcwd_2026,
      color = "#B3001B",
      linewidth = 0.4
    ) +
    annotate(
      "text",
      x = -Inf,
      y = data$pcwd_2026,
      label = "2026",
      color = "#B3001B",
      size = 3,
      hjust = -0.2,
      vjust = -0.4
    ) +
    scale_y_continuous(limits = y_limits, breaks = common_breaks) +
    scale_color_manual(
      name = "Data",
      values = c(
        ensemble_members = "#7EA3CC",
        ensemble_mean = "#255C99",
        rolling_mean = "black",
        era5_land = "#B3001B"
      ),
      labels = c(
        ensemble_members = "EMs - ModE-Sim*",
        ensemble_mean = "Mean - ModE-Sim*",
        rolling_mean = "25-yr rolling\nmean - ModE-Sim*",
        era5_land = "ERA5-Land"
      )
    ) +
    labs(x = "Year", y = "PCWD (mm)", color = "Data") +
    theme_classic(base_size = 14) +
    theme(
      axis.title.y = element_text(margin = margin(r = 10)),
      legend.position = "none",
      legend.title = element_text(face = "bold")
    )
}
```

```{r plot-density}
plot_region_pcwd_density <- function(
  region_code,
  show_legend = region_code == "CHE"
) {
  data <- prepare_region_data(
    region_code,
    corrected_full,
    era5_combined_26
  )

  y_limits <- get_pcwd_y_limits(region_code, data$mode_long$PCWD)
  common_breaks <- get_pcwd_y_breaks(region_code, y_limits)
  record_pcwd <- max(data$mode_long$PCWD, na.rm = TRUE)

  density_plot <- ggplot() +
    geom_density(
      data = data$mode_long,
      aes(
        x = after_stat(density),
        y = PCWD,
        fill = "mode_sim_based"
      ),
      alpha = 0.8,
      linewidth = 0.6,
      trim = FALSE
    ) +
    geom_density(
      data = data$era5_modern,
      aes(
        x = after_stat(density),
        y = PCWD,
        fill = "era5_2000_2026"
      ),
      alpha = 0.6,
      linewidth = 0.6,
      trim = FALSE
    ) +
    geom_hline(
      yintercept = record_pcwd,
      color = "#255C99",
      linewidth = 0.4
    ) +
    geom_hline(
      yintercept = data$pcwd_2026,
      color = "#B3001B",
      linewidth = 0.4
    ) +
    annotate(
      "text",
      x = Inf,
      y = data$pcwd_2026,
      label = "2026",
      color = "#B3001B",
      size = 3,
      hjust = 1.2,
      vjust = -0.4
    ) +
    scale_fill_manual(
      name = NULL,
      breaks = c("era5_2000_2026", "mode_sim_based"),
      values = c(
        mode_sim_based = "#255C99",
        era5_2000_2026 = "#B3001B"
      ),
      labels = c(
        era5_2000_2026 = "Climate reanalysis",
        mode_sim_based = "Simulations"
      )
    ) +
    scale_y_continuous(limits = y_limits, breaks = common_breaks) +
    labs(x = "Density", y = NULL) +
    theme_classic(base_size = 14) +
    theme(
      axis.title.y = element_text(margin = margin(r = 10)),
      legend.key = element_rect(fill = NA, colour = NA),
      legend.position = "none",
      legend.title = element_text(face = "bold"),
      plot.title = element_text(hjust = 0.5, face = "bold"),
      text = element_text(size = 14)
    )

  if (show_legend) {
    density_plot +
      guides(fill = guide_legend(nrow = 1, byrow = TRUE)) +
      theme(
        legend.background = element_blank(),
        legend.direction = "horizontal",
        legend.justification = c(0.5, 0.5),
        legend.key.height = grid::unit(0.35, "cm"),
        legend.key.width = grid::unit(0.45, "cm"),
        legend.margin = margin(0),
        legend.position = "inside",
        legend.position.inside = c(0.5, 0.9),
        legend.spacing.x = grid::unit(0.08, "cm"),
        legend.text = element_text(
          family = "Helvetica",
          size = 10,
          margin = margin(l = 2, r = 8)
        )
      )
  } else {
    density_plot
  }
}
```

## Combine and save panels

```{r build-figure}
region_codes <- c("WCE", "CHE")
panel_labels <- c(
  "WCE – West Central Europe",
  "",
  "CHE – Switzerland",
  ""
)
region_titles <- c(
  WCE = "WCE – West Central Europe",
  CHE = "CHE – Switzerland"
)

build_pcwd_figure <- function(zoom = FALSE) {
  time_series_plots <- map(region_codes, plot_region_pcwd_ts)

  if (zoom) {
    time_series_plots <- map(
      time_series_plots,
      ~ .x +
        coord_cartesian(xlim = c(2000, 2028)) +
        scale_x_continuous(
          "Year",
          minor_breaks = 2000:2028
        ) +
        theme_bw()
    )
  }

  density_plots <- region_codes |>
    map(plot_region_pcwd_density) |>
    map(~ .x + theme(plot.margin = margin(t = 20)))

  panel_plots <- map2(time_series_plots, density_plots, list) |>
    list_flatten()

  plot_grid(
    plotlist = panel_plots,
    ncol = 2,
    align = "h",
    axis = "tb",
    labels = panel_labels,
    label_x = 0.02,
    label_y = 0.975,
    hjust = 0,
    vjust = 1,
    rel_widths = c(2, 1)
  )
}

build_region_pcwd_figure <- function(region_code) {
  stopifnot(region_code %in% names(region_titles))

  time_series_plot <- plot_region_pcwd_ts(region_code)
  density_plot <- plot_region_pcwd_density(
    region_code,
    show_legend = TRUE
  ) +
    theme(plot.margin = margin(t = 20))

  plot_grid(
    time_series_plot,
    density_plot,
    ncol = 2,
    align = "h",
    axis = "tb",
    labels = c(region_titles[[region_code]], ""),
    label_x = 0.02,
    label_y = 0.975,
    hjust = 0,
    vjust = 1,
    rel_widths = c(2, 1)
  )
}

save_pcwd_figure <- function(plot, filename, height = 5) {
  ggsave(
    filename = here("figures", filename),
    plot = plot,
    device = cairo_pdf,
    width = 10,
    height = height,
    dpi = 300,
    bg = "white"
  )
}
```

```{r save-figures}
pcwd_figure <- build_pcwd_figure()
print(pcwd_figure)
save_pcwd_figure(
  pcwd_figure,
  "Annmax_PCWD_Timeseries_Density_WCEW_CHE_Fig.pdf"
)

pcwd_figure_zoom <- build_pcwd_figure(zoom = TRUE)
print(pcwd_figure_zoom)
save_pcwd_figure(
  pcwd_figure_zoom,
  "Annmax_PCWD_Timeseries_Density_WCEW_CHE_Fig_zoom2026.pdf"
)

pcwd_figure_wce <- build_region_pcwd_figure("WCE")
save_pcwd_figure(
  pcwd_figure_wce,
  "PCWD_WCE.pdf",
  height = 3
)

pcwd_figure_che <- build_region_pcwd_figure("CHE")
save_pcwd_figure(
  pcwd_figure_che,
  "PCWD_CHE.pdf",
  height = 3
)
```

## Percentile of the 2026 PCWD

The empirical quantile is the proportion of values in a distribution that
are less than or equal to the ERA5-Land 2026 PCWD. The climate-reanalysis
distribution covers 2000-2026, matching the red density, while the simulation
distribution contains the ModE-Sim* member-year values shown in blue.

```{r calculate-2026-percentiles}
calculate_2026_percentiles <- function(region_code) {
  data <- prepare_region_data(
    region_code,
    corrected_full,
    era5_combined_26
  )

  tibble(
    region = region_code,
    distribution = c(
      "Climate reanalysis (ERA5-Land, 2000-2026)",
      "Simulations (ModE-Sim*, 1421-2009)"
    ),
    number_of_values = c(
      sum(!is.na(data$era5_modern$PCWD)),
      sum(!is.na(data$mode_long$PCWD))
    ),
    pcwd_2026_mm = data$pcwd_2026,
    quantile = c(
      mean(
        data$era5_modern$PCWD <= data$pcwd_2026,
        na.rm = TRUE
      ),
      mean(
        data$mode_long$PCWD <= data$pcwd_2026,
        na.rm = TRUE
      )
    )
  )
}

percentiles_2026 <- region_codes |>
  map(calculate_2026_percentiles) |>
  list_rbind() |>
  mutate(percentile = 100 * quantile)

percentile_table <- percentiles_2026 |>
  transmute(
    Region = recode(region, WCE = "West Central Europe", CHE = "Switzerland"),
    Distribution = distribution,
    N = number_of_values,
    `2026 PCWD (mm)` = sprintf("%.1f", pcwd_2026_mm),
    Quantile = sprintf("%.4f", quantile),
    Percentile = sprintf("%.1f%%", percentile)
  )

knitr::kable(
  percentile_table,
  align = c("l", "l", "r", "r", "r", "r"),
  caption = "Empirical position of the 2026 PCWD in each displayed distribution."
)
```

## Ten highest PCWD values

ERA5-Land entries are annual values. ModE-Sim* entries are individual
ensemble-member and year combinations from the blue distribution.

```{r calculate-top-10-pcwd}
calculate_top_10_pcwd <- function(region_code) {
  data <- prepare_region_data(
    region_code,
    corrected_full,
    era5_combined_26
  )

  era5_top_10 <- data$era5_data |>
    slice_max(order_by = PCWD, n = 10, with_ties = FALSE) |>
    arrange(desc(PCWD)) |>
    transmute(
      region = region_code,
      data_source = "ERA5-Land",
      rank = row_number(),
      year = time,
      ensemble_member = "-",
      pcwd_mm = PCWD
    )

  mode_top_10 <- data$mode_long |>
    slice_max(order_by = PCWD, n = 10, with_ties = FALSE) |>
    arrange(desc(PCWD)) |>
    transmute(
      region = region_code,
      data_source = "ModE-Sim*",
      rank = row_number(),
      year = time,
      ensemble_member = ensemble,
      pcwd_mm = PCWD
    )

  bind_rows(era5_top_10, mode_top_10)
}

top_10_pcwd <- region_codes |>
  map(calculate_top_10_pcwd) |>
  list_rbind()
```

```{r print-top-10-pcwd, results='asis'}
region_names <- c(
  WCE = "West Central Europe",
  CHE = "Switzerland"
)

for (region_code in region_codes) {
  for (source in c("ERA5-Land", "ModE-Sim*")) {
    cat(
      "### ", region_names[[region_code]], " - ", source, "\n\n",
      sep = ""
    )

    table_data <- top_10_pcwd |>
      filter(region == region_code, data_source == source) |>
      transmute(
        Rank = rank,
        Year = year,
        `Ensemble member` = ensemble_member,
        `PCWD (mm)` = round(pcwd_mm, 1)
      )

    print(knitr::kable(table_data, align = c("r", "r", "c", "r")))
    cat("\n\n")
  }
}
```
