---
title: "PCWD values from ModE-Sim vs ERA5Land output"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Bern/Master thesis/R/lastmill_droughts") #set working directory
knitr::opts_knit$set(root.dir = "~/Bern/Master thesis/R/lastmill_droughts")

```

```{r libraries, message=FALSE, echo=FALSE}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(ggplot2)
library(ncdf4)
library(reshape2)
library(ggpubr)
library(maps)
library(terra)
library(rnaturalearth)
library(sf)
library(sp)
library(RColorBrewer)
library(tidyverse)
library(abind)
library(patchwork)
library(gridExtra)
library(ggsci)
library(purrr)
library(qmap)
library(cowplot)
library(extRemes)
library(ggpattern)

```

## Comparison of ModE-Sim and ERA5-Land PCWD data and bias correction - Regional comparisons 

## Compute regional dataset for ERA5Land data: 

```{r}
### read in IPCC Region information:
#Load reference regions and coastlines:

load("data/Reference_regions/IPCC-WGI-reference-regions-v4_R.rda", verbose = TRUE)

#simplify this object by converting it to a SpatialPolygons class object (i.e., only the polygons are retained and their attributes discarded):
refregions <- as(IPCC_WGI_reference_regions_v4, "SpatialPolygons")

# List of regions to loop over --- excludes ocean basins
regions <- c("GIC", "NWN", "NEN", "WNA", "CNA", "ENA", "NCA", "SCA", "CAR", "NWS",
             "NSA", "NES", "SAM", "SWS", "SES", "SSA", "NEU", "WCE", "EEU", "MED",
             "SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG", "RAR", "WSB",
             "ESB", "RFE", "WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "NAU",
             "CAU", "EAU", "SAU", "NZ")
# regions <- c("GIC")

# Define continent assignments
continent_mapping <- list(
  "Europe" = c("NEU", "WCE", "EEU", "MED"),
  "North America" = c("NWN", "NEN", "WNA", "CNA", "ENA", "GIC"),
  "Africa" = c("SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG"),
  "Asia" = c("WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "WSB", "ESB", "RFE", "RAR"),
  "Central America" =c("NCA", "SCA", "CAR"),
  "South America" = c("NWS", "NSA", "NES", "SAM", "SWS", "SES", "SSA"),
  "Australia" = c("NAU", "CAU", "EAU", "SAU", "NZ")
)


regional_results_ERA5Land<- readRDS("data/regionalResults_ERA5Land.RData") #1950 - 2024
### to exclude potential precipitation problems in ERA5-Land only include years from 1970s onwards
regional_results_ERA5Land <- lapply(regional_results_ERA5Land, function(x) x[21:75])

# ### read in ModE-Sim regional results
# regional_results_1420 <- readRDS("~/cwd_global/data/regionalResults_1420_1.RData")
# regional_results_1850 <- readRDS("~/cwd_global/data/regionalResults_1850_1.RData")
# merged_results <- Map(rbind, regional_results_1420, regional_results_1850) # Stitch matrices together row-wise for each region
# 
# ### Remove m001_tidy for areas CAU, EAU, NAU, NZ, SAU, RFE, EAS, SEA
# # List of regions where the first ensemble member should be removed
# regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")
# 
# # Loop through each specified region and remove the first column (m001_tidy) - bias over Australia
# for (region in regions_to_modify) {
#   merged_results[[region]] <- merged_results[[region]][, -1]
# }
# 

```

### Bias correct regional ModE-Sim data

#### Apply Quantile Mapping to already regional average data in order to fit distribution better
- fit quantile mapping for an overlap period
- then apply to the rest of the dataset

Apply bias correction for chosen overlap period (21:50) - excludes 60s with potential precipitation problem in ModE-Sim and somewhat allows for stationary climate

```{r}
# #install.packages("qmap") # uncomment if you haven't installed it
# library(qmap)
# library(Metrics)
# 
# corrected_full <- list()
# 
# for (region in names(regional_results_ERA5Land)) {
#   
#   obs <- regional_results_ERA5Land[[region]][1:30]  # [1:60] - 30 years (1970-1999)
#   sim_overlap <- merged_results[[region]][551:580, ]  # 1970-1999 period for training as 60s have precip problems
#   sim_full <- merged_results[[region]]  # 160 years
#   
#   corrected_matrix <- matrix(NA, nrow = nrow(sim_full), ncol = ncol(sim_full))
#   colnames(corrected_matrix) <- colnames(sim_full)
# 
#   for (i in 1:ncol(sim_full)) {  #loop through ensemble members
#     mod_train <- sim_overlap[, i]  # 60-year overlap
#     mod_full <- sim_full[, i]      # 160-year full run
#     
#     # # Fit QM model on 60-year overlap on a parametric transfer function
#     qm_model <- fitQmapPTF(obs = obs, mod = mod_train, wet.day = FALSE, qstep = 0.001)
#     #
#     # # Apply to full 160-year series
#     corrected_matrix[, i] <- doQmapPTF(mod_full, qm_model)
#   }
# 
#   corrected_full[[region]] <- corrected_matrix
# }
# 
# ###save bias corrected ModE-Sim data: 
# saveRDS(corrected_full, "~/cwd_global/data/regionalResults_ModESim_bc.RData")

corrected_full <- readRDS("data/regionalResults_ModESim_bc.RData")

```



Validation plots: 
- violin plots for each region, showing the values from ModESim with ERA5 overlain to see where the match works well 
- scatterplot where data should lie on 1:1 line if it represents well 

```{r}
# Convert ModESim data to long format
ModESim_long <- bind_rows(lapply(names(merged_results), function(region) {
  data.frame(
    Year = rep(1950:2009, times = 20),  # Repeat years for each ensemble member
    Value = as.vector(merged_results[[region]][531:590,]),  # Flatten matrix
    Region = region,
    Source = "ModESim"
  )
}))

# Convert ModESim data to long format
ModESim_long_bc <- bind_rows(lapply(names(corrected_full), function(region) { #regional_results_ModESim_bc for mean_bias corrected global data
  data.frame(
    Year = rep(1950:2009, times = 20),  # Repeat years for each ensemble member
    Value = as.vector(corrected_full[[region]][531:590,]),  # Flatten matrix
    Region = region,
    Source = "ModESim_bc"
  )
}))

# Convert ERA5Land data to long format (assuming it's stored in a similar list format)
ERA5Land_long <- bind_rows(lapply(names(regional_results_ERA5Land), function(region) {
  data.frame(
    Year = 1970:2009, #change to 2009 for overlap period
    Value = regional_results_ERA5Land[[region]][1:40],
    Region = region,
    Source = "ERA5Land"
  )
}))

# Combine both datasets
combined_data_raw <- bind_rows(ModESim_long, ERA5Land_long)
combined_data_bc <- bind_rows(ModESim_long_bc, ERA5Land_long) #bias corrected ModE-Sim

#### colours x-axis
# Define continent assignments
continent_mapping <- list(
  "Europe" = c("NEU", "WCE", "EEU", "MED"),
  "North America" = c("NWN", "NEN", "WNA", "CNA", "ENA", "GIC"),
  "Africa" = c("SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG"),
  "Asia" = c("WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "WSB", "ESB", "RFE", "RAR"),
  "Central America" =c("NCA", "SCA", "CAR"),
  "South America" = c("NWS", "NSA", "NES", "SAM", "SWS", "SES", "SSA"),
  "Australia" = c("NAU", "CAU", "EAU", "SAU", "NZ")
)
continent_colors <- c(
  "Europe" = "#F5B0CB",
  "North America" = "#236A3F",
  "Africa" = "#91C7B1",
  "Asia" = "#E5973B",
  "Central America" = "#B33951",
  "South America" = "#E3D081",
  "Australia" = "#3C3980",
  "Ocean" = "white"
)

# Extract unique regions from your data
unique_regions <- unique(combined_data_raw$Region)

# Initialize a named character vector for storing formatted labels.
region_color_labels <- setNames(rep(NA, length(unique_regions)), unique_regions)

#install.packages("ggtext")  # if not already installed
library(ggtext)             # must be called in your script


# For each continent, assign the corresponding region labels that match
for (continent in names(continent_mapping)) {
  regions <- continent_mapping[[continent]]
  for (region in regions) {
    if (region %in% unique_regions) {
      # Format the region label with the designated colour from continent_colors
      region_color_labels[region] <- sprintf("<span style='color:%s;'>%s</span>", 
                                               continent_colors[continent], 
                                               region)
    }
  }
}

# In case any regions haven't been matched, assign a default colour (e.g. black)
for (region in unique_regions) {
  if (is.na(region_color_labels[region])) {
    region_color_labels[region] <- sprintf("<span style='color:%s;'>%s</span>", "black", region)
  }
}

# Inspect the mapping to confirm:
#print(region_color_labels)

# Plot the violin plot
ggplot(combined_data_raw, aes(x = Region, y = Value, fill = Source)) +
  geom_violin(scale = "width") +
  theme_minimal() +
  scale_fill_manual(
    values = c( "ERA5Land" = "#087E8B", "ModESim" = "#FF5A5F"),
    labels = c( "ERA5-Land", "ModE-Sim")  # Or customize as you like
  ) +
  scale_x_discrete(labels = region_color_labels) +  # Apply custom coloured labels
  theme(axis.text.x = element_markdown(angle = 90, hjust = 1, size = 12),
    legend.position = "none",
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),) +  # Rotate x-axis labels
  labs(title = "ModE-Sim (RAW) vs ERA5-Land by Region (1950-2009)",
       y = "PCWD Value [mm]",
       x = "Region",
       fill = "Dataset")


# Plot the violin plot
ggplot(combined_data_bc, aes(x = Region, y = Value, fill = Source)) +
  geom_violin(scale = "width") +
  theme_minimal() +
  scale_fill_manual(
    values = c( "ERA5Land" = "#087E8B", "ModESim_bc" = "#FF5A5F"),
    labels = c( "ERA5-Land", "ModE-Sim")  # Or customize as you like
  ) +
  scale_x_discrete(labels = region_color_labels) +  # Apply custom coloured labels
  theme(axis.text.x = element_markdown(angle = 90, hjust = 1, size = 12),
    legend.position = "none",
    axis.text.y = element_text(size = 12), 
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),) +  # Rotate x-axis labels
  labs(title = "ModE-Sim (BC) vs ERA5-Land by Region (1950-2009)",
       y = "PCWD Value [mm]",
       x = "Region",
       fill = "Dataset")

```


```{r}
library(ggExtra)
library(MASS)  # For density estimation

# Merge the datasets by Year and Region
scatter_data <- inner_join(ModESim_long, ERA5Land_long, by = c("Year", "Region"), suffix = c("_ModESim", "_ERA5Land"))
#bias corrected: 
scatter_data_bc <- inner_join(ModESim_long_bc, ERA5Land_long, by = c("Year", "Region"), suffix = c("_ModESim_bc", "_ERA5Land"))

# Compute 2D Kernel Density Estimation
dens <- kde2d(scatter_data$Value_ERA5Land, scatter_data$Value_ModESim, n = 100)
dens_bc <- kde2d(scatter_data_bc$Value_ERA5Land, scatter_data_bc$Value_ModESim_bc, n = 100)

# Find the density values for each point in the scatter plot
scatter_data$density <- dens$z[cbind(
  findInterval(scatter_data$Value_ERA5Land, dens$x), 
  findInterval(scatter_data$Value_ModESim, dens$y)
)]

# Find the density values for each point in the scatter plot
scatter_data_bc$density <- dens_bc$z[cbind(
  findInterval(scatter_data_bc$Value_ERA5Land, dens_bc$x), 
  findInterval(scatter_data_bc$Value_ModESim, dens_bc$y)
)]

# Scatter plot with density-based coloring
ggplot(scatter_data, aes(x = Value_ERA5Land, y = Value_ModESim, color = density)) +
  geom_point(size = 1.5) +  # Regular points, colored by density
  geom_abline(slope = 1, intercept = 0, color = "grey42") +
  scale_color_viridis_c(option = "plasma") +  # Heatmap-like color scheme
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  theme_minimal() +
  labs(title = "ERA5Land vs. ModESim (1950-2009)",
       x = expression(PCWD[ERA5Land]),
       y = expression(PCWD[ModE-Sim]),
       color = "Density") +
  theme(legend.position = "right")

# Scatter plot with density-based coloring
ggplot(scatter_data_bc, aes(x = Value_ERA5Land, y = Value_ModESim_bc, color = density)) +
  geom_point(size = 1.5) +  # Regular points, colored by density
  geom_abline(slope = 1, intercept = 0, color = "grey42") +
  scale_color_viridis_c(option = "plasma") +  # Heatmap-like color scheme
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  theme_minimal() +
  labs(title = "ERA5Land vs. ModESim_bc (QM) (1950-2009)",
       x = expression(PCWD[ERA5Land]),
       y = expression(PCWD[ModE-Sim]),
       color = "Density") +
  theme(legend.position = "right")
```


### Absolute values timeseries of bias corrected ModE-Sim and ERA5-Land

```{r}
### raw ModESim regional data - corrected_full = bias corrected 1850-2009
abs_ModESim <- do.call(rbind,lapply(names(corrected_full), function(region) { #merged_results for raw
  df <- as.data.frame(corrected_full[[region]][431:590,]) ## corrected_full for bc
  df$time <- seq(1850,2009)
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "PCWD")
  return(df_long)
}))

# Add a continent column
abs_ModESim$continent <- NA  # Initialize continent column
for (continent in names(continent_mapping)) {
  abs_ModESim$continent[abs_ModESim$region %in% continent_mapping[[continent]]] <- continent
}

abs_ERA5 <- do.call(rbind,lapply(names(regional_results_ERA5Land), function(region) {
  df <- as.data.frame(regional_results_ERA5Land[[region]])
  df$time <- seq(1970,2024)
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region),  values_to = "PCWD")
  return(df_long)
}))

# Add a continent column
abs_ERA5$continent <- NA  # Initialize continent column
for (continent in names(continent_mapping)) {
  abs_ERA5$continent[abs_ERA5$region %in% continent_mapping[[continent]]] <- continent
}

############### plot for regions
# Filter for selected continent
selected_continent <- "Europe"
mod_filtered <- abs_ModESim %>% filter(continent == selected_continent)
era_filtered <- abs_ERA5 %>% filter(continent == selected_continent)

# Plot rolling mean time series
ggplot() +
  # Plot individual ensemble members as fine grey lines
  geom_line(data = mod_filtered, aes(x = time, y = PCWD, group = ensemble_member, 
            color = factor("Ensemble Members")), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean z-score as a bold black line
  geom_line(data = mod_filtered %>% group_by(time, region) %>% summarise(EM = mean(PCWD, na.rm = TRUE)),
            aes(x = time, y = EM, color = factor("ModESim")), linewidth = 1) +
  geom_line(data = era_filtered, aes(x = time, y = PCWD,
            color = factor("ERA5")), linewidth = 1, show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD [mm]",
    title = paste("Absolute PCWD (QM-BC) for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("Ensemble Members" = "#7EA3CC","ModESim" = "#255C99",
               "ERA5" = "#B3001B" ),
    labels = c("Ensemble Members","PCWD (Mean) - ModESim", 
               "PCWD - ERA5Land" )
  ) +
  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))  # Improve legend readability
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 10),
    strip.background = element_rect(fill = "#F5B0CB"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  
  facet_wrap(~region, scales = "free_y")


# Filter for selected continent
selected_continent <- "Asia"
mod_filtered <- abs_ModESim %>% filter(continent == selected_continent)
era_filtered <- abs_ERA5 %>% filter(continent == selected_continent)

# Plot rolling mean time series
ggplot() +
  # Plot individual ensemble members as fine grey lines
  geom_line(data = mod_filtered, aes(x = time, y = PCWD, group = ensemble_member, 
            color = factor("Ensemble Members")), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean z-score as a bold black line
  geom_line(data = mod_filtered %>% group_by(time, region) %>% summarise(EM = mean(PCWD, na.rm = TRUE)),
            aes(x = time, y = EM, color = factor("ModESim")), linewidth = 1) +
  geom_line(data = era_filtered, aes(x = time, y = PCWD,
            color = factor("ERA5")), linewidth = 1, show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD [mm]",
    title = paste("Absolute PCWD (QM-BC) for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("Ensemble Members" = "#7EA3CC","ModESim" = "#255C99", "ERA5" = "#B3001B" ),
    labels = c("Ensemble Members","PCWD (Mean) - ModESim", "PCWD - ERA5Land" )
  ) +
  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))  # Improve legend readability
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 10),
    strip.background = element_rect(fill = "#E5973B"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  
  facet_wrap(~region, scales = "free_y")


# Filter for selected continent
selected_continent <- "Australia"
mod_filtered <- abs_ModESim %>% filter(continent == selected_continent)
era_filtered <- abs_ERA5 %>% filter(continent == selected_continent)

ggplot() +
  # Plot individual ensemble members as fine grey lines
  geom_line(data = mod_filtered, aes(x = time, y = PCWD, group = ensemble_member, 
            color = factor("Ensemble Members")), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean z-score as a bold black line
  geom_line(data = mod_filtered %>% group_by(time, region) %>% summarise(EM = mean(PCWD, na.rm = TRUE)),
            aes(x = time, y = EM, color = factor("ModESim")), linewidth = 1) +
  geom_line(data = era_filtered, aes(x = time, y = PCWD,
            color = factor("ERA5")), linewidth = 1, show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD [mm]",
    title = paste("Absolute PCWD (QM-BC) for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("Ensemble Members" = "#7EA3CC","ModESim" = "#255C99",
               "ERA5" = "#B3001B" ),
    labels = c("Ensemble Members","PCWD (Mean) - ModESim", 
               "PCWD - ERA5Land" )
  ) +
  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))  # Improve legend readability
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 10, color = "white"),
    strip.background = element_rect(fill = "#3C3980"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  
  facet_wrap(~region, scales = "free_y")

# Filter for selected continent
selected_continent <- "Africa"
mod_filtered <- abs_ModESim %>% filter(continent == selected_continent)
era_filtered <- abs_ERA5 %>% filter(continent == selected_continent)

ggplot() +
  # Plot individual ensemble members as fine grey lines
  geom_line(data = mod_filtered, aes(x = time, y = PCWD, group = ensemble_member, 
            color = factor("Ensemble Members")), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean z-score as a bold black line
  geom_line(data = mod_filtered %>% group_by(time, region) %>% summarise(EM = mean(PCWD, na.rm = TRUE)),
            aes(x = time, y = EM, color = factor("ModESim")), linewidth = 1) +
  geom_line(data = era_filtered, aes(x = time, y = PCWD,
            color = factor("ERA5")), linewidth = 1, show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD [mm]",
    title = paste("Absolute PCWD (QM-BC) for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("Ensemble Members" = "#7EA3CC","ModESim" = "#255C99",
               "ERA5" = "#B3001B" ),
    labels = c("Ensemble Members","PCWD (Mean) - ModESim", 
               "PCWD - ERA5Land" )
  ) +
  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))  # Improve legend readability
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 10),
    strip.background = element_rect(fill = "#91C7B1"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  
  facet_wrap(~region, scales = "free_y")

# Filter for selected continent
selected_continent <- "North America"
mod_filtered <- abs_ModESim %>% filter(continent == selected_continent)
era_filtered <- abs_ERA5 %>% filter(continent == selected_continent)

ggplot() +
  # Plot individual ensemble members as fine grey lines
  geom_line(data = mod_filtered, aes(x = time, y = PCWD, group = ensemble_member, 
            color = factor("Ensemble Members")), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean z-score as a bold black line
  geom_line(data = mod_filtered %>% group_by(time, region) %>% summarise(EM = mean(PCWD, na.rm = TRUE)),
            aes(x = time, y = EM, color = factor("ModESim")), linewidth = 1) +
  geom_line(data = era_filtered, aes(x = time, y = PCWD,
            color = factor("ERA5")), linewidth = 1, show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD [mm]",
    title = paste("Absolute PCWD (QM-BC) for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("Ensemble Members" = "#7EA3CC","ModESim" = "#255C99",
               "ERA5" = "#B3001B" ),
    labels = c("Ensemble Members","PCWD (Mean) - ModESim", 
               "PCWD - ERA5Land" )
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 10, color = "white"),
    strip.background = element_rect(fill = "#236A3F"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  
  facet_wrap(~region, scales = "free_y")

# Filter for selected continent
selected_continent <- "South America"
mod_filtered <- abs_ModESim %>% filter(continent == selected_continent)
era_filtered <- abs_ERA5 %>% filter(continent == selected_continent)

ggplot() +
  # Plot individual ensemble members as fine grey lines
  geom_line(data = mod_filtered, aes(x = time, y = PCWD, group = ensemble_member, 
            color = factor("Ensemble Members")), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean z-score as a bold black line
  geom_line(data = mod_filtered %>% group_by(time, region) %>% summarise(EM = mean(PCWD, na.rm = TRUE)),
            aes(x = time, y = EM, color = factor("ModESim")), linewidth = 1) +
  geom_line(data = era_filtered, aes(x = time, y = PCWD,
            color = factor("ERA5")), linewidth = 1, show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD [mm]",
    title = paste("Absolute PCWD (QM-BC) for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("Ensemble Members" = "#7EA3CC","ModESim" = "#255C99",
               "ERA5" = "#B3001B" ),
    labels = c("Ensemble Members","PCWD (Mean) - ModESim", 
               "PCWD - ERA5Land" )
  ) +
  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))  # Improve legend readability
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 10),
    strip.background = element_rect(fill = "#E3D081"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  
  facet_wrap(~region, scales = "free_y")

# Filter for selected continent
selected_continent <- "Central America"
mod_filtered <- abs_ModESim %>% filter(continent == selected_continent)
era_filtered <- abs_ERA5 %>% filter(continent == selected_continent)

ggplot() +
  # Plot individual ensemble members as fine grey lines
  geom_line(data = mod_filtered, aes(x = time, y = PCWD, group = ensemble_member, 
            color = factor("Ensemble Members")), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean z-score as a bold black line
  geom_line(data = mod_filtered %>% group_by(time, region) %>% summarise(EM = mean(PCWD, na.rm = TRUE)),
            aes(x = time, y = EM, color = factor("ModESim")), linewidth = 1) +
  geom_line(data = era_filtered, aes(x = time, y = PCWD,
            color = factor("ERA5")), linewidth = 1, show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD [mm]",
    title = paste("Absolute PCWD (QM-BC) for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("Ensemble Members" = "#7EA3CC","ModESim" = "#255C99",
               "ERA5" = "#B3001B" ),
    labels = c("Ensemble Members","PCWD (Mean) - ModESim", 
               "PCWD - ERA5Land" )
  ) +
  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))  # Improve legend readability
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 10),
    strip.background = element_rect(fill = "#B33951"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  
  facet_wrap(~region, scales = "free_y")
```


## Regional Extreme Value analysis on bias corrected data

Select entire ERA5 data available to calculate 2 and 5 year events for specific regions: 

```{r}
# ##Read in data: 
# ##ERA5Land regional aggregates:
regional_results_ERA5Land <- readRDS("~/cwd_global/data/regionalResults_ERA5Land.RData")
#exclude 1950s-1970s due to potential precipitation problems
regional_results_ERA5Land <- lapply(regional_results_ERA5Land, function(x) x[21:75])
###########------------------------- TODO

WCE_ERA5 <- regional_results_ERA5Land[["WCE"]]

#ModESim data
WCE_MODE <- corrected_full[["WCE"]]

#attach time dimension
time_info_mode <- seq(from = as.numeric(1420), to = as.numeric(2009))
time_info_era <- seq(from = as.numeric(1970), to = as.numeric(2024))


# Convert ModE dataset to long format
WCE_MODE_long <- WCE_MODE %>%
  as.data.frame() %>%
  mutate(time = time_info_mode) %>%
  pivot_longer(cols = -time, names_to = "ensemble", values_to = "PCWD")

# Compute means
WCE_MODE_mean <- WCE_MODE_long %>%
  group_by(time) %>%
  summarise(PCWD = mean(PCWD, na.rm = TRUE), .groups = "drop")

# Convert ERA5 dataset into long format
WCE_ERA5_df <- data.frame(time = time_info_era, PCWD = WCE_ERA5)


#### ---------------------- Copy for MED ---------------------------------
MED_ERA5 <- regional_results_ERA5Land[["MED"]]

#ModESim data
MED_MODE <- corrected_full[["MED"]]

#attach time dimension
time_info_mode <- seq(from = as.numeric(1420), to = as.numeric(2009))
time_info_era <- seq(from = as.numeric(1970), to = as.numeric(2024))


# Convert ModE dataset to long format
MED_MODE_long <- MED_MODE %>%
  as.data.frame() %>%
  mutate(time = time_info_mode) %>%
  pivot_longer(cols = -time, names_to = "ensemble", values_to = "PCWD")

# Compute means
MED_MODE_mean <- MED_MODE_long %>%
  group_by(time) %>%
  summarise(PCWD = mean(PCWD, na.rm = TRUE), .groups = "drop")

# Convert ERA5 dataset into long format
MED_ERA5_df <- data.frame(time = time_info_era, PCWD = MED_ERA5)

##WNA------------------------------
WNA_ERA5 <- regional_results_ERA5Land[["WNA"]]

#ModESim data
WNA_MODE <- corrected_full[["WNA"]]

#attach time dimension
time_info_mode <- seq(from = as.numeric(1420), to = as.numeric(2009))
time_info_era <- seq(from = as.numeric(1970), to = as.numeric(2024))


# Convert ModE dataset to long format
WNA_MODE_long <- WNA_MODE %>%
  as.data.frame() %>%
  mutate(time = time_info_mode) %>%
  pivot_longer(cols = -time, names_to = "ensemble", values_to = "PCWD")

# Compute means
WNA_MODE_mean <- WNA_MODE_long %>%
  group_by(time) %>%
  summarise(PCWD = mean(PCWD, na.rm = TRUE), .groups = "drop")

# Convert ERA5 dataset into long format
WNA_ERA5_df <- data.frame(time = time_info_era, PCWD = WNA_ERA5)

##ARP -----------------------------------------
ARP_ERA5 <- regional_results_ERA5Land[["ARP"]]

#ModESim data
ARP_MODE <- corrected_full[["ARP"]]

#attach time dimension
time_info_mode <- seq(from = as.numeric(1420), to = as.numeric(2009))
time_info_era <- seq(from = as.numeric(1970), to = as.numeric(2024))


# Convert ModE dataset to long format
ARP_MODE_long <- ARP_MODE %>%
  as.data.frame() %>%
  mutate(time = time_info_mode) %>%
  pivot_longer(cols = -time, names_to = "ensemble", values_to = "PCWD")

# Compute means
ARP_MODE_mean <- ARP_MODE_long %>%
  group_by(time) %>%
  summarise(PCWD = mean(PCWD, na.rm = TRUE), .groups = "drop")

# Convert ERA5 dataset into long format
ARP_ERA5_df <- data.frame(time = time_info_era, PCWD = ARP_ERA5)


##CAU-----------------------------------
CAU_ERA5 <- regional_results_ERA5Land[["CAU"]]

#ModESim data
CAU_MODE <- corrected_full[["CAU"]]

#attach time dimension
time_info_mode <- seq(from = as.numeric(1420), to = as.numeric(2009))
time_info_era <- seq(from = as.numeric(1970), to = as.numeric(2024))


# Convert ModE dataset to long format
CAU_MODE_long <- CAU_MODE %>%
  as.data.frame() %>%
  mutate(time = time_info_mode) %>%
  pivot_longer(cols = -time, names_to = "ensemble", values_to = "PCWD")

# Compute means
CAU_MODE_mean <- CAU_MODE_long %>%
  group_by(time) %>%
  summarise(PCWD = mean(PCWD, na.rm = TRUE), .groups = "drop")

# Convert ERA5 dataset into long format
CAU_ERA5_df <- data.frame(time = time_info_era, PCWD = CAU_ERA5)

```

plot raw data all together: 
```{r}
# # Plot
# ggplot() +
#   geom_line(data = WCE_MODE_long, aes(x = time, y = PCWD, group = ensemble, color = factor("EMs_ModE")), linewidth=0.3,alpha = 0.3, show.legend = TRUE) +
#   geom_line(data = WCE_MODE_mean, aes(x = time, y = PCWD, color = factor("Mean_ModE")), linewidth = 0.8) +
#   geom_line(data = WCE_ERA5_df, aes(x = time, y = PCWD, color = factor("ERA5")), linewidth = 0.8) +
#     # Manual color scale for legend
#   scale_color_manual(
#     values = c("EMs_ModE" = "#7EA3CC","Mean_ModE" = "#255C99", "ERA5" = "#B3001B" ),
#     labels = c("EMs - ModESim","Mean - ModESim", "ERA5Land" )
#   ) +
#   labs(title = "Annmax PCWD Time Series for WCE",
#        x = "Year",
#        y = "PCWD [mm]",
#        color = "Dataset") +
#   theme_minimal()+
#   theme(legend.position = "bottom")

### ------------------- MED ---------------------------------
# Plot
ggplot() +
  geom_line(data = MED_MODE_long, aes(x = time, y = PCWD, group = ensemble, color = factor("EMs_ModE")), linewidth=0.3,alpha = 0.3, show.legend = TRUE) +
  geom_line(data = MED_MODE_mean, aes(x = time, y = PCWD, color = factor("Mean_ModE")), linewidth = 0.8) +
  geom_line(data = MED_ERA5_df, aes(x = time, y = PCWD, color = factor("ERA5")), linewidth = 0.8) +
    # Manual color scale for legend
  scale_color_manual(
    values = c("EMs_ModE" = "#7EA3CC","Mean_ModE" = "#255C99", "ERA5" = "#B3001B" ),
    labels = c("EMs - ModE-Sim*","Mean - ModE-Sim*", "ERA5-Land" )
  ) +
  labs(title = "Annmax PCWD Time Series for MED",
       x = "Year",
       y = "PCWD [mm]",
       color = "Dataset") +
  theme_minimal()+
  theme(legend.position = "right",
         legend.title = element_text(face = "bold"))

# ggsave(
#   filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_legend.pdf",
#   device   = cairo_pdf,    # better text rendering
#   width    = 12,           # in inches
#   height   = 6,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )


### ------------------- WNA ---------------------------------
# Plot
ggplot() +
  geom_line(data = WNA_MODE_long, aes(x = time, y = PCWD, group = ensemble, color = factor("EMs_ModE")), linewidth=0.3,alpha = 0.3, show.legend = TRUE) +
  geom_line(data = WNA_MODE_mean, aes(x = time, y = PCWD, color = factor("Mean_ModE")), linewidth = 0.8) +
  geom_line(data = WNA_ERA5_df, aes(x = time, y = PCWD, color = factor("ERA5")), linewidth = 0.8) +
    # Manual color scale for legend
  scale_color_manual(
    values = c("EMs_ModE" = "#7EA3CC","Mean_ModE" = "#255C99", "ERA5" = "#B3001B" ),
    labels = c("EMs - ModE-Sim*","Mean - ModE-Sim*", "ERA5-Land" )
  ) +
  labs(title = "Annmax PCWD Time Series for WNA",
       x = "Year",
       y = "PCWD [mm]",
       color = "Dataset") +
  theme_minimal()+
  theme(legend.position = "bottom")


### ------------------- ARP ---------------------------------
# Plot
ggplot() +
  geom_line(data = ARP_MODE_long, aes(x = time, y = PCWD, group = ensemble, color = factor("EMs_ModE")), linewidth=0.3,alpha = 0.3, show.legend = TRUE) +
  geom_line(data = ARP_MODE_mean, aes(x = time, y = PCWD, color = factor("Mean_ModE")), linewidth = 0.8) +
  geom_line(data = ARP_ERA5_df, aes(x = time, y = PCWD, color = factor("ERA5")), linewidth = 0.8) +
    # Manual color scale for legend
  scale_color_manual(
    values = c("EMs_ModE" = "#7EA3CC","Mean_ModE" = "#255C99", "ERA5" = "#B3001B" ),
    labels = c("EMs - ModE-Sim*","Mean - ModE-Sim*", "ERA5-Land" )
  ) +
  labs(title = "Annmax PCWD Time Series for ARP",
       x = "Year",
       y = "PCWD [mm]",
       color = "Dataset") +
  theme_minimal()+
  theme(legend.position = "bottom")


### ------------------- CAU ---------------------------------
# Plot
ggplot() +
  geom_line(data = CAU_MODE_long, aes(x = time, y = PCWD, group = ensemble, color = factor("EMs_ModE")), linewidth=0.3,alpha = 0.3, show.legend = TRUE) +
  geom_line(data = CAU_MODE_mean, aes(x = time, y = PCWD, color = factor("Mean_ModE")), linewidth = 0.8) +
  geom_line(data = CAU_ERA5_df, aes(x = time, y = PCWD, color = factor("ERA5")), linewidth = 0.8) +
    # Manual color scale for legend
  scale_color_manual(
    values = c("EMs_ModE" = "#7EA3CC","Mean_ModE" = "#255C99", "ERA5" = "#B3001B" ),
    labels = c("EMs - ModE-Sim*","Mean - ModE-Sim*", "ERA5-Land" )
  ) +
  labs(title = "Annmax PCWD Time Series for CAU",
       x = "Year",
       y = "PCWD [mm]",
       color = "Dataset") +
  theme_minimal()+
  theme(legend.position = "bottom")
```

Include rolling 30-year anomalies as thin black line with higlighted events

```{r}
# Initialize list to store rolling 30-year means
rolling_means <- list()

# Define window size
window_size <- 30

# Compute rolling means for each ensemble member
for (region_name in names(corrected_full)) {
  
  # Extract ensemble members for the region
  ensemble_members <- corrected_full[[region_name]][2:590,]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means[[region_name]] <- rolling_mean_matrix
}

rolling_means_long <-  do.call(rbind, lapply(names(rolling_means), function(region) {
  df <- as.data.frame(rolling_means[[region]])
  df$time <- seq(1436, 1995)  # Adjust time range accordingly
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "rolling_means")
  return(df_long)
}))
```

Calculate 600 year rolling anomalies from bias corrected ModE-Sim data with 1420-1450 as reference period
```{r}

# Initialize lists to store baselines
anom_baseline <- list()

# Compute baseline mean from the first 30 values of rolling mean
for (region_name in names(rolling_means)) {
  rolling_mean_matrix <- rolling_means[[region_name]]
  
  anom_baseline[[region_name]] <- mean(rolling_mean_matrix[30:1,], na.rm = TRUE)  ### for reference period of first 30 values: rolling_mean_matrix[1:30,]
}

# Initialize lists to store anomalies
ensemble_30anom <- list()

# Compute anomalies
for (region_name in names(rolling_means)) {
  rolling_mean_matrix <- rolling_means[[region_name]]
  num_years <- nrow(rolling_mean_matrix)
  
  # Initialize matrix to store anomalies
  anoms <- matrix(NA, nrow=num_years, ncol=ncol(rolling_mean_matrix))
  
  # Compute anomalies using rolling mean baseline
  for (t in 1:num_years) {
    anoms[t, ] <- (rolling_mean_matrix[t, ] - anom_baseline[[region_name]])
  }
  
  # Store results
  ensemble_30anom[[region_name]] <- anoms
}

# Compute mean anomalies across ensemble members
mean_30anom <- list()
for (region_name in names(ensemble_30anom)) {
  mean_30anom[[region_name]] <- rowMeans(ensemble_30anom[[region_name]], na.rm = TRUE)
}

# Compute region-specific pooled standard deviation using only the first 30 years
pooled_sd_by_region <- list()
for (region_name in names(ensemble_30anom)) {
  # Extract only the first 30 years of anomalies
  first_30_years <- ensemble_30anom[[region_name]][1:30, ]
  
  # Compute pooled standard deviation over these values
  pooled_sd_by_region[[region_name]] <- sd(unlist(first_30_years), na.rm = TRUE)
}

# Convert ensemble data into a long format for plotting individual members
anoms_long <- do.call(rbind, lapply(names(ensemble_30anom), function(region) {
  df <- as.data.frame(ensemble_30anom[[region]])
  df$time <- seq(1436, 1995)  # Adjust time range accordingly
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "anomalies")
  return(df_long)
}))

# Add a continent column
anoms_long$continent <- NA  # Initialize continent column
for (continent in names(continent_mapping)) {
  anoms_long$continent[anoms_long$region %in% continent_mapping[[continent]]] <- continent
}


```

Anomaly timeseries plot with bias corrected data: 

```{r}
####-----------------------------------------------------------------------
selected_region <- "MED"
anoms_filtered <- anoms_long %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - pooled_sd,
    upper_bound = first_mean_anomaly + pooled_sd
  )

anom_MED <- ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "",
    y = "PCWD anomalies [mm] w.r.t 1420-1450"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15),
    strip.background = element_rect(fill = "#F5B0CB"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11), 
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11)
  ) +
  
  facet_wrap(~region, scales = "free_y")

#######---------------------------------------------------------------------------------
selected_region <- "WNA"
anoms_filtered <- anoms_long %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - pooled_sd,
    upper_bound = first_mean_anomaly + pooled_sd
  )

anom_WNA <- ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "",
    y = ""
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15, color = "white"),
    strip.background = element_rect(fill = "#236A3F"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11), 
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11)
  ) +
  
  facet_wrap(~region, scales = "free_y")

###########------------------------------------------------------------
selected_region <- "ARP"
anoms_filtered <- anoms_long %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - pooled_sd,
    upper_bound = first_mean_anomaly + pooled_sd
  )

anom_ARP <- ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1420-1450"
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15),
    strip.background = element_rect(fill = "#E5973B"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 11)
  ) +
  
  facet_wrap(~region, scales = "free_y")

###----------------------------------------------------------------------------
selected_region <- "CAU"
anoms_filtered <- anoms_long %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - pooled_sd,
    upper_bound = first_mean_anomaly + pooled_sd
  )

anom_CAU <- ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = ""
    #,title = paste("Rolling 30yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15,color = "white"),
    strip.background = element_rect(fill = "#3C3980"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 11)
  ) +
  
  facet_wrap(~region, scales = "free_y")
```

```{r}
###put together in one plot and save
#### combine plots in grid and save
anom_MED2 <- anom_MED + theme(plot.margin = margin(t = 10, r = 10, b = 5, l = 5))
anom_WNA2 <- anom_WNA + theme(plot.margin = margin(t = 10, r = 10, b = 5, l = 5))
anom_ARP2 <- anom_ARP + theme(plot.margin = margin(t = 0, r = 10, b = 5, l = 5))
anom_CAU2 <- anom_CAU + theme(plot.margin = margin(t = 0, r = 10, b = 5, l = 5))


combined_anom <- plot_grid(
  anom_MED2, anom_WNA2,
  anom_ARP2, anom_CAU2,
  ncol = 2,
  nrow = 2,
  rel_widths = c(1, 1),
  align = "h",
  axis = "tblr",
  labels = c(
    "a)",  "b) ", "c)", "d)"
  ),
  label_size = 15,
  label_x = 0.05,
  hjust = 0,
  label_y = 0.97,
  vjust = 1
)

print(combined_anom)

ggsave(
  filename = "~/cwd_global/manuscript/anomalies.pdf",
  plot     = combined_anom,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 6.5,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

# ggsave(
#   filename = "~/cwd_global/manuscript/anomalies_legend.pdf",
#   plot     = anom_CAU,
#   device   = cairo_pdf,    # better text rendering
#   width    = 10,           # in inches
#   height   = 6.5,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )

```



### Assessing the change in frequency of an ERA5 5-year drought events (Moderate drought)

```{r functions for RP calc}
# From https://geco-bern.github.io/cwd/articles/cwd_example.html: 
# calculate return period as a function of the CWD extreme. Using the two 
# coefficients of the fitted distribution as arguments
calc_return_period <- function(x, loc, scale){
  1 / (1 - exp(-exp(-(x-loc)/scale)))
}

extract_loc <- function(mod){
  loc <- mod$results$par[ "location" ]
  if (!is.null(loc)){
    return(loc)
  } else {
    return(NA)
  }
}

extract_scale <- function(mod){
  scale <- mod$results$par[ "scale" ]
  if (!is.null(scale)){
    return(scale)
  } else {
    return(NA)
  }
}
```


ModESim MED annmax PCWD distribution as context for ERA5Land extremes

```{r}
baseline <- MED_ERA5[1:30] ##1970-1999
fit_ref <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_era_ref_5yr <- round(as.numeric(extRemes::return.level(fit_ref, return.period = 5)))

baseline <- MED_ERA5[31:55] ##1970-1999
fit_mod <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_era_mod_5yr <- round(as.numeric(extRemes::return.level(fit_mod, return.period = 5)))

baseline <- as.vector(MED_MODE[531:560,])  #1970-1999
fit_ref <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_ModESim_ref_5yr <- round(as.numeric(extRemes::return.level(fit_ref, return.period = 5)))

baseline <- as.vector(MED_MODE[1:530,])  #1420-1969
fit_full <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_ModESim_full_5yr <- round(as.numeric(extRemes::return.level(fit_full, return.period = 5)))


# 1) extract 1970-1999 of the wide matrix, then pivot
MED_last60_long <- MED_MODE[(nrow(MED_MODE)-40):(nrow(MED_MODE)-10), ] %>%
  as.data.frame() %>%
  pivot_longer(cols = everything(),
               names_to  = "ensemble",
               values_to = "PCWD")

# Convert ModE dataset to long format 1420 - 1969
MED_MODE_long_sel <- MED_MODE %>%
  as.data.frame() %>%
  mutate(time = time_info_mode) %>%
  filter(time < 1970) %>%
  pivot_longer(cols = -time, names_to = "ensemble", values_to = "PCWD")

# 1) extract last 60 rows of the wide matrix, then pivot
MED_ERA5_long <- MED_ERA5[1:30] %>%
  as.data.frame() %>%
  pivot_longer(cols = everything(),
              names_to  = "year",
               values_to = "PCWD")

MED_ERA5_last30 <- MED_ERA5[31:55] %>%
  as.data.frame() %>%
  pivot_longer(cols = everything(),
              names_to  = "year",
               values_to = "PCWD")

ggplot() +
  # full‑data density
  geom_density(
    data    = MED_MODE_long_sel,
    aes(x   = PCWD,
    color   = factor("ModESim_fill_full"),
    fill    = factor("ModESim_fill_full")),
    alpha   = 0.8,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
  # last‑60 density
  geom_density(
    data    = MED_last60_long,
    aes(x   = PCWD,
    color   = factor("ModESim_fill"),
    fill    = factor("ModESim_fill")),
    alpha   = 0.8,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
    # first‑60 density
  geom_density(
    data    = MED_ERA5_long,
    aes(x   = PCWD,
    color   = factor("ERA5_fill"),
    fill    = factor("ERA5_fill")),
    alpha   = 0.6,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
  geom_density(
    data    = MED_ERA5_last30,
    aes(x   = PCWD,
    color   = factor("ERA5_fill_last30"),
    fill    = factor("ERA5_fill_last30")),
    alpha   = 0.6,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
  geom_vline(aes(xintercept = ModESim_full_5yr, color = factor("ModESim_full")), linetype = "dashed", linewidth=0.9) +
  geom_vline(aes(xintercept = ModESim_ref_5yr, color = factor("ModESim_ref")), linetype = "dashed", linewidth=0.9) +
  geom_vline(aes(xintercept = era_ref_5yr, color = factor("ERA_ref")), linetype = "dashed", linewidth=0.9) +
  geom_vline(aes(xintercept = era_mod_5yr, color = factor("ERA_mod")), linetype = "dashed", linewidth=0.9) +
    labs(
    x = "PCWD [mm]",
    y = "Density",
    title = "MED Extreme Value Distribution"
  ) +
  theme_minimal() +
 scale_fill_manual(
    name = "Data Source",
    values = c("ModESim_fill_full" = "#255C99","ModESim_fill" = "#91C499", "ERA5_fill" = "#A25E8E", "ERA5_fill_last30" = "#B3001B") ,
    labels = c("ModE-Sim* (1420-1969)", "ModE-Sim* (1970-1999)", "ERA5-Land (1970-1999)", "ERA5-Land (2000-2024)")
  )+
  scale_color_manual(
    name = "5-year Event",
    values = c("ModESim_full" = "#17385E", "ModESim_ref" = "#628567", "ERA_ref" = "#E8B5CC", "ERA_mod" = "#870014"),
    labels = c("ModE-Sim* (1420 - 1969)", "ModE-Sim* (1970 - 1999)", "ERA5Land (1970 - 1999)", "ERA5Land (2000 - 2024)")
    #guide = guide_legend(override.aes = list(fill= NA,shape = NA, color = "white"))
  )+

    theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.key = element_rect(fill   = NA,    # transparent fill
                              colour = NA),# no border
    legend.title = element_text(face = "bold")
  )


ggsave(
  filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_legend.pdf",
  device   = cairo_pdf,    # better text rendering
  width    = 12,           # in inches
  height   = 6,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

```



Function that does the plotting for all regions: 

```{r}
# ------------------------------------------------------------------------------
# Helper: prepare the four slices + FEVD-based 5‑yr return levels
# ------------------------------------------------------------------------------
prepare_region_data <- function(region,
                                corrected_full,
                                regional_results_ERA5Land,
                                mode_years = 1420:2009,
                                era5_years = 1970:2024,
                                drop_initial_years = 2) {
  MODE_mat <- corrected_full[[region]]
  ERA5_vec <- regional_results_ERA5Land[[region]]
  
  # Time‐series data
  MODE_long <- MODE_mat %>%
    as.data.frame() %>%
    mutate(time = mode_years) %>%
    pivot_longer(-time, names_to = "ensemble", values_to = "PCWD")
  
  MODE_mean <- MODE_long %>% group_by(time) %>% summarise(PCWD = mean(PCWD, na.rm=TRUE), .groups="drop")
  
  ERA5_df   <- data.frame(time = era5_years, PCWD = ERA5_vec)
  
  # ———————— drop the first `drop_initial_years` rows ————————
  MODE_long <- MODE_long %>% 
    filter(time >= mode_years[drop_initial_years])
  
  MODE_mean <- MODE_mean %>% 
    filter(time >= mode_years[drop_initial_years])
  # ——————————————————————————————————————————————————————————————
  
  # Four density slices
  MODE_full <- MODE_long %>% filter(time < 1970)
  MODE_ref  <- MODE_long %>% filter(time >= 1970 & time <= 1999)
  ERA5_ref  <- ERA5_df   %>% filter(time >= 1970 & time <= 1999)
  ERA5_mod  <- ERA5_df   %>% filter(time >= 2000)
  
  # FEVD + 5‑yr return level
  fit_rl <- function(x) {
    fit <- fevd(x, type="Gumbel", method="MLE", units="years")
    round(as.numeric(return.level(fit, return.period = 5)))
  }
  rl <- list(
    MODE_full    = fit_rl(MODE_full$PCWD),
    MODE_ref     = fit_rl(MODE_ref$PCWD),
    ERA5_ref     = fit_rl(ERA5_ref$PCWD),
    ERA5_mod     = fit_rl(ERA5_mod$PCWD)
  )
  
  list(
    MODE_long   = MODE_long,
    MODE_mean   = MODE_mean,
    ERA5_df     = ERA5_df,
    MODE_full   = MODE_full,
    MODE_ref    = MODE_ref,
    ERA5_ref    = ERA5_ref,
    ERA5_mod    = ERA5_mod,
    rl          = rl
  )
}

# ------------------------------------------------------------------------------
# Main plot function: with manual fill/color scales + legends restored
# ------------------------------------------------------------------------------
plot_region_PCWD <- function(region_code) {
  
  dat <- prepare_region_data(region_code,
                             corrected_full,
                             regional_results_ERA5Land)
  
  # 2) Compute anomalies + bounds + exceedance runs
  anoms_filtered <- anoms_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      mean_anomaly = mean(anomalies, na.rm=TRUE),
      pooled_sd    = pooled_sd_by_region[[region_code]],
      .groups = "drop"
    )
  
    # first-year mean
  first_mean <- anoms_filtered %>%
    arrange(time) %>%
    slice(1) %>%
    pull(mean_anomaly)
  
  anoms_filtered <- anoms_filtered %>%
    mutate(
      lower_bound = first_mean - pooled_sd,
      upper_bound = first_mean + pooled_sd,
      exceeds     = mean_anomaly < lower_bound | mean_anomaly > upper_bound,
      run_id      = data.table::rleid(exceeds)
    )
  
  exceed_periods <- anoms_filtered %>%
    filter(exceeds) %>%
    group_by(run_id) %>%
    summarise(
      start = min(time),
      end   = max(time),
      .groups="drop"
    )
  
    # 3) Filter running means for specified region
  rolling_filtered <- rolling_means_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      mean_rolling = mean(rolling_means, na.rm=TRUE),
      .groups = "drop"
    )
  
  # shared y‐limits & breaks
  y_limits     <- range(c(dat$MODE_long$PCWD+50, dat$ERA5_df$PCWD+50), na.rm=TRUE) #adjust depending on region
  common_breaks <- pretty(y_limits, n=6)
  
  # 1) start with an empty ggplot
  p_ts <- ggplot()
  
  # 2) add the grey bands first (if any)
  if (nrow(exceed_periods) > 0) {
    p_ts <- p_ts + 
      geom_rect(
        data = exceed_periods,
        aes(xmin = start, xmax = end, 
            ymin = y_limits[1], ymax = y_limits[2]),
        inherit.aes = FALSE,
        fill = "grey60", alpha = 0.5
      )
  }
  
  # 3) now add your PCWD & rolling‐mean lines
  p_ts <- p_ts +
    geom_line(data = dat$MODE_long, aes(time, PCWD, group=ensemble, color = factor("EMs_ModE")), linewidth=0.3,alpha = 0.3, show.legend = TRUE) +
    geom_line(data = dat$MODE_mean, aes(time, PCWD, color = factor("Mean_ModE")), linewidth = 0.8) +
    geom_line(data = rolling_filtered, aes(time, mean_rolling, color="rolling_mean"), linewidth=0.7) +   ###needs to be rolling means!
    geom_line(data = dat$ERA5_df, aes(time, PCWD, color = factor("ERA5")), linewidth = 0.8) +
    scale_y_continuous(limits=y_limits, breaks=common_breaks) +
        # Manual color scale for legend
  scale_color_manual(
    values = c("EMs_ModE"  = "#7EA3CC",
               "Mean_ModE" = "#255C99", "rolling_mean"="black", 
               "ERA5"      = "#B3001B"),
    labels = c( "EMs - ModE-Sim*", "Mean - ModE-Sim*","30-yr rolling\nmean - ModE-Sim*", "ERA5-Land")
  )+
  labs(title = "",
       x = "Year", y = "PCWD [mm]", color = "Dataset") +  #change labelling depending on position in multipanel
    theme_minimal(base_size=14) +
    theme(axis.title.y = element_text(margin=margin(r=10)),
          legend.position="none", legend.title = element_text(face="bold"))

  # Density + legend
  p_den <- ggplot() +
    geom_density(data = dat$MODE_full,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ModESim_full", color="ModESim_full"),
                 alpha=0.8, size=0.6) +
    geom_density(data = dat$MODE_ref,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ModESim_ref", color="ModESim_ref"),
                 alpha=0.8, size=0.6) +
    geom_density(data = dat$ERA5_ref,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ERA5_70_99", color="ERA5_70_99"),
                 alpha=0.6, size=0.6) +
    geom_density(data = dat$ERA5_mod,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ERA5_00_24", color="ERA5_00_24"),
                 alpha=0.6, size=0.6) +
    
    geom_hline(aes(yintercept=dat$rl$MODE_full,  color="ModESim_full_lin"),
               linetype="dashed", size=0.8) +
    geom_hline(aes(yintercept=dat$rl$MODE_ref,   color="ModESim_ref_lin"),
               linetype="dashed", size=0.8) +
    geom_hline(aes(yintercept=dat$rl$ERA5_ref,   color="ERA5_70_99_lin"),
               linetype="dashed", size=0.8) +
    geom_hline(aes(yintercept=dat$rl$ERA5_mod,   color="ERA5_00_24_lin"),
               linetype="dashed", size=0.8) +
    
    scale_y_continuous(limits=y_limits, breaks=common_breaks) +
    scale_fill_manual(
      name   = "",
      values = c(
        ModESim_full  = "#255C99",
        ModESim_ref   = "#91C499",
        ERA5_70_99    = "#A25E8E",
        ERA5_00_24    = "#B3001B"
      ),
      labels = c(
        "ModE-Sim* (1420-1969)",
        "ModE-Sim* (1970-1999)",
        "ERA5-Land (1970-1999)",
        "ERA5-Land (2000-2024)"
      )
    ) +
    scale_color_manual(
      name   = "5-year Event",
      values = c(
        ModESim_full_lin  = "#17385E",
        ModESim_ref_lin   = "#628567",
        ERA5_70_99_lin    = "#E8B5CC",
        ERA5_00_24_lin    = "#870014",
        ModESim_full      = "#255C99",
        ModESim_ref       = "#91C499",
        ERA5_70_99        = "#A25E8E",
        ERA5_00_24        = "#B3001B"
      ),
      labels = c(
        "ModE-Sim* (1420-1969)",
        "ModE-Sim* (1970-1999)",
        "ERA5-Land (1970-1999)",
        "ERA5-Land (2000-2024)"
      )
    ) +
    labs(x="Density", y=NULL) +  #change labelling depending on position in multipanel
    theme_minimal(base_size=14) +
    theme(
      plot.title     = element_text(hjust=0.5, face="bold"),
      legend.key     = element_rect(fill=NA, colour=NA),
      legend.title   = element_text(face="bold"),
      text           = element_text(size=14),
      axis.title.y   = element_text(margin=margin(r=10)),
      legend.position = "none"
    )
  
  # Combine side by side
  cowplot::plot_grid(
    p_ts, p_den,
    ncol       = 2,
    rel_widths = c(2.5, 1.),
    align      = "h",
    axis       = "tb",
    labels     = "", 
    # Push labels all the way to the left margin of each panel
    label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
    hjust   = 0,      # left‑justify the label text
    # Tweak vertical positioning if you like
    label_y = 0.98,   # just below the top
    vjust   = 1       # align text at its top
  ) 
}

# Example:
plot_MED <- plot_region_PCWD("MED")
plot_ARP <- plot_region_PCWD("ARP")
plot_WNA <- plot_region_PCWD("WNA")
plot_CAU <- plot_region_PCWD("CAU")

# ##### save legends
# p_ts
# ggsave(
#   filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_legend.pdf",
#   plot     = p_ts,
#   device   = cairo_pdf,    # better text rendering
#   width    = 10,           # in inches
#   height   = 13,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )
# 
# p_den
# ggsave(
#   filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_legend.pdf",
#   plot     = p_den,
#   device   = cairo_pdf,    # better text rendering
#   width    = 10,           # in inches
#   height   = 13,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )

```

Combine multiple panels into one: 

```{r}
##for Fig. 1 in paper: 

combined2all <- plot_grid(
  plot_MED, plot_WNA, plot_ARP, plot_CAU,
  nrow        = 4,
  ncol        = 1,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) MED – Mediterranean", 
    "b) WNA – Western North America", 
    "c) ARP – Arabian Peninsula", 
    "d) CAU – Central Australia" 
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1       # align text at its top
)

print(combined2all)

ggsave(
  filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_all_Fig.pdf",
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```

Plot by continent for supplementary material: 

```{r}

##EUROPE
plot_NEU <- plot_region_PCWD("NEU")
plot_WCE <- plot_region_PCWD("WCE")
plot_EEU <- plot_region_PCWD("EEU")
plot_MED <- plot_region_PCWD("MED")


combined2all <- plot_grid(
  plot_NEU, plot_WCE, plot_EEU, plot_MED,
  nrow        = 4,
  ncol        = 1,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) EUROPE - NEU", 
    "b) EUROPE - WCE", 
    "c) EUROPE - EEU", 
    "d) EUROPE - MED" 
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1       # align text at its top
)

print(combined2all)

ggsave(
  filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_all_Europe.pdf",
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

###NORTH AMERICA
plot_NWN <- plot_region_PCWD("NWN")
plot_NEN <- plot_region_PCWD("NEN")
plot_WNA <- plot_region_PCWD("WNA")
plot_CNA <- plot_region_PCWD("CNA")
plot_ENA <- plot_region_PCWD("ENA")
plot_GIC <- plot_region_PCWD("GIC")


combined2all <- plot_grid(
  plot_NWN, plot_NEN, plot_WNA, plot_CNA,plot_ENA, plot_GIC,
  nrow        = 6,
  ncol        = 1,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) NORTH AMERICA - NWN", 
    "b) NORTH AMERICA - NEN", 
    "c) NORTH AMERICA - WNA", 
    "d) NORTH AMERICA - CNA",
    "e) NORTH AMERICA - ENA",
    "f) NORTH AMERICA - GIC"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1       # align text at its top
)

print(combined2all)

ggsave(
  filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_all_NorthAmerica.pdf",
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


#### AFRICA
plot_SAH <- plot_region_PCWD("SAH")
plot_WAF <- plot_region_PCWD("WAF")
plot_CAF <- plot_region_PCWD("CAF")
plot_NEAF <- plot_region_PCWD("NEAF")
plot_SEAF <- plot_region_PCWD("SEAF")
plot_WSAF <- plot_region_PCWD("WSAF")
plot_ESAF <- plot_region_PCWD("ESAF")
plot_MDG <- plot_region_PCWD("MDG")



combined2all <- plot_grid(
  plot_SAH, plot_WAF, plot_CAF, plot_NEAF,plot_SEAF, plot_WSAF,plot_ESAF, plot_MDG,
  nrow        = 8,
  ncol        = 1,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) AFRICA - SAH", 
    "b) AFRICA - WAF", 
    "c) AFRICA - CAF", 
    "d) AFRICA - NEAF",
    "e) AFRICA - SEAF",
    "f) AFRICA - WSAF", 
    "g) AFRICA - ESAF",
    "h) AFRICA - MDG"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1       # align text at its top
)

print(combined2all)

ggsave(
  filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_all_Africa.pdf",
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 17,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


#### Asia
plot_WCA <- plot_region_PCWD("WCA")
plot_ARP <- plot_region_PCWD("ARP")
plot_WSB <- plot_region_PCWD("WSB")
plot_RAR <- plot_region_PCWD("RAR")
plot_ESB <- plot_region_PCWD("ESB")


combined2all <- plot_grid(
  plot_WCA, plot_ARP, plot_WSB, plot_RAR,plot_ESB,
  nrow        = 5,
  ncol        = 1,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) ASIA - WCA", 
    "b) ASIA - ARP", 
    "c) ASIA - WSB", 
    "d) ASIA - RAR",
    "e) ASIA - ESB"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1       # align text at its top
)

print(combined2all)

ggsave(
  filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_all_Asia1.pdf",
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


###ASIA continued
plot_ECA <- plot_region_PCWD("ECA")
plot_TIB <- plot_region_PCWD("TIB")
plot_SAS <- plot_region_PCWD("SAS")
plot_EAS <- plot_region_PCWD("EAS")
plot_RFE <- plot_region_PCWD("RFE")
plot_SEA <- plot_region_PCWD("SEA")

combined2all <- plot_grid(
  plot_ECA, plot_TIB, plot_SAS, plot_EAS,plot_RFE,plot_SEA,
  nrow        = 6,
  ncol        = 1,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) ASIA - ECA", 
    "b) ASIA - TIB", 
    "c) ASIA - SAS", 
    "d) ASIA - EAS",
    "e) ASIA - RFE", 
    "f) ASIA - SEA"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1       # align text at its top
)

print(combined2all)

ggsave(
  filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_all_Asia2.pdf",
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

####CENTRAL AMERICA
plot_NCA <- plot_region_PCWD("NCA")
plot_SCA <- plot_region_PCWD("SCA")
plot_CAR <- plot_region_PCWD("CAR")

combined2all <- plot_grid(
  plot_NCA, plot_SCA, plot_CAR,
  nrow        = 3,
  ncol        = 1,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) CENTRAL AMERICA - NCA", 
    "b) CENTRAL AMERICA - SCA", 
    "c) CENTRAL AMERICA - CAR"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1       # align text at its top
)

print(combined2all)

ggsave(
  filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_all_CentralAmerica.pdf",
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 11,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


###SOUTH AMERICA
plot_NWS <- plot_region_PCWD("NWS")
plot_NSA <- plot_region_PCWD("NSA")
plot_NES <- plot_region_PCWD("NES")
plot_SAM <- plot_region_PCWD("SAM")
plot_SWS <- plot_region_PCWD("SWS")
plot_SES <- plot_region_PCWD("SES")
plot_SSA <- plot_region_PCWD("SSA")


combined2all <- plot_grid(
  plot_NWS, plot_NSA, plot_NES,plot_SAM, plot_SWS, plot_SES, plot_SSA,
  nrow        = 7,
  ncol        = 1,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) SOUTH AMERICA - NWS", 
    "b) SOUTH AMERICA - NSA", 
    "c) SOUTH AMERICA - NES",
    "d) SOUTH AMERICA - SAM",
    "e) SOUTH AMERICA - SWS",
    "f) SOUTH AMERICA - SES",
    "g) SOUTH AMERICA - SSA"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1       # align text at its top
)

print(combined2all)

ggsave(
  filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_all_SouthAmerica.pdf",
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 17,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


###AUSTRALIA
plot_NAU <- plot_region_PCWD("NAU")
plot_CAU <- plot_region_PCWD("CAU")
plot_EAU <- plot_region_PCWD("EAU")
plot_SAU <- plot_region_PCWD("SAU")
plot_NZ <- plot_region_PCWD("NZ")


combined2all <- plot_grid(
  plot_NAU, plot_CAU, plot_EAU, plot_SAU,plot_NZ,
  nrow        = 5,
  ncol        = 1,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) AUSTRALIA - NAU", 
    "b) AUSTRALIA - CAU", 
    "c) AUSTRALIA - EAU", 
    "d) AUSTRALIA - SAU",
    "e) AUSTRALIA - NZ"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1       # align text at its top
)

print(combined2all)

ggsave(
  filename = "~/cwd_global/manuscript/Annmax_PCWD_Timeseries_Density_all_Australia.pdf",
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

```



#### rolling effective RP for all regions
calculate 5 yr event return periods
```{r}
#################### Calc for ModE-SIm ######################################
use_rp <- 5    # return period of interest
nyears <- 30   # window size

path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim"
folders <- list.files(path, full.names = TRUE, pattern = "m[0-9]{3}_tidy") #m001 - m020 in first set

# Extract unique ensemble member identifiers from folder names
n_ens <- unique(basename(folders))[1:20]

# Store output
EVA_regional_ensemble <- list()

for (region in names(corrected_full)) {
  result_array_region_BL <- corrected_full[[region]][551:580,]  # [time, ensemble]
  result_array_region <- corrected_full[[region]]                         # [590, 20]
  
  # ----- Fit pooled baseline -----
  baseline_pooled <- as.vector(result_array_region_BL)
  fit_ref <- extRemes::fevd(baseline_pooled, type = "Gumbel", method = "MLE", units = "years")
  cwd_ref <- extRemes::return.level(fit_ref, return.period = use_rp)
  
    # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    n_ens <- n_ens[n_ens != "m001_tidy"]  # Skip m001
  } else {
    n_ens <- n_ens
  }

  region_df_list <- list()

  for (ens_member in n_ens) {
    # Initialize return period time series
    rp_window <- rep(NA, nrow(result_array_region) - nyears + 1)

    for (window_start in 1:(nrow(result_array_region) - nyears + 1)) {
      # Extract current 30-year window for this ensemble member
      window_data <- result_array_region[window_start:(window_start + nyears - 1), ens_member]

      # Fit Gumbel distribution
      fit_window <- extRemes::fevd(window_data, type = "Gumbel")

      # Estimate return period of the baseline event in this new fit
      rp_window[window_start] <- calc_return_period(cwd_ref, extract_loc(fit_window), extract_scale(fit_window))
    }

    # Time vector (adjust if needed)
    time_info <- seq(from = as.numeric(1420), to = as.numeric(2009))
    ens_df <- data.frame(
      time = time_info[15:575],
      RP = rp_window,
      region = region,
      member = ens_member
    )

    region_df_list[[ens_member]] <- ens_df
  }

  # Combine all members for this region
  EVA_regional_ensemble[[region]] <- do.call(rbind, region_df_list)
}

# Combine all regions
combined_results_5yr_ens <- do.call(rbind, EVA_regional_ensemble)

# Save to file
write.csv(combined_results_5yr_ens, "~/cwd_global/data/5yr_return_periods_regions_ERA_members.csv", row.names = FALSE) #ModESim all ensemble members

################ now for ERA5 ###############################
######################### --------------------- 5 year events -------------------------------------------------------------
use_rp <- 5    # return period of two years
nyears <- 30   # bin width

# Initialize a list to store results for all regions
EVA_regional <- list()

# Loop through each region in regional_results
for (region in names(regional_results_ERA5Land)) { #swap european_regions for regional_results for all regions
  # Extract the result array for the current region
  result_array_region_BL <- corrected_full[[region]][551:580,]  # [time, ensemble]
  result_array_region <- regional_results_ERA5Land[[region]][20:75] #exclude 60s                         # [590, 20]
  
  # ----- Fit pooled baseline -----
  baseline_pooled <- as.vector(result_array_region_BL)
  fit_ref <- extRemes::fevd(baseline_pooled, type = "Gumbel", method = "MLE", units = "years")
  cwd_ref <- extRemes::return.level(fit_ref, return.period = use_rp)

  # initialise
  rp_window <- rep(NA, length(result_array_region) - nyears + 1)

  for (window_start in 1:(length(result_array_region) - nyears + 1)){

    # Extract data for the current window
    window_data <- result_array_region[window_start:(window_start + nyears - 1)]
  
    # Flatten the matrix into a single vector
    window_data <- as.vector(window_data)
  
    # Fit Gumbel distribution to the window data
    fit_window <- fevd(window_data, type = "Gumbel")

	  # determine return period of cwd_ref based on new fit (How much more/less probably is an event that corresponded to an X-yearly event in the reference period?)
	  rp_window[window_start] <- calc_return_period(cwd_ref, extract_loc(fit_window), extract_scale(fit_window))

  }
  # Create a dataframe for the current region
  time_info <- seq(from = as.numeric(1970), to = as.numeric(2024))  # Replace with actual time if available
  region_df <- data.frame(
    time = time_info[15:41],
    RP = rp_window,
    region = region
  )
  
  # Store the region's dataframe in the results list
  EVA_regional[[region]] <- region_df
}
  
# Combine all region results into a single dataframe
combined_results_5yr_ERA <- do.call(rbind, EVA_regional)

# Save or process the combined dataframe
# Example: Save to a CSV file
write.csv(combined_results_5yr_ERA, "~/cwd_global/data/5yr_return_periods_regions_ERA5Land.csv", row.names = FALSE)

```

### Frequency changes of 5-yr return period events for all IPCC regions (1420 - 2009 epoch):

```{r timeseries and map plots for each region, echo=FALSE, warning=FALSE, message=FALSE}
#plot all ensemble members and ensemble mean: 
combined_df <- read.csv("~/cwd_global/data/5yr_return_periods_regions_ERA_members.csv") #ModESim all ensemble members
combined_era <- read.csv("~/cwd_global/data/5yr_return_periods_regions_ERA5Land.csv")
#combined_df$window_size <- "30yr"

####################################### EUROPE #####################################
##### Step 2: Filter European regions for the map and timeseries
# Filter regions for Europe
selected_continent <- "Europe"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "MED"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# Compute ensemble mean
ensemble_mean_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")

ensemble_band_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(
    RP_mean = mean(RP, na.rm = TRUE),
    RP_lower = quantile(RP, 0.05, na.rm = TRUE),
    RP_upper = quantile(RP, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  # Uncertainty band with legend
  geom_ribbon(data = ensemble_band_df, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim* Ensemble Range (5th-95th)"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = ensemble_band_df, aes(x = time, y = RP_mean, color = "ModE-Sim* Mean (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
  # Labels and theme
  labs(
    x = "Year",
    y = "Effective Return Period",
    title = "Frequency Change of 5-Year Events",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim* Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("ModE-Sim* Ensemble Range (5th-95th)" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 18),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18), 
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    strip.background = element_rect(fill = "#F5B0CB")
  ) +
  facet_wrap(~region, scales = "free_y")


######################################ASIA#######################################
##### Step 2: Filter Asian regions for the map and timeseries
# Filter regions for Asian
selected_continent <- "Asia"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "ARP"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# Compute ensemble mean
ensemble_mean_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")

ensemble_band_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(
    RP_mean = mean(RP, na.rm = TRUE),
    RP_lower = quantile(RP, 0.05, na.rm = TRUE),
    RP_upper = quantile(RP, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  # Uncertainty band with legend
  geom_ribbon(data = ensemble_band_df, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = ensemble_band_df, aes(x = time, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
  # Labels and theme
  labs(
    x = "Year",
    y = "Effective Return Period",
    title = "Frequency Change of 5-Year Events",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 18),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18), 
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    strip.background = element_rect(fill = "#E5973B"),
  ) +
  facet_wrap(~region, scales = "free_y")


###################################### Australia #######################################
##### Step 2: Filter Australian regions for the map and timeseries
# Filter regions for Australia
selected_continent <- "Australia"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "CAU"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# Compute ensemble mean
ensemble_mean_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")

ensemble_band_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(
    RP_mean = mean(RP, na.rm = TRUE),
    RP_lower = quantile(RP, 0.05, na.rm = TRUE),
    RP_upper = quantile(RP, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  # Uncertainty band with legend
  geom_ribbon(data = ensemble_band_df, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = ensemble_band_df, aes(x = time, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
  # Labels and theme
  labs(
    x = "Year",
    y = "Effective Return Period",
    title = "Frequency Change of 5-Year Events",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14, color = "white"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18), 
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    strip.background = element_rect(fill = "#3C3980")
  ) +
  facet_wrap(~region, scales = "free_y")

# ###################################### AFRICA #######################################
# ##### Step 2: Filter African regions for the map and timeseries
# # Filter regions for Africa
# selected_continent <- "Africa"
# sel_regions <- continent_mapping[[selected_continent]]
# #sel_regions <- "NEAF"
# 
# # Filter combined data for timeseries 
# combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
# combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.05, na.rm = TRUE),
#     RP_upper = quantile(RP, 0.95, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# ggplot() +
#   # Uncertainty band with legend
#   geom_ribbon(data = ensemble_band_df, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
#   # Ensemble mean line with legend
#   geom_line(data = ensemble_band_df, aes(x = time, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
#   # ERA5 line with legend
#   geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
#   # Reference line (no legend)
#   geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
#   # Labels and theme
#   labs(
#     x = "Year",
#     y = "Effective Return Period",
#     title = "Frequency Change of 5-Year Events",
#     color = "Line",  # Legend titles
#     fill = "Shading"
#   ) +
#   scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
#   scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
#   theme_classic() +
#   theme(
#     legend.position = "none",
#     plot.title = element_text(hjust = 0.5, face = "bold"),
#     strip.text = element_text(size = 18),
#     axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
#     axis.text.y = element_text(size = 18), 
#     axis.title.x = element_text(size = 18),
#     axis.title.y = element_text(size = 18),
#     legend.title = element_text(size = 14),
#     legend.text = element_text(size = 12),
#     strip.background = element_rect(fill = "#91C7B1"),
#   ) +
#   facet_wrap(~region, scales = "free_y")


###################################### N AMERICA #######################################
##### Step 2: Filter N AMERICA regions for the map and timeseries
# Filter regions for N America
selected_continent <- "North America"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "WNA"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# Compute ensemble mean
ensemble_mean_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")

ensemble_band_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(
    RP_mean = mean(RP, na.rm = TRUE),
    RP_lower = quantile(RP, 0.05, na.rm = TRUE),
    RP_upper = quantile(RP, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

ggplot() +
  # Uncertainty band with legend
  geom_ribbon(data = ensemble_band_df, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = ensemble_band_df, aes(x = time, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
  # Labels and theme
  labs(
    x = "Year",
    y = "Effective Return Period",
    title = "Frequency Change of 5-Year Events",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14, color= "white"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18), 
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    strip.background = element_rect(fill = "#236A3F"),
  ) +
  facet_wrap(~region, scales = "free_y")

# ###################################### C AMERICA #######################################
# ##### Step 2: Filter C AMERICA regions for the map and timeseries
# # Filter regions for C America
# selected_continent <- "Central America"
# sel_regions <- continent_mapping[[selected_continent]]
# #sel_regions <- "NCA"
# 
# # Filter combined data for timeseries 
# combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
# combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.05, na.rm = TRUE),
#     RP_upper = quantile(RP, 0.95, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# ggplot() +
#   # Uncertainty band with legend
#   geom_ribbon(data = ensemble_band_df, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
#   # Ensemble mean line with legend
#   geom_line(data = ensemble_band_df, aes(x = time, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
#   # ERA5 line with legend
#   geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
#   # Reference line (no legend)
#   geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
#   # Labels and theme
#   labs(
#     x = "Year",
#     y = "Effective Return Period",
#     title = "Frequency Change of 5-Year Events",
#     color = "Line",  # Legend titles
#     fill = "Shading"
#   ) +
#   scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
#   scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
#   theme_classic() +
#   theme(
#     legend.position = "none",
#     plot.title = element_text(hjust = 0.5, face = "bold"),
#     strip.text = element_text(size = 18),
#     axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
#     axis.text.y = element_text(size = 18), 
#     axis.title.x = element_text(size = 18),
#     axis.title.y = element_text(size = 18),
#     legend.title = element_text(size = 14),
#     legend.text = element_text(size = 12),
#     strip.background = element_rect(fill = "#B33951")
#   ) +
#   facet_wrap(~region, scales = "free_y")
# 
# 



# ###################################### S AMERICA #######################################
# ##### Step 2: Filter S AMERICA regions for the map and timeseries
# # Filter regions for S America
# selected_continent <- "South America"
# sel_regions <- continent_mapping[[selected_continent]]
# #sel_regions <- "NWS"
# 
# # Filter combined data for timeseries 
# combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
# combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.05, na.rm = TRUE),
#     RP_upper = quantile(RP, 0.95, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# ggplot() +
#   # Uncertainty band with legend
#   geom_ribbon(data = ensemble_band_df, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
#   # Ensemble mean line with legend
#   geom_line(data = ensemble_band_df, aes(x = time, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
#   # ERA5 line with legend
#   geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
#   # Reference line (no legend)
#   geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
#   # Labels and theme
#   labs(
#     x = "Year",
#     y = "Effective Return Period (log10)",
#     title = "Frequency Change of 5-Year Events",
#     color = "Line",  # Legend titles
#     fill = "Shading"
#   ) +
#   scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
#   scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
#   theme_classic() +
#   theme(
#     legend.position = "none",
#     plot.title = element_text(hjust = 0.5, face = "bold"),
#     strip.text = element_text(size = 18),
#     axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
#     axis.text.y = element_text(size = 18), 
#     axis.title.x = element_text(size = 18),
#     axis.title.y = element_text(size = 18),
#     legend.title = element_text(size = 14),
#     legend.text = element_text(size = 12),
#     strip.background = element_rect(fill = "#E3D081")
#   ) +
#   facet_wrap(~region, scales = "free_y")


```

time warped plots: 

```{r}
# Warp function: compress early years, stretch recent years
warp_time <- function(year) {
  ifelse(year < 1950, year * 0.5, 975 + (year - 1950) * 2)
}


#plot all ensemble members and ensemble mean: 
combined_df <- read.csv("~/cwd_global/data/5yr_return_periods_regions_ERA_members.csv")
combined_era <- read.csv("~/cwd_global/data/5yr_return_periods_regions_ERA5Land.csv")
#combined_df$window_size <- "30yr"

####################################### EUROPE #####################################
##### Step 2: Filter European regions for the map and timeseries
# Filter regions for Europe
selected_continent <- "Europe"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "MED"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# Compute ensemble mean
ensemble_mean_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")

ensemble_band_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(
    RP_mean = mean(RP, na.rm = TRUE),
    RP_lower = quantile(RP, 0.05, na.rm = TRUE),
    RP_upper = quantile(RP, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

# Apply to both datasets
ensemble_band_df$time_warped <- warp_time(ensemble_band_df$time)
combined_era_sel$time_warped <- warp_time(combined_era_sel$time)

# 3. Define range for highlight box
highlight_xmin <- warp_time(1950)
highlight_xmax <- warp_time(2010)


rollingRP_MED <- ggplot() +
  # Highlight box for stretched area
# geom_rect(aes(xmin = highlight_xmin, xmax = highlight_xmax, ymin = 0.8, ymax = 110),
#           fill = "white", color = "black", linetype = "solid", alpha = 0.6)+
  geom_vline(xintercept = highlight_xmin, color = "black", linewidth=0.5, linetype= 1)+
  geom_ribbon(data = ensemble_band_df, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper,
                                           fill = "ModE-Sim Ensemble Range (5th-95th)"), alpha = 0.4) +
  geom_line(data = ensemble_band_df, aes(x = time_warped, y = RP_mean,
                                         color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
  geom_line(data = combined_era_sel, aes(x = time_warped, y = RP,
                                         color = "ERA5 (1970-2024)"), linewidth = 1) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
  
  # Custom x-axis ticks: use original years as labels
  scale_x_continuous(
    breaks = warp_time(c( 1500, 1600, 1700, 1800, 1900, 1950, 1970, 1990, 2010)),
    labels = c( "1500", "1600", "1700", "1800", "1900", "1950", "1970", "1990", "2010"),
    name = ""
  ) +
  scale_y_log10(name = expression("Effective Return Period (" * log[10] * ")"),
        breaks = c(1, 2, 5, 10, 20, 50, 100),
  labels = c("1", "2", "5", "10", "20", "50", "100")) +  # Log scale for y-axis
  scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
  labs(title = "", color = "Line", fill = "Shading") +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size=13),
    strip.background = element_rect(fill = "#F5B0CB"),
    axis.title = element_text(size=12)
  ) +
  facet_wrap(~region, scales = "free_y")+
  annotation_logticks()

########################################
selected_continent <- "Asia"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "ARP"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# Compute ensemble mean
ensemble_mean_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")

ensemble_band_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(
    RP_mean = mean(RP, na.rm = TRUE),
    RP_lower = quantile(RP, 0.05, na.rm = TRUE),
    RP_upper = quantile(RP, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

# Apply to both datasets
ensemble_band_df$time_warped <- warp_time(ensemble_band_df$time)
combined_era_sel$time_warped <- warp_time(combined_era_sel$time)

# 3. Define range for highlight box
highlight_xmin <- warp_time(1950)
highlight_xmax <- warp_time(2010)

rollingRP_ARP <- ggplot() +
  # Highlight box for stretched area
  # geom_rect(aes(xmin = highlight_xmin, xmax = highlight_xmax, ymin = 1, ymax = 40),
  #         fill = "white", color = "black", linetype = "solid", alpha = 0.6)+
  geom_vline(xintercept = highlight_xmin, color = "black", linewidth=0.5, linetype= 1)+
  # Uncertainty band with legend
  geom_ribbon(data = ensemble_band_df, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = ensemble_band_df, aes(x = time_warped, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  geom_line(data = combined_era_sel, aes(x = time_warped, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
    # Custom x-axis ticks: use original years as labels
  scale_x_continuous(
    breaks = warp_time(c( 1500, 1600, 1700, 1800, 1900, 1950, 1970, 1990, 2010)),
    labels = c("1500", "1600", "1700", "1800", "1900", "1950", "1970", "1990", "2010"),
    name = "Year"
  ) +
  scale_y_log10(name = expression("Effective Return Period (" * log[10] * ")"),
        breaks = c(2, 5, 10, 20, 30),
  labels = c( "2", "5", "10", "20", "30")) +  # Log scale for y-axis
  # Labels and theme
  labs(
    title = "",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size=13),
    strip.background = element_rect(fill = "#E5973B"),
    axis.title = element_text(size=12)
  ) +
  facet_wrap(~region, scales = "free_y")+
  annotation_logticks()


###################################### Australia #######################################
##### Step 2: Filter Australian regions for the map and timeseries
# Filter regions for Australia
selected_continent <- "Australia"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "CAU"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# Compute ensemble mean
ensemble_mean_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")

ensemble_band_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(
    RP_mean = mean(RP, na.rm = TRUE),
    RP_lower = quantile(RP, 0.05, na.rm = TRUE),
    RP_upper = quantile(RP, 0.95, na.rm = TRUE),
    .groups = "drop"
  )

# Apply to both datasets
ensemble_band_df$time_warped <- warp_time(ensemble_band_df$time)
combined_era_sel$time_warped <- warp_time(combined_era_sel$time)

# 3. Define range for highlight box
highlight_xmin <- warp_time(1950)
highlight_xmax <- warp_time(2010)

rollingRP_CAU <-ggplot() +
  # Highlight box for stretched area
  # geom_rect(aes(xmin = highlight_xmin, xmax = highlight_xmax, ymin = 1.8, ymax = 110),
  #         fill = "white", color = "black", linetype = "solid", alpha = 0.6)+
  geom_vline(xintercept = highlight_xmin, color = "black", linewidth=0.5, linetype= 1)+
  # Uncertainty band with legend
  geom_ribbon(data = ensemble_band_df, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = ensemble_band_df, aes(x = time_warped, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  geom_line(data = combined_era_sel, aes(x = time_warped, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
  scale_x_continuous(
    breaks = warp_time(c(1500, 1600, 1700, 1800, 1900, 1950, 1970, 1990, 2010)),
    labels = c("1500", "1600", "1700", "1800", "1900", "1950", "1970", "1990", "2010"),
    name = "Year"
  ) +
  scale_y_log10(name = expression(""),
        breaks = c(2, 5, 10, 20, 50, 100),
  labels = c("2", "5", "10", "20", "50", "100")) +  # Log scale for y-axis
  # Labels and theme
  labs(
    title = "",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(color = "white"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size=13),
    strip.background = element_rect(fill = "#3C3980"),
    axis.title = element_text(size=12)
  ) +
  facet_wrap(~region, scales = "free_y")+
    annotation_logticks()


###################################### N AMERICA #######################################
##### Step 2: Filter N AMERICA regions for the map and timeseries
# Filter regions for N America
selected_continent <- "North America"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "WNA"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# Compute ensemble mean
ensemble_mean_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")

ensemble_band_df <- combined_df_sel %>%
  group_by(region, time) %>%
  summarise(
    RP_mean = mean(RP, na.rm = TRUE),
    RP_lower = quantile(RP, 0.05, na.rm = TRUE),
    RP_upper = quantile(RP, 0.95, na.rm = TRUE),
    .groups = "drop"
  )
# Apply to both datasets
ensemble_band_df$time_warped <- warp_time(ensemble_band_df$time)
combined_era_sel$time_warped <- warp_time(combined_era_sel$time)

# 3. Define range for highlight box
highlight_xmin <- warp_time(1950)
highlight_xmax <- warp_time(2010)

rollingRP_WNA <-ggplot() +
  # Highlight box for stretched area
  # geom_rect(aes(xmin = highlight_xmin, xmax = highlight_xmax, ymin = 2, ymax = 30),
  #         fill = "white", color = "black", linetype = "solid", alpha = 0.6)+
  geom_vline(xintercept = highlight_xmin, color = "black", linewidth=0.5, linetype= 1)+
  # Uncertainty band with legend
  geom_ribbon(data = ensemble_band_df, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble\nRange (5th-95th)"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = ensemble_band_df, aes(x = time_warped, y = RP_mean, color = "ModE-Sim Mean\n(1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  geom_line(data = combined_era_sel, aes(x = time_warped, y = RP, color = "ERA5-Land (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
  scale_x_continuous(
    breaks = warp_time(c(1500, 1600, 1700, 1800, 1900, 1950, 1970, 1990, 2010)),
    labels = c("1500", "1600", "1700", "1800", "1900", "1950", "1970", "1990", "2010"),
    name = ""
  ) +
  scale_y_log10(name = expression(""),
        breaks = c(2, 5, 10, 20, 30),
  labels = c("2", "5", "10", "20", "30")) +  # Log scale for y-axis
  # Labels and theme
  labs(
    title = "",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim Mean\n(1420-2009)" = "black", "ERA5-Land (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("ModE-Sim Ensemble\nRange (5th-95th)" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(color= "white"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size=13),
    strip.background = element_rect(fill = "#236A3F"),
    axis.title = element_text(size=12),
    legend.title = element_text(face = "bold")
  ) +
  facet_wrap(~region, scales = "free_y")+
    annotation_logticks()

#### combine plots in grid and save
rollingRP_MED2 <- rollingRP_MED + theme(plot.margin = margin(t = 10, r = 10, b = 5, l = 5))
rollingRP_ARP2 <- rollingRP_ARP + theme(plot.margin = margin(t = 0, r = 10, b = 5, l = 5))
rollingRP_CAU2 <- rollingRP_CAU + theme(plot.margin = margin(t = 0, r = 10, b = 5, l = 5))
rollingRP_WNA2 <- rollingRP_WNA + theme(plot.margin = margin(t = 10, r = 10, b = 5, l = 5))


combined_rollingRP <- plot_grid(
  rollingRP_MED2, rollingRP_WNA2,
  rollingRP_ARP2, rollingRP_CAU2,
  ncol = 2,
  nrow = 2,
  rel_widths = c(1, 1),
  align = "h",
  axis = "tblr",
  labels = c(
    "a)",  "b) ", "c)", "d)"
  ),
  label_size = 14,
  label_x = 0.09,
  hjust = 0,
  label_y = 0.95,
  vjust = 1
)

print(combined_rollingRP)

ggsave(
  filename = "~/cwd_global/manuscript/rollingRP.pdf",
  plot     = combined_rollingRP,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 6.5,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

ggsave(
  filename = "~/cwd_global/manuscript/rollingRP_legend.pdf",
  plot     = rollingRP_WNA,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 6.5,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

```


#### Scatterplots for modern times:
first 3: 
1st plot: reference/overlap period *5 yr* events in mm for ERA5 and ModE-Sim: 1970-2009 (exclude problematic 60s of ERA5) 
2nd: y-axis: ERA5_Land 1995-2024 5yr vs x-axis: ERA5 Land 1970-1999
3rd: y-axis: ERA5-Land 1995-2024 5 yr vs x-axis: ModE-SIm BC 1970-1999

```{r}
corrected_full <- readRDS("data/regionalResults_ModESim_bc.RData")

##calculate return level of 5 year events - 30 year chunks for consistency: 
#remove SAH due to unphysically large PCWD values - runaway PCWD 
regional_results_ERA5Land_eva <- regional_results_ERA5Land[ names(regional_results_ERA5Land) != "SAH" ]

use_rp <- 5   # return period of two years

# Modern ERA5-Land (1994-2024) for each region: 
# Initialize a list to store results for all regions
RL_ERA_modern <- list() #save modern 5yr events (in mm) for each region

# Loop through each region in regional_results
for (region in names(regional_results_ERA5Land_eva)) { #swap european_regions for regional_results for all regions
  # Extract the result array for the current region
  result_array_region <- regional_results_ERA5Land_eva[[region]][26:55] #1995-2024
  fit_ref <- extRemes::fevd(result_array_region, type = "Gumbel", method = "MLE", units = "years")
  return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))

  RL_ERA_modern[[region]] <- return_level
  
}

# Modern ERA5-Land (1994-2024) for each region: 
# Initialize a list to store results for all regions
RL_ERA_overlap <- list() #save modern 5yr events (in mm) for each region

# Loop through each region in regional_results
for (region in names(regional_results_ERA5Land_eva)) { #swap european_regions for regional_results for all regions
  # Extract the result array for the current region
  result_array_region <- regional_results_ERA5Land_eva[[region]][1:30] #1970-1999
  fit_ref <- extRemes::fevd(result_array_region, type = "Gumbel", method = "MLE", units = "years")
  return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))

  RL_ERA_overlap[[region]] <- return_level
  
}

##### ModE-Sim 
path <- "data/volc_impact_aggregates/collected_p_pet/"
folders <- list.files(path, full.names = TRUE, pattern = "m[0-9]{3}") #m001 - m020 in first set

# Extract unique ensemble member identifiers from folder names
ensemble_member <- unique(basename(folders))[1:20]
ensemble_members <- paste0(ensemble_member, "_tidy")


corrected_full_eva <- corrected_full[ names(corrected_full) != "SAH" ]

regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")


###Overlap with ERA5-land for each region (1970-2009):
RL_ModE_overlap <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][551:580,]#1970-1999
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_overlap[[region]] <- result_array_region
}

### Pre-industrial for every region (1420-1850):
RL_ModE_preind <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][1:30,]#1420-1450
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_preind[[region]] <- result_array_region
}

### Early-industrial for every region (1851-1950):
RL_ModE_earlyind <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][470:501,]#1890-1920
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_earlyind[[region]] <- result_array_region
}


### Modern ModeSIM for every region (1951-2009):
RL_ModE_modern <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][551:580,]#1970-1999
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_modern[[region]] <- result_array_region
}

```

Plot in scatterplots: 

```{r}
# #### Overlap period ModE-Sim and ERA5-Land
# library(dplyr)
# library(ggplot2)

####### -------------------- Overlap period --------------------------------
# 1) Summarise lists into a data.frame:
df_overlap <- tibble(
  region = names(RL_ERA_overlap),
  era     = unlist(RL_ERA_overlap)
) %>% 
  mutate(
    # Grab the ensemble vector for each region
    mod_vec = RL_ModE_overlap[region],
    # Compute mean, min, max
    mod_mean = map_dbl(mod_vec, ~ mean(.x)),
    mod_min  = map_dbl(mod_vec, ~ min(.x)),
    mod_max  = map_dbl(mod_vec, ~ max(.x))
  ) %>% 
  dplyr::select(-mod_vec)

# Ensure the continent information is consistent
df_overlap <- df_overlap %>%
  mutate(continent = case_when(
    region %in% continent_mapping[["Europe"]] ~ "Europe",
    region %in% continent_mapping[["North America"]] ~ "North America",
    region %in% continent_mapping[["Africa"]] ~ "Africa",
    region %in% continent_mapping[["Asia"]] ~ "Asia",
    region %in% continent_mapping[["Central America"]] ~ "Central America",
    region %in% continent_mapping[["South America"]] ~ "South America",
    region %in% continent_mapping[["Australia"]] ~ "Australia",
    TRUE ~ "Other"
  ))

# 2) Plot with horizontal error bars
scatter_overlap <- ggplot(df_overlap, aes(x = mod_mean, y = era, color = continent)) +
  geom_errorbarh(aes(xmin = mod_min, xmax = mod_max), height = 100) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names
  labs(
    x     = "ModE-Sim* 5yr PCWD [mm]",
    y     = "ERA5-Land 5yr PCWD [mm]",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 2000, 500)) +
  scale_y_continuous(limits = c(0, 2000, 500)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


####### -------------------- Modern_ModE-Sim --------------------------------
# 1) Summarise lists into a data.frame:
df_modern_era <- tibble(
  region = names(RL_ERA_modern),
  era_mod     = unlist(RL_ERA_modern),
  era_overlap = unlist(RL_ERA_overlap)
) 

# Ensure the continent information is consistent
df_modern_era <- df_modern_era %>%
  mutate(continent = case_when(
    region %in% continent_mapping[["Europe"]] ~ "Europe",
    region %in% continent_mapping[["North America"]] ~ "North America",
    region %in% continent_mapping[["Africa"]] ~ "Africa",
    region %in% continent_mapping[["Asia"]] ~ "Asia",
    region %in% continent_mapping[["Central America"]] ~ "Central America",
    region %in% continent_mapping[["South America"]] ~ "South America",
    region %in% continent_mapping[["Australia"]] ~ "Australia",
    TRUE ~ "Other"
  ))

# 2) Plot with horizontal error bars
scatter_era5 <- ggplot(df_modern_era, aes(x = era_overlap, y = era_mod, color = continent)) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names
  labs(
    x     = "ERA5-Land 5yr PCWD [mm] (1970-1999)",
    y     = "ERA5-Land 5yr PCWD [mm] (1995-2024)",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 2000, 500)) +
  scale_y_continuous(limits = c(0, 2000, 500)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


####### -------------------- Modern ModE-SIm --------------------------------
# 1) Summarise lists into a data.frame:
df_modern_ModE <- tibble(
  region = names(RL_ERA_modern),
  era     = unlist(RL_ERA_modern)
) %>% 
  mutate(
    # Grab the ensemble vector for each region
    mod_vec = RL_ModE_overlap[region],
    # Compute mean, min, max
    mod_mean = map_dbl(mod_vec, ~ mean(.x)),
    mod_min  = map_dbl(mod_vec, ~ min(.x)),
    mod_max  = map_dbl(mod_vec, ~ max(.x))
  ) %>% 
  dplyr::select(-mod_vec)

# Ensure the continent information is consistent
df_modern_ModE <- df_modern_ModE %>%
  mutate(continent = case_when(
    region %in% continent_mapping[["Europe"]] ~ "Europe",
    region %in% continent_mapping[["North America"]] ~ "North America",
    region %in% continent_mapping[["Africa"]] ~ "Africa",
    region %in% continent_mapping[["Asia"]] ~ "Asia",
    region %in% continent_mapping[["Central America"]] ~ "Central America",
    region %in% continent_mapping[["South America"]] ~ "South America",
    region %in% continent_mapping[["Australia"]] ~ "Australia",
    TRUE ~ "Other"
  ))

# 2) Plot with horizontal error bars
scatter_era_mod <- ggplot(df_modern_ModE, aes(x = mod_mean, y = era, color = continent)) +
  geom_errorbarh(aes(xmin = mod_min, xmax = mod_max), height = 100) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names
  labs(
    x     = "ModE-Sim* 5yr PCWD [mm] (1970-1999)",
    y     = "ERA5-Land 5yr PCWD [mm] (1995-2024)",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 2000, 500)) +
  scale_y_continuous(limits = c(0, 2000, 500)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
        text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


scatter_overlap_2 <- scatter_overlap + theme(plot.margin = margin(t = 40, r = 20, b = 5, l = 5))
scatter_era5_2 <- scatter_era5 + theme(plot.margin = margin(t = 40, r = 20, b = 5, l = 5))
scatter_era_mod_2 <- scatter_era_mod + theme(plot.margin = margin(t = 40, r = 10, b = 5, l = 5))


combined_val_scatter <- plot_grid(
  scatter_overlap_2, 
  scatter_era5_2, 
  scatter_era_mod_2,
  ncol = 3,
  rel_widths = c(1, 1, 1),
  align = "h",
  axis = "tblr",
  labels = c(
    "a) Overlap period (1970-1999)", 
    "b) Modern ERA5-Land", "c) Modern ERA5-Land vs ModE-Sim"
  ),
  label_size = 14,
  label_x = 0.1,
  hjust = 0,
  label_y = 0.98,
  vjust = 1
)

print(combined_val_scatter)

ggsave(
  filename = "~/cwd_global/manuscript/Scatter_validation_plots.pdf",
  plot     = combined_val_scatter,
  device   = cairo_pdf,    # better text rendering
  width    = 17,           # in inches
  height   = 5.9,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

```

#### Same thing but comparing historic extremes to modern times and *30 yr* events
Scatterplots for 4 periods:
first 2: 
- y-axis: ModE-Sim 5yr extreme event in mm (1420-1490) for each region (70 year period)
- x-axis: ModE-Sim 5yr extreme event in mm 
  - early-industrial (1880-1950) for each region with error denoting spread of ensemble members
  - ERA5-Land "modern" (1954-2024) for each region with error denoting spread of ensemble members

```{r}
##calculate return level of 5 year events - 30 year chunks for consistency: 
#remove SAH due to unphysically large PCWD values - runaway PCWD 
regional_results_ERA5Land_eva <- regional_results_ERA5Land[ names(regional_results_ERA5Land) != "SAH" ]

use_rp <- 30   # return period of two years

# Modern ERA5-Land (1994-2024) for each region: 
# Initialize a list to store results for all regions
RL_ERA_modern <- list() #save modern 5yr events (in mm) for each region

# Loop through each region in regional_results
for (region in names(regional_results_ERA5Land_eva)) { #swap european_regions for regional_results for all regions
  # Extract the result array for the current region
  result_array_region <- regional_results_ERA5Land_eva[[region]][26:55] #1994-2024
  fit_ref <- extRemes::fevd(result_array_region, type = "Gumbel", method = "MLE", units = "years")
  return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))

  RL_ERA_modern[[region]] <- return_level
  
}

##### ModE-Sim 
path <- "data/volc_impact_aggregates/collected_p_pet/"
folders <- list.files(path, full.names = TRUE, pattern = "m[0-9]{3}") #m001 - m020 in first set
# Extract unique ensemble member identifiers from folder names
ensemble_member <- unique(basename(folders))[1:20]
ensemble_members <- paste0(ensemble_member, "_tidy")

corrected_full_eva <- corrected_full[ names(corrected_full) != "SAH" ]
regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")



### Pre-industrial for every region (1420-1800):
RL_ModE_preind <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][1:381,]#1420-1800
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_preind[[region]] <- result_array_region
}

### Early-industrial for every region (1851-1950):
RL_ModE_earlyind <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][382:551,]#1801-1970
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_earlyind[[region]] <- result_array_region
}


### Modern for every region (1970-2009):
RL_ModE_modern <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][552:590,]#1971-2009
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_modern[[region]] <- result_array_region
}


### Full ModESim period for every region (1420-1994):
RL_ModE_full <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][1:575,]#1420-1994
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_full[[region]] <- result_array_region
}

```


Plot in scatterplots: 

```{r}
# #### Overlap period ModE-Sim and ERA5-Land
# library(dplyr)
# library(ggplot2)

####### -------------------- Early-industrial --------------------------------
# 1) Summarise lists into a data.frame:
df_30yr <- tibble(
  region = names(RL_ERA_modern),
  era = unlist(RL_ERA_modern)
) %>% 
  mutate(
    # Grab the ensemble vector for each region
    mod_vec_pre = RL_ModE_preind[region],
    # Compute mean, min, max
    mod_mean_pre = map_dbl(mod_vec_pre, ~ mean(.x)),
    mod_min_pre  = map_dbl(mod_vec_pre, ~ min(.x)),
    mod_max_pre  = map_dbl(mod_vec_pre, ~ max(.x)),
    mod_sd_pre  = map_dbl(mod_vec_pre, ~ sd(.x)),

    # Grab the ensemble vector for each region earlyind
    mod_vec_early = RL_ModE_earlyind[region],
    # Compute mean, min, max
    mod_mean_early = map_dbl(mod_vec_early, ~ mean(.x)),
    mod_min_early  = map_dbl(mod_vec_early, ~ min(.x)),
    mod_max_early  = map_dbl(mod_vec_early, ~ max(.x)),
    mod_sd_early  = map_dbl(mod_vec_early, ~ sd(.x)),

    # Grab the ensemble vector for each region modern
    mod_vec_modern = RL_ModE_modern[region],
    # Compute mean, min, max
    mod_mean_modern = map_dbl(mod_vec_modern, ~ mean(.x)),
    mod_min_modern  = map_dbl(mod_vec_modern, ~ min(.x)),
    mod_max_modern  = map_dbl(mod_vec_modern, ~ max(.x)), 
    mod_sd_modern  = map_dbl(mod_vec_modern, ~ sd(.x)),

    # Grab the ensemble vector for each region full
    mod_vec_full = RL_ModE_full[region],
    # Compute mean, min, max
    mod_mean_full = map_dbl(mod_vec_full, ~ mean(.x)),
    mod_min_full  = map_dbl(mod_vec_full, ~ min(.x)),
    mod_max_full  = map_dbl(mod_vec_full, ~ max(.x)),
    mod_sd_full  = map_dbl(mod_vec_full, ~ sd(.x)),

  ) %>% 
  dplyr::select(-mod_vec_early, -mod_vec_pre, -mod_vec_modern, -mod_vec_full)


#####-------------------------Modern
# Ensure the continent information is consistent
df_30yr <- df_30yr %>%
  mutate(continent = case_when(
    region %in% continent_mapping[["Europe"]] ~ "Europe",
    region %in% continent_mapping[["North America"]] ~ "North America",
    region %in% continent_mapping[["Africa"]] ~ "Africa",
    region %in% continent_mapping[["Asia"]] ~ "Asia",
    region %in% continent_mapping[["Central America"]] ~ "Central America",
    region %in% continent_mapping[["South America"]] ~ "South America",
    region %in% continent_mapping[["Australia"]] ~ "Australia",
    TRUE ~ "Other"
  ))

# 2) Plot with horizontal error bars
scatter_mod <- ggplot(df_30yr, aes(x = mod_mean_modern, y = era, color = continent)) +
  annotate("rect", xmin  = 0, xmax  = 500, ymin  = 0, ymax  = 500, fill  = "grey70", alpha = 0.3) +
  geom_errorbarh(aes(xmin = mod_min_modern, xmax = mod_max_modern), height = 0) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  #geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names

  labs(
    x     = "ModE-Sim* 30yr PCWD [mm] (1971-2009)",
    y     = "ERA5-Land 30yr PCWD [mm] (1995-2024)",
  #  title = "Modern period",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 2500, 500)) +
  scale_y_continuous(limits = c(0, 2500, 500)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


#########-----------------Early industrial

# 2) Plot with horizontal error bars
scatter_early <- ggplot(df_30yr, aes(x = mod_mean_early, y = era, color = continent)) +
    annotate("rect", xmin  = 0, xmax  = 500, ymin  = 0, ymax  = 500, fill  = "grey70", alpha = 0.3) +
geom_errorbarh(aes(xmin = mod_min_early, xmax = mod_max_early), height = 0) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  #geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names

  labs(
    x     = "ModE-Sim* 30yr PCWD [mm] (1801-1970)",
    y     = "ERA5-Land 30yr PCWD [mm] (1995-2024)",
   # title = "Modern vs Early-Industrial Period",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 2500, 500)) +
  scale_y_continuous(limits = c(0, 2500, 500)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


####### -------------------- Preindustrial --------------------------------
# 2) Plot with horizontal error bars
scatter_pre <- ggplot(df_30yr, aes(x =mod_mean_pre , y = era, color = continent)) +
    annotate("rect", xmin  = 0, xmax  = 500, ymin  = 0, ymax  = 500, fill  = "grey70", alpha = 0.3) +
geom_errorbarh(aes(xmin = mod_min_pre, xmax = mod_max_pre), height = 0) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  #geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names

  labs(
    x     = "ModE-Sim* 30yr PCWD [mm] (1420-1800)",
    y     = "ERA5-Land 30yr PCWD [mm] (1995-2024)",
  #  title = "Modern vs Preindustrial Period",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 2500, 500)) +
  scale_y_continuous(limits = c(0, 2500, 500)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )

####### -------------------- Full period --------------------------------
# 2) Plot with horizontal error bars
#scatter_full <- 
  ggplot(df_30yr, aes(x =mod_mean_full , y = era, color = continent)) +
    annotate("rect", xmin  = 0, xmax  = 500, ymin  = 0, ymax  = 500, fill  = "grey70", alpha = 0.3) +
geom_errorbarh(aes(xmin = mod_min_full, xmax = mod_max_full), height = 0) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  #geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names

  labs(
    x     = "ModE-Sim* 30yr PCWD [mm] (1420-1994)",
    y     = "ERA5-Land 30yr PCWD [mm] (1995-2024)",
  #  title = "Modern vs Preindustrial Period",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 2500, 500)) +
  scale_y_continuous(limits = c(0, 2500, 500)) +
  theme_classic() +
  theme(
    legend.position    = "right",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )

ggsave(
  filename = "figures/scatter_legend.pdf",
 # plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   =8,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


##17 regions show exceedance compared to the last 600 years 
## 17 / 43 regions = 0.3953488 --> ~40% of all land regions experience unprecedented extreme droughts


```


"Zoomed in" version of the previous plots: 

```{r}
df_zoom <- df_30yr %>%
  filter(
    mod_mean_modern <= 500,
    era              <= 500
  )

# 2) Plot with horizontal error bars
scatter_zoom_mod <- ggplot(df_zoom, aes(x = mod_mean_modern, y = era, color = continent)) +
annotate("rect", xmin  = 0, xmax  = 500, ymin  = 0, ymax  = 500, fill  = "grey70", alpha = 0.3) +
geom_errorbarh(aes(xmin = mod_min_modern, xmax = mod_max_modern), height = 0) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  #geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names

  labs(
    x     = "ModE-Sim* 30yr PCWD [mm] (1971-2009)",
  #  y     = "ERA5-Land 30yr PCWD [mm] (1995-2024)",
  y = "",
   # title = "Modern period",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 500), breaks = seq(0, 500, 100)) +
  scale_y_continuous(limits = c(0, 500), breaks = seq(0, 500, 100)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


#########-----------------Early industrial

# 2) Plot with horizontal error bars
scatter_zoom_early <- ggplot(df_zoom, aes(x = mod_mean_early, y = era, color = continent)) +
  annotate("rect", xmin  = 0, xmax  = 500, ymin  = 0, ymax  = 500, fill  = "grey70", alpha = 0.3) +
 geom_errorbarh(aes(xmin = mod_min_early, xmax = mod_max_early), height = 0) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  #geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names

  labs(
    x     = "ModE-Sim* 30yr PCWD [mm] (1801-1970)",
   # y     = "ERA5-Land 30yr PCWD [mm] (1995-2024)",
   y= "",
   # title = "Modern vs Early-industrial period",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 500), breaks = seq(0, 500, 100)) +
  scale_y_continuous(limits = c(0, 500), breaks = seq(0, 500, 100)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


####### -------------------- Preindustrial --------------------------------
# 2) Plot with horizontal error bars
scatter_zoom_pre <- ggplot(df_zoom, aes(x =mod_mean_pre , y = era, color = continent)) +
  annotate("rect", xmin  = 0, xmax  = 500, ymin  = 0, ymax  = 500, fill  = "grey70", alpha = 0.3) +
 geom_errorbarh(aes(xmin = mod_min_pre, xmax = mod_max_pre), height = 0) +
  #geom_errorbar(aes(ymin = mod_min_pre, ymax = mod_max_pre), width = 50) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  #geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names

  labs(
    x     = "ModE-Sim* 30yr PCWD [mm] (1420-1800)",
   # y     = "ERA5-Land 30yr PCWD [mm] (1995-2024)",
   y = "",
   # title = "Modern vs Preindustrial period",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 500), breaks = seq(0, 500, 100)) +
  scale_y_continuous(limits = c(0, 500), breaks = seq(0, 500, 100)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


####### -------------------- Full period --------------------------------
# 2) Plot with horizontal error bars
scatter_zoom_full <- 
  ggplot(df_zoom, aes(x =mod_mean_full , y = era, color = continent)) +
  annotate("rect", xmin  = 0, xmax  = 500, ymin  = 0, ymax  = 500, fill  = "grey70", alpha = 0.3) +
 geom_errorbarh(aes(xmin = mod_min_full, xmax = mod_max_full), height = 0) +
  #geom_errorbar(aes(ymin = mod_min_pre, ymax = mod_max_pre), width = 50) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  #geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names

  labs(
    x     = "ModE-Sim* 30yr PCWD [mm] (1420-1994)",
    y     = "ERA5-Land 30yr PCWD [mm] (1995-2024)",
   # title = "Modern vs Preindustrial period",
    color = "Continent"
  ) +
  scale_x_continuous(limits = c(0, 500), breaks = seq(0, 500, 100)) +
  scale_y_continuous(limits = c(0, 500), breaks = seq(0, 500, 100)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )

```
Quantify how many regions are within simulated spread
```{r}
#library(dplyr)

df_30yr %>%
  # 1) flag “outside of [min, max]” in each period
  mutate(
    outside_pre   = era < mod_min_pre   | era > mod_max_pre,
    outside_early = era < mod_min_early | era > mod_max_early,
    outside_mod   = era < mod_min_modern| era > mod_max_modern
  ) %>%
  # 2) tally up how many TRUEs in each flag
  summarise(
    n_regions = n(),
    n_out_pre   = sum(outside_pre),
    n_out_early = sum(outside_early),
    n_out_mod   = sum(outside_mod)
  ) %>%
  # 3) optionally add percentage of total
  mutate(
    pct_out_pre   = 100 * n_out_pre   / n_regions,
    pct_out_early = 100 * n_out_early / n_regions,
    pct_out_mod   = 100 * n_out_mod   / n_regions
  )

###only era larger than past: 
df_30yr %>%
  # 1) flag “outside of [min, max]” in each period
  mutate(
    outside_pre   = era > mod_max_pre,
    outside_early = era > mod_max_early,
    outside_mod   = era > mod_max_modern
  ) %>%
  # 2) tally up how many TRUEs in each flag
  summarise(
    n_regions = n(),
    n_out_pre   = sum(outside_pre),
    n_out_early = sum(outside_early),
    n_out_mod   = sum(outside_mod)
  ) %>%
  # 3) optionally add percentage of total
  mutate(
    pct_out_pre   = 100 * n_out_pre   / n_regions,
    pct_out_early = 100 * n_out_early / n_regions,
    pct_out_mod   = 100 * n_out_mod   / n_regions
  )

```

```{r}
## check values for regions exceeding the natural variability
df_30yr %>% 
  mutate(
    outside_pre   = era > mod_max_pre,
    outside_early = era > mod_max_early,
    outside_mod   = era > mod_max_modern
  ) %>% 
  filter(outside_pre) %>% 
  dplyr::select(region, era, mod_min_pre, mod_max_pre)

```
##### Show which regions are affected

Load reference region info:
```{r}
#Load reference regions and coastlines:

load("data/Reference_regions/IPCC-WGI-reference-regions-v4_R.rda", verbose = TRUE)
# IPCC_WGI_reference_regions_v4 is a Spatial*DataFrame with a column "region" (or similar)
refregions_sf <- st_as_sf(IPCC_WGI_reference_regions_v4)
refregions_sf <- refregions_sf %>% rename(region = Acronym)

temp.dir <- tempdir()
unzip("data/Reference_regions/ne_110m_coastline.zip", exdir = temp.dir)
coastLines <- st_read(dsn = temp.dir, layer = "ne_110m_coastline")

names(names(refregions))

proj4string(refregions)

# Convert `refregions` (SpatialPolygons) to an `sf` object for ggplot compatibility
#refregions_sf <- st_as_sf(refregions)

# Define continent assignments
continent_mapping <- list(
  "Europe" = c("NEU", "WCE", "EEU", "MED"),
  "North America" = c("NWN", "NEN", "WNA", "CNA", "ENA", "GIC"),
  "Africa" = c("SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG"),
  "Asia" = c("WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "WSB", "ESB", "RFE", "RAR"),
  "Central America" =c("NCA", "SCA", "CAR"),
  "South America" = c("NWS", "NSA", "NES", "SAM", "SWS", "SES", "SSA"),
  "Australia" = c("NAU", "CAU", "EAU", "SAU", "NZ")
)

# Define continent assignments
continent_mapping_sel <- list(
  "Europe" = c("NEU"),
  "North America" = c("ENA"),
  "Africa" = c("MDG"),
  "Asia" = c("EAS"),
  "Central America" =c(""),
  "South America" = c("NES"),
  "Australia" = c("EAU")
)

# # Assign continents to regions
# refregions_sf <- refregions_sf %>%
#   mutate(
#     continent = case_when(
#       row.names(refregions) %in% continent_mapping_sel$Europe ~ "Europe",
#       row.names(refregions) %in% continent_mapping_sel$`North America` ~ "North America",
#       row.names(refregions) %in% continent_mapping_sel$Africa ~ "Africa",
#       row.names(refregions) %in% continent_mapping_sel$Asia ~ "Asia",
#       row.names(refregions) %in% continent_mapping_sel$`Central America` ~ "Central America",
#       row.names(refregions) %in% continent_mapping_sel$`South America` ~ "South America",
#       row.names(refregions) %in% continent_mapping_sel$Australia ~ "Australia",
#       TRUE ~ "Ocean"  # For regions not mapped to a continent
#     )
#   ) 


# Assign continents to regions
refregions_sf <- refregions_sf %>%
  mutate(
    continent = case_when(
      row.names(refregions) %in% continent_mapping$Europe ~ "Europe",
      row.names(refregions) %in% continent_mapping$`North America` ~ "North America",
      row.names(refregions) %in% continent_mapping$Africa ~ "Africa",
      row.names(refregions) %in% continent_mapping$Asia ~ "Asia",
      row.names(refregions) %in% continent_mapping$`Central America` ~ "Central America",
      row.names(refregions) %in% continent_mapping$`South America` ~ "South America",
      row.names(refregions) %in% continent_mapping$Australia ~ "Australia",
      TRUE ~ "Ocean"  # For regions not mapped to a continent
    )
  ) 


# Define continent colors
continent_colors <- c(
  "Europe" = "#F5B0CB",
  "North America" = "#236A3F",
  "Africa" = "#91C7B1",
  "Asia" = "#E5973B",
  "Central America" = "#B33951",
  "South America" = "#E3D081",
  "Australia" = "#3C3980",
  "Ocean" = "white"
)

# Extract region centroids for labeling
region_labels <- refregions_sf %>%
  st_centroid() %>%
  mutate(label = row.names(refregions))  # Add region names as labels


# Load land data for filling continents
land <- ne_countries(scale = "medium", returnclass = "sf")

# Plot the map with ggplot2
ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +  # Fill land areas
  geom_sf(data = refregions_sf, aes(fill = continent), color = "black", alpha = 0.65) +  # Transparent overlay
  geom_sf_text(data = region_labels, aes(label = label), size = 3, color = "black", check_overlap = FALSE) +  # Add region names
  scale_fill_manual(values = continent_colors, name = "Continent") +
  theme_minimal() +
  labs(x= "", y=""
       #title = "Map of IPCC Regions by Continent",
       #subtitle = "Regions coloured by assigned continents, region names labeled"
       ) +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 12),
    legend.position = "none"
  )
```


```{r}
# For Pre‑industrial exceedance (any region where era > pre‑industrial max)
regions_pre <- df_30yr %>%
  filter(era > mod_max_pre) %>%
  pull(region) %>%
  unique()

# For Early‑industrial exceedance
regions_early <- df_30yr %>%
  filter(era > mod_max_early) %>%
  pull(region) %>%
  unique()

# For Modern exceedance
regions_modern <- df_30yr %>%
  filter(era > mod_max_modern) %>%
  pull(region) %>%
  unique()


plot_exceed_map <- function(regions_to_show) {
  # subset only those regions
  shown_sf <- refregions_sf %>%
    filter(region %in% regions_to_show)
  
  # centroids for labels
  pts <- shown_sf %>%
    st_centroid() %>%
    mutate(label = region)
  
  # build the map
  ggplot() +
    geom_sf(data = land, fill = "lightgray", color = NA) +
    geom_sf(
      data = shown_sf,
      aes(fill = continent),
      color = "grey20",
      size = 0.2,
      alpha = 0.5
    ) +
    geom_sf_text(
      data = pts,
      aes(label = label),
      size = 3.5,
      colour= "black"
    ) +
    scale_fill_manual(
      values   = continent_colors,
      na.value = "transparent",
      name     = "Continent"
    ) +
    coord_sf(
      xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE)+
    theme_void() +
    theme(
      plot.title      = element_text(face = "bold", hjust = .5),
      plot.margin     = margin(0,0,0,0),
      legend.position = "none"
    ) 
}

map_pre    <- plot_exceed_map(regions_pre)
map_early  <- plot_exceed_map(regions_early)
map_modern <- plot_exceed_map(regions_modern)

map_pre
map_early
map_modern


#--------------- all plots
regions_all <- df_30yr %>%
  filter(
    era > mod_max_pre,
    era > mod_max_early,
    era > mod_max_modern
  ) %>%
  pull(region) %>%
  unique()

map_all <- plot_exceed_map(regions_all) +
  ggtitle("Regions Exceeding All Three Historical Periods") +
  theme(legend.position = "right", legend.title = element_text(face= "bold")) +
  guides(fill = guide_legend(title.position = "top", 
                             title.hjust = 0))

map_all



ggsave(
  filename = "~/cwd_global/manuscript/Map_600_yr_exceedance_legend.pdf",
  plot     = map_all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   =6,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


##17 regions show exceedance compared to the last 600 years 
## 17 / 43 regions = 0.3953488 --> ~40% of all land regions experience unprecedented extreme droughts

```


Figure 2: Scatter plot comparing 30 yr drought magnitudes and exceedance maps

```{r}

scatter_pre2 <- scatter_pre + theme(plot.margin = margin(t = 40, r = 10, b = 5, l = 5))
scatter_zoom_pre2 <- scatter_zoom_pre + theme(plot.margin = margin(t = 15, r = 0, b = 5, l = 5))
map_pre2 <- map_pre + theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0))

scatter_early2 <- scatter_early + theme(plot.margin = margin(t = 15, r = 10, b = 5, l = 5))
scatter_zoom_early2 <- scatter_zoom_early + theme(plot.margin = margin(t = 15, r = 0, b = 5, l = 5))
map_early2 <- map_early + theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0))

scatter_mod2 <- scatter_mod + theme(plot.margin = margin(t = 15, r = 10, b = 5, l = 5))
scatter_zoom_mod2 <- scatter_zoom_mod + theme(plot.margin = margin(t = 15, r = 0, b = 5, l = 5))
map_modern2 <- map_modern + theme(plot.margin = margin(t = 0, r = 0, b = 0, l = 0))



combined_scatter <- plot_grid(
  scatter_pre2, scatter_zoom_pre2, map_pre2,
  scatter_early2, scatter_zoom_early2, map_early2,
  scatter_mod2, scatter_zoom_mod2, map_modern2,
  nrow = 3,
  ncol = 3,
  rel_widths = c(1, 1, 1.7),
  rel_heights = c(1, 1, 1),
  align = "h",
  axis = "tblr",
  labels = c(
    "a) Modern vs Pre-Industrial Period", 
    "b)", "c)", 
    "d) Modern vs Early-Industrial Period", 
    "e)", "f)", 
    "g) Modern Period", 
    "h)", "i)"
  ),
  label_size = 14,
  label_x = 0.07,
  hjust = 0,
  label_y = 0.96,
  vjust = 1
)

print(combined_scatter)

ggsave(
  filename = "~/cwd_global/manuscript/Scatter_plots.pdf",
  plot     = combined_scatter,
  device   = cairo_pdf,    # better text rendering
  width    = 17,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```

```{r}
###Calculate t-value for each region for 4 different period sets (preindustrial, early industrial, modern, full ModESim (til 1994))
regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")

df_tvals <- df_30yr %>%
  mutate(
    n = case_when(
      region %in% regions_to_modify ~ 19,
      TRUE ~ 20
    ),

    # Preindustrial
    delta_pre = abs(era - mod_mean_pre) / sqrt(2),
    SE_mod_pre = mod_sd_pre / sqrt(n),
    SE_delta_pre = SE_mod_pre / sqrt(2),
    t_pre = delta_pre / SE_delta_pre,

    # Early
    delta_early = abs(era - mod_mean_early) / sqrt(2),
    SE_mod_early = mod_sd_early / sqrt(n),
    SE_delta_early = SE_mod_early / sqrt(2),
    t_early = delta_early / SE_delta_early,

    # Modern
    delta_modern = abs(era - mod_mean_modern) / sqrt(2),
    SE_mod_modern = mod_sd_modern / sqrt(n),
    SE_delta_modern = SE_mod_modern / sqrt(2),
    t_modern = delta_modern / SE_delta_modern,

    # Full
    delta_full = abs(era - mod_mean_full) / sqrt(2),
    SE_mod_full = mod_sd_full / sqrt(n),
    SE_delta_full = SE_mod_full / sqrt(2),
    t_full = delta_full / SE_delta_full
  ) %>%
  dplyr::select(region, starts_with("delta_"), starts_with("t_"))

```


```{r}
# Join your t-values to the regions
refregions_sf_plot <- refregions_sf %>%
  left_join(df_tvals %>% dplyr::select(region, t_full, t_pre, t_early, t_modern), by = "region")

refregions_sf_plot <- refregions_sf_plot %>%
  mutate(
    exceed_pre = region %in% regions_pre,
    exceed_early = region %in% regions_early,
    exceed_modern = region %in% regions_modern,
    exceed_all = region %in% regions_all
  )

# Check for NAs if needed:
# summary(refregions_sf$t_full)

# Recompute centroids for labels if needed
region_labels <- refregions_sf_plot %>%
  st_centroid() %>%
  mutate(label = region)

# Load land for basemap
land <- ne_countries(scale = "medium", returnclass = "sf")

plot_tvalue_with_hatch <- function(t_column = "t_full", exceed_col = "exceed_pre") {
  if (!t_column %in% names(refregions_sf_plot)) {
    stop(paste("Column", t_column, "not found in refregions_sf!"))
  }
  if (!exceed_col %in% names(refregions_sf_plot)) {
    stop(paste("Column", exceed_col, "not found in refregions_sf!"))
  }
  
  shown_sf <- refregions_sf_plot %>%
    filter(is.finite(!!sym(t_column)))
  
  pts <- st_point_on_surface(shown_sf) %>% mutate(label = region)
  
  ggplot() +
    geom_sf(data = land, fill = "lightgrey", color = NA) +
    
    # Main fill by t-value
    geom_sf(
      data = shown_sf,
      aes(fill = !!sym(t_column)),
      color = "grey20",
      size = 0.15
    ) +
    scale_fill_viridis_c(
      option = "magma",
      na.value = "transparent",
      name = "t-value",
      limits = c(0, 120), 
      alpha = 0.55,
      direction = -1
      #guide = guide_legend(override.aes = list(alpha = 1)) 
    ) +
    
    # Overlay hatch pattern where exceedance is TRUE
    geom_sf_pattern(
      data = filter(shown_sf, !!sym(exceed_col)),
      fill = NA,
      color = NA,
      pattern = "stripe",
      pattern_fill = "grey50",      # the fill of the stripes
      pattern_colour = NA,       # turn off black edges!
      pattern_angle = 45,
      pattern_spacing   = unit(5, "pt"),
      pattern_density = 0.2,
      size = 0.3,
      show.legend = TRUE
    ) +
    geom_sf_text(
      data = pts,
      aes(label = label),
      size = 3,
      colour = "black"
    ) +
    coord_sf(
      xlim = c(-180, 178.125),
      ylim = c(-60, 90),
      expand = FALSE
    ) +
    theme_void() +
    theme(
      legend.title = element_text(face = "bold"),
      plot.margin = margin(0, 0, 0, 0),
      legend.position = "right"
    )
}


map_full <- plot_tvalue_with_hatch("t_full", "exceed_all")

# ggsave(
#   filename = "~/cwd_global/manuscript/Scatter_plots_tvalue_full.pdf",
#   plot = map_full,
#   device   = cairo_pdf,    # better text rendering
#   width    = 17,           # in inches
#   height   = 13,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )

map_pre <- plot_tvalue_with_hatch("t_pre", "exceed_pre")
map_early <- plot_tvalue_with_hatch("t_early", "exceed_early")
map_modern <- plot_tvalue_with_hatch("t_modern", "exceed_modern")


map_full
```

Put together with scatter plots: 

```{r}

scatter_pre2 <- scatter_pre + theme(plot.margin = margin(t = 40, r = 10, b = 5, l = 5))
scatter_zoom_pre2 <- scatter_zoom_pre + theme(plot.margin = margin(t = 15, r = 0, b = 5, l = 5))
map_pre2 <- map_pre + theme(plot.margin = margin(t = 0, r = 5, b = 0, l = 0))

scatter_early2 <- scatter_early + theme(plot.margin = margin(t = 15, r = 10, b = 5, l = 5))
scatter_zoom_early2 <- scatter_zoom_early + theme(plot.margin = margin(t = 15, r = 0, b = 5, l = 5))
map_early2 <- map_early + theme(plot.margin = margin(t = 0, r = 5, b = 0, l = 0))

scatter_mod2 <- scatter_mod + theme(plot.margin = margin(t = 15, r = 10, b = 5, l = 5))
scatter_zoom_mod2 <- scatter_zoom_mod + theme(plot.margin = margin(t = 15, r = 0, b = 5, l = 5))
map_modern2 <- map_modern + theme(plot.margin = margin(t = 0, r = 5, b = 0, l = 0))



combined_scatter <- plot_grid(
  scatter_pre2, scatter_zoom_pre2, map_pre2,
  scatter_early2, scatter_zoom_early2, map_early2,
  scatter_mod2, scatter_zoom_mod2, map_modern2,
  nrow = 3,
  ncol = 3,
  rel_widths = c(1, 1, 1.7),
  rel_heights = c(1, 1, 1),
  align = "h",
  axis = "tblr",
  labels = c(
    "a) Reference 1420-1800", 
    "b)", "c)", 
    "d) Reference 1801-1970", 
    "e)", "f)", 
    "g) Reference 1971-2009", 
    "h)", "i)"
  ),
  label_size = 14,
  label_x = 0.07,
  hjust = 0,
  label_y = 0.96,
  vjust = 1
)

print(combined_scatter)

ggsave(
  filename = "~/cwd_global/manuscript/Scatter_plots_tvalue.pdf",
  plot     = combined_scatter,
  device   = cairo_pdf,    # better text rendering
  width    = 17,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```


legend for maps: 
```{r}
plot_tvalue_with_hatch <- function(t_column = "t_full", exceed_col = "exceed_pre") {
  shown_sf <- refregions_sf_plot %>%
    filter(is.finite(!!sym(t_column)))
  pts      <- st_point_on_surface(shown_sf) %>% mutate(label = region)

  ggplot() +
    # 1) Base layers (land + t-value fill)
    geom_sf(data = land, fill = "lightgrey", color = NA) +
    geom_sf(
      data = shown_sf,
      aes(fill = !!sym(t_column)),
      color = "grey50", size = 0.2
    ) +

    # 2) Hatch *all* polygons, mapping pattern=exceed_col
    geom_sf_pattern(
      data              = shown_sf,
      aes(pattern = !!sym(exceed_col)),
      fill              = NA,
      color             = NA,
      pattern_fill      = "grey50",
      pattern_colour    = NA,
      pattern_angle     = 45,
      pattern_spacing   = unit(5, "pt"),
      pattern_density   = 0.2,
      size              = 0.3,
      show.legend       = TRUE
    ) +

    # 3) Text labels
    geom_sf_text(data = pts, aes(label = label), size = 3.2) +

    # 4) Continuous t‐value legend
    scale_fill_viridis_c(
      option    = "magma",
      name      = "t-value",
      limits    = c(0, 120),
      alpha     = 0.6,
      direction = -1 
      #guide = guide_legend(override.aes = c(alpha = 1)) 
    ) +

    # 5) Single-entry hatch legend: FALSE → none, TRUE → stripe
    scale_pattern_manual(
      name   = "Exceeds\nreference\nvariability",
      values = c(`FALSE` = "none", `TRUE` = "stripe"),
      breaks = "TRUE",         # only show the TRUE key
     # labels = "exceeding modeled range",
      guide = guide_legend()
    ) +

    # 6) Make that one legend key look right
    guides(
      pattern = guide_legend(
        override.aes = list(
          fill             = NA,
          pattern          = "stripe",
          pattern_fill     = "grey50",
          pattern_colour   = NA,
          pattern_angle    = 45,
          pattern_density  = 0.2
        )
      )
    ) +

    # 7) Coordinate & theme
    coord_sf(
      xlim   = c(-180, 178.125),
      ylim   = c(-60,   90),
      expand = FALSE
    ) +
    theme_void() +
    theme(
      legend.position = "right",
      legend.title    = element_text(face = "bold"), 
      text = element_text(size=13)
    )
}

plot_tvalue_with_hatch("t_full", "exceed_all")


ggsave(
  filename = "figures/Scatter_plots_tvalue_legend.pdf",
  device   = cairo_pdf,    # better text rendering
  width    = 17,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

```

