---
title: "PCWD values from ModE-Sim vs ERA5Land output"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=FALSE}
library(here)
library(readr)
library(dplyr)
library(slider)
library(lubridate)
library(patchwork)
library(ggplot2)
library(ncdf4)
library(reshape2)
library(ggpubr)
library(ggtext)
library(maps)
library(terra)
library(rnaturalearth)
library(sf)
library(sp)
library(RColorBrewer)
library(tidyverse)
library(abind)
library(patchwork)
library(gridExtra)
library(ggsci)
library(purrr)
library(qmap)
library(cowplot)
library(extRemes)
library(ggpattern)

```

## Comparison of ModE-Sim and ERA5-Land PCWD data and bias correction - Regional comparisons 

## Compute regional dataset for ERA5Land data: 

```{r}
### read in IPCC Region information:
#Load reference regions and coastlines:
load(here::here("data", "Reference_regions", "IPCC-WGI-reference-regions-v4_R.rda"))


#simplify this object by converting it to a SpatialPolygons class object (i.e., only the polygons are retained and their attributes discarded):
refregions <- as(IPCC_WGI_reference_regions_v4, "SpatialPolygons")

# List of regions to loop over --- excludes ocean basins
regions <- c("GIC", "NWN", "NEN", "WNA", "CNA", "ENA", "NCA", "SCA", "CAR", "NWS",
             "NSA", "NES", "SAM", "SWS", "SES", "SSA", "NEU", "WCE", "EEU", "MED",
             "SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG", "RAR", "WSB",
             "ESB", "RFE", "WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "NAU",
             "CAU", "EAU", "SAU", "NZ")

# Define continent assignments
continent_mapping <- list(
  "Europe" = c("NEU", "WCE", "EEU", "MED"),
  "North America" = c("NWN", "NEN", "WNA", "CNA", "ENA", "GIC"),
  "Africa" = c("SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG"),
  "Asia" = c("WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "WSB", "ESB", "RFE", "RAR"),
  "Central America" =c("NCA", "SCA", "CAR"),
  "South America" = c("NWS", "NSA", "NES", "SAM", "SWS", "SES", "SSA"),
  "Australia" = c("NAU", "CAU", "EAU", "SAU", "NZ")
)

```

### Load modern ERA5-Land reanalysis data

```{r}

regional_results_ERA5Land<- readRDS(here("data/regionalResults_ERA5Land.RData")) #1950 - 2024
### to exclude potential precipitation problems in ERA5-Land only include years from 1970s onwards
regional_results_ERA5Land <- lapply(regional_results_ERA5Land, function(x) x[21:75])

```


### Bias correct regional ModE-Sim data

#### Apply Quantile Mapping to already regional average data in order to fit distribution better
- fit quantile mapping for an overlap period
- then apply to the rest of the dataset

Apply bias correction for chosen overlap period (21:50) - excludes 60s with potential precipitation problem in ModE-Sim and somewhat allows for stationary climate

```{r}
#---------- Read in raw ModE-Sim regional aggregate data and merge epochs
### read in ModE-Sim regional results
regional_results_1420 <- readRDS(here("data/regionalResults_1420_1.RData"))
regional_results_1850 <- readRDS(here("data/regionalResults_1850_1.RData"))
merged_results <- Map(rbind, regional_results_1420, regional_results_1850) # Stitch matrices together row-wise for each region

### Remove m001_tidy for areas CAU, EAU, NAU, NZ, SAU, RFE, EAS, SEA
# List of regions where the first ensemble member should be removed
regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")

# Loop through each specified region and remove the first column (m001_tidy) - bias over Australia
for (region in regions_to_modify) {
  merged_results[[region]] <- merged_results[[region]][, -1]
}


# saveRDS(regional_results_ERA5Land, file = here::here("data/regional_results_ERA5Land.rds"))


#install.packages("qmap") # uncomment if you haven't installed it
library(qmap)
library(Metrics)

corrected_full <- list()

for (region in names(regional_results_ERA5Land)) {

  obs <- regional_results_ERA5Land[[region]][1:30]  # [1:60] - 30 years (1970-1999)
  sim_overlap <- merged_results[[region]][551:580, ]  # 1970-1999 period for training as 60s have precip problems
  sim_full <- merged_results[[region]]  # 160 years

  corrected_matrix <- matrix(NA, nrow = nrow(sim_full), ncol = ncol(sim_full))
  colnames(corrected_matrix) <- colnames(sim_full)

  for (i in 1:ncol(sim_full)) {  #loop through ensemble members
    mod_train <- sim_overlap[, i]  # 60-year overlap
    mod_full <- sim_full[, i]      # 160-year full run

    # # Fit QM model on 60-year overlap on a parametric transfer function
    qm_model <- fitQmapPTF(obs = obs, mod = mod_train, wet.day = FALSE, qstep = 0.001)
    #
    # # Apply to full 160-year series
    corrected_matrix[, i] <- doQmapPTF(mod_full, qm_model)
  }

  corrected_full[[region]] <- corrected_matrix
}

###save bias corrected ModE-Sim data:
saveRDS(corrected_full, file = here::here("data/regionalResults_ModESim_bc.RData"))

corrected_full <- readRDS(here("data/regionalResults_ModESim_bc.RData"))
```

```{r}
## Bias correction cross-validation: 
library(qmap)
library(Metrics)
library(dplyr)
library(tidyr)
library(ggplot2)

set.seed(123)
n_splits <- 10  # number of CV splits

# Initialize a list to store CV results in long format
cv_long <- list()

for (region in names(regional_results_ERA5Land)) {
  
  obs <- regional_results_ERA5Land[[region]][1:30]
  sim_overlap <- merged_results[[region]][551:580, ]
  
  # Store split results for this region
  region_cv <- data.frame()
  
  for (s in 1:n_splits) {
    # Randomly split 30 years into train (70%) and validation (30%)
    train_idx <- sample(1:30, size = 21)
    val_idx <- setdiff(1:30, train_idx)
    
    obs_train <- obs[train_idx]
    obs_val <- obs[val_idx]
    
    sim_train <- sim_overlap[train_idx, ]
    sim_val <- sim_overlap[val_idx, ]
    
    # Initialize matrices to store RMSE
    rmse_raw <- numeric(ncol(sim_overlap))
    rmse_corrected <- numeric(ncol(sim_overlap))
    
    for (i in 1:ncol(sim_overlap)) {
      # Fit QM on training
      qm_model <- fitQmapPTF(obs = obs_train, mod = sim_train[, i], wet.day = FALSE, qstep = 0.001)
      # Corrected validation
      sim_val_corrected <- doQmapPTF(sim_val[, i], qm_model)
      
      # Compute RMSE for raw and corrected
      rmse_raw[i] <- rmse(obs_val, sim_val[, i])
      rmse_corrected[i] <- rmse(obs_val, sim_val_corrected)
    }
    
    # Combine into long format for this split
    split_df <- data.frame(
      region = region,
      split = s,
      ensemble = 1:ncol(sim_overlap),
      raw = rmse_raw,
      corrected = rmse_corrected
    )
    
    # Pivot longer to have a 'type' column
    split_df_long <- pivot_longer(split_df, cols = c("raw", "corrected"),
                                  names_to = "type", values_to = "rmse")
    
    # Append to region CV
    region_cv <- bind_rows(region_cv, split_df_long)
  }
  
  # Store for all regions
  cv_long[[region]] <- region_cv
}

# Combine all regions into a single data frame
cv_long_df <- bind_rows(cv_long)

regions <- unique(cv_long_df$region)

shade_data <- data.frame(
  region = regions,
  xmin = seq_along(regions) - 0.5,
  xmax = seq_along(regions) + 0.5,
  shade = seq_along(regions) %% 2
)

shade_data <- subset(shade_data, shade == 1)

# Plot RMSE comparison raw vs corrected
RMSE_bc_boxplot <- ggplot(cv_long_df, aes(x = region, y = rmse, fill = type)) +
  # Background alternating bands
  geom_rect(
    data = shade_data,
    inherit.aes = FALSE,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    fill = "grey90"
  ) +
  geom_boxplot(position = position_dodge(width = 0.1), linewidth=0.2, outlier.alpha = 0.4) +
  scale_fill_manual(
    values = c("raw" = "#FFBD00", "corrected" = "#255C99"),
    labels = c("ModE-Sim (raw)", "ModESim* (bc)")  # Or customize as you like
  ) +
  labs(y = "RMSE", x = "Region", fill = "Simulation type",
       title = "") +
  theme_classic()  +
theme(axis.text.x = element_text(angle = 45, hjust = 1, color="black", size=11),
    axis.text.y = element_text(size = 12, color="black"),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    legend.position = "none") + # anchor point of the legend box) +  # Rotate x-axis labels
  labs(title = "",
       y = "RMSE",
       x = "",
       fill = "Data Set")
```

Validation plots: 
- violin plots for each region, showing the values from ModESim with ERA5 overlain to see where the match works well 
- scatterplot where data should lie on 1:1 line if it represents well 


```{r}
# Convert ModESim data to long format
ModESim_long <- bind_rows(lapply(names(merged_results), function(region) {
  mat <- corrected_full[[region]]
  n_members <- ncol(mat)       # number of ensemble members (could be 19 or 20)
  n_years <- length(years)                # 550
  sub <- mat[seq_len(n_years), , drop = FALSE]  # take first 550 rows
  data.frame(
    Year = rep(1970:2009, times = n_members),  # Repeat years for each ensemble member
    Value = as.vector(merged_results[[region]][551:590,]),  # Flatten matrix
    Region = region,
    Source = "ModESim"
  )
}))

# Convert ModESim data to long format
ModESim_long_bc <- bind_rows(lapply(names(corrected_full), function(region) {
  mat <- corrected_full[[region]]
  n_members <- ncol(mat)       # number of ensemble members (could be 19 or 20)
  n_years <- length(years)                # 550
  sub <- mat[seq_len(n_years), , drop = FALSE]  # take first 550 rows
  
  data.frame(
    Year   = rep(1970:2009, times = n_members), # repeat years accordingly
    Value   = as.vector(mat[551:590, ]),           # flatten matrix
    Region = region,
    Source = "ModESim*"
  )
}))

# Convert ERA5Land data to long format (assuming it's stored in a similar list format)
ERA5Land_long <- bind_rows(lapply(names(regional_results_ERA5Land), function(region) {
  data.frame(
    Year = 1970:2009, #change to 2009 for overlap period
    Value = regional_results_ERA5Land[[region]][1:40],
    Region = region,
    Source = "ERA5Land"
  )
}))

# Combine both datasets
combined_data_raw <- bind_rows(ModESim_long, ModESim_long_bc, ERA5Land_long)
#combined_data_bc <- bind_rows(ModESim_long_bc, ERA5Land_long) #bias corrected ModE-Sim

#### colours x-axis
# Define continent assignments
continent_mapping <- list(
  "Europe" = c("NEU", "WCE", "EEU", "MED"),
  "North America" = c("NWN", "NEN", "WNA", "CNA", "ENA", "GIC"),
  "Africa" = c("SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG"),
  "Asia" = c("WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "WSB", "ESB", "RFE", "RAR"),
  "Central America" =c("NCA", "SCA", "CAR"),
  "South America" = c("NWS", "NSA", "NES", "SAM", "SWS", "SES", "SSA"),
  "Australia" = c("NAU", "CAU", "EAU", "SAU", "NZ")
)

continent_colors <- c(
  "Europe" = "#F5B0CB",
  "North America" = "#236A3F",
  "Africa" = "#91C7B1",
  "Asia" = "#E5973B",
  "Central America" = "#B33951",
  "South America" = "#E3D081",
  "Australia" = "#3C3980",
  "Ocean" = "white"
)

# Extract unique regions from your data
unique_regions <- unique(combined_data_bc$Region)

# Initialize a named character vector for storing formatted labels.
region_color_labels <- setNames(rep(NA, length(unique_regions)), unique_regions)


# For each continent, assign the corresponding region labels that match
for (continent in names(continent_mapping)) {
  regions <- continent_mapping[[continent]]
  for (region in regions) {
    if (region %in% unique_regions) {
      # Format the region label with the designated colour from continent_colors
      region_color_labels[region] <- sprintf('<span style="color:%s;">%s</span>', 
                                        continent_colors[continent], 
                                        region)
    }
  }
}

# In case any regions haven't been matched, assign a default colour (e.g. black)
for (region in unique_regions) {
  if (is.na(region_color_labels[region])) {
    region_color_labels[region] <- sprintf('<span style="color:%s;">%s</span>', "black", region)
  }
}


library(ggtext)  # <-- needed to render markdown/html in ggplot

# Create numeric index for Regions
regions <- unique(combined_data_raw$Region)

shade_data <- data.frame(
  Region = regions,
  xmin = seq_along(regions) - 0.5,
  xmax = seq_along(regions) + 0.5,
  shade = seq_along(regions) %% 2  # alternate
)

shade_data <- subset(shade_data, shade == 1)

# Plot the violin plot
bc_violin <- ggplot(combined_data_raw, aes(x = Region, y = Value, fill = Source)) +
    # Add alternating background bands
  geom_rect(
    data = shade_data,
    inherit.aes = FALSE,
    aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
    fill = "grey90"
  ) +
  geom_violin(scale = "width", linewidth=0.2, adjust = .6) +
  theme_classic() +
  scale_fill_manual(
    values = c( "ERA5Land" = "#B3001B", "ModESim" = "#FFBD00", "ModESim*" = "#255C99"),
    labels = c(  "ERA5-Land", "ModE-Sim (raw)", "ModESim* (bc)")  # Or customize as you like
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color="black", size=11),
    axis.text.y = element_text(size = 12, color="black"),
    axis.title.x = element_text(size = 13),
    axis.title.y = element_text(size = 13),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12),
    legend.position = c(0.2, 0.95),  # top-right corner inside plot
    legend.justification = c("right", "top")) + # anchor point of the legend box) +  # Rotate x-axis labels
  labs(title = "",
       y = "PCWD [mm]",
       x = "Region",
       fill = "Data Set")


```

```{r}
## combine bc plots for SI: 
combined_bc_plots <- plot_grid(
  RMSE_bc_boxplot, bc_violin,
  ncol = 1,
  nrow = 2,
  rel_widths = c(1, 1),
  align = "h",
  axis = "tblr",
  labels = c(
    "a)",  "b) "
  )
  # label_size = 14,
  # label_x = 0.09,
  # hjust = 0,
  # label_y = 0.95,
  # vjust = 1
)

combined_bc_plots

ggsave(
  filename = here::here("figures/combined_bc_plots.pdf"),
  plot     = combined_bc_plots,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 7.5,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```


#### Internal variability vs forced signal in ModE-Sim
```{r}
library(tidyverse)

# List of regions where the first ensemble member should be removed
regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")

# Loop through each specified region and remove the first column (m001_tidy) - bias over Australia
for (region in regions_to_modify) {
  corrected_full[[region]] <- corrected_full[[region]][, -1]
}

variability_summary <- map_dfr(names(corrected_full), function(region) {
  mat <- corrected_full[[region]]  # 590 x 20
  
  ensemble_mean <- rowMeans(mat)           # forced signal (590,)
  residuals     <- mat - ensemble_mean     # internal variability (590 x 20)
  
  tibble(
    region        = region,
    mean_pcwd     = mean(ensemble_mean),
    sd_internal   = mean(apply(residuals, 1, sd)),  # avg internal spread
    cv            = sd(rowMeans(mat)) / mean(rowMeans(mat)),
    snr           = mean(ensemble_mean) / mean(apply(residuals, 1, sd))
  )
})

# Define continent assignments
continent_mapping <- list(
  "Europe" = c("NEU", "WCE", "EEU", "MED"),
  "North America" = c("NWN", "NEN", "WNA", "CNA", "ENA", "GIC"),
  "Africa" = c("SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG"),
  "Asia" = c("WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "WSB", "ESB", "RFE", "RAR"),
  "Central America" =c("NCA", "SCA", "CAR"),
  "South America" = c("NWS", "NSA", "NES", "SAM", "SWS", "SES", "SSA"),
  "Australia" = c("NAU", "CAU", "EAU", "SAU", "NZ")
)

continent_colors <- c(
  "Europe" = "#F5B0CB",
  "North America" = "#236A3F",
  "Africa" = "#91C7B1",
  "Asia" = "#E5973B",
  "Central America" = "#B33951",
  "South America" = "#E3D081",
  "Australia" = "#3C3980",
  "Ocean" = "white"
)

# Convert list to a lookup data frame
continent_df <- stack(continent_mapping) %>%
  rename(region = values, continent = ind)

# Join onto your variability_summary
variability_summary <- variability_summary %>%
  left_join(continent_df, by = "region")

# Assuming variability_summary has columns: region, snr, continent
variability_summary %>%
  arrange(snr) %>%
  mutate(region = factor(region, levels = region)) %>%   # fix order
  ggplot(aes(x = snr, y = region, color = continent)) +
  
  # Reference line at SNR = 1 (noise-dominated threshold)
  #geom_vline(xintercept = 5, linetype = "dashed", color = "grey60", linewidth = 0.4) +
  
  # Lollipop stem
  geom_segment(aes(x = 0, xend = snr, y = region, yend = region),
               linewidth = 0.5, alpha = 0.8) +
  
  # Lollipop head
  geom_point(size = 3) +
  
  scale_color_manual(values = continent_colors) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  
  labs(
    x     = "SNR",
    y     = NULL,
    color = "Continent",
  ) +
  
  theme_classic() +
  theme(
    axis.text.y      = element_text(size = 10),
    legend.position  = c(0.8, 0.2),
    legend.title     = element_text(face = "bold"),
    text             = element_text(size = 12),
    axis.title.x     = element_text(margin = margin(t = 10)),
    plot.caption     = element_text(color = "grey50", size = 9)
  )

ggsave(
  filename = here::here("figures/SNR_regions.pdf"),
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 7.5,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```


#### Quantitative analysis 
How many 2000 - 2024 ERA5Land values lie outside the EM range of ModE-Sim (i.e. are unprecedented?)
Check how many regions lie outside of 2SD when considering whole ensemble member range? 

Calculate how many modern PCWD values are truly unprecedented (max(ERA5) > max(ref))
Calculate how many modern PCWD values are record shattering (max(ERA5) > (max(ref)+2sd*(ref))

```{r}
#------------reference values ERA5 - 2000-2024
#Convert ERA5Land data to long format 
ERA5Land_modern <- bind_rows(lapply(names(regional_results_ERA5Land), function(region) {
  data.frame(
    Year = 2000:2024, #change to 2009 for overlap period
    PCWD = regional_results_ERA5Land[[region]][31:55], #indexing year 2000-2024
    Region = region,
    Source = "ERA5Land"
  )
}))


#-----------reference values long context ModESim - 1420-1969
# Convert ModESim data to long format
ModESim_long <- bind_rows(lapply(names(corrected_full), function(region) {
  mat <- corrected_full[[region]]
  n_members <- ncol(mat)       # number of ensemble members (could be 19 or 20)
  n_years <- length(years)                # 550
  sub <- mat[seq_len(n_years), , drop = FALSE]  # take first 550 rows
  
  data.frame(
    Year   = rep(1420:1969, times = n_members), # repeat years accordingly
    PCWD   = as.vector(mat[1:550, ]),           # flatten matrix
    Ensemble = rep(colnames(mat), each = n_years),
    Region = region,
    Source = "ModESim"
  )
}))


#reference values short context ModESim - 1970-1999
ModESim_short <- bind_rows(lapply(names(corrected_full), function(region) {
  mat <- corrected_full[[region]]
  n_members <- ncol(mat)       # number of ensemble members (could be 19 or 20)
  n_years <- length(years)                # 550
  sub <- mat[seq_len(n_years), , drop = FALSE]  # take first 550 rows
  
  data.frame(
    Year   = rep(1970:1999, times = n_members), # repeat years accordingly
    PCWD   = as.vector(mat[551:580, ]),           # flatten matrix
    Ensemble = rep(colnames(mat), each = n_years),
    Region = region,
    Source = "ModESim"
  )
}))

##------------------ Calculate whether PCWD is unprecedented
#Get reference ranges (1420–1969) for each region
ref_range_long <- ModESim_long %>%
  group_by(Region) %>%
  summarise(ref_max = max(PCWD, na.rm = TRUE),
            ref_maxsd = max(PCWD, na.rm = TRUE) + 2 * sd(PCWD, na.rm = TRUE),
            .groups = "drop")

#Get reference ranges (1970–1999) for each region
ref_range_short <- ModESim_short %>%
  group_by(Region) %>%
  summarise(ref_max = max(PCWD, na.rm = TRUE),
            ref_maxsd = max(PCWD, na.rm = TRUE) + 2 * sd(PCWD, na.rm = TRUE),
            .groups = "drop")

# 2. Compare modern period (2000–2024) to reference range
########Long reference
modern_outside_long <- ERA5Land_modern %>%
  left_join(ref_range_long, by = "Region") %>%
  filter(PCWD > ref_max) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_outside_long)
n_regions_outside_long <- nrow(modern_outside_long)
print(paste("In ", n_regions_outside_long, "out of 44 regions, PCWD is unprecendent in the long reference"))

##########Short reference
modern_outside_short <- ERA5Land_modern %>%
  left_join(ref_range_short, by = "Region") %>%
  filter(PCWD > ref_max) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_outside_short)
n_regions_outside_short <- nrow(modern_outside_short)
print(paste("In ", n_regions_outside_short, "out of 44 regions, PCWD is unprecendent in the short reference"))



##------------------ Calculate whether PCWD is "record-shattering"
# 2. Compare modern period (2000–2024) to reference range
########Long reference
modern_shattering_long <- ERA5Land_modern %>%
  left_join(ref_range_long, by = "Region") %>%
  filter(PCWD > ref_maxsd) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_shattering_long)
n_regions_shattering_long <- nrow(modern_shattering_long)
print(paste("In ", n_regions_shattering_long, "out of 44 regions, PCWD is record-shattering in the long reference"))

##########Short reference
modern_shattering_short <- ERA5Land_modern %>%
  left_join(ref_range_short, by = "Region") %>%
  filter(PCWD > ref_maxsd) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_shattering_short)
n_regions_shattering_short <- nrow(modern_shattering_short)
print(paste("In ", n_regions_shattering_short, "out of 44 regions, PCWD is record-shattering in the short reference"))



```
Compute the same for 25-year rolling means - general shift in PCWD distribution? 

```{r}
Modesim_25_long <- ModESim_long %>%
  group_by(Region, Ensemble) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(PCWD25 = slide_dbl(PCWD, mean, .before = 29, .complete = TRUE)) %>%
  ungroup()

Modesim_25_short <- ModESim_short %>%
  group_by(Region, Ensemble) %>%
  arrange(Year, .by_group = TRUE) %>%
  mutate(PCWD25 = slide_dbl(PCWD, mean, .before = 29, .complete = TRUE)) %>%
  ungroup()


ERA5Land_means <- ERA5Land_modern %>%
  group_by(Region) %>%
  summarise(PCWD_mean = mean(PCWD, na.rm = TRUE), .groups = "drop")


#----------------Calculate whether ERA5Land means are above reference running means
#Get reference ranges (1420–1969) for each region
ref_range_long25 <- Modesim_25_long %>%
  group_by(Region) %>%
  summarise(ref_max = max(PCWD25, na.rm = TRUE),
            .groups = "drop")

#Get reference ranges (1970–1999) for each region
ref_range_short25 <- Modesim_25_short %>%
  group_by(Region) %>%
  summarise(ref_max = max(PCWD25, na.rm = TRUE),
            .groups = "drop")

# 2. Compare modern period (2000–2024) to reference range
########Long reference
modern_outside_long25 <- ERA5Land_means %>%
  left_join(ref_range_long25, by = "Region") %>%
  filter(PCWD_mean > ref_max) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_outside_long25)
n_regions_outside_long25 <- nrow(modern_outside_long25)
print(paste("In ", n_regions_outside_long25, "out of 44 regions, mean PCWD is unprecendent in the long reference"))

##########Short reference
modern_outside_short25 <- ERA5Land_means %>%
  left_join(ref_range_short25, by = "Region") %>%
  filter(PCWD_mean > ref_max) %>%
  distinct(Region)

# 3. Count how many regions
print(modern_outside_short25)
n_regions_outside_short25 <- nrow(modern_outside_short25)
print(paste("In ", n_regions_outside_short25, "out of 44 regions, mean PCWD is unprecendent in the short reference"))

```

### Absolute values timeseries of bias corrected ModE-Sim and ERA5-Land - visualisation

Include rolling 25-year anomalies as thin black line with highlighted events

```{r}
# Initialize list to store rolling 30-year means
rolling_means <- list()

# Define window size
window_size <- 25
half_window <- (window_size - 1) / 2

start_year <- 1420
end_year   <- 2009

centers <- seq(start_year, end_year - window_size + 1) + half_window

# Compute rolling means for each ensemble member
for (region_name in names(corrected_full)) {
  
  # Extract ensemble members for the region
  ensemble_members <- corrected_full[[region_name]][2:590,]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means[[region_name]] <- rolling_mean_matrix
}

rolling_means_long <-  do.call(rbind, lapply(names(rolling_means), function(region) {
  df <- as.data.frame(rolling_means[[region]])
  df$time <- centers[1:nrow(df)]
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "rolling_means")
  return(df_long)
}))


```


Calculate 600 year rolling anomalies from bias corrected ModE-Sim data with complete data as reference period
```{r}

# Initialize lists to store baselines
anom_baseline <- list()

# Compute baseline mean from all values of rolling mean
for (region_name in names(rolling_means)) {
  rolling_mean_matrix <- rolling_means[[region_name]]
  
  anom_baseline[[region_name]] <- mean(rolling_mean_matrix, na.rm = TRUE)  ### for reference period of first 25 values: rolling_mean_matrix[1:25,]
}

# Initialize lists to store anomalies
ensemble_25anom <- list()

# Compute anomalies
for (region_name in names(rolling_means)) {
  rolling_mean_matrix <- rolling_means[[region_name]]
  num_years <- nrow(rolling_mean_matrix)
  
  # Initialize matrix to store anomalies
  anoms <- matrix(NA, nrow=num_years, ncol=ncol(rolling_mean_matrix))
  
  # Compute anomalies using rolling mean baseline
  for (t in 1:num_years) {
    anoms[t, ] <- (rolling_mean_matrix[t, ] - anom_baseline[[region_name]])
  }
  
  # Store results
  ensemble_25anom[[region_name]] <- anoms
}

# Compute mean anomalies across ensemble members
mean_25anom <- list()
for (region_name in names(ensemble_25anom)) {
  mean_25anom[[region_name]] <- rowMeans(ensemble_25anom[[region_name]], na.rm = TRUE)
}

# Compute region-specific pooled standard deviation using only the first 25 years
pooled_sd_by_region <- list()
for (region_name in names(ensemble_25anom)) {
  # Extract only the first 25 years of anomalies
  first_25_years <- ensemble_25anom[[region_name]]
  
  # Compute pooled standard deviation over these values
  pooled_sd_by_region[[region_name]] <- sd(unlist(first_25_years), na.rm = TRUE)
}

# Convert ensemble data into a long format for plotting individual members
anoms_long <- do.call(rbind, lapply(names(ensemble_25anom), function(region) {
  df <- as.data.frame(ensemble_25anom[[region]])
  df$time <- centers[1:nrow(df)]
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "anomalies")
  return(df_long)
}))

# Add a continent column
anoms_long$continent <- NA  # Initialize continent column
for (continent in names(continent_mapping)) {
  anoms_long$continent[anoms_long$region %in% continent_mapping[[continent]]] <- continent
}


```


Anomaly timeseries plot with bias corrected data for selected regions: 

```{r}
####-----------------------------------------------------------------------
selected_region <- "MED"
anoms_filtered <- anoms_long %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - pooled_sd,
    upper_bound = first_mean_anomaly + pooled_sd
  )

anom_MED <- ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.4, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "",
    y = "PCWD anomalies [mm] w.r.t 1420-1445"
    #,title = paste("Rolling 25yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15),
    strip.background = element_rect(fill = "#F5B0CB"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11), 
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11)
  ) +
  
  facet_wrap(~region, scales = "free_y")

#######---------------------------------------------------------------------------------
selected_region <- "WNA"
anoms_filtered <- anoms_long %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - pooled_sd,
    upper_bound = first_mean_anomaly + pooled_sd
  )

anom_WNA <- ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "",
    y = ""
    #,title = paste("Rolling 25yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15, color = "white"),
    strip.background = element_rect(fill = "#236A3F"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11), 
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11)
  ) +
  
  facet_wrap(~region, scales = "free_y")

###########------------------------------------------------------------
selected_region <- "ARP"
anoms_filtered <- anoms_long %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - pooled_sd,
    upper_bound = first_mean_anomaly + pooled_sd
  )

anom_ARP <- ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = "PCWD anomalies [mm] w.r.t 1420-1445"
    #,title = paste("Rolling 25yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15),
    strip.background = element_rect(fill = "#E5973B"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 11)
  ) +
  
  facet_wrap(~region, scales = "free_y")

###----------------------------------------------------------------------------
selected_region <- "CAU"
anoms_filtered <- anoms_long %>% filter(region == selected_region)

# Compute mean anomalies per (time, region)
mean_anomalies_filtered <- anoms_filtered %>%
  group_by(time, region) %>%
  summarise(
    mean_anomaly = mean(anomalies, na.rm = TRUE),
    pooled_sd = pooled_sd_by_region[[unique(region)]]
  ) %>%
  ungroup()

# Extract the first year's mean anomaly per region
first_year_mean <- mean_anomalies_filtered %>%
  group_by(region) %>%
  arrange(time) %>%
  slice_head(n = 1) %>%  # Take only the first year
  dplyr::select(region, first_mean_anomaly = mean_anomaly) %>%
  ungroup()

# Join back and compute fixed bounds
mean_anomalies_filtered <- mean_anomalies_filtered %>%
  left_join(first_year_mean, by = "region") %>%
  mutate(
    lower_bound = first_mean_anomaly - pooled_sd,
    upper_bound = first_mean_anomaly + pooled_sd
  )

anom_CAU <- ggplot() +
  geom_hline(yintercept = 0, color= "grey40", linetype="dashed")+

  # Plot individual ensemble members with distinct colors
  geom_line(data = anoms_filtered, aes(x = time, y = anomalies, group = ensemble_member, color = "Ensemble Members"), alpha = 0.5, linewidth = 0.3, show.legend = TRUE) +
  # Plot mean anomaly as a bold black line
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = mean_anomaly, color = "PCWD"), linewidth = 1, show.legend = TRUE) +
  # Dashed lines for ±2σ bounds
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = lower_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  geom_line(data = mean_anomalies_filtered, aes(x = time, y = upper_bound, color = "2σ"), linetype = "dashed", show.legend = TRUE) +
  labs(
    x = "Year",
    y = ""
    #,title = paste("Rolling 25yr PCWD anomalies for", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("2σ" = "red", "Ensemble Members" = "grey60", "PCWD" = "black"),
    labels = c("± 2σ (reference)", "Ensemble Members", "PCWD (Mean)")
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "none",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15,color = "white"),
    strip.background = element_rect(fill = "#3C3980"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 11)
  ) +
  
  facet_wrap(~region, scales = "free_y")
```

Test - ModESim MED annmax PCWD distribution as context for ERA5Land extremes

```{r}
## select MED region data
MED_ERA5 <- regional_results_ERA5Land[["MED"]]

#ModESim data
MED_MODE <- corrected_full[["MED"]]

#attach time dimension
time_info_mode <- seq(from = as.numeric(1420), to = as.numeric(2009))
time_info_era <- seq(from = as.numeric(1970), to = as.numeric(2024))


# Convert ModE dataset to long format
MED_MODE_long <- MED_MODE %>%
  as.data.frame() %>%
  mutate(time = time_info_mode) %>%
  pivot_longer(cols = -time, names_to = "ensemble", values_to = "PCWD")

# Compute means
MED_MODE_mean <- MED_MODE_long %>%
  group_by(time) %>%
  summarise(PCWD = mean(PCWD, na.rm = TRUE), .groups = "drop")

# Convert ERA5 dataset into long format
MED_ERA5_df <- data.frame(time = time_info_era, PCWD = MED_ERA5)

## Calculation return period values 

baseline <- MED_ERA5[1:30] ##1970-1999
fit_ref <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_era_ref_5yr <- round(as.numeric(extRemes::return.level(fit_ref, return.period = 5)))

baseline <- MED_ERA5[31:55] ##1970-1999
fit_mod <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_era_mod_5yr <- round(as.numeric(extRemes::return.level(fit_mod, return.period = 5)))

baseline <- as.vector(MED_MODE[531:560,])  #1970-1999
fit_ref <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_ModESim_ref_5yr <- round(as.numeric(extRemes::return.level(fit_ref, return.period = 5)))

baseline <- as.vector(MED_MODE[1:530,])  #1420-1969
fit_full <- extRemes::fevd(baseline, type = "Gumbel", method = "MLE", units = "years")
MED_ModESim_full_5yr <- round(as.numeric(extRemes::return.level(fit_full, return.period = 5)))


# 1) extract 1970-1999 of the wide matrix, then pivot
MED_last60_long <- MED_MODE[(nrow(MED_MODE)-40):(nrow(MED_MODE)-10), ] %>%
  as.data.frame() %>%
  pivot_longer(cols = everything(),
               names_to  = "ensemble",
               values_to = "PCWD")

# Convert ModE dataset to long format 1420 - 1969
MED_MODE_long_sel <- MED_MODE %>%
  as.data.frame() %>%
  mutate(time = time_info_mode) %>%
  filter(time < 1970) %>%
  pivot_longer(cols = -time, names_to = "ensemble", values_to = "PCWD")

# 1) extract last 60 rows of the wide matrix, then pivot
MED_ERA5_long <- MED_ERA5[1:30] %>%
  as.data.frame() %>%
  pivot_longer(cols = everything(),
              names_to  = "year",
               values_to = "PCWD")

MED_ERA5_last30 <- MED_ERA5[31:55] %>%
  as.data.frame() %>%
  pivot_longer(cols = everything(),
              names_to  = "year",
               values_to = "PCWD")

ggplot() +
  # full‑data density
  geom_density(
    data    = MED_MODE_long_sel,
    aes(x   = PCWD,
    color   = factor("ModESim_fill_full"),
    fill    = factor("ModESim_fill_full")),
    alpha   = 0.8,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
  # last‑60 density
  geom_density(
    data    = MED_last60_long,
    aes(x   = PCWD,
    color   = factor("ModESim_fill"),
    fill    = factor("ModESim_fill")),
    alpha   = 0.8,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
    # first‑60 density
  geom_density(
    data    = MED_ERA5_long,
    aes(x   = PCWD,
    color   = factor("ERA5_fill"),
    fill    = factor("ERA5_fill")),
    alpha   = 0.6,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
  geom_density(
    data    = MED_ERA5_last30,
    aes(x   = PCWD,
    color   = factor("ERA5_fill_last30"),
    fill    = factor("ERA5_fill_last30")),
    alpha   = 0.6,
    linewidth    = 0.6,
    show.legend = TRUE
  ) +
  geom_vline(aes(xintercept = MED_ModESim_full_5yr, color = factor("ModESim_full")), linetype = "dashed", linewidth=0.9) +
  geom_vline(aes(xintercept = MED_ModESim_ref_5yr, color = factor("ModESim_ref")), linetype = "dashed", linewidth=0.9) +
  geom_vline(aes(xintercept = MED_era_ref_5yr, color = factor("ERA_ref")), linetype = "dashed", linewidth=0.9) +
  geom_vline(aes(xintercept = MED_era_mod_5yr, color = factor("ERA_mod")), linetype = "dashed", linewidth=0.9) +
    labs(
    x = "PCWD [mm]",
    y = "Density",
    title = "MED Extreme Value Distribution"
  ) +
  theme_minimal() +
 scale_fill_manual(
    name = "Data Source",
    values = c("ModESim_fill_full" = "#255C99","ModESim_fill" = "#91C499", "ERA5_fill" = "#A25E8E", "ERA5_fill_last30" = "#B3001B") ,
    labels = c("ModE-Sim* (1420-1969)", "ModE-Sim* (1970-1999)", "ERA5-Land (1970-1999)", "ERA5-Land (2000-2024)")
  )+
  scale_color_manual(
    name = "5-year Event",
    values = c("ModESim_full" = "#17385E", "ModESim_ref" = "#628567", "ERA_ref" = "#E8B5CC", "ERA_mod" = "#870014"),
    labels = c("ModE-Sim* (1420 - 1969)", "ModE-Sim* (1970 - 1999)", "ERA5Land (1970 - 1999)", "ERA5Land (2000 - 2024)")
    #guide = guide_legend(override.aes = list(fill= NA,shape = NA, color = "white"))
  )+

    theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.key = element_rect(fill   = NA,    # transparent fill
                              colour = NA),# no border
    legend.title = element_text(face = "bold")
  )


# ggsave(
#   filename = here::here("manuscript/Annmax_PCWD_Timeseries_Density_legend.pdf"),
#   device   = cairo_pdf,    # better text rendering
#   width    = 12,           # in inches
#   height   = 6,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )

```


Function that does the plotting for all regions: 

```{r}
# ------------------------------------------------------------------------------
# Helper: prepare the four slices + FEVD-based 5‑yr return levels
# ------------------------------------------------------------------------------
prepare_region_data <- function(region,
                                corrected_full,
                                regional_results_ERA5Land,
                                mode_years = 1420:2009,
                                era5_years = 1970:2024,
                                drop_initial_years = 2) {
  MODE_mat <- corrected_full[[region]]
  ERA5_vec <- regional_results_ERA5Land[[region]]
  
  # Time‐series data
  MODE_long <- MODE_mat %>%
    as.data.frame() %>%
    mutate(time = mode_years) %>%
    pivot_longer(-time, names_to = "ensemble", values_to = "PCWD")
  
  MODE_mean <- MODE_long %>% group_by(time) %>% summarise(PCWD = mean(PCWD, na.rm=TRUE), .groups="drop")
  
  ERA5_df   <- data.frame(time = era5_years, PCWD = ERA5_vec)
  
  # ———————— drop the first `drop_initial_years` rows ————————
  MODE_long <- MODE_long %>% 
    filter(time >= mode_years[drop_initial_years])
  
  MODE_mean <- MODE_mean %>% 
    filter(time >= mode_years[drop_initial_years])
  # ——————————————————————————————————————————————————————————————
  
  # Four density slices
  MODE_full <- MODE_long %>% filter(time < 1970)
  MODE_ref  <- MODE_long %>% filter(time >= 1970 & time <= 1999)
  ERA5_ref  <- ERA5_df   %>% filter(time >= 1970 & time <= 1999)
  ERA5_mod  <- ERA5_df   %>% filter(time >= 2000)
  
  # FEVD + 5‑yr return level
  fit_rl <- function(x) {
    fit <- fevd(x, type="Gumbel", method="MLE", units="years")
    round(as.numeric(return.level(fit, return.period = 5)))
  }
  rl <- list(
    MODE_full    = fit_rl(MODE_full$PCWD),
    MODE_ref     = fit_rl(MODE_ref$PCWD),
    ERA5_ref     = fit_rl(ERA5_ref$PCWD),
    ERA5_mod     = fit_rl(ERA5_mod$PCWD)
  )
  
  list(
    MODE_long   = MODE_long,
    MODE_mean   = MODE_mean,
    ERA5_df     = ERA5_df,
    MODE_full   = MODE_full,
    MODE_ref    = MODE_ref,
    ERA5_ref    = ERA5_ref,
    ERA5_mod    = ERA5_mod,
    rl          = rl
  )
}

```

```{r}
#Timeseries function alone: 
plot_region_PCWD_ts <- function(region_code) {
  
  dat <- prepare_region_data(region_code,
                             corrected_full,
                             regional_results_ERA5Land)
  
  # Compute anomalies + bounds + exceedance runs
  anoms_filtered <- anoms_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      mean_anomaly = mean(anomalies, na.rm=TRUE),
      pooled_sd    = pooled_sd_by_region[[region_code]],
      .groups = "drop"
    )
  
  first_mean <- anoms_filtered %>%
    arrange(time) %>%
    slice(1) %>%
    pull(mean_anomaly)
  
  anoms_filtered <- anoms_filtered %>%
    mutate(
      lower_bound = first_mean - pooled_sd,
      upper_bound = first_mean + pooled_sd,
      exceeds     = mean_anomaly < lower_bound | mean_anomaly > upper_bound,
      direction   = case_when(
        mean_anomaly > upper_bound ~ "drier",
        mean_anomaly < lower_bound ~ "wetter",
        TRUE ~ NA_character_
      ),
      run_id      = data.table::rleid(exceeds)
    )
  
  exceed_periods <- anoms_filtered %>%
    filter(exceeds) %>%
    group_by(run_id, direction) %>%
    summarise(
      start = min(time),
      end   = max(time),
      .groups="drop"
    )
  
  # Filter rolling means
  rolling_filtered <- rolling_means_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      mean_rolling = mean(rolling_means, na.rm=TRUE),
      .groups = "drop"
    )
  
  # y-limits
  y_limits     <- range(c(dat$MODE_long$PCWD, dat$ERA5_df$PCWD), na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  # Compute max of each ensemble for dotted lines
  ensemble_max <- dat$MODE_long %>%
    group_by(ensemble) %>%
    summarise(max_PCWD = max(PCWD, na.rm=TRUE), .groups = "drop") %>%
    summarise(max_max = max(max_PCWD, na.rm = TRUE), .groups="drop") %>%
    mutate(label = "Ens_maximum")
  
  # Build ggplot
  p_ts <- ggplot()
  
  # Exceedance rectangles colored by direction
  if (nrow(exceed_periods) > 0) {
    p_ts <- p_ts + 
      geom_rect(
        data = exceed_periods,
        aes(xmin = start, xmax = end, 
            ymin = y_limits[1], ymax = y_limits[2],
            fill = direction),
        inherit.aes = FALSE,
        alpha = 0.4
      ) 
  }
  p_ts <- p_ts +
    geom_line(data = dat$MODE_long, aes(time, PCWD, group=ensemble, color = factor("EMs_ModE")), linewidth=0.3, alpha=0.4) +
    geom_line(data = dat$MODE_mean, aes(time, PCWD, color = factor("Mean_ModE")), linewidth = 0.8) +
    geom_line(data = rolling_filtered, aes(time, mean_rolling, color="rolling_mean"), linewidth=0.7) +
    geom_line(data = dat$ERA5_df, aes(time, PCWD, color = factor("ERA5")), linewidth = 0.8) +
    geom_hline(data = ensemble_max, aes(yintercept = max_max, color= factor("Ens_maximum")),linetype="dotted", linewidth=0.5)+
    scale_y_continuous(limits=y_limits, breaks=common_breaks) +
    scale_color_manual(name = "Data",
      values = c(
        "EMs_ModE"  = "#7EA3CC",
        "Mean_ModE" = "#255C99",
        "rolling_mean"="black", 
        "ERA5"      = "#B3001B", 
        "Ens_maximum"= "grey30"
      ),
      labels = c(
        "EMs - ModE-Sim*", "Mean - ModE-Sim*", "25-yr rolling\nmean - ModE-Sim*", "ERA5-Land", "ModE-Sim* - max PCWD")
      ) +
      scale_fill_manual(name ="Exceedances", values = c("wetter" = "#136E40", "drier" = "#F4DE70" ),
      labels = c("wetter", "drier")
    ) +
    labs(x = "Year", y = "PCWD [mm]", color = "Data") +
    theme_classic(base_size = 14) +
    theme(axis.title.y = element_text(margin=margin(r=10)),
          legend.position="none", legend.title = element_text(face="bold"))
  
  return(p_ts)
}

ts_MED <- plot_region_PCWD_ts("MED")
ts_MED
```

```{r}
#density plot function alone: 
plot_region_PCWD_density <- function(region_code) {
  
  dat <- prepare_region_data(region_code,
                             corrected_full,
                             regional_results_ERA5Land)
  
  # y-limits
  y_limits     <- range(c(dat$MODE_long$PCWD, dat$ERA5_df$PCWD), na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  p_den <- ggplot() +
    geom_density(data = dat$MODE_full,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ModESim_full"),
                 alpha=0.8, size=0.6, trim = FALSE) +
    geom_density(data = dat$MODE_ref,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ModESim_ref"),
                 alpha=0.8, size=0.6, trim = FALSE) +
    geom_density(data = dat$ERA5_ref,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ERA5_70_99"),
                 alpha=0.6, size=0.6, trim = FALSE) +
    geom_density(data = dat$ERA5_mod,
                 aes(y=PCWD, x=after_stat(density),
                     fill="ERA5_00_24"),
                 alpha=0.6, size=0.6, trim = FALSE) +
    # 
    # geom_hline(aes(yintercept=dat$rl$MODE_full,  color="ModESim_full_lin"),
    #            linetype="dashed", size=0.8) +
    # geom_hline(aes(yintercept=dat$rl$MODE_ref,   color="ModESim_ref_lin"),
    #            linetype="dashed", size=0.8) +
    # geom_hline(aes(yintercept=dat$rl$ERA5_ref,   color="ERA5_70_99_lin"),
    #            linetype="dashed", size=0.8) +
    # geom_hline(aes(yintercept=dat$rl$ERA5_mod,   color="ERA5_00_24_lin"),
    #            linetype="dashed", size=0.8) +
    
    scale_fill_manual(
      name   = "Period",
      breaks = c("ModESim_full", "ModESim_ref", "ERA5_70_99", "ERA5_00_24"),
      values = c(
        ModESim_full  = "#255C99",
        ModESim_ref   = "#91C499",
        ERA5_70_99    = "#A25E8E",
        ERA5_00_24    = "#B3001B"
      ),
      labels = c(
        "ModE-Sim* (1420-1969)",
        "ModE-Sim* (1970-1999)",
        "ERA5-Land (1970-1999)",
        "ERA5-Land (2000-2024)"
      )
    ) +
    # scale_color_manual(
    #   name   = "5-year Event",
    #   breaks = c("ModESim_full_lin", "ModESim_ref_lin", "ERA5_70_99_lin", "ERA5_00_24_lin"),
    #   values = c(
    #     ModESim_full_lin  = "#17385E",
    #     ModESim_ref_lin   = "#628567",
    #     ERA5_70_99_lin    = "#E8B5CC",
    #     ERA5_00_24_lin    = "#870014"
    #   ),
    #   labels = c(
    #     "ModE-Sim* (1420-1969)",
    #     "ModE-Sim* (1970-1999)",
    #     "ERA5-Land (1970-1999)",
    #     "ERA5-Land (2000-2024)"
    #   ))
    # +
    labs(x="Density", y=NULL) +
    theme_classic(base_size=14) +
    theme(
      plot.title     = element_text(hjust=0.5, face="bold"),
      legend.key     = element_rect(fill=NA, colour=NA),
      legend.title   = element_text(face="bold"),
      text           = element_text(size=14),
      axis.title.y   = element_text(margin=margin(r=10)),
      legend.position = "right"
    )+
    scale_y_continuous(limits = y_limits, breaks = common_breaks)
  
  return(p_den)
}

den_MED <- plot_region_PCWD_density("MED")
den_MED
```
```{r}

##### save legends
# ts_MED
# ggsave(
#   filename = here::here("figures/Annmax_PCWD_Timeseries_legend_wet.pdf"),
#   plot     = ts_MED,
#   device   = cairo_pdf,    # better text rendering
#   width    = 10,           # in inches
#   height   = 8,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )
# 
# den_MED
ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_legend.pdf"),
  plot     = den_MED,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```


Combine multiple panels into one and save: 

```{r}
##for Fig. 1 in paper: 

# Example:
plot_MED_ts <- plot_region_PCWD_ts("MED")
plot_MED_ts <- plot_MED_ts + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_ARP_ts <- plot_region_PCWD_ts("ARP")
plot_WNA_ts <- plot_region_PCWD_ts("WNA")
plot_CAU_ts <- plot_region_PCWD_ts("CAU")

plot_MED_den <- plot_region_PCWD_density("MED")
plot_MED_den <- plot_MED_den + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_ARP_den <- plot_region_PCWD_density("ARP")
plot_WNA_den <- plot_region_PCWD_density("WNA")
plot_CAU_den <- plot_region_PCWD_density("CAU")


combined2all <- plot_grid(
  plot_MED_ts,plot_MED_den, plot_WNA_ts, plot_WNA_den, plot_ARP_ts,plot_ARP_den, plot_CAU_ts,
     plot_CAU_den,
  nrow        = 4,
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) MED – Mediterranean",  "e)", 
    "b) WNA – Western North America", "f)",
    "c) ARP – Arabian Peninsula", "g)",
    "d) CAU – Central Australia","h)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(2, 1),
  expand      = TRUE  
)

print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_all_Fig.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```

Plot by continent for supplementary material: 

```{r}

##EUROPE
plot_NEU_ts <- plot_region_PCWD_ts("NEU")
plot_NEU_ts <- plot_NEU_ts + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_WCE_ts <- plot_region_PCWD_ts("WCE")
plot_EEU_ts <- plot_region_PCWD_ts("EEU")
plot_MED_ts <- plot_region_PCWD_ts("MED")

plot_NEU_den <- plot_region_PCWD_density("NEU")
plot_NEU_den <- plot_NEU_den + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_WCE_den <- plot_region_PCWD_density("WCE")
plot_EEU_den <- plot_region_PCWD_density("EEU")
plot_MED_den <- plot_region_PCWD_density("MED")

combined2all <- plot_grid(
  plot_NEU_ts, plot_NEU_den, plot_WCE_ts,plot_WCE_den, plot_EEU_ts, plot_EEU_den,plot_MED_ts,plot_MED_den,
  nrow        = 4,
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) EUROPE - NEU", "e)",
    "b) EUROPE - WCE", "f)",
    "c) EUROPE - EEU", "g)",
    "d) EUROPE - MED" , "h)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(2, 1)
)

print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_all_Europe.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

###NORTH AMERICA
plot_NWN <- plot_region_PCWD_ts("NWN")
plot_NWN <- plot_NWN+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_NEN <- plot_region_PCWD_ts("NEN")
plot_WNA <- plot_region_PCWD_ts("WNA")
plot_CNA <- plot_region_PCWD_ts("CNA")
plot_ENA <- plot_region_PCWD_ts("ENA")
plot_GIC <- plot_region_PCWD_ts("GIC")

plot_NWN_den <- plot_region_PCWD_density("NWN")
plot_NWN_den <- plot_NWN_den + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_NEN_den <- plot_region_PCWD_density("NEN")
plot_WNA_den <- plot_region_PCWD_density("WNA")
plot_CNA_den <- plot_region_PCWD_density("CNA")
plot_ENA_den <- plot_region_PCWD_density("ENA")
plot_GIC_den <- plot_region_PCWD_density("GIC")


combined2all <- plot_grid(
  plot_NWN,plot_NWN_den, plot_NEN,plot_NEN_den, plot_WNA,plot_WNA_den, plot_CNA,plot_CNA_den,plot_ENA, plot_ENA_den,plot_GIC,plot_GIC_den,
  nrow        = 6,
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) NORTH AMERICA - NWN", "g)",
    "b) NORTH AMERICA - NEN", "h)",
    "c) NORTH AMERICA - WNA", "i)",
    "d) NORTH AMERICA - CNA", "j)",
    "e) NORTH AMERICA - ENA", "k)",
    "f) NORTH AMERICA - GIC", "l)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.978,   # just below the top
  vjust   = 1,       # align text at its top,
  rel_widths = c(2, 1)
)

print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_all_NorthAmerica.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


#### AFRICA
plot_SAH <- plot_region_PCWD_ts("SAH")
plot_SAH <- plot_SAH+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_WAF <- plot_region_PCWD_ts("WAF")
plot_CAF <- plot_region_PCWD_ts("CAF")
plot_NEAF <- plot_region_PCWD_ts("NEAF")
plot_SEAF <- plot_region_PCWD_ts("SEAF")
plot_WSAF <- plot_region_PCWD_ts("WSAF")
plot_ESAF <- plot_region_PCWD_ts("ESAF")
plot_MDG <- plot_region_PCWD_ts("MDG")

plot_SAH_den <- plot_region_PCWD_density("SAH")
plot_SAH_den <- plot_SAH_den+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_WAF_den <- plot_region_PCWD_density("WAF")
plot_CAF_den <- plot_region_PCWD_density("CAF")
plot_NEAF_den <- plot_region_PCWD_density("NEAF")
plot_SEAF_den <- plot_region_PCWD_density("SEAF")
plot_WSAF_den <- plot_region_PCWD_density("WSAF")
plot_ESAF_den <- plot_region_PCWD_density("ESAF")
plot_MDG_den <- plot_region_PCWD_density("MDG")



combined2all <- plot_grid(
  plot_SAH,plot_SAH_den, plot_WAF, plot_WAF_den,plot_CAF,plot_CAF_den, plot_NEAF,plot_NEAF_den,
  plot_SEAF, plot_SEAF_den,plot_WSAF,plot_WSAF_den,plot_ESAF,plot_ESAF_den, plot_MDG,plot_MDG_den,
  nrow        = 8,
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) AFRICA - SAH", "i)",
    "b) AFRICA - WAF", "j)",
    "c) AFRICA - CAF", "k)",
    "d) AFRICA - NEAF", "l)",
    "e) AFRICA - SEAF", "m)",
    "f) AFRICA - WSAF", "n)",
    "g) AFRICA - ESAF", "o)",
    "h) AFRICA - MDG", "p)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(2, 1)
)

print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_all_Africa.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 17,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


#### Asia
plot_WCA <- plot_region_PCWD_ts("WCA")
plot_WCA <- plot_WCA+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_ARP <- plot_region_PCWD_ts("ARP")
plot_WSB <- plot_region_PCWD_ts("WSB")
plot_RAR <- plot_region_PCWD_ts("RAR")
plot_ESB <- plot_region_PCWD_ts("ESB")

plot_WCA_den <- plot_region_PCWD_density("WCA")
plot_WCA_den <- plot_WCA_den+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_ARP_den <- plot_region_PCWD_density("ARP")
plot_WSB_den <- plot_region_PCWD_density("WSB")
plot_RAR_den <- plot_region_PCWD_density("RAR")
plot_ESB_den <- plot_region_PCWD_density("ESB")


combined2all <- plot_grid(
  plot_WCA, plot_WCA_den,plot_ARP,plot_ARP_den, plot_WSB, plot_WSB_den,plot_RAR,plot_RAR_den,plot_ESB,plot_ESB_den,
  nrow        = 5,
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) ASIA - WCA", "f)",
    "b) ASIA - ARP", "g)",
    "c) ASIA - WSB", "h)",
    "d) ASIA - RAR", "i)",
    "e) ASIA - ESB", "j)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1  ,     # align text at its top
rel_widths = c(2, 1)
)

print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_all_Asia1.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


###ASIA continued
plot_ECA <- plot_region_PCWD_ts("ECA")
plot_ECA <- plot_ECA+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_TIB <- plot_region_PCWD_ts("TIB")
plot_SAS <- plot_region_PCWD_ts("SAS")
plot_EAS <- plot_region_PCWD_ts("EAS")
plot_RFE <- plot_region_PCWD_ts("RFE")
plot_SEA <- plot_region_PCWD_ts("SEA")

plot_ECA_den <- plot_region_PCWD_density("ECA")
plot_ECA_den <- plot_ECA_den+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_TIB_den <- plot_region_PCWD_density("TIB")
plot_SAS_den <- plot_region_PCWD_density("SAS")
plot_EAS_den <- plot_region_PCWD_density("EAS")
plot_RFE_den <- plot_region_PCWD_density("RFE")
plot_SEA_den <- plot_region_PCWD_density("SEA")

combined2all <- plot_grid(
  plot_ECA, plot_ECA_den, plot_TIB,plot_TIB_den, plot_SAS,plot_SAS_den, plot_EAS,plot_EAS_den, plot_RFE,plot_RFE_den,plot_SEA,plot_SEA_den,
  nrow        = 6,
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) ASIA - ECA", "g)",
    "b) ASIA - TIB", "h)",
    "c) ASIA - SAS", "i)",
    "d) ASIA - EAS", "j)",
    "e) ASIA - RFE", "k)",
    "f) ASIA - SEA", "l)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1 ,      # align text at its top
  rel_widths = c(2, 1)
)

print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_all_Asia2.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

####CENTRAL AMERICA
plot_NCA <- plot_region_PCWD_ts("NCA")
plot_NCA <- plot_NCA+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_SCA <- plot_region_PCWD_ts("SCA")
plot_CAR <- plot_region_PCWD_ts("CAR")

plot_NCA_den <- plot_region_PCWD_density("NCA")
plot_NCA_den <- plot_NCA_den+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_SCA_den <- plot_region_PCWD_density("SCA")
plot_CAR_den <- plot_region_PCWD_density("CAR")

combined2all <- plot_grid(
  plot_NCA, plot_NCA_den,plot_SCA,plot_SCA_den, plot_CAR,plot_CAR_den,
  nrow        = 3,
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) CENTRAL AMERICA - NCA", "d)",
    "b) CENTRAL AMERICA - SCA", "e)",
    "c) CENTRAL AMERICA - CAR", "f)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
rel_widths = c(2, 1)
)

print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_all_CentralAmerica.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 11,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


###SOUTH AMERICA
plot_NWS <- plot_region_PCWD_ts("NWS")
plot_NWS <- plot_NWS+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_NSA <- plot_region_PCWD_ts("NSA")
plot_NES <- plot_region_PCWD_ts("NES")
plot_SAM <- plot_region_PCWD_ts("SAM")
plot_SWS <- plot_region_PCWD_ts("SWS")
plot_SES <- plot_region_PCWD_ts("SES")
plot_SSA <- plot_region_PCWD_ts("SSA")

plot_NWS_den <- plot_region_PCWD_density("NWS")
plot_NWS_den <- plot_NWS_den+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_NSA_den <- plot_region_PCWD_density("NSA")
plot_NES_den <- plot_region_PCWD_density("NES")
plot_SAM_den <- plot_region_PCWD_density("SAM")
plot_SWS_den <- plot_region_PCWD_density("SWS")
plot_SES_den <- plot_region_PCWD_density("SES")
plot_SSA_den <- plot_region_PCWD_density("SSA")


combined2all <- plot_grid(
  plot_NWS,plot_NWS_den, plot_NSA, plot_NSA_den,plot_NES,plot_NES_den,plot_SAM,plot_SAM_den, plot_SWS,plot_SWS_den, plot_SES,plot_SES_den, plot_SSA,plot_SSA_den,
  nrow        = 7,
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) SOUTH AMERICA - NWS", "h)",
    "b) SOUTH AMERICA - NSA", "i)",
    "c) SOUTH AMERICA - NES", "j)",
    "d) SOUTH AMERICA - SAM", "k)",
    "e) SOUTH AMERICA - SWS", "l)",
    "f) SOUTH AMERICA - SES", "m)",
    "g) SOUTH AMERICA - SSA", "n)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(2, 1)
)

print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_all_SouthAmerica.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 17,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)


###AUSTRALIA
plot_NAU <- plot_region_PCWD_ts("NAU")
plot_NAU <- plot_NAU+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_CAU <- plot_region_PCWD_ts("CAU")
plot_EAU <- plot_region_PCWD_ts("EAU")
plot_SAU <- plot_region_PCWD_ts("SAU")
plot_NZ <- plot_region_PCWD_ts("NZ")

plot_NAU_den <- plot_region_PCWD_density("NAU")
plot_NAU_den <- plot_NAU_den+ theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
plot_CAU_den <- plot_region_PCWD_density("CAU")
plot_EAU_den <- plot_region_PCWD_density("EAU")
plot_SAU_den <- plot_region_PCWD_density("SAU")
plot_NZ_den <- plot_region_PCWD_density("NZ")

combined2all <- plot_grid(
  plot_NAU, plot_NAU_den,plot_CAU,plot_CAU_den, plot_EAU, plot_EAU_den,plot_SAU,plot_SAU_den, plot_NZ,plot_NZ_den,
  nrow        = 5,
  ncol        = 2,
  align       = "h",
  axis        = "tb",
  labels      = c(
    "a) AUSTRALIA - NAU", "f)",
    "b) AUSTRALIA - CAU", "g)",
    "c) AUSTRALIA - EAU", "h)",
    "d) AUSTRALIA - SAU", "i)",
    "e) AUSTRALIA - NZ", "j)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.0,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(2, 1)
)

print(combined2all)

ggsave(
  filename = here::here("figures/Annmax_PCWD_Timeseries_Density_all_Australia.pdf"),
  plot     = combined2all,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

```

## Assessing the change in frequency of an ERA5 30-year drought events (Moderate drought) for context

```{r functions for RP calc}
# From https://geco-bern.github.io/cwd/articles/cwd_example.html: 
# calculate return period as a function of the CWD extreme. Using the two 
# coefficients of the fitted distribution as arguments
calc_return_period <- function(x, loc, scale){
  1 / (1 - exp(-exp(-(x-loc)/scale)))
}

extract_loc <- function(mod){
  loc <- mod$results$par[ "location" ]
  if (!is.null(loc)){
    return(loc)
  } else {
    return(NA)
  }
}

extract_scale <- function(mod){
  scale <- mod$results$par[ "scale" ]
  if (!is.null(scale)){
    return(scale)
  } else {
    return(NA)
  }
}
```


#### rolling effective RP for all regions
calculate 30 yr event return periods
```{r}
# #################### Calc for ModE-SIm ######################################
# use_rp <- 30    # return period of interest
# nyears <- 25   # window size
# 
# path <- here::here("data/volc_impact_aggregates/collected_p_pet")
# folders <- list.files(path, full.names = TRUE, pattern = "m[0-9]{3}") #m001 - m020 in first set
# 
# # Extract unique ensemble member identifiers from folder names
# n_ens <- unique(basename(folders))[1:20]
# # Add "_tidy" to each element
# n_ens <- paste0(n_ens, "_tidy")
# all_members <- n_ens
# 
# # Store output
# EVA_regional_ensemble <- list()
# 
# for (region in names(corrected_full)) {
# 
#   # Full matrix for this region
#   result_array_region <- corrected_full[[region]]  # [590, 20]
# 
#   # Determine which ensemble members to include
#   if (region %in% regions_to_modify) {
#     ens_cols <- all_members[all_members != "m001_tidy"]
#   } else {
#     ens_cols <- all_members
#   }
# 
#   # Baseline: last 30 years, pooled across selected members
#   result_array_region_BL <- result_array_region[551:580, ens_cols]
#   baseline_pooled <- as.vector(result_array_region_BL)
#   fit_ref <- fevd(baseline_pooled, type = "Gumbel", method = "MLE", units = "years")
#   cwd_ref <- return.level(fit_ref, return.period = use_rp)
# 
#   # Initialize return period vector (pooled across ensemble)
#   rp_window <- rep(NA, nrow(result_array_region) - nyears + 1)
# 
#   for (window_start in 1:(nrow(result_array_region) - nyears + 1)) {
#     # Pool all selected ensemble members in this window
#     window_data <- as.vector(result_array_region[window_start:(window_start + nyears - 1), ens_cols])
#     
#     # Fit Gumbel distribution to pooled window
#     fit_window <- fevd(window_data, type = "Gumbel")
#     
#     # Calculate return period of baseline event
#     rp_window[window_start] <- calc_return_period(cwd_ref,
#                                                   extract_loc(fit_window),
#                                                   extract_scale(fit_window))
#   }
# 
#   # Time vector
#   time_info <- seq(from = 1420, to = 2009)
#   region_df <- data.frame(
#     time = time_info[12:577],  # adjust to match rolling windows
#     RP = rp_window,
#     region = region
#   )
# 
#   EVA_regional_ensemble[[region]] <- region_df
# }
# 
# # Combine all regions
# combined_results_25yr_pooled <- do.call(rbind, EVA_regional_ensemble)
# # Save to file
# write.csv(combined_results_25yr_pooled, file = here::here("data/25yr_pooled_return_periods_regions_ERA_members_test.csv"), row.names = FALSE) #ModESim all ensemble members

```

```{r}
### test confidence interval estimation for ModE-Sim data also: 
use_rp <- 30
nyears <- 25
path <- here::here("data/volc_impact_aggregates/collected_p_pet")
folders <- list.files(path, full.names = TRUE, pattern = "m[0-9]{3}")
n_ens <- unique(basename(folders))[1:20]
n_ens <- paste0(n_ens, "_tidy")
all_members <- n_ens

EVA_regional_ensemble <- list()

for (region in names(corrected_full)) {
  
  result_array_region <- corrected_full[[region]]  # [590, 20]
  
  # Determine which ensemble members to include
  if (region %in% regions_to_modify) {
    ens_cols <- all_members[all_members != "m001_tidy"]
  } else {
    ens_cols <- all_members
  }
  
  # Baseline: pooled across selected members
  result_array_region_BL <- result_array_region[551:580, ens_cols]
  baseline_pooled <- as.vector(result_array_region_BL)
  fit_ref <- fevd(baseline_pooled, type = "Gumbel", method = "MLE", units = "years")
  cwd_ref <- return.level(fit_ref, return.period = use_rp)
  
  n_windows <- nrow(result_array_region) - nyears + 1
  rp_window <- rep(NA, n_windows)
  RP_lower  <- rep(NA, n_windows)
  RP_upper  <- rep(NA, n_windows)
  
  for (window_start in 1:n_windows) {
    
    # Pool all selected ensemble members in this window
    window_data <- as.vector(result_array_region[window_start:(window_start + nyears - 1), ens_cols])
    n_pool      <- length(window_data)  # nyears * n_ensemble_members
    
    fit_window <- tryCatch(
      fevd(window_data, type = "Gumbel", method = "MLE"),
      error = function(e) NULL
    )
    if (is.null(fit_window)) next
    
    loc_hat   <- extract_loc(fit_window)
    scale_hat <- extract_scale(fit_window)
    
    # Point estimate
    rp_window[window_start] <- calc_return_period(cwd_ref, loc_hat, scale_hat)
    
    # Parametric bootstrap
    R       <- 250
    boot_rp <- numeric(R)
    
    for (b in seq_len(R)) {
      # Draw bootstrap sample of same size as pooled window (nyears * n_members)
      boot_sample <- revd(n_pool, loc = loc_hat, scale = scale_hat, shape = 0, type = "GEV")
      
      fit_boot <- tryCatch(
        fevd(boot_sample, type = "Gumbel", method = "MLE"),
        error = function(e) NULL
      )
      if (is.null(fit_boot)) { boot_rp[b] <- NA; next }
      
      boot_rp[b] <- calc_return_period(cwd_ref, extract_loc(fit_boot), extract_scale(fit_boot))
    }
    
    RP_lower[window_start] <- quantile(boot_rp, 0.025, na.rm = TRUE)
    RP_upper[window_start] <- min(quantile(boot_rp, 0.975, na.rm = TRUE), 200)
  }
  
  time_info <- seq(from = 1420, to = 2009)

  window_size <- 25
  half_window <- (window_size - 1) / 2  # 12

  n_out <- length(time_info) - window_size + 1

  region_df <- data.frame(
    time     = time_info[(1:n_out) + half_window], #midpoints
    RP       = rp_window,
    RP_lower = RP_lower,
    RP_upper = RP_upper,
    region   = region
  )
  
  EVA_regional_ensemble[[region]] <- region_df
}

combined_results_25yr_pooled <- do.call(rbind, EVA_regional_ensemble)
write.csv(combined_results_25yr_pooled, 
          file = here::here("data/25yr_pooled_return_periods_regions_ERA_members_bootstrap.csv"), 
          row.names = FALSE)

```



```{r}
## test chunk to test different uncertainty sampling
use_rp <- 30    # return period of two years
nyears <- 25   # bin width

# Initialize a list to store results for all regions
EVA_regional <- list()

for (region in names(regional_results_ERA5Land)) {
  
  # Extract the result array for the current region
  result_array_region_BL <- corrected_full[[region]][551:580, ]  # baseline [time, ensemble]
  result_array_region <- regional_results_ERA5Land[[region]]     # full time series [time, ensemble]
  
  # ----- Fit pooled baseline -----
  baseline_pooled <- as.vector(result_array_region_BL)
  fit_ref <- fevd(baseline_pooled, type = "Gumbel", method = "MLE", units = "years")
  cwd_ref <- return.level(fit_ref, return.period = use_rp)
  
  # Initialize storage vectors
  rp_window <- rep(NA, length(result_array_region) - nyears + 1)
  RP_lower <- rep(NA, length(result_array_region) - nyears + 1)
  RP_upper <- rep(NA, length(result_array_region) - nyears + 1)
  
  for (window_start in 1:(length(result_array_region) - nyears + 1)) {
  
    window_data <- result_array_region[window_start:(window_start + nyears - 1)]
    window_data <- as.vector(window_data)
    fit_window  <- fevd(window_data, type = "Gumbel", method = "MLE")
    
    loc_hat   <- extract_loc(fit_window)
    scale_hat <- extract_scale(fit_window)
    
    # Point estimate
    rp_window[window_start] <- calc_return_period(cwd_ref, loc_hat, scale_hat)
    
    # --- Parametric bootstrap over the WINDOW fit ---
    # Sample new parameter sets from the fitted Gumbel, then ask:
    # "what is the return period of cwd_ref under each bootstrap replicate?"
    R <- 250
    boot_rp <- numeric(R)
    
    n_pool <- length(window_data)   # same sample size as the original window
    
    for (b in seq_len(R)) {
      # Draw a bootstrap sample from the fitted Gumbel
      boot_sample <- revd(n_pool, loc = loc_hat, scale = scale_hat, shape = 0, type = "GEV")
      
      # Refit Gumbel to bootstrap sample
      fit_boot <- tryCatch(
        fevd(boot_sample, type = "Gumbel", method = "MLE"),
        error = function(e) NULL
      )
      if (is.null(fit_boot)) {
        boot_rp[b] <- NA
        next
      }
      
      # Compute return period of the FIXED baseline event under this bootstrap fit
      boot_rp[b] <- calc_return_period(
        cwd_ref,
        extract_loc(fit_boot),
        extract_scale(fit_boot)
      )
    }
    
    # 95% CI from bootstrap distribution
    RP_lower[window_start] <- quantile(boot_rp, 0.025, na.rm = TRUE)
    RP_upper[window_start] <- min(quantile(boot_rp, 0.975, na.rm = TRUE), 200) #set upper limit of return period
  }
  
    # Time vector (match rolling windows)
  time_info <- seq(from = 1970, to = 2024)
  window_size <- 25
  half_window <- (window_size - 1) / 2  # 12

  n_out <- length(time_info) - window_size + 1
  region_df <- data.frame(
    time = time_info[(1:n_out) + half_window],
    RP = rp_window,
    RP_lower = RP_lower,
    RP_upper = RP_upper,
    region = region
  )
  
  EVA_regional[[region]] <- region_df
}

# Combine all regions into one dataframe
combined_results_25yr_ERA <- do.call(rbind, EVA_regional)

# Save or process the combined dataframe
# Example: Save to a CSV file
write.csv(combined_results_25yr_ERA, here::here("data/25yr_pooled_return_periods_regions_ERA5Land_bootstrap.csv"), row.names = FALSE)


```


```{r}
# # Test with one window before running everything
# some_window_data <- as.vector(regional_results_ERA5Land$WNA[21:50])
# test_fit <- fevd(some_window_data, type = "Gumbel", method = "MLE")
# ci_100  <- ci(test_fit, type = "return.level", return.period = 30, method = "boot", R = 100)
# ci_250  <- ci(test_fit, type = "return.level", return.period = 30, method = "boot", R = 250)
# ci_500  <- ci(test_fit, type = "return.level", return.period = 30, method = "boot", R = 500)
# ci_1000 <- ci(test_fit, type = "return.level", return.period = 30, method = "boot", R = 1000)
# print(rbind(ci_100,ci_250, ci_500, ci_1000))
```



test: 
```{r}
library(ggplot2)
library(dplyr)

# Filter for region MED
med_data <- combined_results_25yr_ERA %>% 
  filter(region == "MED")

# Plot
ggplot(med_data, aes(x = time, y = RP)) +
  geom_line(color = "blue", linewidth = 1) +                        # main RP line
  geom_ribbon(aes(ymin = RP_lower, ymax = RP_upper),           # uncertainty band
              fill = "lightblue", alpha = 0.3) +
  labs(
    title = "Rolling 30-Year Return Period for MED",
    x = "Year",
    y = "Return Period (years)"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5)
  )

```

```{r}
#### ModE-Sim test plot: 

# Filter for region MED
med_data <- combined_results_25yr_pooled %>% 
  filter(region == "MED")

# Plot
ggplot(med_data, aes(x = time, y = RP)) +
  geom_line(color = "blue", linewidth = 1) +                        # main RP line
  geom_ribbon(aes(ymin = RP_lower, ymax = RP_upper),           # uncertainty band
              fill = "lightblue", alpha = 0.3) +
  labs(
    title = "Rolling 25-Year Return Period for MED - ModE-Sim",
    x = "Year",
    y = "Return Period (years)"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    plot.title = element_text(hjust = 0.5)
  )
```

### Frequency changes of 30-yr return period events for all IPCC regions (1420 - 2009 epoch):

```{r timeseries and map plots for each region, echo=FALSE, warning=FALSE, message=FALSE}
#plot all ensemble members and ensemble mean: 
combined_df <- read.csv(here::here("data/25yr_pooled_return_periods_regions_ERA_members_bootstrap.csv")) #ModESim all ensemble members
combined_era <- read.csv(here::here("data/25yr_pooled_return_periods_regions_ERA5Land_bootstrap.csv"))


####################################### EUROPE #####################################
##### Step 2: Filter European regions for the map and timeseries
# Filter regions for Europe
selected_continent <- "Europe"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "MED"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) 
# %>%
  # summarise(
  #   RP_mean = mean(RP, na.rm = TRUE),
  #   RP_lower = quantile(RP, 0.05, na.rm = TRUE),
  #   RP_upper = quantile(RP, 0.95, na.rm = TRUE),
  #   .groups = "drop"
  # )

ggplot() +
  # Uncertainty band with legend
  geom_ribbon(data = combined_df_sel, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = combined_df_sel, aes(x = time, y = RP, color = "ModE-Sim* (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  # Uncertainty band with legend
  geom_ribbon(data = combined_era_sel, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 30, linetype = "dashed", color = "grey40") +
  # Labels and theme
  labs(
    x = "Year",
    y = "Effective Return Period",
    title = "Frequency Change of 30-Year Events",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim* (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("95% CI" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 18),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18), 
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    strip.background = element_rect(fill = "#F5B0CB")
  ) +
  facet_wrap(~region, scales = "free_y")


######################################ASIA#######################################
##### Step 2: Filter Asian regions for the map and timeseries
# Filter regions for Asian
selected_continent <- "Asia"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "ARP"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.05, na.rm = TRUE),
#     RP_upper = quantile(RP, 0.95, na.rm = TRUE),
#     .groups = "drop"
#   )

ggplot() +
  # Uncertainty band with legend
  geom_ribbon(data = combined_df_sel, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = combined_df_sel, aes(x = time, y = RP, color = "ModE-Sim* (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  # Uncertainty band with legend
  geom_ribbon(data = combined_era_sel, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 30, linetype = "dashed", color = "grey40") +
  # Labels and theme
  labs(
    x = "Year",
    y = "Effective Return Period",
    title = "Frequency Change of 30-Year Events",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim* (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("95% CI" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 18),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18), 
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    strip.background = element_rect(fill = "#E5973B"),
  ) +
  facet_wrap(~region, scales = "free_y")


###################################### Australia #######################################
##### Step 2: Filter Australian regions for the map and timeseries
# Filter regions for Australia
selected_continent <- "Australia"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "CAU"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.05, na.rm = TRUE),
#     RP_upper = quantile(RP, 0.95, na.rm = TRUE),
#     .groups = "drop"
#   )

ggplot() +
  # Uncertainty band with legend
  geom_ribbon(data = combined_df_sel, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = combined_df_sel, aes(x = time, y = RP, color = "ModE-Sim* (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  # Uncertainty band with legend
  geom_ribbon(data = combined_era_sel, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 30, linetype = "dashed", color = "grey40") +
  # Labels and theme
  labs(
    x = "Year",
    y = "Effective Return Period",
    title = "Frequency Change of 30-Year Events",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim* (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("95% CI" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14, color = "white"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18), 
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    strip.background = element_rect(fill = "#3C3980")
  ) +
  facet_wrap(~region, scales = "free_y")

# ###################################### AFRICA #######################################
# ##### Step 2: Filter African regions for the map and timeseries
# # Filter regions for Africa
# selected_continent <- "Africa"
# sel_regions <- continent_mapping[[selected_continent]]
# #sel_regions <- "NEAF"
# 
# # Filter combined data for timeseries 
# combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
# combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.05, na.rm = TRUE),
#     RP_upper = quantile(RP, 0.95, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# ggplot() +
#   # Uncertainty band with legend
#   geom_ribbon(data = ensemble_band_df, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
#   # Ensemble mean line with legend
#   geom_line(data = ensemble_band_df, aes(x = time, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
#   # ERA5 line with legend
#   geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
#   # Reference line (no legend)
#   geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
#   # Labels and theme
#   labs(
#     x = "Year",
#     y = "Effective Return Period",
#     title = "Frequency Change of 5-Year Events",
#     color = "Line",  # Legend titles
#     fill = "Shading"
#   ) +
#   scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
#   scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
#   theme_classic() +
#   theme(
#     legend.position = "none",
#     plot.title = element_text(hjust = 0.5, face = "bold"),
#     strip.text = element_text(size = 18),
#     axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
#     axis.text.y = element_text(size = 18), 
#     axis.title.x = element_text(size = 18),
#     axis.title.y = element_text(size = 18),
#     legend.title = element_text(size = 14),
#     legend.text = element_text(size = 12),
#     strip.background = element_rect(fill = "#91C7B1"),
#   ) +
#   facet_wrap(~region, scales = "free_y")


###################################### N AMERICA #######################################
##### Step 2: Filter N AMERICA regions for the map and timeseries
# Filter regions for N America
selected_continent <- "North America"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "WNA"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.05, na.rm = TRUE),
#     RP_upper = quantile(RP, 0.95, na.rm = TRUE),
#     .groups = "drop"
#   )

ggplot() +
  # Uncertainty band with legend
  geom_ribbon(data = combined_df_sel, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = combined_df_sel, aes(x = time, y = RP, color = "ModE-Sim* (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  # Uncertainty band with legend
  geom_ribbon(data = combined_era_sel, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 30, linetype = "dashed", color = "grey40") +
  # Labels and theme
  labs(
    x = "Year",
    y = "Effective Return Period",
    title = "Frequency Change of 30-Year Events",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim* (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("95% CI" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 14, color= "white"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
    axis.text.y = element_text(size = 18), 
    axis.title.x = element_text(size = 18),
    axis.title.y = element_text(size = 18),
    legend.title = element_text(size = 14),
    legend.text = element_text(size = 12),
    strip.background = element_rect(fill = "#236A3F"),
  ) +
  facet_wrap(~region, scales = "free_y")

# ###################################### C AMERICA #######################################
# ##### Step 2: Filter C AMERICA regions for the map and timeseries
# # Filter regions for C America
# selected_continent <- "Central America"
# sel_regions <- continent_mapping[[selected_continent]]
# #sel_regions <- "NCA"
# 
# # Filter combined data for timeseries 
# combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
# combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.05, na.rm = TRUE),
#     RP_upper = quantile(RP, 0.95, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# ggplot() +
#   # Uncertainty band with legend
#   geom_ribbon(data = ensemble_band_df, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
#   # Ensemble mean line with legend
#   geom_line(data = ensemble_band_df, aes(x = time, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
#   # ERA5 line with legend
#   geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
#   # Reference line (no legend)
#   geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
#   # Labels and theme
#   labs(
#     x = "Year",
#     y = "Effective Return Period",
#     title = "Frequency Change of 5-Year Events",
#     color = "Line",  # Legend titles
#     fill = "Shading"
#   ) +
#   scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
#   scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
#   theme_classic() +
#   theme(
#     legend.position = "none",
#     plot.title = element_text(hjust = 0.5, face = "bold"),
#     strip.text = element_text(size = 18),
#     axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
#     axis.text.y = element_text(size = 18), 
#     axis.title.x = element_text(size = 18),
#     axis.title.y = element_text(size = 18),
#     legend.title = element_text(size = 14),
#     legend.text = element_text(size = 12),
#     strip.background = element_rect(fill = "#B33951")
#   ) +
#   facet_wrap(~region, scales = "free_y")
# 
# 



# ###################################### S AMERICA #######################################
# ##### Step 2: Filter S AMERICA regions for the map and timeseries
# # Filter regions for S America
# selected_continent <- "South America"
# sel_regions <- continent_mapping[[selected_continent]]
# #sel_regions <- "NWS"
# 
# # Filter combined data for timeseries 
# combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
# combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.05, na.rm = TRUE),
#     RP_upper = quantile(RP, 0.95, na.rm = TRUE),
#     .groups = "drop"
#   )
# 
# ggplot() +
#   # Uncertainty band with legend
#   geom_ribbon(data = ensemble_band_df, aes(x = time, ymin = RP_lower, ymax = RP_upper, fill = "ModE-Sim Ensemble Range (5th-95th)"),alpha = 0.4) +
#   # Ensemble mean line with legend
#   geom_line(data = ensemble_band_df, aes(x = time, y = RP_mean, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
#   # ERA5 line with legend
#   geom_line(data = combined_era_sel, aes(x = time, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
#   # Reference line (no legend)
#   geom_hline(yintercept = 5, linetype = "dashed", color = "grey40") +
#   # Labels and theme
#   labs(
#     x = "Year",
#     y = "Effective Return Period (log10)",
#     title = "Frequency Change of 5-Year Events",
#     color = "Line",  # Legend titles
#     fill = "Shading"
#   ) +
#   scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
#   scale_fill_manual(values = c("ModE-Sim Ensemble Range (5th-95th)" = "grey70")) +
#   theme_classic() +
#   theme(
#     legend.position = "none",
#     plot.title = element_text(hjust = 0.5, face = "bold"),
#     strip.text = element_text(size = 18),
#     axis.text.x = element_text(angle = 45, hjust = 1, size = 18),
#     axis.text.y = element_text(size = 18), 
#     axis.title.x = element_text(size = 18),
#     axis.title.y = element_text(size = 18),
#     legend.title = element_text(size = 14),
#     legend.text = element_text(size = 12),
#     strip.background = element_rect(fill = "#E3D081")
#   ) +
#   facet_wrap(~region, scales = "free_y")


```

time warped plots: 

```{r}
# Warp function: compress early years, stretch recent years
warp_time <- function(year) {
  ifelse(year < 1950, year * 0.5, 975 + (year - 1950) * 2)
}


#plot all ensemble members and ensemble mean: 
combined_df <- read.csv(here::here("data/25yr_pooled_return_periods_regions_ERA_members_bootstrap.csv"))
combined_era <- read.csv(here::here("data/25yr_pooled_return_periods_regions_ERA5Land_bootstrap.csv"))

####################################### EUROPE #####################################
##### Step 2: Filter European regions for the map and timeseries
# Filter regions for Europe
selected_continent <- "Europe"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "MED"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.0, na.rm = TRUE),
#     RP_upper = quantile(RP, 1, na.rm = TRUE),
#     .groups = "drop"
#   )

# Apply to both datasets
combined_df_sel$time_warped <- warp_time(combined_df_sel$time)
combined_era_sel$time_warped <- warp_time(combined_era_sel$time)

# 3. Define range for highlight box
highlight_xmin <- warp_time(1950)
highlight_xmax <- warp_time(2010)


rollingRP_MED <- ggplot() +
  # Highlight box for stretched area
# geom_rect(aes(xmin = highlight_xmin, xmax = highlight_xmax, ymin = 0.8, ymax = 110),
#           fill = "white", color = "black", linetype = "solid", alpha = 0.6)+
  geom_vline(xintercept = highlight_xmin, color = "black", linewidth=0.5, linetype= 1)+
  geom_ribbon(data = combined_df_sel, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper,
                                           fill = "95% CI"), alpha = 0.4) +
  geom_line(data = combined_df_sel, aes(x = time_warped, y = RP,
                                         color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
  geom_ribbon(data = combined_era_sel, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper,
                                           fill = "95% CI"), alpha = 0.4) +
  geom_line(data = combined_era_sel, aes(x = time_warped, y = RP,
                                         color = "ERA5 (1970-2024)"), linewidth = 1) +
  geom_hline(yintercept = 30, linetype = "dashed", color = "grey40") +
  
  # Custom x-axis ticks: use original years as labels
  scale_x_continuous(
    breaks = warp_time(c( 1500, 1600, 1700, 1800, 1900, 1950, 1970, 1990, 2010)),
    labels = c( "1500", "1600", "1700", "1800", "1900", "1950", "1970", "1990", "2010"),
    name = ""
  ) +
  scale_y_log10(name = expression("Effective Return Period"),
        breaks = c(1, 2, 5, 10, 20, 50, 100, 150),
  labels = c("1", "2", "5", "10", "20", "50", "100", "150")) +  # Log scale for y-axis
  scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("95% CI" = "grey70")) +
  labs(title = "", color = "Line", fill = "Shading") +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size=13),
    strip.background = element_rect(fill = scales::alpha("#F5B0CB", 0.6)),
    axis.title = element_text(size=12)
  ) +
  facet_wrap(~region, scales = "free_y")+
  annotation_logticks()

########################################
selected_continent <- "Asia"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "ARP"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.0, na.rm = TRUE),
#     RP_upper = quantile(RP, 1, na.rm = TRUE),
#     .groups = "drop"
#   )

# Apply to both datasets
combined_df_sel$time_warped <- warp_time(combined_df_sel$time)
combined_era_sel$time_warped <- warp_time(combined_era_sel$time)

# 3. Define range for highlight box
highlight_xmin <- warp_time(1950)
highlight_xmax <- warp_time(2010)

rollingRP_ARP <- ggplot() +
  # Highlight box for stretched area
  # geom_rect(aes(xmin = highlight_xmin, xmax = highlight_xmax, ymin = 1, ymax = 40),
  #         fill = "white", color = "black", linetype = "solid", alpha = 0.6)+
  geom_vline(xintercept = highlight_xmin, color = "black", linewidth=0.5, linetype= 1)+
  # Uncertainty band with legend
  geom_ribbon(data = combined_df_sel, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = combined_df_sel, aes(x = time_warped, y = RP, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
  geom_ribbon(data = combined_era_sel, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper,
                                           fill = "95% CI"), alpha = 0.4) +
  # ERA5 line with legend
  geom_line(data = combined_era_sel, aes(x = time_warped, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 30, linetype = "dashed", color = "grey40") +
    # Custom x-axis ticks: use original years as labels
  scale_x_continuous(
    breaks = warp_time(c( 1500, 1600, 1700, 1800, 1900, 1950, 1970, 1990, 2010)),
    labels = c("1500", "1600", "1700", "1800", "1900", "1950", "1970", "1990", "2010"),
    name = "Year"
  ) +
  scale_y_log10(name = expression("Effective Return Period"),
        breaks = c(2, 5, 10, 20, 30, 50, 70),
  labels = c( "2", "5", "10", "20", "30", "50", "70")) +  # Log scale for y-axis
  # Labels and theme
  labs(
    title = "",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("95% CI" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size=13),
    strip.background = element_rect(fill =  scales::alpha("#E5973B", 0.6)),
    axis.title = element_text(size=12)
  ) +
  facet_wrap(~region, scales = "free_y")+
  annotation_logticks()


###################################### Australia #######################################
##### Step 2: Filter Australian regions for the map and timeseries
# Filter regions for Australia
selected_continent <- "Australia"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "CAU"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.0, na.rm = TRUE),
#     RP_upper = quantile(RP, 1, na.rm = TRUE),
#     .groups = "drop"
#   )

# Apply to both datasets
combined_df_sel$time_warped <- warp_time(combined_df_sel$time)
combined_era_sel$time_warped <- warp_time(combined_era_sel$time)

# 3. Define range for highlight box
highlight_xmin <- warp_time(1950)
highlight_xmax <- warp_time(2010)

rollingRP_CAU <-ggplot() +
  # Highlight box for stretched area
  # geom_rect(aes(xmin = highlight_xmin, xmax = highlight_xmax, ymin = 1.8, ymax = 110),
  #         fill = "white", color = "black", linetype = "solid", alpha = 0.6)+
  geom_vline(xintercept = highlight_xmin, color = "black", linewidth=0.5, linetype= 1)+
  # Uncertainty band with legend
  geom_ribbon(data = combined_df_sel, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = combined_df_sel, aes(x = time_warped, y = RP, color = "ModE-Sim Mean (1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  geom_ribbon(data = combined_era_sel, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper,
                                           fill = "95% CI"), alpha = 0.4) +
  geom_line(data = combined_era_sel, aes(x = time_warped, y = RP, color = "ERA5 (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 30, linetype = "dashed", color = "grey40") +
  scale_x_continuous(
    breaks = warp_time(c(1500, 1600, 1700, 1800, 1900, 1950, 1970, 1990, 2010)),
    labels = c("1500", "1600", "1700", "1800", "1900", "1950", "1970", "1990", "2010"),
    name = "Year"
  ) +
  scale_y_log10(name = expression(""),
        breaks = c(2, 5, 10, 20, 50, 100, 150),
  labels = c("2", "5", "10", "20", "50", "100", "150")) +  # Log scale for y-axis
  # Labels and theme
  labs(
    title = "",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim Mean (1420-2009)" = "black", "ERA5 (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("95% CI" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "none",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(color = "white"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size=13),
    strip.background = element_rect(fill =  scales::alpha("#3C3980", 0.6)),
    axis.title = element_text(size=12)
  ) +
  facet_wrap(~region, scales = "free_y")+
    annotation_logticks()


###################################### N AMERICA #######################################
##### Step 2: Filter N AMERICA regions for the map and timeseries
# Filter regions for N America
selected_continent <- "North America"
sel_regions <- continent_mapping[[selected_continent]]
sel_regions <- "WNA"

# Filter combined data for timeseries 
combined_df_sel <- combined_df %>% filter(region %in% sel_regions)
combined_era_sel <- combined_era %>% filter(region %in% sel_regions)
# # Compute ensemble mean
# ensemble_mean_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(RP_mean = mean(RP, na.rm = TRUE), .groups = "drop")
# 
# ensemble_band_df <- combined_df_sel %>%
#   group_by(region, time) %>%
#   summarise(
#     RP_mean = mean(RP, na.rm = TRUE),
#     RP_lower = quantile(RP, 0.0, na.rm = TRUE),
#     RP_upper = quantile(RP, 1, na.rm = TRUE),
#     .groups = "drop"
#   )
# Apply to both datasets
combined_df_sel$time_warped <- warp_time(combined_df_sel$time)
combined_era_sel$time_warped <- warp_time(combined_era_sel$time)

# 3. Define range for highlight box
highlight_xmin <- warp_time(1950)
highlight_xmax <- warp_time(2010)

rollingRP_WNA <-ggplot() +
  # Highlight box for stretched area
  # geom_rect(aes(xmin = highlight_xmin, xmax = highlight_xmax, ymin = 2, ymax = 30),
  #         fill = "white", color = "black", linetype = "solid", alpha = 0.6)+
  geom_vline(xintercept = highlight_xmin, color = "black", linewidth=0.5, linetype= 1)+
  # Uncertainty band with legend
  geom_ribbon(data = combined_df_sel, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper, fill = "95% CI"),alpha = 0.4) +
  # Ensemble mean line with legend
  geom_line(data = combined_df_sel, aes(x = time_warped, y = RP, color = "ModE-Sim Mean\n(1420-2009)"), linewidth = 1) +
  # ERA5 line with legend
  geom_ribbon(data = combined_era_sel, aes(x = time_warped, ymin = RP_lower, ymax = RP_upper,
                                           fill = "95% CI"), alpha = 0.4) +
  geom_line(data = combined_era_sel, aes(x = time_warped, y = RP, color = "ERA5-Land (1970-2024)"), linewidth = 1) +
  # Reference line (no legend)
  geom_hline(yintercept = 30, linetype = "dashed", color = "grey40") +
  scale_x_continuous(
    breaks = warp_time(c(1500, 1600, 1700, 1800, 1900, 1950, 1970, 1990, 2010)),
    labels = c("1500", "1600", "1700", "1800", "1900", "1950", "1970", "1990", "2010"),
    name = ""
  ) +
  scale_y_log10(name = expression(""),
        breaks = c(2, 5, 10, 20, 30, 50, 70),
  labels = c("2", "5", "10", "20", "30", "50", "70")) +  # Log scale for y-axis
  # Labels and theme
  labs(
    title = "",
    color = "Line",  # Legend titles
    fill = "Shading"
  ) +
  scale_color_manual(values = c("ModE-Sim Mean\n(1420-2009)" = "black", "ERA5-Land (1970-2024)" = "#B3001B")) +
  scale_fill_manual(values = c("95% CI" = "grey70")) +
  theme_classic() +
  theme(
    legend.position = "right",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(color= "white"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    text = element_text(size=13),
    strip.background = element_rect(fill =  scales::alpha("#236A3F", 0.6)),
    axis.title = element_text(size=12),
    legend.title = element_text(face = "bold")
  ) +
  facet_wrap(~region, scales = "free_y")+
    annotation_logticks()

rollingRP_MED 
rollingRP_ARP
rollingRP_CAU 
rollingRP_WNA


#### combine plots in grid and save
rollingRP_MED2 <- rollingRP_MED + theme(plot.margin = margin(t = 10, r = 10, b = 5, l = 5))
rollingRP_ARP2 <- rollingRP_ARP + theme(plot.margin = margin(t = 0, r = 10, b = 5, l = 5))
rollingRP_CAU2 <- rollingRP_CAU + theme(plot.margin = margin(t = 0, r = 10, b = 5, l = 5))
rollingRP_WNA2 <- rollingRP_WNA + theme(plot.margin = margin(t = 10, r = 10, b = 5, l = 5))


combined_rollingRP <- plot_grid(
  rollingRP_MED2, rollingRP_WNA2,
  rollingRP_ARP2, rollingRP_CAU2,
  ncol = 2,
  nrow = 2,
  rel_widths = c(1, 1),
  align = "h",
  axis = "tblr",
  labels = c(
    "a)",  "b) ", "c)", "d)"
  ),
  label_size = 14,
  label_x = 0.09,
  hjust = 0,
  label_y = 0.95,
  vjust = 1
)

print(combined_rollingRP)

ggsave(
  filename = here::here("figures/rollingRP.pdf"),
  plot     = combined_rollingRP,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 6.5,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

# ggsave(
#   filename = here::here("figures/rollingRP_legend.pdf"),
#   plot     = rollingRP_WNA,
#   device   = cairo_pdf,    # better text rendering
#   width    = 10,           # in inches
#   height   = 6.5,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )

```

##### DO THIS ANALYSIS FOR THE SAME PERIODS AS FOR THE INTENSITY 

How many regions experience a doubling in return period (below every 15) in the period 2000 - 2024? 

```{r}
## remove SAH from quantitative results
combined_era <- combined_era %>% filter(region != "SAH")

regions_with_high_RP <- combined_era %>%
  filter(time >= 2000, time <= 2024, RP <= 15) %>%
  distinct(region)

regions_with_high_RP

combined_era %>%
  filter(time >= 2000, time <= 2024, RP <= 15) %>%
  distinct(region) %>%
  count()
```
How many regions lie outside the range of return periods covered in the last 600 years? 
```{r}
## remove Sahara: 
combined_df <- combined_df %>% filter(region != "SAH")

# 1. Get reference ranges (1434–1999) for each region
ref_range_short <- combined_df %>%
  filter(time >= 1975, time <= 1999) %>%
  group_by(region) %>%
  summarise(ref_min = min(RP, na.rm = TRUE),
            ref_max = max(RP, na.rm = TRUE),
            .groups = "drop")
  
ref_range_medium <- combined_df %>%
  filter(time >= 1850, time <= 1999) %>%
  group_by(region) %>%
  summarise(ref_min = min(RP, na.rm = TRUE),
            ref_max = max(RP, na.rm = TRUE),
            .groups = "drop")
  
ref_range_long <- combined_df %>%
  filter(time <= 1999) %>%
  group_by(region) %>%
  summarise(ref_min = min(RP, na.rm = TRUE),
            ref_max = max(RP, na.rm = TRUE),
            .groups = "drop")

# 2. Compare modern period (2000–2024) to reference range
modern_outside_short <- combined_era %>%
  filter(time >= 2000, time <= 2024) %>%
  left_join(ref_range_short, by = "region") %>%
  filter(RP < ref_min) %>%
  distinct(region)

modern_outside_medium <- combined_era %>%
  filter(time >= 2000, time <= 2024) %>%
  left_join(ref_range_medium, by = "region") %>%
  filter(RP < ref_min) %>%
  distinct(region)

modern_outside_long <- combined_era %>%
  filter(time >= 2000, time <= 2024) %>%
  left_join(ref_range_long, by = "region") %>%
  filter(RP < ref_min) %>%
  distinct(region)

# 3. Count how many regions
n_regions_outside_short <- nrow(modern_outside_short)
n_regions_outside_medium <- nrow(modern_outside_medium)
n_regions_outside_long <- nrow(modern_outside_long)

print(paste0("Short context: have ", n_regions_outside_short, " regions outside the ref (1975-1999)."))
print(paste0("Medium context: have ", n_regions_outside_medium, " regions outside the ref (1850-1999)."))
print(paste0("Long context: have ", n_regions_outside_long, " regions outside the ref (1420-1999)."))

modern_outside_long
```

##### CREATE MAP SHOWING WHERE FREQUENCY IS OUTSIDE OF REFERENCE BEST ESTIMATE

Plot which regions are unprecedented and which record shattering

```{r}
# Load reference regions and coastlines:
load(here::here("data", "Reference_regions", "IPCC-WGI-reference-regions-v4_R.rda"))
refregions_sf <- st_as_sf(IPCC_WGI_reference_regions_v4)
refregions_sf <- refregions_sf %>% rename(region = Acronym)
temp.dir <- tempdir()
unzip(here::here("data/Reference_regions/ne_110m_coastline.zip"), exdir = temp.dir)
coastLines <- st_read(dsn = temp.dir, layer = "ne_110m_coastline")

# Load land data for filling continents
land <- ne_countries(scale = "medium", returnclass = "sf")

colors_map <- c("exceeding long term\nreference (1420-1999)" = "#D46273",
                "within long term\nreference (1420-1999)"  = "#301934")

# Get the unique regions that actually appear in your data
valid_regions <- unique(combined_df$region)

# Build color_cat — only for regions present in combined_df
region_cats <- refregions_sf %>%
  st_drop_geometry() %>%
  dplyr::select(region) %>%
  filter(region %in% valid_regions) %>%          # ← drop ocean/unused regions
  mutate(color_cat = ifelse(
    region %in% modern_outside_long$region,
    "exceeding long term\nreference (1420-1999)",                          # ← must match colors_map key
    "within long term\nreference (1420-1999)"                            # ← must match colors_map key
  ))

# Join categories onto the spatial data (only matched regions survive)
shown_sf <- refregions_sf %>%
  inner_join(region_cats %>% dplyr::select(region, color_cat), by = "region")

# Centroids for labels
pts <- shown_sf %>%
  st_centroid() %>%
  mutate(label = region)

# Build the map
map_RP <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour = "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Return Period category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin          = margin(0, 0, 0, 0),
    legend.position      = c(0.05, 0.05),
    legend.justification = c("left", "bottom"),
    legend.background    = element_blank(),
    legend.title         = element_text(face = "bold")
  )

map_RP
```

Combine Rolling RP plots and map into one
```{r}
# Step 2: stack the panel grid on top of the map
final_plot <- plot_grid(
  combined_rollingRP,
  map_RP,
  ncol = 1,
  labels = c("", "e)"),      # leave panels labelled as-is, add e) to map
  label_size = 14,
  label_x = 0.01,
  hjust = 0,
  rel_heights = c(2, 1)      # panels take 2/3, map takes 1/3 — adjust to taste
)

final_plot

ggsave(
  filename = here::here("figures/RP_rolling_map.pdf"),
  plot     = final_plot,
  device   = cairo_pdf,    # better text rendering
  width    = 10,           # in inches
  height   = 9,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```


#### Scatterplots for modern times:
first 3: 
1st plot: reference/overlap period *30 yr* events in mm for ERA5 and ModE-Sim: 1970-2009 (exclude problematic 60s of ERA5) 
2nd: y-axis: ERA5_Land 1995-2024 30 yr vs x-axis: ERA5 Land 1970-1999
3rd: y-axis: ERA5-Land 1995-2024 30 yr vs x-axis: ModE-SIm BC 1970-1999

```{r}
#corrected_full <- readRDS(here::here("data/regionalResults_ModESim_bc.RData"))

##calculate return level of 5 year events - 30 year chunks for consistency: 
#remove SAH due to unphysically large PCWD values - runaway PCWD 
regional_results_ERA5Land_eva <- regional_results_ERA5Land[ names(regional_results_ERA5Land) != "SAH" ]

use_rp <- 30   # return period of two years

# Modern ERA5-Land (1994-2024) for each region: 
# Initialize a list to store results for all regions
RL_ERA_modern <- list() #save modern 5yr events (in mm) for each region

# Loop through each region in regional_results
for (region in names(regional_results_ERA5Land_eva)) { #swap european_regions for regional_results for all regions
  # Extract the result array for the current region
  result_array_region <- regional_results_ERA5Land_eva[[region]][26:55] #1995-2024
  fit_ref <- extRemes::fevd(result_array_region, type = "Gumbel", method = "MLE", units = "years")
  return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))

  RL_ERA_modern[[region]] <- return_level
  
}

# Modern ERA5-Land (1994-2024) for each region: 
# Initialize a list to store results for all regions
RL_ERA_overlap <- list() #save modern 5yr events (in mm) for each region

# Loop through each region in regional_results
for (region in names(regional_results_ERA5Land_eva)) { #swap european_regions for regional_results for all regions
  # Extract the result array for the current region
  result_array_region <- regional_results_ERA5Land_eva[[region]][1:30] #1970-1999
  fit_ref <- extRemes::fevd(result_array_region, type = "Gumbel", method = "MLE", units = "years")
  return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))

  RL_ERA_overlap[[region]] <- return_level
  
}

##### ModE-Sim 
path <- here::here("data/volc_impact_aggregates/collected_p_pet/")
folders <- list.files(path, full.names = TRUE, pattern = "m[0-9]{3}") #m001 - m020 in first set

# Extract unique ensemble member identifiers from folder names
ensemble_member <- unique(basename(folders))[1:20]
ensemble_members <- paste0(ensemble_member, "_tidy")


corrected_full_eva <- corrected_full[ names(corrected_full) != "SAH" ]

regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")


###Overlap with ERA5-land for each region (1970-2009):
RL_ModE_overlap <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][551:580,]#1970-1999
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_overlap[[region]] <- result_array_region
}

### Pre-industrial for every region (1420-1850):
RL_ModE_preind <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][1:30,]#1420-1450
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_preind[[region]] <- result_array_region
}

### Early-industrial for every region (1851-1950):
RL_ModE_earlyind <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][470:501,]#1890-1920
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_earlyind[[region]] <- result_array_region
}


### Modern ModeSIM for every region (1951-2009):
RL_ModE_modern <- list()

# Loop over all regions
for (region in names(corrected_full_eva)) {
  # Extract the result array for the current region
  result_array_region <- corrected_full_eva[[region]][551:580,]#1970-1999
  
  # Choose ensemble members dynamically - exclude m001 for Australia and South East Asia 
  if (region %in% regions_to_modify) {
    ems <- ensemble_members[ensemble_members != "m001_tidy"]  # Skip m001
  } else {
    ems <- ensemble_members
  }
  
  RL_list <- list()
  
  for (em in ems) {
    # Extract the result array for the current em
    result_array_em <- result_array_region[,em]
    fit_ref <- extRemes::fevd(result_array_em, type = "Gumbel", method = "MLE", units = "years")
    return_level <- round(as.numeric(extRemes::return.level(fit_ref, return.period = use_rp)))
    RL_list[[em]] <- return_level
 }
    # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, RL_list)
  # Store the result for the current region
  RL_ModE_modern[[region]] <- result_array_region
}

```

Plot in scatterplots: 

```{r}
# #### Overlap period ModE-Sim and ERA5-Land

####### -------------------- Overlap period --------------------------------
# 1) Summarise lists into a data.frame:
df_overlap <- tibble(
  region = names(RL_ERA_overlap),
  era     = unlist(RL_ERA_overlap)
) %>% 
  mutate(
    # Grab the ensemble vector for each region
    mod_vec = RL_ModE_overlap[region],
    # Compute mean, min, max
    mod_mean = map_dbl(mod_vec, ~ mean(.x)),
    mod_min  = map_dbl(mod_vec, ~ min(.x)),
    mod_max  = map_dbl(mod_vec, ~ max(.x))
  ) %>% 
  dplyr::select(-mod_vec)

# Ensure the continent information is consistent
df_overlap <- df_overlap %>%
  mutate(continent = case_when(
    region %in% continent_mapping[["Europe"]] ~ "Europe",
    region %in% continent_mapping[["North America"]] ~ "North America",
    region %in% continent_mapping[["Africa"]] ~ "Africa",
    region %in% continent_mapping[["Asia"]] ~ "Asia",
    region %in% continent_mapping[["Central America"]] ~ "Central America",
    region %in% continent_mapping[["South America"]] ~ "South America",
    region %in% continent_mapping[["Australia"]] ~ "Australia",
    TRUE ~ "Other"
  ))

# 2) Plot with horizontal error bars
scatter_overlap <- ggplot(df_overlap, aes(x = mod_mean, y = era, color = continent)) +
  geom_errorbarh(aes(xmin = mod_min, xmax = mod_max), height = 100) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names
  labs(
    x     = "ModE-Sim* 30yr PCWD [mm]",
    y     = "ERA5-Land 30yr PCWD [mm]",
    color = "Continent"
  ) +
  scale_x_continuous(breaks = c(0, 500, 1000, 1500, 2000), limits = c(0, 2000)) +
  scale_y_continuous(breaks = c(0, 500, 1000, 1500, 2000),limits = c(0, 2000)) +
  theme_classic() +
  theme(
    legend.position =  c(0.2, 0.75),
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


####### -------------------- Modern_ModE-Sim --------------------------------
# 1) Summarise lists into a data.frame:
df_modern_era <- tibble(
  region = names(RL_ERA_modern),
  era_mod     = unlist(RL_ERA_modern),
  era_overlap = unlist(RL_ERA_overlap)
) 

# Ensure the continent information is consistent
df_modern_era <- df_modern_era %>%
  mutate(continent = case_when(
    region %in% continent_mapping[["Europe"]] ~ "Europe",
    region %in% continent_mapping[["North America"]] ~ "North America",
    region %in% continent_mapping[["Africa"]] ~ "Africa",
    region %in% continent_mapping[["Asia"]] ~ "Asia",
    region %in% continent_mapping[["Central America"]] ~ "Central America",
    region %in% continent_mapping[["South America"]] ~ "South America",
    region %in% continent_mapping[["Australia"]] ~ "Australia",
    TRUE ~ "Other"
  ))

# 2) Plot with horizontal error bars
scatter_era5 <- ggplot(df_modern_era, aes(x = era_overlap, y = era_mod, color = continent)) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names
  labs(
    x     = "ERA5-Land 30yr PCWD [mm] (1970-1999)",
    y     = "ERA5-Land 30yr PCWD [mm] (1995-2024)",
    color = "Continent"
  ) +
  scale_x_continuous(breaks = c(0, 500, 1000, 1500, 2000), limits = c(0, 2000)) +
  scale_y_continuous(breaks = c(0, 500, 1000, 1500, 2000),limits = c(0, 2000)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
    text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


####### -------------------- Modern ModE-SIm --------------------------------
# 1) Summarise lists into a data.frame:
df_modern_ModE <- tibble(
  region = names(RL_ERA_modern),
  era     = unlist(RL_ERA_modern)
) %>% 
  mutate(
    # Grab the ensemble vector for each region
    mod_vec = RL_ModE_overlap[region],
    # Compute mean, min, max
    mod_mean = map_dbl(mod_vec, ~ mean(.x)),
    mod_min  = map_dbl(mod_vec, ~ min(.x)),
    mod_max  = map_dbl(mod_vec, ~ max(.x))
  ) %>% 
  dplyr::select(-mod_vec)

# Ensure the continent information is consistent
df_modern_ModE <- df_modern_ModE %>%
  mutate(continent = case_when(
    region %in% continent_mapping[["Europe"]] ~ "Europe",
    region %in% continent_mapping[["North America"]] ~ "North America",
    region %in% continent_mapping[["Africa"]] ~ "Africa",
    region %in% continent_mapping[["Asia"]] ~ "Asia",
    region %in% continent_mapping[["Central America"]] ~ "Central America",
    region %in% continent_mapping[["South America"]] ~ "South America",
    region %in% continent_mapping[["Australia"]] ~ "Australia",
    TRUE ~ "Other"
  ))

# 2) Plot with horizontal error bars
scatter_era_mod <- ggplot(df_modern_ModE, aes(x = mod_mean, y = era, color = continent)) +
  geom_errorbarh(aes(xmin = mod_min, xmax = mod_max), height = 100) +
  geom_point(size = 2) +
  scale_color_manual(values = continent_colors) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_smooth(method = "lm", color = "red", se = FALSE, linewidth= 0.5) +  # Trend line
  #geom_text(aes(label = region), hjust = 0.5, vjust = -0.5, size = 3, colour = "black") + # Annotate with region names
  labs(
    x     = "ModE-Sim* 30yr PCWD [mm] (1970-1999)",
    y     = "ERA5-Land 30yr PCWD [mm] (1995-2024)",
    color = "Continent"
  ) +
  scale_x_continuous(breaks = c(0, 500, 1000, 1500, 2000), limits = c(0, 2000)) +
  scale_y_continuous(breaks = c(0, 500, 1000, 1500, 2000),limits = c(0, 2000)) +
  theme_classic() +
  theme(
    legend.position    = "none",
    plot.title         = element_text(hjust = 0.5, face = "bold"),
    legend.title       = element_text(face = "bold"),
        text = element_text(size= 13),
    axis.title.y = element_text(
      margin = margin(r = 10)),
    axis.title.x = element_text(
      margin = margin(t = 10))
  )


scatter_overlap_2 <- scatter_overlap + theme(plot.margin = margin(t = 40, r = 20, b = 5, l = 5))
scatter_era5_2 <- scatter_era5 + theme(plot.margin = margin(t = 40, r = 20, b = 5, l = 5))
scatter_era_mod_2 <- scatter_era_mod + theme(plot.margin = margin(t = 40, r = 10, b = 5, l = 5))


combined_val_scatter <- plot_grid(
  scatter_overlap_2, 
  scatter_era5_2, 
  scatter_era_mod_2,
  ncol = 3,
  rel_widths = c(1, 1, 1),
  align = "h",
  axis = "tblr",
  labels = c(
    "a) Overlap period (1970-1999)", 
    "b) Modern ERA5-Land", "c) Modern ERA5-Land vs ModE-Sim"
  ),
  label_size = 14,
  label_x = 0.1,
  hjust = 0,
  label_y = 0.98,
  vjust = 1
)

print(combined_val_scatter)

ggsave(
  filename = here::here("figures/Scatter_validation_plots.pdf"),
  plot     = combined_val_scatter,
  device   = cairo_pdf,    # better text rendering
  width    = 17,           # in inches
  height   = 5.9,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

```


```{r}
# Calculate the correlation and mean bias between ERA5-Land and ModE-Sim* 5-year events during overlap 
summary_stats <- tibble(
  correlation = cor(df_overlap$era, df_overlap$mod_mean, use = "complete.obs"),
  mean_bias   = mean(df_overlap$mod_mean - df_overlap$era, na.rm = TRUE)
)

summary_stats


fit_era <- lm(era_mod ~ era_overlap, data = df_modern_era)
summary(fit_era)
coef(fit_era)

fit_mod <- lm(era ~ mod_mean, data = df_modern_ModE)
summary(fit_mod)
coef(fit_mod)

```


## Scatterplots comparing annual PCWD over time to 30yr extremes over time
### Approach

Determining whether 21st century (2000-2024) droughts are unprecedented and quantifying their "extremeness" depends on the reference. Here, we determine these aspects based on alternative choices of the reference:

1. Recent past climate reanalysis (ERA5) data (1975-1999)
2. Recent past climate model simulation, across all ensemble members (1975-1999)
3. Historical climate model simulation, across all ensemble members (1850-1999)
4. Multi-centennial preindustrial climate model simulation, across all ensemble members (1420-1999)

The magnitude of 21st century droughts are estimated based on ERA5 climate reanalysis data for years 2000-2024.

## Read data

### ERA5

Interpreting the objects into nice and tidy data frames.
```{r}
list_era5 <- read_rds(here("data/regional_results_ERA5Land.rds"))

nregions <- length(list_era5)

make_df_era5 <- function(vec){
  tibble(
    year = 1970:2024,
    pcwd = vec
  )
}

df_era5 <- purrr::map(
  list_era5, 
  ~make_df_era5(.)
) |> 
  bind_rows(.id = "region")

df_era5 <- df_era5 |> 
  filter(region != "SAH") #exclude Sahara Desert as unphysical values
```

### ModE-Sim

```{r}
extract_ensemble_number <- function(char){
  char |> 
    str_remove("_tidy") |> 
    str_remove("m0") |> 
    as.integer()
}

make_df_modesim <- function(df){
  as_tibble(df) |> 
    # not sure this is correct. Better to take years forward as part of the workflow. 
    mutate(year = 1420:2009) |> 
    pivot_longer(
      cols = ends_with("_tidy"), 
      names_to = "member", 
      values_to = "pcwd", 
      names_transform = extract_ensemble_number
      )
}

list_modesim <- read_rds(here("data/regional_results_ModESim_bc.rds"))

df_modesim <- purrr::map(
  list_modesim,
  ~make_df_modesim(.)
) |> 
  bind_rows(.id = "region")

df_modesim <- df_modesim |> 
  filter(region != "SAH")
```


## Record annual values

Create objects that contain records for different references in different datasets.

### ERA5

1. ERA5: Record PCWD by region in years 1975-1999 and in years 2000-2024.
```{r}
df_record_era5 <- df_era5 |> 
  filter(year %in% 1975:1999) |> 
  group_by(region) |> 
  summarise(pcwd_max0 = max(pcwd)) |> 
  left_join(
    df_era5 |> 
      filter(year %in% 2000:2024) |> 
      group_by(region) |> 
      summarise(pcwd_max1 = max(pcwd)),
    by = join_by(region)
  )

```

Plot.
```{r}
# calculate SD of pcwd_max0
sd0 <- df_era5 |> 
  filter(year %in% 1975:1999) |> 
  group_by(region) |> 
  summarise(sd_pcwd = sd(pcwd, na.rm = TRUE)) |> 
  ungroup()

tmp <- df_record_era5 |>
  left_join(sd0, by = "region") |> 
  mutate(
    is_greater = pcwd_max1 > pcwd_max0,
    is_shattering = pcwd_max1 > pcwd_max0 + 2 * sd_pcwd,
    color_cat = case_when(
      is_shattering ~ "shattering",
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    )
  )

# counts
n_greater <- sum(tmp$is_greater, na.rm = TRUE)
n_shattering <- sum(tmp$is_shattering, na.rm = TRUE)

# plot
gg1 <- tmp |>
  ggplot(aes(pcwd_max0, pcwd_max1, color = color_cat)) +
  geom_point(size = 0.9, show.legend = TRUE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
  scale_color_manual(name = "Event category", values = c("within" = "#301934",
                                "unprecedented" = "#91C499",
                                "shattering" = "#B3001B")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "ERA5 1975-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater) ~ ", " ~ 
                      italic(N[plain("record-shattering")]) == .(n_shattering)),
    title = ""
  )+  theme(
    legend.position = c(0.05, 0.95),   # inside top-left
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

gg1
```
Plot which regions are unprecedented and which record shattering

```{r}
#Load reference regions and coastlines:
load(here::here("data", "Reference_regions", "IPCC-WGI-reference-regions-v4_R.rda"))

# IPCC_WGI_reference_regions_v4 is a Spatial*DataFrame with a column "region" (or similar)
refregions_sf <- st_as_sf(IPCC_WGI_reference_regions_v4)
refregions_sf <- refregions_sf %>% rename(region = Acronym)

temp.dir <- tempdir()
unzip(here::here("data/Reference_regions/ne_110m_coastline.zip"), exdir = temp.dir)
coastLines <- st_read(dsn = temp.dir, layer = "ne_110m_coastline")

names(names(refregions))

proj4string(refregions)


# Load land data for filling continents
land <- ne_countries(scale = "medium", returnclass = "sf")
```


```{r}
colors_map <- c("within" = "#301934", 
                "unprecedented"     = "#91C499", 
                "shattering"  = "#B3001B")

# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_recent <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Event category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin     = margin(0,0,0,0),
    legend.position = c(0.0, 0.0),   # inside top-left
    legend.justification = c("left", "bottom"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

map_recent

```


### ModE-Sim recent (1975-1999)

Record PCWD by region in ERA5, years 2000-2024 vs. in years 1975-1999 and all ensemble members in ModE-Sim.
```{r}
df_record_era5_modesim_recent <- df_modesim |> 
  filter(year %in% 1975:1999) |> 
  group_by(region, member) |> 
  summarise(pcwd_max0 = max(pcwd)) |> 
  ungroup() |> 
  group_by(region) |> 
  summarise(pcwd_max0_left = min(pcwd_max0), pcwd_max0_right = max(pcwd_max0)) |> 
  left_join(
    df_era5 |> 
      filter(year %in% 2000:2024) |> 
      group_by(region) |> 
      summarise(pcwd_max1 = max(pcwd)),
    by = join_by(region)
  )
```

Plot.
```{r}
# calculate SD of pcwd_max0
sd1 <- df_modesim |> 
  filter(year %in% 1975:1999) |> 
  group_by(region) |> 
  summarise(sd_pcwd = sd(pcwd, na.rm = TRUE)) |> 
  ungroup()

tmp <- df_record_era5_modesim_recent |> 
  left_join(sd1, by = "region") |> 
  mutate(
    is_greater = pcwd_max1 > pcwd_max0_right,
    is_shattering = pcwd_max1 > pcwd_max0_right + 2 * sd_pcwd,
    color_cat = case_when(
      is_shattering ~ "shattering",
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    )
  )

# counts
n_greater <- sum(tmp$is_greater, na.rm = TRUE)
n_shattering <- sum(tmp$is_shattering, na.rm = TRUE)


gg2 <- tmp |> 
  ggplot(aes(x = pcwd_max0_right, xend = pcwd_max0_left, y = pcwd_max1, color = color_cat)) +
  geom_segment(show.legend = FALSE, linewidth = 0.75) +
  geom_point(show.legend = FALSE, size = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
  scale_color_manual(values = c("within" = "#301934",
                                "unprecedented" = "#91C499",
                                "shattering" = "#B3001B")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "ModE-Sim 1975-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater) ~ ", " ~ 
                      italic(N[plain("record-shattering")]) == .(n_shattering)),
    title = "   "
  )

gg2
```


Plot which regions are unprecedented and which record shattering
```{r}
colors_map <- c("within" = "#301934", 
                "unprecedented"     = "#91C499", 
                "shattering"  = "#B3001B")

# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_mode_recent <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.title      = element_text(face = "bold", hjust = .5),
    plot.margin     = margin(0,0,0,0),
    legend.position = "none"
  )

map_mode_recent
```



### ModE-Sim historical (1850-1999)

Record PCWD by region in ERA5, years 2000-2024 vs. in years 1850-1999 and all ensemble members in ModE-Sim.
```{r}
df_record_era5_modesim_historical <- df_modesim |> 
  filter(year %in% 1850:1999) |> 
  group_by(region, member) |> 
  summarise(pcwd_max0 = max(pcwd)) |> 
  ungroup() |> 
  group_by(region) |> 
  summarise(pcwd_max0_left = min(pcwd_max0), pcwd_max0_right = max(pcwd_max0)) |> 
  left_join(
    df_era5 |> 
      filter(year %in% 2000:2024) |> 
      group_by(region) |> 
      summarise(pcwd_max1 = max(pcwd)),
    by = join_by(region)
  )
```

Plot.
```{r}
# calculate SD of pcwd_max0
sd2 <- df_modesim |> 
  filter(year %in% 1850:1999) |> 
  group_by(region) |> 
  summarise(sd_pcwd = sd(pcwd, na.rm = TRUE)) |> 
  ungroup()

tmp <- df_record_era5_modesim_historical |> 
  left_join(sd2, by = "region") |> 
  mutate(
    is_greater = pcwd_max1 > pcwd_max0_right,
    is_shattering = pcwd_max1 > pcwd_max0_right + 2 * sd_pcwd,
    color_cat = case_when(
      is_shattering ~ "shattering",
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    )
  )

# counts
n_greater <- sum(tmp$is_greater, na.rm = TRUE)
n_shattering <- sum(tmp$is_shattering, na.rm = TRUE)


gg3 <- tmp |> 
  ggplot(aes(x = pcwd_max0_right, xend = pcwd_max0_left, y = pcwd_max1, color = color_cat)) +
  geom_segment(show.legend = FALSE, linewidth = 0.75) +
  geom_point(show.legend = FALSE, size = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
  scale_color_manual(values = c("within" = "#301934",
                                "unprecedented" = "#91C499",
                                "shattering" = "#B3001B")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "ModE-Sim 1850-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater) ~ ", " ~ 
                      italic(N[plain("record-shattering")]) == .(n_shattering)),
    title = "   "
  )

gg3
```


Plot which regions are unprecedented and which record shattering


```{r}
# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_historical <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.title      = element_text(face = "bold", hjust = .5),
    plot.margin     = margin(0,0,0,0),
    legend.position = "none"
  )

map_historical
```
### ModE-Sim multi-century (1420-1999)

Record PCWD by region in ERA5, years 2000-2024 vs. in years 1850-1999 and all ensemble members in ModE-Sim.
```{r}
df_record_era5_modesim_multicent <- df_modesim |> 
  filter(year %in% 1420:1999) |> 
  group_by(region, member) |> 
  summarise(pcwd_max0 = max(pcwd)) |> 
  ungroup() |> 
  group_by(region) |> 
  summarise(pcwd_max0_left = min(pcwd_max0), pcwd_max0_right = max(pcwd_max0)) |> 
  left_join(
    df_era5 |> 
      filter(year %in% 2000:2024) |> 
      group_by(region) |> 
      summarise(pcwd_max1 = max(pcwd)),
    by = join_by(region)
  )
```

Plot.
```{r}
# calculate SD of pcwd_max0
sd3 <- df_modesim |> 
  filter(year %in% 1420:1999) |> 
  group_by(region) |> 
  summarise(sd_pcwd = sd(pcwd, na.rm = TRUE)) |> 
  ungroup()

tmp <- df_record_era5_modesim_multicent |> 
  left_join(sd3, by = "region") |> 
  mutate(
    is_greater = pcwd_max1 > pcwd_max0_right,
    is_shattering = pcwd_max1 > pcwd_max0_right + 2 * sd_pcwd,
    color_cat = case_when(
      is_shattering ~ "shattering",
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    )
  )

# counts
n_greater <- sum(tmp$is_greater, na.rm = TRUE)
n_shattering <- sum(tmp$is_shattering, na.rm = TRUE)

gg4 <- tmp |> 
  ggplot(aes(x = pcwd_max0_right, xend = pcwd_max0_left, y = pcwd_max1, color = color_cat)) +
  geom_segment(show.legend = FALSE, linewidth = 0.75) +
  geom_point(show.legend = FALSE, size = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
  scale_color_manual(name = "Event type", values = c("within" = "#301934",
                                "unprecedented" = "#91C499",
                                "shattering" = "#B3001B")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "ModE-Sim 1420-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater) ~ ", " ~ 
                      italic(N[plain("record-shattering")]) == .(n_shattering)),
    title = "   "
  )

gg4
#ggsave(here("figures/category_scatter_legend.pdf"), plot = gg4, width = 6, height = 4)
```

Plot which regions are unprecedented and which record shattering

```{r}
# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_multicen <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Event type"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(legend.position = "none")

map_multicen

#ggsave(here("figures/category_maps_legend.pdf"), plot = map_multicen, width = 8, height = 4)
```

## Record 25-year means

Calculate rolling means of all data, separated by region (ModE-Sim also separated by ensemble member).
```{r}
df_era5_25 <- df_era5 |>
  group_by(region) |>
  arrange(year) |>
  mutate(
    pcwd_25 = slide_dbl(pcwd, mean, .before = 24, .complete = TRUE)
  ) |>
  ungroup()

df_modesim_25 <- df_modesim |> 
  group_by(region, member) |> 
  nest() |> 
  mutate(
    data = map(data, ~ .x  |> 
      mutate(
        roll_mean = slide_dbl(
          .x$pcwd,
          mean,
          .before = 24,     # 2 before = 3-point window (includes current)
          .complete = TRUE  # require full window, so first 2 will be NA
        )
      )
    )
  )
```


### ERA5

```{r}
df_record25_era5 <- df_era5_25 |>
  filter(year %in% c(1999, 2024)) |>
  dplyr::select(region, year, pcwd_25) |>
  pivot_wider(names_from = "year", values_from = "pcwd_25") |>
  rename(
    pcwd_max0 = `1999`,
    pcwd_max1 = `2024`
  )

```


```{r}
tmp <- df_record25_era5 |>
  mutate(
    is_greater = pcwd_max1 > pcwd_max0,
    color_cat = case_when(
    is_greater ~ "unprecedented",
      TRUE ~ "within"
    )
  )

# counts
n_greater <- sum(tmp$is_greater, na.rm = TRUE)
```

Plot.
```{r}

gg5 <- tmp |> 
  ggplot(aes(pcwd_max0, pcwd_max1, color = color_cat)) +
  geom_point(size = 0.9, show.legend = TRUE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
  scale_color_manual(name = "Event category", values = c("within" = "#301934",
                                                 "unprecedented"     = "#91C499")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
    labs(
    x = "ERA5 1975-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater)),
    title = ""
  )+  theme(
    legend.position = c(0.05, 0.95),   # inside top-left
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

gg5
```
```{r}

colors_map <- c("within" = "#301934", 
       "unprecedented"     = "#91C499")

# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_modern_25 <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Event category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin     = margin(0,0,0,0),
    legend.position = c(0.0, 0.0),   # inside top-left
    legend.justification = c("left", "bottom"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

map_modern_25
```
### ModE-Sim recent (1975-1999)

```{r}
get_record <- function(df){
  df |> 
    filter(year %in% 1975:1999) |> 
    filter(roll_mean == max(roll_mean)) |> 
    pull(roll_mean)
}

df_record25_era5_modesim_recent <- df_modesim_25 |> 
  mutate(pcwd_max0 = purrr::map_dbl(data, ~get_record(.))) |> 
  dplyr::select(-data) |> 
  ungroup() |> 
  group_by(region) |> 
  summarise(pcwd_max0_left = min(pcwd_max0), pcwd_max0_right = max(pcwd_max0)) |> 
  left_join(
    df_record25_era5 |> 
      dplyr::select(region, pcwd_max1),
    by = join_by(region)
  )
```

Plot.
```{r}
tmp <- df_record25_era5_modesim_recent |> 
  mutate(
    is_greater = pcwd_max1 > pcwd_max0_right,
    color_cat = case_when(
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    )
  )

# counts
n_greater <- sum(tmp$is_greater, na.rm = TRUE)

gg6 <- tmp |> 
  ggplot(aes(x = pcwd_max0_right, xend = pcwd_max0_left, y = pcwd_max1, color = color_cat)) +
  geom_segment(show.legend = FALSE, linewidth = 0.75) +
  geom_point(size = 0.75, show.legend = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
  scale_color_manual(name = "Event category", values = c("within" = "#301934",
                                                 "unprecedented"     = "#91C499")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
    labs(
    x = "ModE-Sim 1975-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater)),
    title = ""
  )+  theme(
    legend.position = c(0.05, 0.95),   # inside top-left
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

gg6
```

```{r}
# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_recent_25 <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Event category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin     = margin(0,0,0,0),
    legend.position = "none",   # inside top-left
    legend.justification = c("left", "bottom"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

map_recent_25
```

### ModE-Sim historical (1850-1999)

```{r}
get_record <- function(df){
  df |> 
    filter(year %in% 1850:1999) |> 
    filter(roll_mean == max(roll_mean)) |> 
    pull(roll_mean)
}

df_record25_era5_modesim_historical <- df_modesim_25 |> 
  mutate(pcwd_max0 = purrr::map_dbl(data, ~get_record(.))) |> 
  dplyr::select(-data) |> 
  ungroup() |> 
  group_by(region) |> 
  summarise(pcwd_max0_left = min(pcwd_max0), pcwd_max0_right = max(pcwd_max0)) |> 
  left_join(
    df_record25_era5 |> 
      dplyr::select(region, pcwd_max1),
    by = join_by(region)
  )
```

Plot.
```{r}

tmp <- df_record25_era5_modesim_historical |> 
  mutate(
    is_greater = pcwd_max1 > pcwd_max0_right,
    color_cat = case_when(
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    )
  )

# counts
n_greater <- sum(tmp$is_greater, na.rm = TRUE)

gg7 <- tmp |> 
  ggplot(aes(x = pcwd_max0_right, xend = pcwd_max0_left, y = pcwd_max1, color = color_cat)) +
  geom_segment(show.legend = FALSE, linewidth = 0.75) +
  geom_point(size = 0.75, show.legend = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
  scale_color_manual(name = "Event category", values = c("within" = "#301934",
                                                "unprecedented"     = "#91C499")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
    labs(
    x = "ModE-Sim 1850-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater)),
    title = ""
  )+  theme(
    legend.position = c(0.05, 0.95),   # inside top-left
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

gg7
```
```{r}
# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_historical_25 <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Event category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin     = margin(0,0,0,0),
    legend.position = "none",   # inside top-left
    legend.justification = c("left", "bottom"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

map_historical_25
```
### ModE-Sim multicentennial (1420-1999)

```{r}
get_record <- function(df){
  df |> 
    filter(year %in% 1444:1999) |> 
    filter(roll_mean == max(roll_mean)) |> 
    pull(roll_mean)
}

df_record25_era5_modesim_multicent <- df_modesim_25 |> 
  mutate(pcwd_max0 = purrr::map_dbl(data, ~get_record(.))) |> 
  dplyr::select(-data) |> 
  ungroup() |> 
  group_by(region) |> 
  summarise(pcwd_max0_left = min(pcwd_max0), pcwd_max0_right = max(pcwd_max0)) |> 
  left_join(
    df_record25_era5 |> 
      dplyr::select(region, pcwd_max1),
    by = join_by(region)
  )
```

Plot.
```{r}
tmp <- df_record25_era5_modesim_multicent |> 
  mutate(
    is_greater = pcwd_max1 > pcwd_max0_right,
    color_cat = case_when(
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    )
  )

# counts
n_greater <- sum(tmp$is_greater, na.rm = TRUE)

gg8 <- tmp |> 
  ggplot(aes(x = pcwd_max0_right, xend = pcwd_max0_left, y = pcwd_max1, color = color_cat)) +
  geom_segment(show.legend = FALSE, linewidth = 0.75) +
  geom_point(size = 0.75, show.legend = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
  scale_color_manual(name = "Event category", values = c("within" = "#301934",
                                                "unprecedented"     = "#91C499")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
    labs(
    x = "ModE-Sim 1420-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater)),
    title = ""
  )+  theme(
    legend.position = c(0.05, 0.95),   # inside top-left
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

gg8
```
```{r}
# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_multicent_25 <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Event category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin     = margin(0,0,0,0),
    legend.position = "none",   # inside top-left
    legend.justification = c("left", "bottom"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

map_multicent_25
```

## 30-year extremes (pooled)

Calculate rolling 30-year extremes of all data, separated by region (ModE-Sim also separated by ensemble member).
```{r}
# Function: fit Gumbel with extRemes and estimate given year return level
fit_gumbel_return <- function(vec, return_period = 30) {
  # Fit Gumbel (GEV with shape = 0)
  fit <- extRemes::fevd(vec, type = "Gumbel")
  
  # Estimate T-year return level
  rl <- unname(c(return.level(fit, return.period = return_period)))

  return(rl)
}

calc_30x <- function(df){
  fit_gumbel_return(df$pcwd)
}

calc_30x_across_members <- function(df){
  # Collapse selected columns into vector
  vec <- unlist(df %>% dplyr::select(-year))
  vec <- vec[is.finite(vec)]
  fit_gumbel_return(vec, return_period = 30)
}

## 25-year sliding windows
apply_sliding <- function(df){
  df %>%
    mutate(return_level = slide_index_dbl(
      seq_len(nrow(.)),
      .i = year,
      .before = 24,
      .complete = TRUE,
      .f = ~ calc_30x_across_members(df[.x, ])
    ))
}

apply_pooled <- function(df, years){
  df |> 
    filter(year %in% years) |> 
    calc_30x_across_members()
}

# fit two extreme value distributions on the two ERA5 periods and extract the 30-year extreme value.
df_era5_30x <- df_era5 |> 
  mutate(period = ifelse(year %in% 1975:1999, "pcwd_x30_ref0", ifelse(year %in% 2000:2024, "pcwd_x30_ref1", NA))) |> 
  group_by(region, period) |> 
  nest() |> 
  drop_na(period) |> 
  mutate(pcwd_30x = purrr::map_dbl(data, ~calc_30x(.))) |> 
  pivot_wider(id_cols = -data, values_from = "pcwd_30x", names_from = "period")

# this takes long to compute
nmembers <- length(unique(df_modesim$member))

# fit extreme value distributions on sliding windows of 25 years, gathering values from all ensemble members
# warning: this takes a few minutes
df_modesim_30x_sliding <- df_modesim |> 
  mutate(member = paste0("m", as.integer(member))) |> 
  pivot_wider(names_from = "member", values_from = "pcwd") |> 
  group_by(region) |> 
  nest() |> 
  mutate(data = map(data, apply_sliding)) |> 
  mutate(data = map(data, ~dplyr::select(., year, pcwd_30x = return_level))) |> 
  unnest(data)

# fit extreme value distributions on data pooled from all years in different periods
df_modesim_30x_pooled <- df_modesim |> 
  mutate(member = paste0("m", as.integer(member))) |> 
  pivot_wider(names_from = "member", values_from = "pcwd") |> 
  group_by(region) |> 
  nest()
  
df_modesim_30x_pooled_multicent <- df_modesim_30x_pooled |> 
  mutate(pcwd_30x = map_dbl(data, ~apply_pooled(., years = 1420:1999)))

df_modesim_30x_pooled_historical <- df_modesim_30x_pooled |> 
  mutate(pcwd_30x = map_dbl(data, ~apply_pooled(., years = 1850:1999)))

df_modesim_30x_pooled_recent <- df_modesim_30x_pooled |> 
  mutate(pcwd_30x = map_dbl(data, ~apply_pooled(., years = 1975:1999)))
```

### ERA5

```{r}
tmp <- df_era5_30x |> 
  mutate(is_greater = pcwd_x30_ref0 < pcwd_x30_ref1,
  color_cat = case_when(
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    ))

n_greater <- tmp |> 
  ungroup() |> 
  summarise(n_greater = sum(is_greater)) |> 
  pull(n_greater)

gg9 <- tmp |> 
  ggplot(aes(pcwd_x30_ref0, pcwd_x30_ref1, color = color_cat)) +
  geom_point(size = 0.9, show.legend = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
    scale_color_manual(name = "Event category", values = c("within" = "#301934",
                                                "unprecedented"     = "#91C499")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "ERA5 1975-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater)),
    title = ""
  ) + theme(
    legend.position = c(0.05, 0.95),   # inside top-left
    legend.justification = c("left", "top"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold"))

gg9
```

```{r}

colors_map <- c("within" = "#301934", 
                                "unprecedented"     = "#91C499")

# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_modern_30yr <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Event category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin     = margin(0,0,0,0),
    legend.position = "none",   # inside top-left
    legend.justification = c("left", "bottom"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

map_modern_30yr
```


### ModE-Sim recent (1975-1999)

```{r}
get_minmax <- function(df){
  df |> 
    filter(year %in% 1975:1999) |> 
    ungroup() |> 
    summarise(pcwd_30x_ref0_max = max(pcwd_30x, na.rm = TRUE), pcwd_30x_ref0_min = min(pcwd_30x, na.rm = TRUE))
}

df_30x_sliding_recent <- df_modesim_30x_sliding |> 
  group_by(region) |> 
  nest() |> 
  mutate(data = purrr::map(data, ~get_minmax(.))) |> 
  unnest(data) |> 
  left_join(
    df_era5_30x |> 
      dplyr::select(region, pcwd_x30_ref1),
    by = join_by(region)
  )
```

```{r}
tmp <- df_30x_sliding_recent |> 
  mutate(is_greater = pcwd_x30_ref1 > pcwd_30x_ref0_max,
  color_cat = case_when(
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    ))

n_greater <- tmp |> 
  ungroup() |> 
  summarise(n_greater = sum(is_greater)) |> 
  pull(n_greater)

gg10 <- tmp |> 
  ggplot(aes(x = pcwd_30x_ref0_max, xend = pcwd_30x_ref0_min, y = pcwd_x30_ref1, color = color_cat)) +
  geom_segment(show.legend = FALSE, linewidth = 0.75) +
  geom_point(show.legend = FALSE, size = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
    scale_color_manual(name = "Event category", values = c("within" = "#301934",
                                                "unprecedented"     = "#91C499")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "ModE-Sim 1975-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater)),
    title = "   "
  )

gg10
```

```{r}

colors_map <- c("within" = "#301934", 
                                "unprecedented"     = "#91C499")

# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_recent_30yr <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Event category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin     = margin(0,0,0,0),
    legend.position = "none",   # inside top-left
    legend.justification = c("left", "bottom"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

map_recent_30yr
```


### ModE-Sim historical (1850-1999)

```{r}
get_minmax <- function(df){
  df |> 
    filter(year %in% 1850:1999) |> 
    ungroup() |> 
    summarise(pcwd_30x_ref0_max = max(pcwd_30x, na.rm = TRUE), pcwd_30x_ref0_min = min(pcwd_30x, na.rm = TRUE))
}

df_30x_sliding_historical <- df_modesim_30x_sliding |> 
  group_by(region) |> 
  nest() |> 
  mutate(data = purrr::map(data, ~get_minmax(.))) |> 
  unnest(data) |> 
  left_join(
    df_era5_30x |> 
      dplyr::select(region, pcwd_x30_ref1),
    by = join_by(region)
  )
```

```{r}
tmp <- df_30x_sliding_historical |> 
  mutate(is_greater = pcwd_x30_ref1 > pcwd_30x_ref0_max,
  color_cat = case_when(
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    ))

n_greater <- tmp |> 
  ungroup() |> 
  summarise(n_greater = sum(is_greater)) |> 
  pull(n_greater)

gg11 <- tmp |> 
  ggplot(aes(x = pcwd_30x_ref0_max, xend = pcwd_30x_ref0_min, y = pcwd_x30_ref1, color = color_cat)) +
  geom_segment(show.legend = FALSE, linewidth = 0.75) +
  geom_point(show.legend = FALSE, size = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
    scale_color_manual(name = "Event category", values = c("within" = "#301934",
                                                "unprecedented"     = "#91C499")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "ModE-Sim 1850-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater)),
    title = "   "
  )

gg11
```

```{r}

colors_map <- c("within" = "#301934", 
                                "unprecedented"     = "#91C499")

# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_historic_30yr <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Event category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin     = margin(0,0,0,0),
    legend.position = "none",   # inside top-left
    legend.justification = c("left", "bottom"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

map_historic_30yr
```

### ModE-Sim multi-century (1420-1999)

```{r}
get_minmax <- function(df){
  df |> 
    filter(year %in% 1420:1999) |> 
    ungroup() |> 
    summarise(pcwd_30x_ref0_max = max(pcwd_30x, na.rm = TRUE), pcwd_30x_ref0_min = min(pcwd_30x, na.rm = TRUE))
}

df_30x_sliding_multicent <- df_modesim_30x_sliding |> 
  group_by(region) |> 
  nest() |> 
  mutate(data = purrr::map(data, ~get_minmax(.))) |> 
  unnest(data) |> 
  left_join(
    df_era5_30x |> 
      dplyr::select(region, pcwd_x30_ref1),
    by = join_by(region)
  )
```

```{r}
tmp <- df_30x_sliding_multicent |> 
  mutate(is_greater = pcwd_x30_ref1 > pcwd_30x_ref0_max,
  color_cat = case_when(
      is_greater ~ "unprecedented",
      TRUE ~ "within"
    ))

n_greater <- tmp |> 
  ungroup() |> 
  summarise(n_greater = sum(is_greater)) |> 
  pull(n_greater)

gg12 <- tmp |> 
  ggplot(aes(x = pcwd_30x_ref0_max, xend = pcwd_30x_ref0_min, y = pcwd_x30_ref1, color = color_cat)) +
  geom_segment(show.legend = FALSE, linewidth = 0.75) +
  geom_point(show.legend = FALSE, size = 0.75) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  coord_equal() +
    scale_color_manual(name = "Event category", values = c("within" = "#301934",
                                                "unprecedented"     = "#91C499")) +
  theme_classic() +
  scale_x_log10() +
  scale_y_log10() +
  labs(
    x = "ModE-Sim 1420-1999 (mm)",
    y = "ERA5 2000-2024 (mm)",
    subtitle = bquote(italic(N[plain("unprecedented")]) == .(n_greater)),
    title = "   "
  )

gg12
```
```{r}

colors_map <- c("within" = "#301934", 
                                "unprecedented"     = "#91C499")

# Join categories onto the spatial data
shown_sf <- refregions_sf %>% 
  inner_join(tmp %>% dplyr::select(region, color_cat), by = "region")

# centroids for labels
pts <- shown_sf %>% 
  st_centroid() %>% 
  mutate(label = region)

# build the map
map_multicent_30yr <- ggplot() +
  geom_sf(data = land, fill = "lightgray", color = NA) +
  geom_sf(
    data = shown_sf,
    aes(fill = color_cat),   # <-- use color_cat instead of continent
    color = "grey20",
    size = 0.2,
    alpha = 0.4
  ) +
  geom_sf_text(
    data = pts,
    aes(label = label),
    size = 2.6,
    colour= "black"
  ) +
  scale_fill_manual(
    values   = colors_map,
    na.value = "transparent",
    name     = "Event category"
  ) +
  coord_sf(
    xlim = c(-180, 178.125), ylim = c(-60, 90), expand = FALSE
  ) +
  theme_void() +
  theme(
    plot.margin     = margin(0,0,0,0),
    legend.position = "none",   # inside top-left
    legend.justification = c("left", "bottom"),
    legend.background = element_blank(),
    legend.title = element_text(face = "bold")
  )

map_multicent_30yr
```


```{r}
#map_recent <- map_recent + theme(plot.margin = margin(0, 0, 0, 0))
col1 <- plot_grid(
  gg1, gg2, gg3, gg4, 
  ncol = 1,
  nrow =4,
  rel_widths = c(1,1, 1,1), # map takes smaller horizontal space
  align = "v",
    labels = c(
    "a) Record annual PCWD", "b)", "c)", "d)"),
    # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  expand      = TRUE 
)

col2 <- plot_grid(
  gg5, gg6, gg7, gg8, 
  ncol = 1,
  nrow =4,
  rel_widths = c(1,1,1, 1), # map takes smaller horizontal space
  align = "v",
    labels = c("e) Record 25-year mean PCWD", "f)", "g)", "h)"),
    # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  expand      = TRUE 
  )

col3 <- plot_grid(
  gg9, gg10,  gg11, gg12,
  ncol = 1,
  nrow =4,
  rel_widths = c(1,1,1, 1), # map takes smaller horizontal space
  align = "v",
  labels = c("i) 30-year extreme PCWD", "j)", "k)", "l)"),
    # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  expand      = TRUE 
  )
```

## All combined plot

```{r}
gg_combined <- cowplot::plot_grid(
  col1, col2, col3, 
  ncol=3
  )
ggsave(here("figures/combined_record_pcwd.pdf"), plot = gg_combined, width = 12, height = 15)
```

## Combine all maps 

```{r}
map_recent <- map_recent + theme(plot.margin = margin(t = 0, r = 10, b = 0, l = 0))
map_mode_recent <- map_mode_recent + theme(plot.margin = margin(t = 0, r = 10, b = 0, l = 0))
map_historical <- map_historical + theme(plot.margin = margin(t = 0, r = 10, b = 0, l = 0))
map_multicen <- map_multicen + theme(plot.margin = margin(t = 0, r = 10, b = 0, l = 0))

map_modern_25 <- map_modern_25 + theme(plot.margin = margin(t = 0, r = 10, b = 0, l = 0))
map_recent_25 <- map_recent_25 + theme(plot.margin = margin(t = 0, r = 10, b = 0, l = 0))
map_historical_25 <- map_historical_25 + theme(plot.margin = margin(t = 0, r = 10, b = 0, l = 0))
map_multicent_25 <- map_multicent_25 + theme(plot.margin = margin(t = 0, r = 10, b = 0, l = 0))

row1 <- plot_grid(
  map_recent, map_mode_recent, map_historical, map_multicen, 
  ncol = 1,
  nrow =4,
  rel_widths = c(1,1, 1,1), # map takes smaller horizontal space
  align = "v",
    labels = c(
    "a) Record annual PCWD mapped", "b)", "c)", "d)"),
    # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  expand      = TRUE 
)

row2 <- plot_grid(
  map_modern_25, map_recent_25, map_historical_25, map_multicent_25, 
  ncol = 1,
  nrow =4,
  rel_widths = c(1,1,1, 1), # map takes smaller horizontal space
  align = "v",
    labels = c("e) Record 25-year mean PCWD mapped", "f)", "g)", "h)"),
    # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  expand      = TRUE 
  )

row3 <- plot_grid(
  map_modern_30yr, map_recent_30yr,  map_historic_30yr, map_multicent_30yr,
  ncol = 1,
  nrow =4,
  rel_widths = c(1,1,1, 1), # map takes smaller horizontal space
  align = "v",
  labels = c("i) 30-year extreme PCWD", "j)", "k)", "l)"),
    # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  expand      = TRUE 
  )
```

```{r}
gg_combined_maps <- cowplot::plot_grid(
  row1, row2, row3, 
  nrow=1, ncol=3
  )

ggsave(here("figures/combined_record_pcwd_maps.pdf"), plot = gg_combined_maps, width = 15, height = 12)
```

