---
title: "Maps_trends"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(extRemes)
library(ggplot2)
library(cwd)
library(ncdf4)
library(reshape2)
library(ggpubr)
library(maps)
library(raster)
library(rnaturalearth)
library(sf)
library(rgdal)
library(sp)
library(RColorBrewer)
library(rJava)
library(loadeR.java)
library(transformeR)
library(loadeR)
library(visualizeR)
library(geoprocessoR)
library(tidyverse)
library(abind)
library(patchwork)
library(gridExtra)
```
# Trend calculation and spatial maps


## Maps showing variability of droughts per gridcell 

- 1420 mean epoch variability for each gridcell
- 1850 mean epoch variabiltiy for each gridcell
- 50 year variability for 1420-1470, 1700-1750, 1950-2000
- which regions show high variability? Is it the same for each time period? 

Read in lat, lon and time info: 
```{r}
#lat lon and time info from netcdf files
##read in data - stored on Ubelix server 
input_file_1850 <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850/PCWD_ANNMAX.nc"
input_file_1420 <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1420/PCWD_ANNMAX.nc"

nc_pwcd_1850 <- nc_open(input_file_1850)
pcwd_annmax_1850 = ncvar_get(nc_pwcd_1850, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_1850, varid="lon")
lat = ncvar_get(nc_pwcd_1850, varid="lat")
time_1850 = ncvar_get(nc_pwcd_1850, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates_1850 <- reference_date + time_1850

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1850)

#1420 file for 1420 time
nc_pwcd_1420 <- nc_open(input_file_1420)
pcwd_annmax_1420 = ncvar_get(nc_pwcd_1420, varid="pcwd_annmax")
time_1420 = ncvar_get(nc_pwcd_1420, varid="time")
# Convert to actual dates (days since 2001-01-01)
time_dates_1420 <- reference_date + time_1420

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1420)
```

Aggregate all 1850 files: 

```{r}
# Define the base path
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)

# Extract ensemble member identifiers from folder names
ensemble_members <- basename(folders)

# Identify the target files
target_files <- file.path(folders, "04_result_1850/PCWD_ANNMAX.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included

# Filter ensemble members to match target files
ensemble_members <- ensemble_members[file.exists(target_files)]

# Initialize a list to store data for each ensemble member
data_by_ensemble_1850 <- list()

# Extract dimensions from the first file to initialize the array size for consistency
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "pcwd_annmax"))[1]
dim_y <- dim(ncvar_get(nc, "pcwd_annmax"))[2]
dim_z <- dim(ncvar_get(nc, "pcwd_annmax"))[3]
nc_close(nc)

# Loop through all the target files and store the data in a list
for (i in seq_along(target_files)) {
  file <- target_files[i]
  member <- ensemble_members[i]
  
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "pcwd_annmax")  # Adjust "pcwd_annmax" to your variable name
  
  # Store the data in the list, keyed by the ensemble member
  data_by_ensemble_1850[[member]] <- data
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the data list
print(names(data_by_ensemble_1850))  # Check stored ensemble members
print(dim(data_by_ensemble_1850[[1]]))  # Check dimensions of the data for the first ensemble member

```

Aggregate all 1420 files: 

```{r}
# Define the base path
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)

# Extract ensemble member identifiers from folder names
ensemble_members <- basename(folders)

# Identify the target files
target_files <- file.path(folders, "04_result_1420/PCWD_ANNMAX.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included
target_files <- target_files[-1] #remove m001 file
# Filter ensemble members to match target files
ensemble_members <- ensemble_members[file.exists(target_files)]
ensemble_members <- ensemble_members[-1]

# Initialize a list to store data for each ensemble member
data_by_ensemble_1420 <- list()

# Extract dimensions from the first file to initialize the array size for consistency
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "pcwd_annmax"))[1]
dim_y <- dim(ncvar_get(nc, "pcwd_annmax"))[2]
dim_z <- dim(ncvar_get(nc, "pcwd_annmax"))[3]
nc_close(nc)

# Loop through all the target files and store the data in a list
for (i in seq_along(target_files)) {
  file <- target_files[i]
  member <- ensemble_members[i]
  
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "pcwd_annmax")  # Adjust "pcwd_annmax" to your variable name
  
  # Store the data in the list, keyed by the ensemble member
  data_by_ensemble_1420[[member]] <- data
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the data list
print(names(data_by_ensemble_1420))  # Check stored ensemble members
print(dim(data_by_ensemble_1420[[1]]))  # Check dimensions of the data for the first ensemble member

```


```{r}
# Combine the list into a 4D array
ensemble_array_1420 <- abind(data_by_ensemble_1420, along = 4)

# Check dimensions of the resulting array
dim(ensemble_array_1420)  # Should be [192, 96, 431, 19]

# 2. Calculate the mean across all ensemble members (4th dimension)
# This reduces the array to [192, 96, 431]
ensemble_mean_1420 <- apply(ensemble_array_1420, c(1, 2, 3), mean, na.rm = TRUE)

EM_1420 <- apply(ensemble_mean_1420, c(1, 2), mean, na.rm = TRUE)


# 3. Calculate the standard deviation over the 431 years (3rd dimension)
# This reduces the array to [192, 96] (longitude × latitude)
std_dev_over_time_1420 <- apply(ensemble_mean_1420, c(1, 2), sd, na.rm = TRUE)

#### 1850 epoch ######

# Combine the list into a 4D array
ensemble_array_1850 <- abind(data_by_ensemble_1850, along = 4)

# Check dimensions of the resulting array
dim(ensemble_array_1850)  # Should be [192, 96, 431, 19]

# 2. Calculate the mean across all ensemble members (4th dimension)
# This reduces the array to [192, 96, 431]
ensemble_mean_1850 <- apply(ensemble_array_1850, c(1, 2, 3), mean, na.rm = TRUE)
EM_1850 <- apply(ensemble_mean_1850, c(1, 2), mean, na.rm = TRUE)
# 3. Calculate the standard deviation over the 431 years (3rd dimension)
# This reduces the array to [192, 96] (longitude × latitude)
std_dev_over_time_1850 <- apply(ensemble_mean_1850, c(1, 2), sd, na.rm = TRUE)

#### epoch mean
##Create a new list to store the combined data
data_combined <- list()

# Loop over each ensemble member
for (i in names(data_by_ensemble_1420)) {
  # Remove last year from 1420 dataset (assumes 3rd dimension is time)
  data_1420_trimmed <- data_by_ensemble_1420[[i]][,,1:(dim(data_by_ensemble_1420[[i]])[3] - 1)]

  # Bind with 1850 data (all years)
  data_combined[[i]] <- abind::abind(
    data_1420_trimmed,
    data_by_ensemble_1850[[i]],
    along = 3
  )
}

mean_ensemble_combined <- array(NA, dim = c(192, 96, 590))
# Compute the mean over all ensemble members
ensemble_array <- simplify2array(data_combined)  # Convert list to array
mean_ensemble_combined <- apply(ensemble_array, c(1,2,3), mean, na.rm = TRUE) #time mean
mean_over_time_comb <- apply(mean_ensemble_combined, c(1,2), mean)


```
Plot mean values for African regions in certain time periods

```{r}
options(scipen = 999) #surpress scientific notation
#data to plot:
data <- mean_over_time_comb

#plot
# Convert first year's data into a raster object
r <- raster(t(data), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(0, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, 1200, 2000, Inf)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Updated custom color palette with two additional colors
custom_colors <- c("#004f5a",  # New color: a darker turquoise at the beginning
                   "#006d75", "#1a9e96", "#55c1ab", "#85c68b", "#d1c54b", "#e8b249", 
                   "#e3842a", "#d45321", "#ba2e1f", "#9e2d2a", "#803c3d", "#63363f", 
                   "#9c276e",  # New color: a deeper shade in the pink tones
                   "#d238a5")

break_labels <- c(
  as.character(scales::label_comma(accuracy = 1)(custom_bins[-length(custom_bins)])),
  "Ocean"
)

# Define bounding box for Africa
#bbox <- c(xmin = -30, xmax = 60, ymin = -35, ymax = 30)  # Approximate bounding box for Oceania


# Define the ggplot2 object with the gradient scale
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "PCWD [mm]", values = custom_colors, na.value = "grey50", 
                    labels = break_labels) +  # Label ocean
  labs(title = "Mean Annmax PCWD over 600 years ", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "none", legend.title = element_text(size=18), legend.text = element_text(size=18), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PCWD [mm]", reverse = TRUE))  # Adjust legend for better presentation

#ggsave("PCWD_1420.png", width = 20, height = 12, units = "cm")

```



Plot standard deviation of mean PCWD for both epochs

```{r}
###### 1420 epoch #######

# Convert first year's data into a raster object
r <- raster(t(std_dev_over_time_1420), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(0, 10, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, Inf)  # No bins beyond 900

# Define a custom color palette ranging from white to light yellow to dark red
custom_colors <- colorRampPalette(c("lightyellow", "orange", "red", "darkred"))(length(custom_bins) - 1)


# Define the numeric labels for the breaks
break_labels <- c(as.character(scales::label_number(accuracy = 1)(custom_bins[-length(custom_bins)])), "Ocean")

# Define the ggplot2 object with the gradient scale
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "5 yr return level", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "PCWD ANNMAX STD over 1420 epoch (EM)", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "STD (mm)", reverse = TRUE))  # Adjust legend for better presentation


###### 1850 epoch #######

# Convert first year's data into a raster object
r <- raster(t(std_dev_over_time_1850), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(0, 10, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, Inf)  # No bins beyond 900

# Define a custom color palette ranging from white to light yellow to dark red
custom_colors <- colorRampPalette(c("lightyellow", "orange", "red", "darkred"))(length(custom_bins) - 1)


# Define the numeric labels for the breaks
break_labels <- c(as.character(scales::label_number(accuracy = 1)(custom_bins[-length(custom_bins)])), "Ocean")

# Define the ggplot2 object with the gradient scale
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "5 yr return level", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "PCWD ANNMAX STD over 1850 epoch (EM)", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "STD (mm)", reverse = TRUE))  # Adjust legend for better presentation


```


## Investigate trends entire epoch

- calculate linear trends for each ensemble member

```{r}
################### 1420-1849 ###############################

##loop through ensemble member and calculate linear trend for each one
# Load necessary libraries
library(dplyr)
library(tidyr)
library(purrr)

# Initialize an empty list to store results
trend_results <- list()

# Iterate through each ensemble member
for (ensemble_name in names(data_by_ensemble_1420)) {
  # Extract the data for the current ensemble member (3D array)
  ensemble_data <- data_by_ensemble_1420[[ensemble_name]]  #trends over the entire epoch
  
  # Get dimensions
  lon_dim <- dim(ensemble_data)[1]
  lat_dim <- dim(ensemble_data)[2]
  time_dim <- dim(ensemble_data)[3]
  
  # Create a data frame for regression: Reshape data into `grid_cell x time`
  reshaped_data <- as.data.frame(
    array(ensemble_data, dim = c(lon_dim * lat_dim, time_dim))
  )
  colnames(reshaped_data) <- 1:time_dim  # Name columns as time indices
  
  # Add grid cell indices
  reshaped_data <- reshaped_data %>%
    mutate(grid_id = 1:(lon_dim * lat_dim)) %>%
    pivot_longer(
      cols = -grid_id,
      names_to = "time",
      values_to = "value"
    ) %>%
    mutate(time = as.numeric(time))
  
  # Perform linear regression for each grid cell
  trends <- reshaped_data %>%
    group_by(grid_id) %>%
    summarize(
      trend = coef(lm(value ~ time))[["time"]],  # Extract slope (trend)
      .groups = "drop"
    )
  
  # Reshape trend data back into 2D array (lon x lat)
  trend_matrix <- matrix(trends$trend, nrow = lon_dim, ncol = lat_dim)
  
  # Store the trend matrix in the results list
  trend_results[[ensemble_name]] <- trend_matrix
}

# `trend_results` now contains a 2D trend array (lon x lat) for each ensemble member


#### 1850 - 2009 trends

# Initialize an empty list to store results
trend_results_1850 <- list()

# Define the number of years to include in trend calculation
last_n_years <- 59  # Modify this if needed


# Iterate through each ensemble member
for (ensemble_name in names(data_by_ensemble_1850)) {
  # Extract the data for the current ensemble member (3D array)
  ensemble_data <- data_by_ensemble_1850[[ensemble_name]]
  
  # Get dimensions
  lon_dim <- dim(ensemble_data)[1]
  lat_dim <- dim(ensemble_data)[2]
  time_dim <- dim(ensemble_data)[3]
  
  # Determine the time indices for the last 50 years
  last_time_indices <- max(1, time_dim - last_n_years + 1):time_dim
  
  # Create a data frame for regression: Reshape data into `grid_cell x time`
  reshaped_data <- as.data.frame(
    array(ensemble_data, dim = c(lon_dim * lat_dim, time_dim))
  )
  colnames(reshaped_data) <- 1:time_dim  # Name columns as time indices
  
  # Add grid cell indices
  reshaped_data <- reshaped_data %>%
    mutate(grid_id = 1:(lon_dim * lat_dim)) %>%
    pivot_longer(
      cols = -grid_id,
      names_to = "time",
      values_to = "value"
    ) %>%
    mutate(time = as.numeric(time)) %>%
  # Filter only the last 50 years
    filter(time %in% last_time_indices)
  
# Perform linear regression for each grid cell and extract slope & p-value
trends <- reshaped_data %>%
  group_by(grid_id) %>%
  summarize(
    trend = coef(lm(value ~ time))[["time"]],  # Extract slope (trend)
    p_value = summary(lm(value ~ time))$coefficients["time", "Pr(>|t|)"],  # Extract p-value
    .groups = "drop"
  )
  
  # Reshape trend data back into 2D array (lon x lat)
  trend_matrix <- matrix(trends$trend, nrow = lon_dim, ncol = lat_dim)
  
  # Reshape p-value data back into 2D array (lon x lat)
  p_matrix <- matrix(trends$p_value, nrow = lon_dim, ncol = lat_dim)
  
  # Store in results list
  trend_results_1850[[ensemble_name]] <- list(trend = trend_matrix, p_value = p_matrix)
}


```

Ensemble means and significance masking:
```{r}
### calculate mean of all trends for plotting: 
# Calculate the mean trend across all ensemble members for each grid cell
mean_trend_1420 <- Reduce("+", trend_results) / length(trend_results)

#range(mean_trend_1420)

# Compute the mean trend
mean_trend_1850 <- Reduce("+", lapply(trend_results_1850, `[[`, "trend")) / length(trend_results_1850)

# Compute the mean p-value
mean_p_1850 <- Reduce("+", lapply(trend_results_1850, `[[`, "p_value")) / length(trend_results_1850)

# Create a significance mask (TRUE for significant, FALSE otherwise)
sig_mask <- mean_p_1850 < 0.05

```


Plot EM trend for 1420-1850:
```{r}
# Convert first year's data into a raster object
r <- raster(t(mean_trend_1420), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations


# Create the plot
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = layer), na.rm = TRUE) +  # Use the binned layer for color
  scale_fill_distiller(
    palette = "RdBu", direction = -1,
    limits = c(-7, 7), name = "trends", na.value = "lightgrey", values = scales::rescale(c(-1, -0.1, 0, 0.1, 1))
  ) +
  labs(title = "Mean ANNMAX PCWD Trend (1420-1849)", x = "Longitude", y = "Latitude") +
  theme_classic() +
  
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  
  theme(
    legend.title = element_text(size = 12), 
    legend.text = element_text(size = 10)
  ) +
  guides(fill = guide_colorbar(title = "mm/yr"))

```


Plot trends within the last epoch:
```{r}
# Convert first year's data into a raster object
r <- raster(t(mean_trend_1850), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Convert significance mask into a raster (same dimensions as trend)
sig_r <- raster(t(sig_mask), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(sig_r) <- "+proj=longlat +datum=WGS84"
sig_r <- flip(sig_r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))
sig_masked <- mask(sig_r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

sig_df <- as.data.frame(sig_masked, xy = TRUE)
colnames(sig_df) <- c("x", "y", "sig")

# Filter non-significant grid cells
sig_df <- sig_df %>% filter(sig == 1)  # Retain only non-significant areas

# Create the plot
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = layer), na.rm = TRUE) +  # Continuous fill
  
  # # Continuous diverging blue-red color scale centered at zero with nonlinear scaling
  # scale_fill_gradient2(
  #   low = "blue", mid = "white", high = "red", midpoint = 0, 
  #   limits = c(-7, 7), name = "trends", na.value = "lightgrey"
  # ) +
  
  # Continuous diverging blue-red color scale centered at zero with improved scaling
  scale_fill_distiller(
    palette = "RdBu", direction = -1,
  #   limits = c(-25, 25), name = "trends", na.value = "grey50", values = scales::rescale(c(-1, -0.1, 0, 0.1, 1))
  # ) +
      limits = c(-25, 25), name = "trends", na.value = "grey50", values = scales::rescale(c(-10, -1, 0, 1, 10))
  ) +
  
  labs(title = "Mean ANNMAX PCWD Trend (1950-2009)", x = "Longitude", y = "Latitude") +
  theme_classic() +
  
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  # Overlay significant areas with hatching or stippling
  geom_point(data = sig_df, aes(x = x, y = y, shape = "Significant"), 
             size = 1.5, color = "#15B34C") +

  scale_shape_manual(name = "", values = c("Significant" = 4)) +  # shape 4 = 'x'
  
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.title = element_text(size=12), legend.text = element_text(size=12), 
        axis.text = element_text(size=13), axis.title = element_text(size=13)) +
  guides(fill = guide_colorbar(title = "mm/yr"))
```

