---
title: "PCWD_maxlen_analysis"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=FALSE}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(extRemes)
library(ggplot2)
library(cwd)
library(ncdf4)
library(reshape2)
library(ggpubr)
library(maps)
library(raster)
library(rnaturalearth)
library(sf)
library(rgdal)
library(sp)
library(RColorBrewer)
library(rJava)
library(loadeR.java)
library(transformeR)
library(loadeR)
library(visualizeR)
library(geoprocessoR)
library(tidyverse)
library(abind)
library(patchwork)
library(gridExtra)
```

# Calculating regional averages of drought duration values 

```{r regions setup, echo=FALSE, warning=FALSE, message=FALSE}
#Load reference regions and coastlines:

load("/storage/homefs/ph23v078/Reference_regions/IPCC-WGI-reference-regions-v4_R.rda", verbose = TRUE)

#simplify this object by converting it to a SpatialPolygons class object (i.e., only the polygons are retained and their attributes discarded):
refregions <- as(IPCC_WGI_reference_regions_v4, "SpatialPolygons")

temp.dir <- tempdir()
unzip("/storage/homefs/ph23v078/Reference_regions/ne_110m_coastline.zip", exdir = temp.dir)
coastLines <- readOGR(dsn = temp.dir, layer = "ne_110m_coastline")

names(names(refregions))

proj4string(refregions)
WCE <- refregions[c("WCE")] #spatial polygon for Western Central Europe

```

Load IPCC reference region information
```{r map of regions, echo=FALSE, warning=FALSE, message=FALSE}
# Convert `refregions` (SpatialPolygons) to an `sf` object for ggplot compatibility
refregions_sf <- st_as_sf(refregions)

# Define continent assignments
continent_mapping <- list(
  "Europe" = c("NEU", "WCE", "EEU", "MED"),
  "North America" = c("NWN", "NEN", "WNA", "CNA", "ENA", "GIC"),
  "Africa" = c("SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG"),
  "Asia" = c("WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "WSB", "ESB", "RFE", "RAR"),
  "Central America" =c("NCA", "SCA", "CAR"),
  "South America" = c("NWS", "NSA", "NES", "SAM", "SWS", "SES", "SSA"),
  "Australia" = c("NAU", "CAU", "EAU", "SAU", "NZ")
)

# Assign continents to regions
refregions_sf <- refregions_sf %>%
  mutate(
    continent = case_when(
      row.names(refregions) %in% continent_mapping$Europe ~ "Europe",
      row.names(refregions) %in% continent_mapping$`North America` ~ "North America",
      row.names(refregions) %in% continent_mapping$Africa ~ "Africa",
      row.names(refregions) %in% continent_mapping$Asia ~ "Asia",
      row.names(refregions) %in% continent_mapping$`Central America` ~ "Central America",
      row.names(refregions) %in% continent_mapping$`South America` ~ "South America",
      row.names(refregions) %in% continent_mapping$Australia ~ "Australia",
      TRUE ~ "Other"  # For regions not mapped to a continent
    )
  ) 
# Define continent colors
continent_colors <- c(
  "Europe" = "pink",
  "North America" = "lightgreen",
  "Africa" = "purple",
  "Asia" = "orange",
  "Central America" = "lightblue",
  "South America" = "yellow",
  "Australia" = "blue",
  "Other" = "white"
)

# Extract region centroids for labeling
region_labels <- refregions_sf %>%
  st_centroid() %>%
  mutate(label = row.names(refregions))  # Add region names as labels


# Load land data for filling continents
land <- ne_countries(scale = "medium", returnclass = "sf")
```


Spatial average for WCE only for development:
```{r read in EMs WCE only, echo=FALSE, message=FALSE, warning=FALSE}
#here reading in only for WCE region for development
#aggregate EM for region i.e. 2 for loops, 1 looping over EMs, second looping over region, aggregating into one file

###############################read in data##################################
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim"
folders <- list.files(path, full.names = TRUE, pattern = "m[0-9]{3}_tidy")

# Identify the target files
target_files <- file.path(folders, "02_2_pcwd_result_1850/PCWD_maxlen.nc")
target_files <- target_files[file.exists(target_files)]


#################################all EMS#######################################
#e.g. loop for EMs: ### append data i.e. get rid of time dimension and just append to exisiting data
# Extract unique ensemble member identifiers from folder names
ensemble_members <- unique(basename(folders))

# Initialize an empty list to store spatial averages for each file
spatial_avg_list <- list()

# Loop over all ensemble member folders
for (em in ensemble_members) {
  # Construct the folder path for the current ensemble member
  folder <- file.path(path, em)
  # Construct file paths for the two time periods
  file_1420 <- file.path(folder, "02_2_pcwd_result_1420/PCWD_maxlen.nc")
  file_1850 <- file.path(folder, "02_2_pcwd_result_1850/PCWD_maxlen.nc")
  
  # Check if both files exist
  if (file.exists(file_1420) && file.exists(file_1850)) {
    # Load grid data for each time period
    grid_1420 <- loadGridData(dataset = file_1420, var = "pcwd_maxlen")
    grid_1850 <- loadGridData(dataset = file_1850, var = "pcwd_maxlen")
    
    # Set spatial projection
    grid_1420 <- setGridProj(grid = grid_1420, proj = proj4string(refregions))
    grid_1850 <- setGridProj(grid = grid_1850, proj = proj4string(refregions))
    
    # Perform spatial overlay
    grid.eu_1420 <- overGrid(grid_1420, WCE)
    grid.eu_1850 <- overGrid(grid_1850, WCE)
    
    # Extract data arrays
    data_array_1420 <- grid.eu_1420$Data[1:430,,]
    data_array_1850 <- grid.eu_1850$Data
    
    # Combine data arrays along the time dimension
    combined_data_array <- abind::abind(data_array_1420, data_array_1850, along = 1)
    
    # Compute spatial average for each time step
    spatial_avg <- apply(combined_data_array, 1, function(slice) {
      mean(slice, na.rm = TRUE) # Compute mean for the spatial dimensions, ignoring NA
    })
    
    # Store the spatial average in the list
    spatial_avg_list[[em]] <- spatial_avg
  } else {
    warning(paste("Missing data for ensemble member:", em))
  }
}

# Combine all spatial averages into a single array with dimensions 160 x EMs
result_array.wce <- do.call(cbind, spatial_avg_list)

# Check the result
#dim(result_array.wce) # Should be 160 x EMs
```

```{r plot all EMs as spaghetti}
# Assuming your years range from 1420 to 2009
years <- 1420:2009

# Convert the matrix to a data frame
result_df <- as.data.frame(result_array.wce)

# Add a column for years
result_df$Year <- years


# Reshape to long format
result_long <- pivot_longer(
  result_df,
  cols = -Year,  # All columns except 'Year' are to be gathered
  names_to = "Ensemble_Member",
  values_to = "Duration"
)

# # Create the spaghetti plot
# ggplot(result_long, aes(x = Year, y = Duration, color = Ensemble_Member)) +
#   geom_line() +
#   labs(
#     title = "Drought Duration for WCE",
#     x = "Year",
#     y = "Max Drought Duration (days)",
#     color = "Ensemble Member"
#   ) +
#   theme_classic() +
#   ylim(0, 2000) +  # Set y-axis limits
#   theme(legend.position = "right")

# Specify the ensemble member you want to plot
ensemble_to_plot <- "m015_tidy"

# Filter the data for the chosen ensemble member
filtered_data <- result_long %>%
  filter(Ensemble_Member == ensemble_to_plot)

# Create the plot for the single ensemble member
ggplot(filtered_data, aes(x = Year, y = Duration)) +
  geom_line(color = "blue") +  # Set the line color
  labs(
    title = paste("Plot of", ensemble_to_plot, "Over Time"),
    x = "Year",
    y = "Max Drought Duration (days)"
  ) +
  theme_classic()

```

Calculate regional averages of maxlen data:
```{r read in EMS all regions, eval=FALSE, echo=FALSE}
###############################read in data##################################
#as before but loop over all regions also, saving everything in an array
path <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim"
folders <- list.files(path, full.names = TRUE, pattern = "m[0-9]{3}_tidy")

# Identify the target files
target_files <- file.path(folders, "02_2_pcwd_result_1850/PCWD_maxlen.nc")
target_files <- target_files[file.exists(target_files)]

# Extract unique ensemble member identifiers from folder names
ensemble_members <- unique(basename(folders))

# # List of regions to loop over --- remove ocean basins
# regions <- c("GIC", "NWN", "NEN", "WNA", "CNA", "ENA", "NCA", "SCA", "CAR", "NWS",
#              "NSA", "NES", "SAM", "SWS", "SES", "SSA", "NEU", "WCE", "EEU", "MED",
#              "SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG", "RAR", "WSB",
#              "ESB", "RFE", "WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "NAU",
#              "CAU", "EAU", "SAU", "NZ")
regions <- c("NZ")
# Initialize a list to store results for all regions
regional_results <- list()

# Loop over all regions
for (region in regions) {
  # Extract the spatial object corresponding to the current region
  region_object <- refregions[c(region)]
  
  # Check if the subset is valid
  if (is.null(region_object)) {
    warning(paste("Region not found:", region))
    next
  }
  
  # Initialize a list to store spatial averages for each file in the current region
  spatial_avg_list <- list()
  
  # Loop over all ensemble member folders
  # Loop over all ensemble member folders
  for (em in ensemble_members) {
    # Construct the folder path for the current ensemble member
    folder <- file.path(path, em)
   # Construct file paths for the two time periods
   file_1420 <- file.path(folder, "02_2_pcwd_result_1420/PCWD_maxlen.nc")
   file_1850 <- file.path(folder, "02_2_pcwd_result_1850/PCWD_maxlen.nc")
  
   # Check if both files exist
   if (file.exists(file_1420) && file.exists(file_1850)) {
    # Load grid data for each time period
    grid_1420 <- loadGridData(dataset = file_1420, var = "pcwd_maxlen")
    grid_1850 <- loadGridData(dataset = file_1850, var = "pcwd_maxlen")
      
      # Set spatial projection
      grid_1420 <- setGridProj(grid = grid_1420, proj = proj4string(refregions))
      grid_1850 <- setGridProj(grid = grid_1850, proj = proj4string(refregions))
      
      # Perform spatial overlay for the current region
      grid_region_1420 <- overGrid(grid_1420, region_object)
      grid_region_1850 <- overGrid(grid_1850, region_object)
      
      # Extract data arrays
      data_array_1420 <- grid_region_1420$Data[1:430,,]
      data_array_1850 <- grid_region_1850$Data
      
      # Combine data arrays along the time dimension
      combined_data_array <- abind::abind(data_array_1420, data_array_1850, along = 1)
      
      # Compute spatial average for each time step
      spatial_avg <- apply(combined_data_array, 1, function(slice) {
        mean(slice, na.rm = TRUE) # Compute mean for the spatial dimensions, ignoring NA
      })
      
      # Store the spatial average in the list
      spatial_avg_list[[em]] <- spatial_avg
    } else {
      warning(paste("Missing data for ensemble member:", em, "in region:", region))
    }
  }
  
  # Combine all spatial averages into a single array for the current region
  result_array_region <- do.call(cbind, spatial_avg_list)
  
  # Store the result for the current region
  regional_results[[region]] <- result_array_region
}

#save calculated list
saveRDS(regional_results, file="~/cwd_global/data/regionalResults_duration.RData")

#Load with
regional_results_test <- readRDS("~/cwd_global/data/regionalResults_duration.RData")

```

## Trend analysis of duration values

```{r}
#remove m001: 
# List of regions where the first ensemble member should be removed
regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")

# Loop through each specified region and remove the first column (m001_tidy)
for (region in regions_to_modify) {
  regional_results_test[[region]] <- regional_results_test[[region]][, -1]
}
```

Calculate running window trends
```{r}
#running window trend analysis with 31 or 101 year windows - 1420 epoch

# Define parameters
years <- 1420:(1420 + 590 - 1)  # Year range based on data length
window_size <- 31  # Moving window size
# window_size <- 101  # Moving window size

# Initialize a list to store results
ensemble_trends_dur <- list()

for (region in names(regional_results_test)) {
  # Extract data for the current region
  data <- regional_results_test[[region]]  # Each column is an ensemble member
  n_years <- nrow(data)
  n_ensembles <- ncol(data)
  
  # Initialize a matrix to store trends (rows = years, cols = ensemble members)
  trends_matrix <- matrix(NA, nrow = n_years - window_size + 1, ncol = n_ensembles)
  midpoint_years <- years[1:(n_years - window_size + 1)] + (window_size - 1) / 2
  rownames(trends_matrix) <- midpoint_years
  colnames(trends_matrix) <- paste0("Member_", 1:n_ensembles)
  
  # Loop through each ensemble member
  for (i in 1:n_ensembles) {
    ensemble_data <- data[, i]  # Extract the ensemble member data
    
    # Calculate moving window trends
    for (start_year in 1:(n_years - window_size + 1)) {
      end_year <- start_year + window_size - 1
      window_years <- years[start_year:end_year]
      window_data <- ensemble_data[start_year:end_year]
      
      # Fit a linear model to the moving window
      if (all(!is.na(window_data))) {  # Check for missing values
        model <- lm(window_data ~ window_years)
        trends_matrix[start_year, i] <- coef(model)[2]  # Extract the slope
      }
    }
  }
  
  # Save the trend matrix for the region
  ensemble_trends_dur[[region]] <- trends_matrix
}

saveRDS(ensemble_trends_dur, file="~/cwd_global/data/droughtLEN_30yr_trends.RData")  ## 30 year trends

```

Example: Visualize trends for one ensemble member in one region
```{r}
region <- "MED"  # Specify region
ensemble <- 1  # Specify ensemble member

plot(as.numeric(rownames(ensemble_trends_dur[[region]])),
     ensemble_trends_dur[[region]][, ensemble], type = "l",
     main = paste("30-Year Moving Trends for", region, "- Member", ensemble),
     xlab = "Year", ylab = "Trend (mm/yr)")
```
Calculate ensemble mean trends: 

```{r}
# Initialize lists to store results
ensemble_means_dur <- list()  # Mean for each region

# Loop over regions
for (region_name in names(ensemble_trends_dur)) {
  
  # Get the trends matrix for the region (time steps x ensemble members)
  trends_matrix <- ensemble_trends_dur[[region_name]]
  
  # Calculate mean and standard deviation across ensemble members
  # Mean across rows (time steps)
  ensemble_means_dur[[region_name]] <- rowMeans(trends_matrix, na.rm = TRUE)

}
```

Plot time series of duration trends

```{r}
#create dataframe

# Define continent colors
continent_colors <- c(
  "Europe" = "#F5B0CB",
  "North America" = "#236A3F",
  "Africa" = "#91C7B1",
  "Asia" = "#E5973B",
  "Central America" = "#B33951",
  "South America" = "#E3D081",
  "Australia" = "#3C3980",
  "Ocean" = "white"
)

# Define continent assignments
continent_mapping <- list(
  "Europe" = c("NEU", "WCE", "EEU", "MED"),
  "North America" = c("NWN", "NEN", "WNA", "CNA", "ENA", "GIC"),
  "Africa" = c("SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG"),
  "Asia" = c("WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "WSB", "ESB", "RFE", "RAR"),
  "Central America" =c("NCA", "SCA", "CAR"),
  "South America" = c("NWS", "NSA", "NES", "SAM", "SWS", "SES", "SSA"),
  "Australia" = c("NAU", "CAU", "EAU", "SAU", "NZ")
)

# Convert ensemble data into a long format for plotting individual members
duration_trends_long <- do.call(rbind, lapply(names(ensemble_trends_dur), function(region) {
  df <- as.data.frame(ensemble_trends_dur[[region]])
  df$time <- seq(1435, 1994)  # Adjust time range accordingly
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region),names_to = "ensemble_member", values_to = "trend")
  return(df_long)
}))

# Add a continent column
duration_trends_long$continent <- NA  # Initialize continent column
for (continent in names(continent_mapping)) {
  duration_trends_long$continent[duration_trends_long$region %in% continent_mapping[[continent]]] <- continent
}

```

```{r}
### Plot for each region
# Filter for selected continent
selected_continent <- "Europe"
duration_trends_long_fil <- duration_trends_long %>% filter(continent == selected_continent)


# Compute mean anomalies per (time, region)
mean_duration_trends_long_fil<- duration_trends_long_fil %>%
  group_by(time, region) %>%
  summarise(
    mean_trends = mean(trend, na.rm = TRUE),
  ) %>%
  ungroup()


ggplot() +
  geom_hline(yintercept = 0, color= "darkgrey", linetype="dashed")+
  # Plot individual ensemble members with distinct colors
  #geom_line(data = duration_trends_long_fil, aes(x = time, y = trend, group = ensemble_member, color = factor("Ensemble Members")), alpha = 0.5, linewidth = 0.3, show.legend = TRUE)+
  # Plot mean anomaly as a bold line
  geom_line(data = mean_duration_trends_long_fil, aes(x = time, y = mean_trends, color = factor("ModESim")), linewidth = 1, show.legend = TRUE) +
  labs(
    x = "Year",
    y = "drought length 30-year trends",
    title = paste("30yr rolling duration trends", selected_continent)
  ) +
  # Manual color scale for legend
  scale_color_manual(
    values = c("Ensemble Members" = "#7EA3CC","ModESim" = "#255C99" ),
    labels = c("Ensemble Members","Length (Mean) - ModESim" )
  ) +

  # Customize the legend
  guides(
    color = guide_legend(title = "Legend", override.aes = list(alpha = 1, linewidth = 1))
  ) +
  theme_classic() +
  theme(
    legend.position = "top",
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 10),
    strip.background = element_rect(fill = "#F5B0CB"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +

  facet_wrap(~region, scales = "free_y")
```
