---
title: "PDSI_PCWD_comparison"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries, message=FALSE, echo=FALSE}
library(here)
library(readr)
library(dplyr)
library(slider)
library(lubridate)
library(patchwork)
library(ggplot2)
library(ncdf4)
library(reshape2)
library(ggpubr)
library(ggtext)
library(maps)
library(terra)
library(rnaturalearth)
library(sf)
library(sp)
library(RColorBrewer)
library(tidyverse)
library(abind)
library(patchwork)
library(gridExtra)
library(ggsci)
library(purrr)
library(qmap)
library(cowplot)
library(extRemes)
library(ggpattern)


### load relevant libraries:
library(dplyr)
library(rJava)
library(loadeR.java)
library(transformeR)
library(loadeR)
library(visualizeR)
#library(geoprocessoR)

```

```{r}
## test read in one ensemble member
##read in data - stored on Ubelix server
PDSI_recon_01 <- "/Users/phelpap/Documents/lastmilldroughts_data/scPDSI_AWC_recon/ModE-Sim_m001_PDSI_1420-2009.nc" ### monthly data - we have annual max PCWD so one value per year - keep monthly here to evaluate the timing aspect of PCWD

nc_test <- nc_open(PDSI_recon_01)
var = ncvar_get(nc_test, varid="pdsi")
lon = ncvar_get(nc_test, varid="longitude")
lat = ncvar_get(nc_test, varid="latitude")
time = ncvar_get(nc_test, varid="time")
# Convert to actual dates (days since 1970-1-1)
reference_date <- as.Date("1970-01-01")
time_dates <- reference_date + time

# # Print the resulting dates
print(nc_test)

nc_close(nc_test)
```



```{r}
### Create regional areas for PDSI 

########### Script for data post-processing to create regional averages (see IPCC regions, AR6) ###########
### read in IPCC Region information:
#Load reference regions and coastlines:
load(here::here("data", "Reference_regions", "IPCC-WGI-reference-regions-v4_R.rda"))

#simplify this object by converting it to a SpatialPolygons class object (i.e., only the polygons are retained and their attributes discarded):
refregions <- as(IPCC_WGI_reference_regions_v4, "SpatialPolygons")

# # List of regions to loop over --- excludes ocean basins
# regions <- c("GIC", "NWN", "NEN", "WNA", "CNA", "ENA", "NCA", "SCA", "CAR", "NWS",
#              "NSA", "NES", "SAM", "SWS", "SES", "SSA", "NEU", "WCE", "EEU", "MED",
#              "SAH", "WAF", "CAF", "NEAF", "SEAF", "WSAF", "ESAF", "MDG", "RAR", "WSB",
#              "ESB", "RFE", "WCA", "ECA", "TIB", "EAS", "ARP", "SAS", "SEA", "NAU",
#              "CAU", "EAU", "SAU", "NZ")
regions <- c("GIC") # for testing

#as before but loop over all regions also, saving everything in an array
path <- "/Users/phelpap/Documents/lastmilldroughts_data/scPDSI_AWC_recon"
files <- list.files(path, full.names = TRUE, pattern = "m[0-9]{3}") #m001 - m020 in first set

# Extract unique ensemble member identifiers from folder names
#ensemble_members <- regmatches(basename(files), regexpr("m[0-9]{3}", basename(files)))
ensemble_members <- "m001"

```

```{r}
### sanity check:
library(sp)
library(ggplot2)

# Load one grid and get coordinates
grid <- loadGridData(dataset = files[1], var = "pdsi")
grid <- setGridProj(grid = grid, proj = proj4string(refregions))
coords <- getCoordinates(grid)

# Get region name for each grid point (returns NA if outside all regions)
region_hit <- over(points_sp, refregions)  # returns the polygon attribute/name
grid_pts$region <- as.character(region_hit)  # NA for ocean/outside

# Extract timestep 1 PDSI
timestep1 <- grid$Data[1, , ]
grid_pts$pdsi <- as.vector(t(timestep1))

# Plot 1: which region each point belongs to
ggplot() +
  geom_point(data = grid_pts, aes(x = lon, y = lat, color = region), size = 0.8) +
  geom_polygon(data = region_df, aes(x = long, y = lat, group = group),
               fill = NA, color = "black", linewidth = 0.4) +
  coord_quickmap() +
  theme_classic() +
  theme(legend.position = "none") +  # too many regions for a readable legend
  labs(title = "Grid points coloured by IPCC AR6 region")

# Plot 2: PDSI field with all region boundaries
ggplot() +
  geom_point(data = grid_pts, aes(x = lon, y = lat, color = pdsi), size = 1.2) +
  geom_polygon(data = region_df, aes(x = long, y = lat, group = group),
               fill = NA, color = "black", linewidth = 0.4) +
  scale_color_distiller(palette = "RdBu", na.value = "grey90",
                        name = "PDSI", limits = c(-4, 4)) +
  coord_quickmap() +
  theme_classic() +
  labs(title = "PDSI timestep 1 with all IPCC AR6 region boundaries",
       x = "Longitude", y = "Latitude")
```


```{r}
# ##################################### Code verification

# Load relevant libraries
library(dplyr)
library(rJava)
library(loadeR.java)
library(transformeR)
library(loadeR)
library(visualizeR)
library(raster)
library(terra)
library(ncdf4)
library(rnaturalearth)
library(sf)
library(ggplot2)
library(viridis) # For better color scale
#
# total_cells <- length(land_sea_mask)
# ocean_cells <- sum(!land_sea_mask)
# land_cells <- sum(land_sea_mask)
#
# cat("Total Grid Cells:", total_cells, "\n")
# cat("Land Cells:", land_cells, "(", round(land_cells / total_cells * 100, 2), "% )\n")
# cat("Ocean Cells:", ocean_cells, "(", round(ocean_cells / total_cells * 100, 2), "% )\n")
#

# Select one ensemble member for testing
test_em <- ensemble_members[1]

test_file = files[1]

## apply land sea mask on one time step for testing
# Load data using loadGridData (keeping consistency with your approach)
grid_test <- loadGridData(dataset = test_file, var = "pdsi")
grid_array_test <- grid_test$Data
test_array <- grid_array_test[1,,]
test_array[!land_sea_mask] <- NA


# Create a copy of the original data to store masked values
masked_grid_array <- grid_array_test

# Loop over the time dimension
for (t in 1:dim(grid_array_test)[1]) {
  masked_grid_array[t, , ][!land_sea_mask] <- NA
}

# The masked data retains the same 3D structure
masked_grid_array


test_array_2 <- apply_land_mask(grid_test, land_sea_mask)
test_array_plot <- test_array_2$Data
plot_data <- test_array_plot[1,,]


### test with raw pdsi - maybe already land masked? 

r <- raster(t(var[,,4]), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84" # Set CRS for raster
#r <- flip(r, direction='y') #Flip the raster to correct orientation
#Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs=st_crs(r)) #Ensure CRS alignment
#Mask the raster with land polygons to remove ocean values
#r_masked <- mask(r, as(land, "Spatial"))
#Convert the masked raster to a dataframe for ggplot
r_df <- as.data.frame(r, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")

# Create ggplot
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = layer), na.rm = TRUE) +
  scale_fill_viridis_c(option = "C", na.value = "white") +  # Use Viridis color scale
  labs(title = "PDSI", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +  # Overlay coastlines
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12), legend.text = element_text(size = 10)) +
  guides(fill = guide_colorbar(title = "PDSI", reverse = FALSE))
```


Analyse final time series for the four regions considered: 
```{r}
## read in produced regional file 
regional_test_PDSI_AWCrecon <- readRDS(here::here("data/regionalResults_scPDSI_1420_2009_AWCrecon.RData"))
regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")
regions_present <- intersect(names(regional_test_PDSI_AWCrecon), regions_to_modify)

regional_test_PDSI_AWCrecon[regions_present] <- lapply(
  regional_test_PDSI_AWCrecon[regions_present],
  function(x) x[, colnames(x) != "m001", drop = FALSE]
)

regional_test_PDSI_AWC100 <- readRDS(here::here("data/regionalResults_scPDSI_1420_2009_AWC100_test.RData"))
regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")
regions_present <- intersect(names(regional_test_PDSI_AWC100), regions_to_modify)

regional_test_PDSI_AWC100[regions_present] <- lapply(
  regional_test_PDSI_AWC100[regions_present],
  function(x) x[, colnames(x) != "m001", drop = FALSE]
)
```


```{r}
### read in the not-bias corrected PCWD data: 

### read in ModE-Sim regional results
regional_results_1420 <- readRDS(here::here("data/regionalResults_1420_1.RData"))
regional_results_1850 <- readRDS(here::here("data/regionalResults_1850_1.RData"))
merged_results <- Map(rbind, regional_results_1420, regional_results_1850) # Stitch matrices together row-wise for each region

regions <- c("GIC", "ARP", "MED", "WNA", "CAU")
regional_PCWD <- merged_results[regions]

regions_to_modify <- c("CAU", "EAU", "NAU", "NZ", "SAU", "RFE", "EAS", "SEA")
regions_present <- intersect(names(regional_PCWD), regions_to_modify)

regional_PCWD[regions_present] <- lapply(
  regional_PCWD[regions_present],
  function(x) x[, colnames(x) != "m001_tidy", drop = FALSE]
)



### calculate anomalies for highlighting wet and dry periods
# Initialize list to store rolling 30-year means
rolling_means <- list()

# Define window size
window_size <- 25
half_window <- (window_size - 1) / 2

start_year <- 1420
end_year   <- 2009

centers <- seq(start_year, end_year - window_size + 1) + half_window

# Compute rolling means for each ensemble member
for (region_name in names(regional_PCWD)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_PCWD[[region_name]][2:590,]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means[[region_name]] <- rolling_mean_matrix
}

rolling_means_PCWD_long <-  do.call(rbind, lapply(names(rolling_means), function(region) {
  df <- as.data.frame(rolling_means[[region]])
  df$time <- centers[1:nrow(df)]
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "rolling_means")
  return(df_long)
}))

```

600 year anomalies with full period as baseline 

```{r}
# Initialize lists to store baselines
anom_baseline <- list()

# Compute baseline mean over all values of rolling mean
for (region_name in names(rolling_means)) {
  rolling_mean_matrix <- rolling_means[[region_name]]
  
  anom_baseline[[region_name]] <- mean(rolling_mean_matrix, na.rm = TRUE)  ### for reference period of first 30 values: rolling_mean_matrix[1:30,]
}

# Initialize lists to store anomalies
ensemble_30anom <- list()

# Compute anomalies
for (region_name in names(rolling_means)) {
  rolling_mean_matrix <- rolling_means[[region_name]]
  num_years <- nrow(rolling_mean_matrix)
  
  # Initialize matrix to store anomalies
  anoms <- matrix(NA, nrow=num_years, ncol=ncol(rolling_mean_matrix))
  
  # Compute anomalies using rolling mean baseline
  for (t in 1:num_years) {
    anoms[t, ] <- (rolling_mean_matrix[t, ] - anom_baseline[[region_name]])
  }
  
  # Store results
  ensemble_30anom[[region_name]] <- anoms
}

# Compute mean anomalies across ensemble members
mean_30anom <- list()
for (region_name in names(ensemble_30anom)) {
  mean_30anom[[region_name]] <- rowMeans(ensemble_30anom[[region_name]], na.rm = TRUE)
}

# Compute region-specific pooled standard deviation all years
pooled_sd_by_region <- list()
for (region_name in names(ensemble_30anom)) {
  # Extract all years of anomalies
  all_years <- ensemble_30anom[[region_name]]
  
  # Compute pooled standard deviation over these values
  pooled_sd_by_region[[region_name]] <- sd(unlist(all_years), na.rm = TRUE)
}

# Convert ensemble data into a long format for plotting individual members
anoms_long <- do.call(rbind, lapply(names(ensemble_30anom), function(region) {
  df <- as.data.frame(ensemble_30anom[[region]])
  df$time <- centers[1:nrow(df)]
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "anomalies")
  return(df_long)
}))

```

```{r}

# ------------------------------------------------------------------------------
# Helper: prepare data
# ------------------------------------------------------------------------------
prepare_region_data <- function(region,
                                regional_PCWD,
                                mode_years = 1420:2009,
                                drop_initial_years = 2) {
  MODE_mat <- regional_PCWD[[region]]
  
  # Time‐series data
  MODE_long <- MODE_mat %>%
    as.data.frame() %>%
    mutate(time = mode_years) %>%
    pivot_longer(-time, names_to = "ensemble", values_to = "PCWD")
  
  MODE_mean <- MODE_long %>% group_by(time) %>% summarise(PCWD = mean(PCWD, na.rm=TRUE), .groups="drop")
  
  # ———————— drop the first `drop_initial_years` rows ————————
  MODE_long <- MODE_long %>% 
    filter(time >= mode_years[drop_initial_years])
  
  MODE_mean <- MODE_mean %>% 
    filter(time >= mode_years[drop_initial_years])
  # ——————————————————————————————————————————————————————————————
    list(
    MODE_long   = MODE_long,
    MODE_mean   = MODE_mean
  )
}


#Timeseries function alone: 
plot_region_PCWD_ts <- function(region_code) {
  
  dat <- prepare_region_data(region_code,
                             regional_PCWD)
  
  # Compute anomalies + bounds + exceedance runs
  anoms_filtered <- anoms_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      mean_anomaly = mean(anomalies, na.rm=TRUE),
      pooled_sd    = pooled_sd_by_region[[region_code]],
      .groups = "drop"
    )
  
  first_mean <- anoms_filtered %>%
    arrange(time) %>%
    slice(1) %>%
    pull(mean_anomaly)
  
  anoms_filtered <- anoms_filtered %>%
    mutate(
      lower_bound = first_mean - pooled_sd,
      upper_bound = first_mean + pooled_sd,
      exceeds     = mean_anomaly < lower_bound | mean_anomaly > upper_bound,
      direction   = case_when(
        mean_anomaly > upper_bound ~ "drier",
        mean_anomaly < lower_bound ~ "wetter",
        TRUE ~ NA_character_
      ),
      run_id      = data.table::rleid(exceeds)
    )
  
  exceed_periods <- anoms_filtered %>%
    filter(exceeds) %>%
    group_by(run_id, direction) %>%
    summarise(
      start = min(time),
      end   = max(time),
      .groups="drop"
    )
  
  # Filter rolling means
  rolling_filtered <- rolling_means_PCWD_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      mean_rolling = mean(rolling_means, na.rm=TRUE),
      .groups = "drop"
    )
  
  # y-limits
  y_limits     <- range(dat$MODE_long$PCWD, na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  # Compute max of each ensemble for dotted lines
  ensemble_max <- dat$MODE_long %>%
    group_by(ensemble) %>%
    summarise(max_PCWD = max(PCWD, na.rm=TRUE), .groups = "drop") %>%
    summarise(max_max = max(max_PCWD, na.rm = TRUE), .groups="drop") %>%
    mutate(label = "Ens_maximum")
  
  # Build ggplot
  p_ts <- ggplot()
  
  # Exceedance rectangles colored by direction
  if (nrow(exceed_periods) > 0) {
    p_ts <- p_ts + 
      geom_rect(
        data = exceed_periods,
        aes(xmin = start, xmax = end, 
            ymin = y_limits[1], ymax = y_limits[2],
            fill = direction),
        inherit.aes = FALSE,
        alpha = 0.4
      ) 
  }
  p_ts <- p_ts +
    geom_line(data = dat$MODE_long, aes(time, PCWD, group=ensemble, color = factor("EMs_ModE")), linewidth=0.3, alpha=0.4) +
    geom_line(data = dat$MODE_mean, aes(time, PCWD, color = factor("Mean_ModE")), linewidth = 0.8) +
    geom_line(data = rolling_filtered, aes(time, mean_rolling, color="rolling_mean"), linewidth=0.7) +
    # geom_hline(data = ensemble_max, aes(yintercept = max_max, color= factor("Ens_maximum")),linetype="dotted", linewidth=0.5)+
    scale_y_continuous(limits=y_limits, breaks=common_breaks) +
    scale_x_continuous(breaks=c(1400, 1600, 1800, 2000)) +
    scale_color_manual(name = "Data",
      values = c(
        "EMs_ModE"  = "#7EA3CC",
        "Mean_ModE" = "#255C99",
        "rolling_mean"="black" 
       # "Ens_maximum"= "grey30"
      ),
      labels = c(
        "Individual Members", "Ensemble Mean", "25-yr rolling\nEnsemble Mean"
        #, "max PCWD"
        )
      ) +
      scale_fill_manual(name ="Exceedances", values = c("wetter" = "#136E40", "drier" = "#F4DE70" ),
      labels = c("wetter", "drier")
    ) +
    labs(x = "", y = "PCWD [mm]", color = "Data") +
    theme_classic(base_size = 14) +
    theme(axis.title.y = element_text(margin=margin(r=10)),
          legend.position="right", legend.title = element_text(face="bold"))
  
  return(p_ts)
}

ts_MED <- plot_region_PCWD_ts("MED")
ts_MED

ts_WNA <- plot_region_PCWD_ts("WNA")
ts_WNA

ts_ARP <- plot_region_PCWD_ts("ARP")
ts_ARP

ts_CAU <- plot_region_PCWD_ts("CAU")
ts_CAU

ggsave(
  filename = here::here("figures/PCWD_PDSI_comp_legend.pdf"),
  device   = cairo_pdf,    # better text rendering
  width    = 12,           # in inches
  height   = 6,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

```

Plot comparison for PDSI: 
- 25-year (monthly data!) rolling means as in PCWD - define period exceedances based on those (but can take absolute value as threshold as already defined as wet (exceeds positive 1) and dry (exceeds negative 1))

```{r}
rolling_means_PDSI_AWC100 <- list()
# Define window size
window_size <- 25*12 # 30-year rolling windows on monthly data
half_window <- floor((window_size - 1) / 2)  # = 149

centers <- time_dates[(1:(length(time_dates) - window_size + 1)) + half_window]

# Compute rolling means for each ensemble member
for (region_name in names(regional_test_PDSI_AWC100)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_test_PDSI_AWC100[[region_name]][13:7080,]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means_PDSI_AWC100[[region_name]] <- rolling_mean_matrix
}

rolling_means_PDSI_AWC100_long <-  do.call(rbind, lapply(names(rolling_means_PDSI_AWC100), function(region) {
  df <- as.data.frame(rolling_means_PDSI_AWC100[[region]])
  df$time <- centers[1:nrow(df)]  # Adjust time range accordingly: -180 on both ends as window is 300 and also -12 as start in the second year as in PCWD
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "rolling_means")
  return(df_long)
}))


rolling_means_PDSI_AWCrecon <- list()

# Compute rolling means for each ensemble member
for (region_name in names(regional_test_PDSI_AWCrecon)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_test_PDSI_AWCrecon[[region_name]][13:7080,]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means_PDSI_AWCrecon[[region_name]] <- rolling_mean_matrix
}

rolling_means_PDSI_AWCrecon_long <-  do.call(rbind, lapply(names(rolling_means_PDSI_AWCrecon), function(region) {
  df <- as.data.frame(rolling_means_PDSI_AWCrecon[[region]])
  df$time <- centers[1:nrow(df)]  # Adjust time range accordingly: -180 on both ends as window is 360 and also -12 as start in the second year as in PCWD
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "rolling_means")
  return(df_long)
}))
```

```{r}
 ------------------------------------------------------------------------------
# Helper: prepare data
# ------------------------------------------------------------------------------
prepare_region_data_PDSI <- function(region,
                                regional_test_PDSI_AWC100,
                                drop_initial_years = 2) {
  MODE_mat <- regional_test_PDSI_AWC100[[region]]

  MODE_long <- MODE_mat %>%
    as.data.frame() %>%
    mutate(time = as.Date(time_dates)) %>%
    pivot_longer(!all_of("time"), names_to = "ensemble", values_to = "PDSI")

  MODE_mean <- MODE_long %>%
    group_by(time) %>%
    summarise(PDSI = mean(PDSI, na.rm = TRUE), .groups = "drop")

  cutoff <- as.Date(time_dates[drop_initial_years * 12 + 1])

  MODE_long <- MODE_long %>% filter(time >= cutoff)
  MODE_mean <- MODE_mean %>% filter(time >= cutoff)

  list(
    MODE_long = MODE_long,
    MODE_mean = MODE_mean
  )
}


#Timeseries function alone: 
plot_region_PDSI_AWC100_ts <- function(region_code) {
  
  dat <- prepare_region_data_PDSI(region_code,
                             regional_test_PDSI_AWC100)
  
  # Compute anomalies + bounds + exceedance runs
  means_filtered <- rolling_means_PDSI_AWC100_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      EM = mean(rolling_means, na.rm=TRUE),
      .groups = "drop"
    )
  
  first_mean <- means_filtered %>%
    arrange(time) %>%
    slice(1) %>%
    pull(EM)
  
  means_filtered <- means_filtered %>%
    mutate(
      lower_bound = -1,
      upper_bound = 1,
      exceeds     = EM < lower_bound | EM > upper_bound,
      direction   = case_when(
        EM > upper_bound ~ "wetter",
        EM < lower_bound ~ "drier",
        TRUE ~ NA_character_
      ),
      run_id      = data.table::rleid(exceeds)
    )
  
  exceed_periods <- means_filtered %>%
    filter(exceeds) %>%
    group_by(run_id, direction) %>%
    summarise(
      start = min(time),
      end   = max(time),
      .groups="drop"
    )
  
  # y-limits
  y_limits     <- range(dat$MODE_long$PDSI, na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  # # Compute max of each ensemble for dotted lines
  # ensemble_max <- dat$MODE_long %>%
  #   group_by(ensemble) %>%
  #   summarise(max_PCWD = max(PCWD, na.rm=TRUE), .groups = "drop") %>%
  #   summarise(max_max = max(max_PCWD, na.rm = TRUE), .groups="drop") %>%
  #   mutate(label = "Ens_maximum")
  
  # Build ggplot
  p_ts <- ggplot()
  
  # Exceedance rectangles colored by direction
  if (nrow(exceed_periods) > 0) {
    p_ts <- p_ts + 
      geom_rect(
        data = exceed_periods,
        aes(xmin = start, xmax = end, 
            ymin = y_limits[1], ymax = y_limits[2],
            fill = direction),
        inherit.aes = FALSE,
        alpha = 0.4
      ) 
  }
  p_ts <- p_ts +
    geom_line(data = dat$MODE_long, aes(time, PDSI, group=ensemble, color = factor("EMs_ModE")), linewidth=0.3, alpha=0.4) +
    geom_line(data = dat$MODE_mean, aes(time, PDSI, color = factor("Mean_ModE")), linewidth = 0.8) +
    geom_line(data = means_filtered, aes(time, EM, color="rolling_mean"), linewidth=0.7) +
    # geom_hline(data = ensemble_max, aes(yintercept = max_max, color= factor("Ens_maximum")),linetype="dotted", linewidth=0.5)+
    scale_y_continuous(limits=rev(y_limits), breaks=common_breaks, trans = "reverse") +
    scale_x_date(
    breaks = as.Date(c("1400-01-01", "1600-01-01", "1800-01-01", "2000-01-01")),
    date_labels = "%Y"
  )+
    scale_color_manual(name = "Data",
      values = c(
        "EMs_ModE"  = "#7EA3CC",
        "Mean_ModE" = "#255C99",
        "rolling_mean"="black" 
        # "Ens_maximum"= "grey30"
      ),
      labels = c(
        "EMs - PDSI AWC_100", "Mean - PDSI AWC_100", "25-yr rolling\nmean - PDSI AWC_100"
        #, "ModE-Sim* - max PCWD"
        )
      ) +
      scale_fill_manual(name ="Exceedances", values = c("wetter" = "#136E40", "drier" = "#F4DE70" ),
      labels = c("wetter", "drier")
    ) +
    labs(x = "Year", y = "PDSI", color = "Data") +
    theme_classic(base_size = 14) +
    theme(axis.title.y = element_text(margin=margin(r=10)),
          legend.position="none", legend.title = element_text(face="bold"))
  
  return(p_ts)
}

ts_MED_PDSI_AWC100 <- plot_region_PDSI_AWC100_ts("MED")
ts_MED_PDSI_AWC100

ts_WNA_PDSI_AWC100 <- plot_region_PDSI_AWC100_ts("WNA")
ts_WNA_PDSI_AWC100

ts_ARP_PDSI_AWC100 <- plot_region_PDSI_AWC100_ts("ARP")
ts_ARP_PDSI_AWC100

ts_CAU_PDSI_AWC100 <- plot_region_PDSI_AWC100_ts("CAU")
ts_CAU_PDSI_AWC100

```

```{r}

#Timeseries function alone: 
plot_region_PDSI_AWCrecon_ts <- function(region_code) {
  
  dat <- prepare_region_data_PDSI(region_code,
                             regional_test_PDSI_AWCrecon)
  
  # Compute anomalies + bounds + exceedance runs
  means_filtered <- rolling_means_PDSI_AWCrecon_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      EM = mean(rolling_means, na.rm=TRUE),
      .groups = "drop"
    )
  
  first_mean <- means_filtered %>%
    arrange(time) %>%
    slice(1) %>%
    pull(EM)
  
  means_filtered <- means_filtered %>%
    mutate(
      lower_bound = -1,
      upper_bound = 1,
      exceeds     = EM < lower_bound | EM > upper_bound,
      direction   = case_when(
        EM > upper_bound ~ "wetter",
        EM < lower_bound ~ "drier",
        TRUE ~ NA_character_
      ),
      run_id      = data.table::rleid(exceeds)
    )
  
  exceed_periods <- means_filtered %>%
    filter(exceeds) %>%
    group_by(run_id, direction) %>%
    summarise(
      start = min(time),
      end   = max(time),
      .groups="drop"
    )
  
  # y-limits
  y_limits     <- range(dat$MODE_long$PDSI, na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  # # Compute max of each ensemble for dotted lines
  # ensemble_max <- dat$MODE_long %>%
  #   group_by(ensemble) %>%
  #   summarise(max_PCWD = max(PCWD, na.rm=TRUE), .groups = "drop") %>%
  #   summarise(max_max = max(max_PCWD, na.rm = TRUE), .groups="drop") %>%
  #   mutate(label = "Ens_maximum")
  
  # Build ggplot
  p_ts <- ggplot()
  
  # Exceedance rectangles colored by direction
  if (nrow(exceed_periods) > 0) {
    p_ts <- p_ts + 
      geom_rect(
        data = exceed_periods,
        aes(xmin = start, xmax = end, 
            ymin = y_limits[1], ymax = y_limits[2],
            fill = direction),
        inherit.aes = FALSE,
        alpha = 0.4
      ) 
  }
  p_ts <- p_ts +
    geom_line(data = dat$MODE_long, aes(time, PDSI, group=ensemble, color = factor("EMs_ModE")), linewidth=0.3, alpha=0.4) +
    geom_line(data = dat$MODE_mean, aes(time, PDSI, color = factor("Mean_ModE")), linewidth = 0.8) +
    geom_line(data = means_filtered, aes(time, EM, color="rolling_mean"), linewidth=0.7) +
    # geom_hline(data = ensemble_max, aes(yintercept = max_max, color= factor("Ens_maximum")),linetype="dotted", linewidth=0.5)+
    scale_y_continuous(limits=rev(y_limits), breaks=common_breaks, trans = "reverse") +    
    scale_x_date(
    breaks = as.Date(c("1400-01-01", "1600-01-01", "1800-01-01", "2000-01-01")),
    date_labels = "%Y"
  )+
    scale_color_manual(name = "Data",
      values = c(
        "EMs_ModE"  = "#7EA3CC",
        "Mean_ModE" = "#255C99",
        "rolling_mean"="black" 
        # "Ens_maximum"= "grey30"
      ),
      labels = c(
        "EMs - PDSI AWC_recon", "Mean - PDSI AWC_recon", "30-yr rolling\nmean - PDSI AWC_recon"
        #, "ModE-Sim* - max PCWD"
        )
      ) +
      scale_fill_manual(name ="Exceedances", values = c("wetter" = "#136E40", "drier" = "#F4DE70" ),
      labels = c("wetter", "drier")
    ) +
    labs(x = "", y = "PDSI", color = "Data") +
    theme_classic(base_size = 14) +
    theme(axis.title.y = element_text(margin=margin(r=10)),
          legend.position="none", legend.title = element_text(face="bold"))
  
  return(p_ts)
}

ts_MED_PDSI_AWCrecon <- plot_region_PDSI_AWCrecon_ts("MED")
ts_MED_PDSI_AWCrecon

ts_WNA_PDSI_AWCrecon <- plot_region_PDSI_AWCrecon_ts("WNA")
ts_WNA_PDSI_AWCrecon

ts_ARP_PDSI_AWCrecon <- plot_region_PDSI_AWCrecon_ts("ARP")
ts_ARP_PDSI_AWCrecon

ts_CAU_PDSI_AWCrecon <- plot_region_PDSI_AWCrecon_ts("CAU")
ts_CAU_PDSI_AWCrecon


```

```{r}
### save plots
ts_MED <- ts_MED + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_MED_PDSI_AWC100 <- ts_MED_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_MED_PDSI_AWCrecon <- ts_MED_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

combined_MED <- plot_grid(
  ts_MED,ts_MED_PDSI_AWC100, ts_MED_PDSI_AWCrecon, 
  nrow        = 3,
  ncol        = 1,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a) MED – Annmax PCWD", 
    "b) MED – PDSI AWC 100mm",
    "c) MED – PDSI AWC recon"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

print(combined_MED)

ggsave(
  filename = here::here("figures/PCWD_PDSI_comp_MED.pdf"),
  plot     = combined_MED,
  device   = cairo_pdf,    # better text rendering
  width    = 8,           # in inches
  height   = 10,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

#####################
### save plots
ts_WNA <- ts_WNA + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_WNA_PDSI_AWC100 <- ts_WNA_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_WNA_PDSI_AWCrecon <- ts_WNA_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

combined_WNA <- plot_grid(
  ts_WNA,ts_WNA_PDSI_AWC100, ts_WNA_PDSI_AWCrecon, 
  nrow        = 3,
  ncol        = 1,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a) WNA – Annmax PCWD", 
    "b) WNA – PDSI AWC 100mm",
    "c) WNA – PDSI AWC recon"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

print(combined_WNA)

ggsave(
  filename = here::here("figures/PCWD_PDSI_comp_WNA.pdf"),
  plot     = combined_WNA,
  device   = cairo_pdf,    # better text rendering
  width    = 8,           # in inches
  height   = 10,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

#########################
### save plots
ts_ARP <- ts_ARP + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_ARP_PDSI_AWC100 <- ts_ARP_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_ARP_PDSI_AWCrecon <- ts_ARP_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

combined_ARP <- plot_grid(
  ts_ARP,ts_ARP_PDSI_AWC100, ts_ARP_PDSI_AWCrecon, 
  nrow        = 3,
  ncol        = 1,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a) ARP – Annmax PCWD", 
    "b) ARP – PDSI AWC 100mm",
    "c) ARP – PDSI AWC recon"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

print(combined_ARP)

ggsave(
  filename = here::here("figures/PCWD_PDSI_comp_ARP.pdf"),
  plot     = combined_ARP,
  device   = cairo_pdf,    # better text rendering
  width    = 8,           # in inches
  height   = 10,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

############################
### save plots
ts_CAU <- ts_CAU + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_CAU_PDSI_AWC100 <- ts_CAU_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_CAU_PDSI_AWCrecon <- ts_CAU_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

combined_CAU <- plot_grid(
  ts_CAU,ts_CAU_PDSI_AWC100, ts_CAU_PDSI_AWCrecon, 
  nrow        = 3,
  ncol        = 1,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a) CAU – Annmax PCWD", 
    "b) CAU – PDSI AWC 100mm",
    "c) CAU – PDSI AWC recon"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

print(combined_CAU)

ggsave(
  filename = here::here("figures/PCWD_PDSI_comp_CAU.pdf"),
  plot     = combined_CAU,
  device   = cairo_pdf,    # better text rendering
  width    = 8,           # in inches
  height   = 10,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```



Repeat the same for PDSI using annual maximum PDSI values

```{r}
regional_test_PDSI_AWC100_annual_mean <- lapply(regional_test_PDSI_AWC100, function(mat) {
  
  # Add time and pivot long
  df <- mat %>%
    as.data.frame() %>%
    mutate(time = time_dates) %>%
    pivot_longer(!all_of("time"), names_to = "ensemble", values_to = "PDSI") %>%
    mutate(year = as.integer(format(time, "%Y"))) %>%
    
    # Annual mean per ensemble member
    group_by(year, ensemble) %>%
    summarise(PDSI = mean(PDSI, na.rm = TRUE), .groups = "drop") %>%
    
    # Pivot back to wide (rows = years, cols = ensemble members)
    pivot_wider(names_from = ensemble, values_from = PDSI) %>%
    arrange(year)
  
  # Return as matrix with same structure as input
  mat_out <- df %>% dplyr::select(-year) %>% as.matrix()
 # rownames(mat_out) <- df$year
  mat_out
})

regional_test_PDSI_AWC100_annual_min <- lapply(regional_test_PDSI_AWC100, function(mat) {
  
  # Add time and pivot long
  df <- mat %>%
    as.data.frame() %>%
    mutate(time = time_dates) %>%
    pivot_longer(!all_of("time"), names_to = "ensemble", values_to = "PDSI") %>%
    mutate(year = as.integer(format(time, "%Y"))) %>%
    
    # Annual mean per ensemble member
    group_by(year, ensemble) %>%
    summarise(PDSI = min(PDSI, na.rm = TRUE), .groups = "drop") %>%
    
    # Pivot back to wide (rows = years, cols = ensemble members)
    pivot_wider(names_from = ensemble, values_from = PDSI) %>%
    arrange(year)
  
  # Return as matrix with same structure as input
  mat_out <- df %>% dplyr::select(-year) %>% as.matrix()
 # rownames(mat_out) <- df$year
  mat_out
})


rolling_means_AWC100_annual_mean <- list()

# Define window size
window_size <- 30

# Compute rolling means for each ensemble member
for (region_name in names(regional_test_PDSI_AWC100_annual_mean)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_test_PDSI_AWC100_annual_mean[[region_name]][2:590,]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means_AWC100_annual_mean[[region_name]] <- rolling_mean_matrix
}

rolling_means_AWC100_annual_mean_long <-  do.call(rbind, lapply(names(rolling_means_AWC100_annual_mean), function(region) {
  df <- as.data.frame(rolling_means_AWC100_annual_mean[[region]])
  df$time <- seq(1436, 1995)  # Adjust time range accordingly
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "rolling_means")
  return(df_long)
}))


rolling_means_AWC100_annual_min <- list()

# Define window size
window_size <- 30

# Compute rolling means for each ensemble member
for (region_name in names(regional_test_PDSI_AWC100_annual_min)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_test_PDSI_AWC100_annual_min[[region_name]][2:590,]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means_AWC100_annual_min[[region_name]] <- rolling_mean_matrix
}

rolling_means_AWC100_annual_min_long <-  do.call(rbind, lapply(names(rolling_means_AWC100_annual_min), function(region) {
  df <- as.data.frame(rolling_means_AWC100_annual_min[[region]])
  df$time <- seq(1436, 1995)  # Adjust time range accordingly
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "rolling_means")
  return(df_long)
}))

```

Repeat for PDSI AWC recon: 

```{r}
regional_test_PDSI_AWCrecon_annual_mean <- lapply(regional_test_PDSI_AWCrecon, function(mat) {
  
  # Add time and pivot long
  df <- mat %>%
    as.data.frame() %>%
    mutate(time = time_dates) %>%
    pivot_longer(!all_of("time"), names_to = "ensemble", values_to = "PDSI") %>%
    mutate(year = as.integer(format(time, "%Y"))) %>%
    
    # Annual mean per ensemble member
    group_by(year, ensemble) %>%
    summarise(PDSI = mean(PDSI, na.rm = TRUE), .groups = "drop") %>%
    
    # Pivot back to wide (rows = years, cols = ensemble members)
    pivot_wider(names_from = ensemble, values_from = PDSI) %>%
    arrange(year)
  
  # Return as matrix with same structure as input
  mat_out <- df %>% dplyr::select(-year) %>% as.matrix()
 # rownames(mat_out) <- df$year
  mat_out
})

regional_test_PDSI_AWCrecon_annual_min <- lapply(regional_test_PDSI_AWCrecon, function(mat) {
  
  # Add time and pivot long
  df <- mat %>%
    as.data.frame() %>%
    mutate(time = time_dates) %>%
    pivot_longer(!all_of("time"), names_to = "ensemble", values_to = "PDSI") %>%
    mutate(year = as.integer(format(time, "%Y"))) %>%
    
    # Annual mean per ensemble member
    group_by(year, ensemble) %>%
    summarise(PDSI = min(PDSI, na.rm = TRUE), .groups = "drop") %>%
    
    # Pivot back to wide (rows = years, cols = ensemble members)
    pivot_wider(names_from = ensemble, values_from = PDSI) %>%
    arrange(year)
  
  # Return as matrix with same structure as input
  mat_out <- df %>% dplyr::select(-year) %>% as.matrix()
 # rownames(mat_out) <- df$year
  mat_out
})


rolling_means_AWCrecon_annual_mean <- list()

# Define window size
window_size <- 30

# Compute rolling means for each ensemble member
for (region_name in names(regional_test_PDSI_AWCrecon_annual_mean)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_test_PDSI_AWCrecon_annual_mean[[region_name]][2:590,]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means_AWCrecon_annual_mean[[region_name]] <- rolling_mean_matrix
}

rolling_means_AWCrecon_annual_mean_long <-  do.call(rbind, lapply(names(rolling_means_AWCrecon_annual_mean), function(region) {
  df <- as.data.frame(rolling_means_AWCrecon_annual_mean[[region]])
  df$time <- seq(1436, 1995)  # Adjust time range accordingly
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "rolling_means")
  return(df_long)
}))


rolling_means_AWCrecon_annual_min <- list()

# Define window size
window_size <- 30

# Compute rolling means for each ensemble member
for (region_name in names(regional_test_PDSI_AWCrecon_annual_min)) {
  
  # Extract ensemble members for the region
  ensemble_members <- regional_test_PDSI_AWCrecon_annual_min[[region_name]][2:590,]
  num_years <- nrow(ensemble_members)
  
  # Initialize matrix to store rolling means
  rolling_mean_matrix <- matrix(NA, nrow=num_years - window_size + 1, ncol=ncol(ensemble_members))
  
  # Compute rolling means
  for (t in 1:(num_years - window_size + 1)) {
    rolling_mean_matrix[t, ] <- colMeans(ensemble_members[t:(t+window_size-1), ], na.rm = TRUE)
  }
  
  # Store rolling means
  rolling_means_AWCrecon_annual_min[[region_name]] <- rolling_mean_matrix
}

rolling_means_AWCrecon_annual_min_long <-  do.call(rbind, lapply(names(rolling_means_AWCrecon_annual_min), function(region) {
  df <- as.data.frame(rolling_means_AWCrecon_annual_min[[region]])
  df$time <- seq(1436, 1995)  # Adjust time range accordingly
  df$region <- region
  df_long <- pivot_longer(df, cols = -c(time, region), names_to = "ensemble_member", values_to = "rolling_means")
  return(df_long)
}))
```

Plotting the annual mean values: 
```{r}
prepare_region_data_PDSI <- function(region,
                                regional_test_PDSI_AWC100_annual_mean,
                                mode_years = 1420:2009,
                                drop_initial_years = 2) {
  MODE_mat <- regional_test_PDSI_AWC100_annual_mean[[region]]

  MODE_long <- MODE_mat %>%
    as.data.frame() %>%
    mutate(time = mode_years) %>%
    pivot_longer(-time, names_to = "ensemble", values_to = "PDSI")

  MODE_mean <- MODE_long %>%
    group_by(time) %>%
    summarise(PDSI = mean(PDSI, na.rm = TRUE), .groups = "drop")
  # ———————— drop the first `drop_initial_years` rows ————————
  MODE_long <- MODE_long %>% 
    filter(time >= mode_years[drop_initial_years])
  
  MODE_mean <- MODE_mean %>% 
    filter(time >= mode_years[drop_initial_years])

  list(
    MODE_long = MODE_long,
    MODE_mean = MODE_mean
  )
}


#Timeseries function alone: 
plot_region_PDSI_AWC100_ts <- function(region_code) {
  
  dat <- prepare_region_data_PDSI(region_code,
                             regional_test_PDSI_AWC100_annual_mean)
  
  # Compute anomalies + bounds + exceedance runs
  means_filtered <- rolling_means_AWC100_annual_mean_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      EM = mean(rolling_means, na.rm=TRUE),
      .groups = "drop"
    )
  
  first_mean <- means_filtered %>%
    arrange(time) %>%
    slice(1) %>%
    pull(EM)
  
  means_filtered <- means_filtered %>%
    mutate(
      lower_bound = -1,
      upper_bound = 1,
      exceeds     = EM < lower_bound | EM > upper_bound,
      direction   = case_when(
        EM > upper_bound ~ "wetter",
        EM < lower_bound ~ "drier",
        TRUE ~ NA_character_
      ),
      run_id      = data.table::rleid(exceeds)
    )
  
  exceed_periods <- means_filtered %>%
    filter(exceeds) %>%
    group_by(run_id, direction) %>%
    summarise(
      start = min(time),
      end   = max(time),
      .groups="drop"
    )
  
  # y-limits
  y_limits     <- range(dat$MODE_long$PDSI, na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  # # Compute max of each ensemble for dotted lines
  # ensemble_max <- dat$MODE_long %>%
  #   group_by(ensemble) %>%
  #   summarise(max_PCWD = max(PCWD, na.rm=TRUE), .groups = "drop") %>%
  #   summarise(max_max = max(max_PCWD, na.rm = TRUE), .groups="drop") %>%
  #   mutate(label = "Ens_maximum")
  
  # Build ggplot
  p_ts <- ggplot()
  
  # Exceedance rectangles colored by direction
  if (nrow(exceed_periods) > 0) {
    p_ts <- p_ts + 
      geom_rect(
        data = exceed_periods,
        aes(xmin = start, xmax = end, 
            ymin = y_limits[1], ymax = y_limits[2],
            fill = direction),
        inherit.aes = FALSE,
        alpha = 0.4
      ) 
  }
  p_ts <- p_ts +
    geom_line(data = dat$MODE_long, aes(time, PDSI, group=ensemble, color = factor("EMs_ModE")), linewidth=0.3, alpha=0.4) +
    geom_line(data = dat$MODE_mean, aes(time, PDSI, color = factor("Mean_ModE")), linewidth = 0.8) +
    geom_line(data = means_filtered, aes(time, EM, color="rolling_mean"), linewidth=0.7) +
    # geom_hline(data = ensemble_max, aes(yintercept = max_max, color= factor("Ens_maximum")),linetype="dotted", linewidth=0.5)+
    scale_y_continuous(limits=rev(y_limits), breaks=common_breaks, trans = "reverse") +    
    scale_color_manual(name = "Data",
      values = c(
        "EMs_ModE"  = "#7EA3CC",
        "Mean_ModE" = "#255C99",
        "rolling_mean"="black" 
        # "Ens_maximum"= "grey30"
      ),
      labels = c(
        "EMs - PDSI AWC_100\nann_mean", "Mean - PDSI AWC_100\nann_mean", "30-yr rolling\nmean - PDSI AWC_100\nann_mean"
        #, "ModE-Sim* - max PCWD"
        )
      ) +
      scale_fill_manual(name ="Exceedances", values = c("wetter" = "#136E40", "drier" = "#F4DE70" ),
      labels = c("wetter", "drier")
    ) +
    labs(x = "", y = "PDSI", color = "Data") +
    theme_classic(base_size = 14) +
    theme(axis.title.y = element_text(margin=margin(r=10)),
          legend.position="none", legend.title = element_text(face="bold"))
  
  return(p_ts)
}

ts_MED_PDSI_AWC100_annmean <- plot_region_PDSI_AWC100_ts("MED")
ts_MED_PDSI_AWC100_annmean

ts_WNA_PDSI_AWC100_annmean <- plot_region_PDSI_AWC100_ts("WNA")
ts_WNA_PDSI_AWC100_annmean

ts_ARP_PDSI_AWC100_annmean <- plot_region_PDSI_AWC100_ts("ARP")
ts_ARP_PDSI_AWC100_annmean

ts_CAU_PDSI_AWC100_annmean <- plot_region_PDSI_AWC100_ts("CAU")
ts_CAU_PDSI_AWC100_annmean

```

```{r}
prepare_region_data_PDSI <- function(region,
                                regional_test_PDSI_AWCrecon_annual_mean,
                                mode_years = 1420:2009,
                                drop_initial_years = 2) {
  MODE_mat <- regional_test_PDSI_AWCrecon_annual_mean[[region]]

  MODE_long <- MODE_mat %>%
    as.data.frame() %>%
    mutate(time = mode_years) %>%
    pivot_longer(-time, names_to = "ensemble", values_to = "PDSI")

  MODE_mean <- MODE_long %>%
    group_by(time) %>%
    summarise(PDSI = mean(PDSI, na.rm = TRUE), .groups = "drop")
  # ———————— drop the first `drop_initial_years` rows ————————
  MODE_long <- MODE_long %>% 
    filter(time >= mode_years[drop_initial_years])
  
  MODE_mean <- MODE_mean %>% 
    filter(time >= mode_years[drop_initial_years])

  list(
    MODE_long = MODE_long,
    MODE_mean = MODE_mean
  )
}


#Timeseries function alone: 
plot_region_PDSI_AWCrecon_ts <- function(region_code) {
  
  dat <- prepare_region_data_PDSI(region_code,
                             regional_test_PDSI_AWCrecon_annual_mean)
  
  # Compute anomalies + bounds + exceedance runs
  means_filtered <- rolling_means_AWCrecon_annual_mean_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      EM = mean(rolling_means, na.rm=TRUE),
      .groups = "drop"
    )
  
  first_mean <- means_filtered %>%
    arrange(time) %>%
    slice(1) %>%
    pull(EM)
  
  means_filtered <- means_filtered %>%
    mutate(
      lower_bound = -1,
      upper_bound = 1,
      exceeds     = EM < lower_bound | EM > upper_bound,
      direction   = case_when(
        EM > upper_bound ~ "wetter",
        EM < lower_bound ~ "drier",
        TRUE ~ NA_character_
      ),
      run_id      = data.table::rleid(exceeds)
    )
  
  exceed_periods <- means_filtered %>%
    filter(exceeds) %>%
    group_by(run_id, direction) %>%
    summarise(
      start = min(time),
      end   = max(time),
      .groups="drop"
    )
  
  # y-limits
  y_limits     <- range(dat$MODE_long$PDSI, na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  # # Compute max of each ensemble for dotted lines
  # ensemble_max <- dat$MODE_long %>%
  #   group_by(ensemble) %>%
  #   summarise(max_PCWD = max(PCWD, na.rm=TRUE), .groups = "drop") %>%
  #   summarise(max_max = max(max_PCWD, na.rm = TRUE), .groups="drop") %>%
  #   mutate(label = "Ens_maximum")
  
  # Build ggplot
  p_ts <- ggplot()
  
  # Exceedance rectangles colored by direction
  if (nrow(exceed_periods) > 0) {
    p_ts <- p_ts + 
      geom_rect(
        data = exceed_periods,
        aes(xmin = start, xmax = end, 
            ymin = y_limits[1], ymax = y_limits[2],
            fill = direction),
        inherit.aes = FALSE,
        alpha = 0.4
      ) 
  }
  p_ts <- p_ts +
    geom_line(data = dat$MODE_long, aes(time, PDSI, group=ensemble, color = factor("EMs_ModE")), linewidth=0.3, alpha=0.4) +
    geom_line(data = dat$MODE_mean, aes(time, PDSI, color = factor("Mean_ModE")), linewidth = 0.8) +
    geom_line(data = means_filtered, aes(time, EM, color="rolling_mean"), linewidth=0.7) +
    # geom_hline(data = ensemble_max, aes(yintercept = max_max, color= factor("Ens_maximum")),linetype="dotted", linewidth=0.5)+
    scale_y_continuous(limits=rev(y_limits), breaks=common_breaks, trans = "reverse") +    
    scale_color_manual(name = "Data",
      values = c(
        "EMs_ModE"  = "#7EA3CC",
        "Mean_ModE" = "#255C99",
        "rolling_mean"="black" 
        # "Ens_maximum"= "grey30"
      ),
      labels = c(
        "EMs - PDSI AWC_recon\nann_mean", "Mean - PDSI AWC_recon\nann_mean", "30-yr rolling\nmean - PDSI AWC_recon\nann_mean"
        #, "ModE-Sim* - max PCWD"
        )
      ) +
      scale_fill_manual(name ="Exceedances", values = c("wetter" = "#136E40", "drier" = "#F4DE70" ),
      labels = c("wetter", "drier")
    ) +
    labs(x = "", y = "PDSI", color = "Data") +
    theme_classic(base_size = 14) +
    theme(axis.title.y = element_text(margin=margin(r=10)),
          legend.position="none", legend.title = element_text(face="bold"))
  
  return(p_ts)
}

ts_MED_PDSI_AWCrecon_annmean <- plot_region_PDSI_AWCrecon_ts("MED")
ts_MED_PDSI_AWCrecon_annmean

ts_WNA_PDSI_AWCrecon_annmean <- plot_region_PDSI_AWCrecon_ts("WNA")
ts_WNA_PDSI_AWCrecon_annmean

ts_ARP_PDSI_AWCrecon_annmean <- plot_region_PDSI_AWCrecon_ts("ARP")
ts_ARP_PDSI_AWCrecon_annmean

ts_CAU_PDSI_AWCrecon_annmean <- plot_region_PDSI_AWCrecon_ts("CAU")
ts_CAU_PDSI_AWCrecon_annmean

```



Min values for AWC_100: reflecting driest values each year

```{r}
prepare_region_data_PDSI <- function(region,
                                regional_test_PDSI_AWC100_annual_min,
                                mode_years = 1420:2009,
                                drop_initial_years = 2) {
  MODE_mat <- regional_test_PDSI_AWC100_annual_min[[region]]

  MODE_long <- MODE_mat %>%
    as.data.frame() %>%
    mutate(time = mode_years) %>%
    pivot_longer(-time, names_to = "ensemble", values_to = "PDSI")

  MODE_mean <- MODE_long %>%
    group_by(time) %>%
    summarise(PDSI = mean(PDSI, na.rm = TRUE), .groups = "drop")
  # ———————— drop the first `drop_initial_years` rows ————————
  MODE_long <- MODE_long %>% 
    filter(time >= mode_years[drop_initial_years])
  
  MODE_mean <- MODE_mean %>% 
    filter(time >= mode_years[drop_initial_years])

  list(
    MODE_long = MODE_long,
    MODE_mean = MODE_mean
  )
}


#Timeseries function alone: 
plot_region_PDSI_AWC100_ts <- function(region_code) {
  
  dat <- prepare_region_data_PDSI(region_code,
                             regional_test_PDSI_AWC100_annual_min)
  
  # Compute anomalies + bounds + exceedance runs
  means_filtered <- rolling_means_AWC100_annual_min_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      EM = mean(rolling_means, na.rm=TRUE),
      .groups = "drop"
    )
  
  first_mean <- means_filtered %>%
    arrange(time) %>%
    slice(1) %>%
    pull(EM)
  
  means_filtered <- means_filtered %>%
    mutate(
      lower_bound = -1,
      upper_bound = 1,
      exceeds     = EM < lower_bound | EM > upper_bound,
      direction   = case_when(
        EM > upper_bound ~ "wetter",
        EM < lower_bound ~ "drier",
        TRUE ~ NA_character_
      ),
      run_id      = data.table::rleid(exceeds)
    )
  
  exceed_periods <- means_filtered %>%
    filter(exceeds) %>%
    group_by(run_id, direction) %>%
    summarise(
      start = min(time),
      end   = max(time),
      .groups="drop"
    )
  
  # y-limits
  y_limits     <- range(dat$MODE_long$PDSI, na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  # # Compute max of each ensemble for dotted lines
  # ensemble_max <- dat$MODE_long %>%
  #   group_by(ensemble) %>%
  #   summarise(max_PCWD = max(PCWD, na.rm=TRUE), .groups = "drop") %>%
  #   summarise(max_max = max(max_PCWD, na.rm = TRUE), .groups="drop") %>%
  #   mutate(label = "Ens_maximum")
  
  # Build ggplot
  p_ts <- ggplot()
  
  # Exceedance rectangles colored by direction
  if (nrow(exceed_periods) > 0) {
    p_ts <- p_ts + 
      geom_rect(
        data = exceed_periods,
        aes(xmin = start, xmax = end, 
            ymin = y_limits[1], ymax = y_limits[2],
            fill = direction),
        inherit.aes = FALSE,
        alpha = 0.4
      ) 
  }
  p_ts <- p_ts +
    geom_line(data = dat$MODE_long, aes(time, PDSI, group=ensemble, color = factor("EMs_ModE")), linewidth=0.3, alpha=0.4) +
    geom_line(data = dat$MODE_mean, aes(time, PDSI, color = factor("Mean_ModE")), linewidth = 0.8) +
    geom_line(data = means_filtered, aes(time, EM, color="rolling_mean"), linewidth=0.7) +
    # geom_hline(data = ensemble_max, aes(yintercept = max_max, color= factor("Ens_maximum")),linetype="dotted", linewidth=0.5)+
    scale_y_continuous(limits=rev(y_limits), breaks=common_breaks, trans = "reverse") +    
    scale_color_manual(name = "Data",
      values = c(
        "EMs_ModE"  = "#7EA3CC",
        "Mean_ModE" = "#255C99",
        "rolling_mean"="black" 
        # "Ens_maximum"= "grey30"
      ),
      labels = c(
        "EMs - PDSI AWC_100\nann_min", "Mean - PDSI AWC_100\nann_min", "30-yr rolling\nmin - PDSI AWC_100\nann_min"
        #, "ModE-Sim* - max PCWD"
        )
      ) +
      scale_fill_manual(name ="Exceedances", values = c("wetter" = "#136E40", "drier" = "#F4DE70" ),
      labels = c("wetter", "drier")
    ) +
    labs(x = "Year", y = "PDSI", color = "Data") +
    theme_classic(base_size = 14) +
    theme(axis.title.y = element_text(margin=margin(r=10)),
          legend.position="none", legend.title = element_text(face="bold"))
  
  return(p_ts)
}

ts_MED_PDSI_AWC100_annmin <- plot_region_PDSI_AWC100_ts("MED")
ts_MED_PDSI_AWC100_annmin

ts_WNA_PDSI_AWC100_annmin <- plot_region_PDSI_AWC100_ts("WNA")
ts_WNA_PDSI_AWC100_annmin

ts_ARP_PDSI_AWC100_annmin <- plot_region_PDSI_AWC100_ts("ARP")
ts_ARP_PDSI_AWC100_annmin

ts_CAU_PDSI_AWC100_annmin <- plot_region_PDSI_AWC100_ts("CAU")
ts_CAU_PDSI_AWC100_annmin
```


```{r}
prepare_region_data_PDSI <- function(region,
                                regional_test_PDSI_AWCrecon_annual_min,
                                mode_years = 1420:2009,
                                drop_initial_years = 2) {
  MODE_mat <- regional_test_PDSI_AWCrecon_annual_min[[region]]

  MODE_long <- MODE_mat %>%
    as.data.frame() %>%
    mutate(time = mode_years) %>%
    pivot_longer(-time, names_to = "ensemble", values_to = "PDSI")

  MODE_mean <- MODE_long %>%
    group_by(time) %>%
    summarise(PDSI = mean(PDSI, na.rm = TRUE), .groups = "drop")
  # ———————— drop the first `drop_initial_years` rows ————————
  MODE_long <- MODE_long %>% 
    filter(time >= mode_years[drop_initial_years])
  
  MODE_mean <- MODE_mean %>% 
    filter(time >= mode_years[drop_initial_years])

  list(
    MODE_long = MODE_long,
    MODE_mean = MODE_mean
  )
}


#Timeseries function alone: 
plot_region_PDSI_AWCrecon_ts <- function(region_code) {
  
  dat <- prepare_region_data_PDSI(region_code,
                             regional_test_PDSI_AWCrecon_annual_min)
  
  # Compute anomalies + bounds + exceedance runs
  means_filtered <- rolling_means_AWCrecon_annual_min_long %>%
    filter(region == region_code) %>%
    group_by(time) %>%
    summarise(
      EM = mean(rolling_means, na.rm=TRUE),
      .groups = "drop"
    )
  
  first_mean <- means_filtered %>%
    arrange(time) %>%
    slice(1) %>%
    pull(EM)
  
  means_filtered <- means_filtered %>%
    mutate(
      lower_bound = -1,
      upper_bound = 1,
      exceeds     = EM < lower_bound | EM > upper_bound,
      direction   = case_when(
        EM > upper_bound ~ "wetter",
        EM < lower_bound ~ "drier",
        TRUE ~ NA_character_
      ),
      run_id      = data.table::rleid(exceeds)
    )
  
  exceed_periods <- means_filtered %>%
    filter(exceeds) %>%
    group_by(run_id, direction) %>%
    summarise(
      start = min(time),
      end   = max(time),
      .groups="drop"
    )
  
  # y-limits
  y_limits     <- range(dat$MODE_long$PDSI, na.rm=TRUE)
  common_breaks <- pretty(y_limits, n=6)
  
  # # Compute max of each ensemble for dotted lines
  # ensemble_max <- dat$MODE_long %>%
  #   group_by(ensemble) %>%
  #   summarise(max_PCWD = max(PCWD, na.rm=TRUE), .groups = "drop") %>%
  #   summarise(max_max = max(max_PCWD, na.rm = TRUE), .groups="drop") %>%
  #   mutate(label = "Ens_maximum")
  
  # Build ggplot
  p_ts <- ggplot()
  
  # Exceedance rectangles colored by direction
  if (nrow(exceed_periods) > 0) {
    p_ts <- p_ts + 
      geom_rect(
        data = exceed_periods,
        aes(xmin = start, xmax = end, 
            ymin = y_limits[1], ymax = y_limits[2],
            fill = direction),
        inherit.aes = FALSE,
        alpha = 0.4
      ) 
  }
  p_ts <- p_ts +
    geom_line(data = dat$MODE_long, aes(time, PDSI, group=ensemble, color = factor("EMs_ModE")), linewidth=0.3, alpha=0.4) +
    geom_line(data = dat$MODE_mean, aes(time, PDSI, color = factor("Mean_ModE")), linewidth = 0.8) +
    geom_line(data = means_filtered, aes(time, EM, color="rolling_mean"), linewidth=0.7) +
    # geom_hline(data = ensemble_max, aes(yintercept = max_max, color= factor("Ens_maximum")),linetype="dotted", linewidth=0.5)+
    scale_y_continuous(limits=rev(y_limits), breaks=common_breaks, trans = "reverse") +    
    scale_color_manual(name = "Data",
      values = c(
        "EMs_ModE"  = "#7EA3CC",
        "Mean_ModE" = "#255C99",
        "rolling_mean"="black" 
        # "Ens_maximum"= "grey30"
      ),
      labels = c(
        "EMs - PDSI AWC_recon\nann_min", "Mean - PDSI AWC_recon\nann_min", "30-yr rolling\nmin - PDSI AWC_recon\nann_min"
        #, "ModE-Sim* - max PCWD"
        )
      ) +
      scale_fill_manual(name ="Exceedances", values = c("wetter" = "#136E40", "drier" = "#F4DE70" ),
      labels = c("wetter", "drier")
    ) +
    labs(x = "Year", y = "PDSI", color = "Data") +
    theme_classic(base_size = 14) +
    theme(axis.title.y = element_text(margin=margin(r=10)),
          legend.position="none", legend.title = element_text(face="bold"))
  
  return(p_ts)
}

ts_MED_PDSI_AWCrecon_annmin <- plot_region_PDSI_AWCrecon_ts("MED")
ts_MED_PDSI_AWCrecon_annmin

ts_WNA_PDSI_AWCrecon_annmin <- plot_region_PDSI_AWCrecon_ts("WNA")
ts_WNA_PDSI_AWCrecon_annmin

ts_ARP_PDSI_AWCrecon_annmin <- plot_region_PDSI_AWCrecon_ts("ARP")
ts_ARP_PDSI_AWCrecon_annmin

ts_CAU_PDSI_AWCrecon_annmin <- plot_region_PDSI_AWCrecon_ts("CAU")
ts_CAU_PDSI_AWCrecon_annmin
```


```{r}
### save plots
ts_MED <- ts_MED + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_MED_PDSI_AWC100 <- ts_MED_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_MED_PDSI_AWC100_annmean <- ts_MED_PDSI_AWC100_annmean + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_MED_PDSI_AWC100_annmin <- ts_MED_PDSI_AWC100_annmin + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_MED_PDSI_AWCrecon <- ts_MED_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_MED_PDSI_AWCrecon_annmean <- ts_MED_PDSI_AWCrecon_annmean + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_MED_PDSI_AWCrecon_annmin <- ts_MED_PDSI_AWCrecon_annmin + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))



combined_MED_ALL <- plot_grid(
  ts_MED,ts_MED,ts_MED_PDSI_AWCrecon, ts_MED_PDSI_AWC100, ts_MED_PDSI_AWCrecon_annmean,ts_MED_PDSI_AWC100_annmean, ts_MED_PDSI_AWCrecon_annmin,ts_MED_PDSI_AWC100_annmin,
  nrow        = 4,
  ncol        = 2,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a) MED – Annmax PCWD", "e) as in a)",
    "b) MED – PDSI AWC recon","f) MED – PDSI AWC 100mm",
    "c) MED – PDSI AWC recon - ANNUAL MEAN",    "g) MED – PDSI AWC 100mm - ANNUAL MEAN",
    "d) MED – PDSI AWC recon - ANNUAL MIN",    "h) MED – PDSI AWC 100mm - ANNUAL MIN"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

print(combined_MED_ALL)

ggsave(
  filename = here::here("figures/PCWD_PDSI_all_comp_MED.pdf"),
  plot     = combined_MED_ALL,
  device   = cairo_pdf,    # better text rendering
  width    = 11,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```

```{r}
### save plots
ts_WNA <- ts_WNA + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_WNA_PDSI_AWC100 <- ts_WNA_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_WNA_PDSI_AWC100_annmean <- ts_WNA_PDSI_AWC100_annmean + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_WNA_PDSI_AWC100_annmin <- ts_WNA_PDSI_AWC100_annmin + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_WNA_PDSI_AWCrecon <- ts_WNA_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_WNA_PDSI_AWCrecon_annmean <- ts_WNA_PDSI_AWCrecon_annmean + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_WNA_PDSI_AWCrecon_annmin <- ts_WNA_PDSI_AWCrecon_annmin + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))



combined_WNA_ALL <- plot_grid(
  ts_WNA,ts_WNA,ts_WNA_PDSI_AWCrecon, ts_WNA_PDSI_AWC100, ts_WNA_PDSI_AWCrecon_annmean,ts_WNA_PDSI_AWC100_annmean, ts_WNA_PDSI_AWCrecon_annmin,ts_WNA_PDSI_AWC100_annmin,
  nrow        = 4,
  ncol        = 2,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a) WNA – Annmax PCWD", "e) as in a)",
    "b) WNA – PDSI AWC recon","f) WNA – PDSI AWC 100mm",
    "c) WNA – PDSI AWC recon - ANNUAL MEAN",    "g) WNA – PDSI AWC 100mm - ANNUAL MEAN",
    "d) WNA – PDSI AWC recon - ANNUAL MIN",    "h) WNA – PDSI AWC 100mm - ANNUAL MIN"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

print(combined_WNA_ALL)

ggsave(
  filename = here::here("figures/PCWD_PDSI_all_comp_WNA.pdf"),
  plot     = combined_WNA_ALL,
  device   = cairo_pdf,    # better text rendering
  width    = 11,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```

```{r}
### save plots
ts_ARP <- ts_ARP + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_ARP_PDSI_AWC100 <- ts_ARP_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_ARP_PDSI_AWC100_annmean <- ts_ARP_PDSI_AWC100_annmean + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_ARP_PDSI_AWC100_annmin <- ts_ARP_PDSI_AWC100_annmin + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_ARP_PDSI_AWCrecon <- ts_ARP_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_ARP_PDSI_AWCrecon_annmean <- ts_ARP_PDSI_AWCrecon_annmean + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_ARP_PDSI_AWCrecon_annmin <- ts_ARP_PDSI_AWCrecon_annmin + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))



combined_ARP_ALL <- plot_grid(
  ts_ARP,ts_ARP,ts_ARP_PDSI_AWCrecon, ts_ARP_PDSI_AWC100, ts_ARP_PDSI_AWCrecon_annmean,ts_ARP_PDSI_AWC100_annmean, ts_ARP_PDSI_AWCrecon_annmin,ts_ARP_PDSI_AWC100_annmin,
  nrow        = 4,
  ncol        = 2,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a) ARP – Annmax PCWD", "e) as in a)",
    "b) ARP – PDSI AWC recon","f) ARP – PDSI AWC 100mm",
    "c) ARP – PDSI AWC recon - ANNUAL MEAN",    "g) ARP – PDSI AWC 100mm - ANNUAL MEAN",
    "d) ARP – PDSI AWC recon - ANNUAL MIN",    "h) ARP – PDSI AWC 100mm - ANNUAL MIN"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

print(combined_ARP_ALL)

ggsave(
  filename = here::here("figures/PCWD_PDSI_all_comp_ARP.pdf"),
  plot     = combined_ARP_ALL,
  device   = cairo_pdf,    # better text rendering
  width    = 11,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```

```{r}
### save plots
ts_CAU <- ts_CAU + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_CAU_PDSI_AWC100 <- ts_CAU_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_CAU_PDSI_AWC100_annmean <- ts_CAU_PDSI_AWC100_annmean + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_CAU_PDSI_AWC100_annmin <- ts_CAU_PDSI_AWC100_annmin + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_CAU_PDSI_AWCrecon <- ts_CAU_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_CAU_PDSI_AWCrecon_annmean <- ts_CAU_PDSI_AWCrecon_annmean + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_CAU_PDSI_AWCrecon_annmin <- ts_CAU_PDSI_AWCrecon_annmin + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))



combined_CAU_ALL <- plot_grid(
  ts_CAU,ts_CAU,ts_CAU_PDSI_AWCrecon, ts_CAU_PDSI_AWC100, ts_CAU_PDSI_AWCrecon_annmean,ts_CAU_PDSI_AWC100_annmean, ts_CAU_PDSI_AWCrecon_annmin,ts_CAU_PDSI_AWC100_annmin,
  nrow        = 4,
  ncol        = 2,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a) CAU – Annmax PCWD", "e) as in a)",
    "b) CAU – PDSI AWC recon","f) CAU – PDSI AWC 100mm",
    "c) CAU – PDSI AWC recon - ANNUAL MEAN",    "g) CAU – PDSI AWC 100mm - ANNUAL MEAN",
    "d) CAU – PDSI AWC recon - ANNUAL MIN",    "h) CAU – PDSI AWC 100mm - ANNUAL MIN"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

print(combined_CAU_ALL)

ggsave(
  filename = here::here("figures/PCWD_PDSI_all_comp_CAU.pdf"),
  plot     = combined_CAU_ALL,
  device   = cairo_pdf,    # better text rendering
  width    = 11,           # in inches
  height   = 13,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```

```{r}
### Stack plots differently for SI with monthly PDSI and PCWD
ts_MED <- ts_MED + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_MED_PDSI_AWC100 <- ts_MED_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_MED_PDSI_AWCrecon <- ts_MED_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

ts_WNA <- ts_WNA + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_WNA_PDSI_AWC100 <- ts_WNA_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_WNA_PDSI_AWCrecon <- ts_WNA_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

ts_ARP <- ts_ARP + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_ARP_PDSI_AWC100 <- ts_ARP_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_ARP_PDSI_AWCrecon <- ts_ARP_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

ts_CAU <- ts_CAU + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_CAU_PDSI_AWC100 <- ts_CAU_PDSI_AWC100 + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
ts_CAU_PDSI_AWCrecon <- ts_CAU_PDSI_AWCrecon + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))

## build plot clusters
cmp_MED <- cowplot::plot_grid(
  ts_MED, ts_MED_PDSI_AWCrecon, ts_MED_PDSI_AWC100,
  nrow        = 3,
  ncol        = 1,
  align       = "v",
  axis        = "l",
  labels      = c(
    "MED – Annmax PCWD", "MED – PDSI AWC recon","MED – PDSI AWC 100mm"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.05,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

cmp_WNA <- cowplot::plot_grid(
  ts_WNA, ts_WNA_PDSI_AWCrecon, ts_WNA_PDSI_AWC100,
  nrow        = 3,
  ncol        = 1,
  align       = "v",
  axis        = "l",
  labels      = c(
    "WNA – Annmax PCWD", "WNA – PDSI AWC recon","WNA – PDSI AWC 100mm"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.05,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

cmp_ARP <- cowplot::plot_grid(
  ts_ARP, ts_ARP_PDSI_AWCrecon, ts_ARP_PDSI_AWC100,
  nrow        = 3,
  ncol        = 1,
  align       = "v",
  axis        = "l",
  labels      = c(
    "ARP – Annmax PCWD", "ARP – PDSI AWC recon","ARP – PDSI AWC 100mm"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.05,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

cmp_CAU <- cowplot::plot_grid(
  ts_CAU, ts_CAU_PDSI_AWCrecon, ts_CAU_PDSI_AWC100,
  nrow        = 3,
  ncol        = 1,
  align       = "v",
  axis        = "l",
  labels      = c(
    "CAU – Annmax PCWD", "CAU – PDSI AWC recon","CAU – PDSI AWC 100mm"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.05,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.975,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

### combine all components: 


cmp_MED <- cmp_MED + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
cmp_WNA <- cmp_WNA + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
cmp_ARP <- cmp_ARP + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))
cmp_CAU <- cmp_CAU + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 0))


all_ts_PDSI <- cowplot::plot_grid(
  cmp_MED, cmp_WNA, cmp_ARP, cmp_CAU,
  nrow        = 2,
  ncol        = 2,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a)", "b)","c)", "d)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.02,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
  # Tweak vertical positioning if you like
  label_y = 0.99,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

print(all_ts_PDSI)


ggsave(
  filename = here::here("figures/all_ts_PDSI_PCWD.pdf"),
  plot     = all_ts_PDSI,
  device   = cairo_pdf,    # better text rendering
  width    = 12,           # in inches
  height   = 12,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)

```



Scatterplots of PDSI vs PCWD metrics 

```{r}
regions <- c("MED","ARP","CAU","WNA")

library(tidyverse)

pcwd_df <- map_dfr(regions, function(r){

  mat <- regional_PCWD[[r]]

  df <- as.data.frame(mat)

  df$year <- 1420:(1420 + nrow(df) - 1)   # adjust if needed

  df %>%
    pivot_longer(-year, names_to = "ensemble", values_to = "PCWD") %>%
    mutate(
      region = r,
      ensemble = gsub("_tidy","",ensemble)
    )
})

pdsi_df_recon <- map_dfr(regions, \(r)
  regional_test_PDSI_AWCrecon[[r]] |>
    as.data.frame() |>
    mutate(time = row_number()) |>
    pivot_longer(-time, names_to = "ensemble", values_to = "PDSI") |>
    mutate(
      region = r,
      year  = 1420 + (time - 1) %/% 12
    ) |>
    group_by(region, year, ensemble) |>
    summarise(
      PDSI_mean = mean(PDSI, na.rm = TRUE),
      PDSI_min  = min(PDSI, na.rm = TRUE),
      .groups = "drop"
    )
)

combined_recon <- left_join(
  pcwd_df,
  pdsi_df_recon,
  by = c("region","year", "ensemble")
)

plot_df_recon <- combined_recon %>%
  pivot_longer(
    cols = c(PDSI_mean, PDSI_min),
    names_to = "PDSI_type",
    values_to = "PDSI"
  )

# Compute correlation per region × PDSI_type
cor_df_recon <- plot_df_recon %>%
  group_by(region, PDSI_type) %>%
  summarise(
    r = cor(PDSI, PCWD, use = "complete.obs"),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("r = ", round(r, 2))
  )


## for the 100mm AWC: 

pdsi_df_100 <- map_dfr(regions, \(r)
  regional_test_PDSI_AWC100[[r]] |>
    as.data.frame() |>
    mutate(time = row_number()) |>
    pivot_longer(-time, names_to = "ensemble", values_to = "PDSI") |>
    mutate(
      region = r,
      year  = 1420 + (time - 1) %/% 12
    ) |>
    group_by(region, year, ensemble) |>
    summarise(
      PDSI_mean = mean(PDSI, na.rm = TRUE),
      PDSI_min  = min(PDSI, na.rm = TRUE),
      .groups = "drop"
    )
)

combined_100 <- left_join(
  pcwd_df,
  pdsi_df_100,
  by = c("region","year","ensemble")
)

plot_df_100 <- combined_100 %>%
  pivot_longer(
    cols = c(PDSI_mean, PDSI_min),
    names_to = "PDSI_type",
    values_to = "PDSI"
  )
# Compute correlation per region × PDSI_type
cor_df_100 <- plot_df_100 %>%
  group_by(region, PDSI_type) %>%
  summarise(
    r = cor(PDSI, PCWD, use = "complete.obs"),
    .groups = "drop"
  ) %>%
  mutate(
    label = paste0("r = ", round(r, 2))
  )

```


```{r}
library(ggh4x)
#install.packages("ggrastr")
library(ggrastr)
library(cowplot)

region_cols <- c(
  ARP = "#E5973B",
  CAU = "#3C3980",
    MED = "#F5B0CB",
  WNA = "#236A3F"
)

positions_recon <- plot_df_recon %>%
  group_by(region) %>%
  summarise(
    x_max = max(PDSI, na.rm = TRUE),
    y_max = max(PCWD, na.rm = TRUE)
  )

cor_df_recon_plot <- cor_df_recon %>%
  left_join(positions_recon, by = "region") %>%
  group_by(region) %>%
  mutate(
    x_pos = x_max,
    y_pos = y_max - (row_number() - 2) * 0.05 * y_max  # stagger vertically
  ) %>%
  ungroup()

scatter_PDSIrecon <- 
  ggplot(plot_df_recon, aes(x = PDSI, y = PCWD, shape = PDSI_type, color = PDSI_type)) +
  geom_point(alpha = 0.5, size = 1.5) +  # rasterized points
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  facet_wrap2(
    ~region,
    scales = "free",
    strip = strip_themed(
      background_x = elem_list_rect(fill = scales::alpha(region_cols, 0.6))
    )
  ) +
  scale_color_manual(
    name = "Annual scPDSI",
    values = c(
      PDSI_min = "#C06C3E",
      PDSI_mean = "#3B6FB6"
    ),
    labels = c(
      PDSI_min = "min scPDSI",
      PDSI_mean = "mean scPDSI"
    )
  ) +
  scale_shape_manual(
    name = "Annual scPDSI",
    values = c(
      PDSI_min = 17,
      PDSI_mean = 16
    ),
    labels = c(
      PDSI_min = "min scPDSI",
      PDSI_mean = "mean scPDSI"
    )
  ) +
    geom_text(
    data = cor_df_recon_plot,
    aes(x = x_pos, y = y_pos, label = label, color = PDSI_type),
    inherit.aes = FALSE,
    fontface = "bold",
    hjust = 1,
    size = 4
  )+
  theme_classic() +
  theme(
    legend.position = "none",
    legend.text = element_text(size=11),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15, color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11)
  ) +
  labs(
    x = "scPDSI_AWCrecon",
    y = "PCWD (mm)"
  )

###
  
positions_100 <- plot_df_100 %>%
  group_by(region) %>%
  summarise(
    x_max = max(PDSI, na.rm = TRUE),
    y_max = max(PCWD, na.rm = TRUE)
  )

cor_df_100_plot <- cor_df_100 %>%
  left_join(positions_100, by = "region") %>%
  group_by(region) %>%
  mutate(
    x_pos = x_max,
    y_pos = y_max - (row_number() - 2) * 0.05 * y_max  # stagger vertically
  ) %>%
  ungroup()


scatter_PDSI100 <- ggplot(plot_df_100, aes(x = PDSI, y = PCWD, shape = PDSI_type, color = PDSI_type)) +
  geom_point(alpha = 0.5, size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1.2) +
  facet_wrap2(
    ~region,
    scales = "free",
    strip = strip_themed(
      background_x = elem_list_rect(fill = scales::alpha(region_cols, 0.6))
    )
  ) +
  scale_color_manual(
    name = "Annual scPDSI",
    values = c(
      PDSI_min = "#C06C3E",
      PDSI_mean = "#3B6FB6"
    ),
    labels = c(
      PDSI_min = "min scPDSI",
      PDSI_mean = "mean scPDSI"
    )
  ) +
  scale_shape_manual(
    name = "Annual scPDSI",
    values = c(
      PDSI_min = 17,
      PDSI_mean = 16
    ),
    labels = c(
      PDSI_min = "min scPDSI",
      PDSI_mean = "mean scPDSI"
    )
  ) +geom_text(
    data = cor_df_100_plot,
    aes(x = x_pos, y = y_pos, label = label, color = PDSI_type),
    inherit.aes = FALSE,
    fontface = "bold",
    hjust = 1,
    size = 4
  )+
  theme_classic() +
  theme(
    legend.position = "none",
    legend.text = element_text(size=11),
    legend.title = element_text(face = "bold"),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(size = 15, color = "black"),
    axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
    axis.text.y = element_text(size = 11),
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11)
  ) +
  labs(
    x = "scPDSI_AWC100",
    y = ""
  )

# ## save legend with larger scatter point size
# ggsave(
#   filename = here::here("figures/PDSI_scatter_legend.pdf"),
#   device   = cairo_pdf,    # better text rendering
#   width    = 10,           # in inches
#   height   = 6.5,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )

combined_PDSI_scatter <- plot_grid(
  scatter_PDSIrecon, scatter_PDSI100,
  ncol = 2,
  nrow = 1,
  rel_widths = c(1, 1),
  align = "h",
  axis = "tblr",
  labels = c(
    "a)",  "b) "
  )
  # label_size = 14,
  # label_x = 0.09,
  # hjust = 0,
  # label_y = 0.95,
  # vjust = 1
)

print(combined_PDSI_scatter)

ggsave(
  filename = here::here("figures/combined_scatterPDSI.pdf"),
  plot     = combined_PDSI_scatter,
  device   = cairo_pdf,    # better text rendering
  width    = 12,           # in inches
  height   = 7,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```

