---
title: "PDSI_fLUE"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Comparison of PCWD and PDSI on fluxnet sites in comparison to soil moisture droughts

```{r message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(here)
library(lubridate)
library(patchwork)
library(extRemes)
library(ggplot2)
library(cwd)
library(visdat)
library(recipes)
library(readxl)
library(zoo)
library(ggpubr)
library(scPDSI)
```

A potential cumulative water deficit can be calculated using net radiation and the *potential* evapotranspiration (PET). Here, we calculate PET based on net radiation and the approach by Priestley & Taylor (1972) as implemented by Davis et al. (2017). Necessary functions are part of the {cwd} package.

## Prepare data

Read data from the example FLUNXET file contained in this repository.
```{r}
df <- readRDS("/Users/phelpap/Library/CloudStorage/OneDrive-Personal/Dokumente/Bern/Master thesis/R/cwd/data/df_ch-lae.rds") |>
   # convert to from kPA to Pa (SI units are used for inputs to functions in the package)
   mutate(PA_F = 1e3 * PA_F) |>
   # Convert TIMESTAMP to Date format
   mutate(TIMESTAMP = ymd(TIMESTAMP))

# 
# df <- readRDS("/Users/phelpap/Library/CloudStorage/OneDrive-Personal/Dokumente/Bern/Master thesis/R/cwd/data/df_fr-pue.rds") |> 
#    # convert to from kPA to Pa (SI units are used for inputs to functions in the package)
#    mutate(PA_F = 1e3 * PA_F) |>
#    # Convert TIMESTAMP to Date format
#    mutate(TIMESTAMP = ymd(TIMESTAMP))

visdat::vis_miss(df)
```

Some net radiation data is missing. Impute missing values by KNN.
```{r}
pp <- recipes::recipe(
  NETRAD ~ SW_IN_F_MDS + LW_IN_F_MDS + TA_F_MDS,
  data = df |> 
    drop_na(SW_IN_F_MDS, LW_IN_F_MDS, TA_F_MDS)
) |>
  recipes::step_center(
    recipes::all_numeric(),
    -recipes::all_outcomes()
  ) |>
  recipes::step_scale(
    recipes::all_numeric(),
    -recipes::all_outcomes()
  ) |>
  recipes::step_impute_knn(
    recipes::all_outcomes(),
    neighbors = 5
  )

pp_prep <- recipes::prep(
  pp,
  training = df |> 
    drop_na(SW_IN_F_MDS, LW_IN_F_MDS, TA_F_MDS)
)

df_baked <- recipes::bake(
  pp_prep,
  new_data = df
)

# fill missing with gap-filled
df <- df |>
  dplyr::bind_cols(
    df_baked |>
      dplyr::select(
        NETRAD_filled = NETRAD)
  ) |>
  dplyr::mutate(
    NETRAD = ifelse(is.na(NETRAD), NETRAD_filled, NETRAD)
    #qc = ifelse(is.na(netrad), TRUE, FALSE)
  ) |>
  dplyr::select(
    -NETRAD_filled
  )

#visdat::vis_miss(df)
```


## Apply PET function

... and convert from units of mm s-1 to mm d-1.
```{r}
# tested: identical results are obtained with:
# bigleaf::potential.ET(Tair = TA_F_MDS, pressure = PA_F*1e-3, Rn = NETRAD, approach = "Priestley-Taylor")$ET_pot * 60 * 60 * 24
df <- df |> 
  mutate(pet = 60 * 60 * 24 * pet(NETRAD, TA_F_MDS, PA_F))
```

## Visualise, contrasting to observed ET after conversion of energy to mass units

Convert latent heat flux (W/m2) to evapotranspiration in mass units (mm/d).
```{r}
# tested: identical results are obtained with:
# bigleaf::LE.to.ET(LE_F_MDS, TA_F_MDS)* 60 * 60 * 24
le_to_et <- function(le, tc, patm){
  1000 * 60 * 60 * 24 * le / (cwd::calc_enthalpy_vap(tc) * cwd::calc_density_h2o(tc, patm))
}

df <- df |>
  mutate(et = le_to_et(LE_F_MDS, TA_F_MDS, PA_F))
```

Plot mean seasonal cycle.
```{r}
# df |>
#   mutate(TIMESTAMP = lubridate::as_date(as_datetime(TIMESTAMP_END)))

df |> 
  mutate(doy = lubridate::yday(TIMESTAMP)) |> 
  group_by(doy) |> 
  summarise(
    #et = mean(et),
    pet = mean(pet),
    P_F = mean(P_F)  ) |> 
  ggplot() +
  geom_line(aes(doy, P_F, color = "P"))+
  geom_line(aes(doy, pet, color = "PET")) + 
  scale_color_manual(values = c("P" = "#6B74DB", "PET" = "#F59430")) +

  labs(
    x = "Day of year",
    y = expression(paste("Flux (mm d"^-1, ")")), 
    color = "Flux"
  ) +
  theme_classic()+
    theme(axis.text.x = element_text(size = 14),
    legend.position = "none",
    axis.text.y = element_text(size = 14), 
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12)) 

### select specific year to plot seasonal cycle for: 
#flux_2013 <- 
df |> 
  filter(year(TIMESTAMP) == 2013) |> 
  mutate(doy = yday(TIMESTAMP)) |> 
  ggplot() +
  # geom_line(aes(doy, et, color = "ET")) +
  geom_line(aes(doy, P_F, color = "P")) +
  geom_line(aes(doy, pet, color = "PET")) +
  scale_color_manual(values = c("P" = "#6B74DB", "PET" = "#F59430")) +
  labs(
    x = "Day of year",
    y = expression(paste("Flux (mm d"^-1, ")")),
    color = "Flux"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.position = c(0.1,0.85),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12)
  )

#over entire period: 

df |> 
  ggplot() +
  # geom_line(aes(doy, et, color = "ET")) +
  geom_line(aes(TIMESTAMP, P_F, color = "P")) +
  geom_line(aes(TIMESTAMP, pet, color = "PET")) +
  scale_color_manual(values = c("P" = "#6B74DB", "PET" = "#F59430")) +
  labs(
    x = "Time",
    y = expression(paste("Flux (mm d"^-1, ")")),
    color = "Flux"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 14),
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.position = c(0.1,0.85),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12)
  )
```

## Cumulating PET - *P*

Check annual totals.
```{r}
adf <- df |>
  mutate(day = day(TIMESTAMP)) |>
  group_by(day) |>
  summarise(pet = sum(pet), prec = sum(P_F))

adf |>
  tidyr::pivot_longer(cols = c(pet, prec), names_to = "Flux") |> 
  ggplot(aes(x = day, y = value, color = Flux)) +
  geom_line() +
  labs(y = "Flux (mm/yr)") +
  theme_classic()
```


Get the potential cumulative water deficit time series and individual *events*. Note that we use the argument `doy_reset` here to force a re-setting of the potential cumulative water deficit on that same day each year.
A relative and an absolute option are possible for determining the threshold to which a CWD has to be reduced to to terminate the event: 
1) `threshold_terminate`: Set the threshold relative to maximum CWD attained. Defaults to 0, meaning that the CWD has to be fully compensated by water infiltration into the soil to terminate a CWD event.
2) `threshold_terminate_absolute`: threshold determining end of event as `thresh_terminate` but in absolute terms (in mm d-1). Default set to `NA`.
If no option is given, the function defaults to the relative threshold.
Option 2: CWD calculation using an absolute threshold, here 10mm. The deficit reset is not prescribed through a given DOY here but resets once the deficit falls below the threshold. Absolute thresholds should be selected based on the regional water balance. 
```{r}
df <- df |> 
  mutate(pwbal = P_F - pet)

out_cwd <- cwd(
  df,
  varname_wbal = "pwbal",
  varname_date = "TIMESTAMP", 
  thresh_terminate_absolute = 10,
  thresh_drop = 0.0
  )
```

Retain only events of a minimum length of 20 days.
```{r}
out_cwd$inst <- out_cwd$inst |>
  filter(len >= 20)
```

```{r}
##calculate the maximum PCWD of each year
get_maxcwd <- function(df){
    df$inst |>
    mutate(year = lubridate::year(date_start)) |>
    group_by(year) |>
    summarise(max_cwd = max(deficit, na.rm = TRUE)) |>  # Select only the relevant columns
    mutate(max_cwd = max_cwd)
   # return(out_cwd)
}

out_maxcwd <- get_maxcwd(out_cwd)
```


Plot the potential cumulative water deficit time series and events.
```{r}
#pcwd_2013 <- 
ggplot() +
  geom_rect(
    data = out_cwd$inst,
    aes(xmin = date_start, xmax = date_end, ymin = 0, ymax = max( out_cwd$df$deficit)), 
    fill = "grey80",
    color = NA) +
  geom_line(data  =  out_cwd$df, aes(TIMESTAMP, deficit), color = "#DB5A39") +
  scale_x_date(limits = as.Date(c("2013-01-01", "2014-01-01"))) +
  ylim(0, max(out_cwd$df$deficit)) + 
  labs(
    x = "Date",    y = "Potential cumulative water deficit (mm)"
    )+
   theme_classic()+
    theme(axis.text.x = element_text(size = 14),
    legend.position = "right",
    axis.text.y = element_text(size = 14), 
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12)) 

ggplot() +
  geom_line(data  =  out_maxcwd, aes(year, max_cwd), color = "#DB5A39") +
  scale_x_continuous(limits = c(2001, 2014)) +
  labs(
    x = "Date",    y = "Annmax PCWD (mm)"
    )+
   theme_classic()+
    theme(axis.text.x = element_text(size = 14),
    legend.position = "right",
    axis.text.y = element_text(size = 14), 
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12)) 
```

```{r}
library(cowplot)
## merge flux plot and pcwd plot
flux_2013 <- flux_2013 + theme(plot.margin = margin(t = 5, r = 0, b = 0, l = 5))
pcwd_2013 <- pcwd_2013 + theme(plot.margin = margin(t = 5, r = 20, b = 0, l = 5))

combined_fluxes <- plot_grid(
  flux_2013 ,pcwd_2013,
  nrow        = 1,
  ncol        = 2,
  align       = "v",
  axis        = "l",
  labels      = c(
    "a)", 
    "b)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.01,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
    # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(1, 1),
  expand      = TRUE  
)

combined_fluxes

# ggsave(
#   filename = here::here("figures/US_ton_fluxes.pdf"),
#   plot     = combined_fluxes,
#   device   = cairo_pdf,    # better text rendering
#   width    = 12,           # in inches
#   height   = 5,            # in inches
#   units    = "in",
#   dpi      = 300,          # for raster elements (if any)
#   bg       = "white"       # white background
# )
```


For comparison: Calculate PDSI timeseries for this area
```{r}
## convert to monthly values: 
df_monthly <- df |>
  mutate(month = floor_date(TIMESTAMP, "month")) |>
  group_by(month) |>
  summarise(
    P = sum(P_F, na.rm = TRUE),
    PET = sum(pet, na.rm = TRUE),
    .groups = "drop"
  )

PDSI <- pdsi(df_monthly$P, df_monthly$PET, AWC = 100, sc = TRUE)

spinup <- length(PDSI$X) - nrow(df_monthly)

pdsi_df <- data.frame(
  TIMESTAMP = df_monthly$month,
  PDSI = PDSI$X[(spinup + 1):length(PDSI$X)]
)
```

```{r}
ggplot() +
  geom_line(data  =  out_cwd$df, aes(TIMESTAMP, deficit), color = "#DB5A39") +
  scale_x_date(limits = as.Date(c("2005-01-01", "2014-01-01"))) +
  ylim(0, max(out_cwd$df$deficit)) + 
  labs(
    x = "Date",    y = "Potential cumulative water deficit (mm)"
    )+
   theme_classic()+
    theme(axis.text.x = element_text(size = 14),
    legend.position = "right",
    axis.text.y = element_text(size = 14), 
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12)) 

ggplot() +
  geom_line(data  =  pdsi_df, aes(TIMESTAMP, PDSI), color = "blue") +
  scale_x_date(limits = as.Date(c("2005-01-01", "2014-01-01"))) +
  labs(
    x = "Date",    y = "scPDSI"
    )+
   theme_classic()+
    theme(axis.text.x = element_text(size = 14),
    legend.position = "right",
    axis.text.y = element_text(size = 14), 
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12)) 
```
```{r}
## read in fLUE dataset that marks soil moisture droughts: 

fLUE <- read_csv("~/Downloads/flue_stocker18nphyt.csv") |>
  filter(site == "US-Ton")

drought_periods <- fLUE |>
  filter(is_flue_drought) |>
  arrange(date) |>
  mutate(group = cumsum(c(TRUE, diff(date) != 1))) |>
  group_by(group) |>
  summarise(
    start = min(date),
    end   = max(date),
    .groups = "drop"
  )

fLUE_PCWD <- ggplot() +
  geom_rect(
    data = drought_periods,
    aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
    fill = "gold4",
    alpha = 0.3
  ) +
  geom_line(
    data = out_cwd$df,
    aes(TIMESTAMP, deficit),
    color = "#DB5A39"
  ) +
    scale_x_date(
  limits = as.Date(c("2001-01-01","2014-01-01")),
  breaks = as.Date(c(
    "2001-01-01","2005-01-01","2009-01-01","2013-01-01"
  ))
)+
  ylim(0, max(out_cwd$df$deficit)) +
  labs(
    x = "",
    y = "PCWD (mm)"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(size = 14),
    legend.position = "right",
    axis.text.y = element_text(size = 14),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 12)
  )

out_maxcwd$year_date <- as.Date(paste0(out_maxcwd$year, "-01-01"))
ggplot() +
  geom_rect(
    data = drought_periods,
    aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
    fill = "lightblue",
    alpha = 0.3
  ) +
  geom_line(data  =  out_maxcwd, aes(year_date, max_cwd), color = "#DB5A39") +
  scale_x_date(
  limits = as.Date(c("2001-01-01","2014-01-01")),
  breaks = as.Date(c(
    "2001-01-01","2005-01-01","2009-01-01","2013-01-01"
  ))
)+
  labs(
    x = "Date",    y = "Annmax PCWD (mm)"
    )+
   theme_classic()+
    theme(axis.text.x = element_text(size = 14),
    legend.position = "right",
    axis.text.y = element_text(size = 14), 
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12)) 


fLUE_PDSI <- ggplot() +
    geom_rect(
    data = drought_periods,
    aes(xmin = start, xmax = end, ymin = -Inf, ymax = Inf),
    fill = "gold4",
    alpha = 0.3
  ) +
  geom_line(data  =  pdsi_df, aes(TIMESTAMP, PDSI), color = "blue") +
  scale_x_date(
  limits = as.Date(c("2001-01-01","2014-01-01")),
  breaks = as.Date(c(
    "2001-01-01","2005-01-01","2009-01-01","2013-01-01"
  ))
)+
  labs(
    x = "Date",    y = "scPDSI"
    )+
   theme_classic()+
    theme(axis.text.x = element_text(size = 14),
    legend.position = "right",
    axis.text.y = element_text(size = 14), 
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 14),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size=12)) 
```

```{r}
## calculate the values of PCWD and PDSI based on whether the fLUE drought is true or false: 
# Add year-month to match PDSI
fLUE2 <- fLUE %>%
  mutate(yearmon = floor_date(date, "month"))

pdsi2 <- pdsi_df %>%
  mutate(yearmon = floor_date(TIMESTAMP, "month"))

cwd2 <- out_cwd$df %>%
  rename(date = TIMESTAMP)

# Merge everything
merged_df <- fLUE2 %>%
  left_join(pdsi2 %>% select(yearmon, PDSI), by = "yearmon") %>%
  left_join(cwd2 %>% select(date, deficit), by = "date")

merged_df <- merged_df %>%
  filter(!is.na(PDSI), !is.na(deficit), !is.na(is_flue_drought))

t_pdsi <- t.test(PDSI ~ is_flue_drought, data = merged_df)
t_def  <- t.test(deficit ~ is_flue_drought, data = merged_df)

format_p <- function(p) {
  if (p < 0.00001) {
    "P = 0"
  } else {
    paste0("P = ", signif(p, 2))
  }
}

p_pdsi <- format_p(t_pdsi$p.value)
p_def  <- format_p(t_def$p.value)


box_pdsi <- ggplot(merged_df, aes(x = is_flue_drought, y = PDSI)) +
  geom_boxplot(width = 0.5, fill = "grey80", color = "black", linewidth = 0.4, outlier.shape = NA, staplewidth = 0.5) +
  stat_summary(fun = median, geom = "segment", aes(xend = is_flue_drought, yend = PDSI), linewidth = 0.8) +
  labs(
    x = "fLUE drought",
    y = "PDSI"
  ) +
  annotate("text", x = 1.5, y = max(merged_df$PDSI, na.rm = TRUE), label = p_pdsi, size = 4) +
  theme_classic() +
  theme(axis.text = element_text(size = 10), axis.title = element_text(size = 11), plot.title = element_text(size = 11, face = "bold"))+ 
  scale_x_discrete(labels = c("FALSE", "TRUE")) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  )


box_pcwd <- ggplot(merged_df, aes(x = is_flue_drought, y = deficit)) +
  geom_boxplot(width = 0.5, fill = "grey80", color = "black", linewidth = 0.4, outlier.shape = NA, staplewidth = 0.5) +
  stat_summary(fun = median, geom = "segment", aes(xend = is_flue_drought, yend = deficit), linewidth = 0.8) +
  labs(
    x = "",
    y = "PCWD [mm]"
  ) +
    annotate("text", x = 1.5, y = max(merged_df$deficit, na.rm = TRUE), label = p_def, size = 4) +
  theme_classic() +
  theme(axis.text = element_text(size = 10), axis.title = element_text(size = 11), plot.title = element_text(size = 11, face = "bold"))+ 
  scale_x_discrete(labels = c("FALSE", "TRUE")) +
  theme(
    panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5),
    axis.line = element_line(color = "black"),
    axis.ticks = element_line(color = "black")
  )

```


merge: 
```{r}
## merge flux plot and pcwd plot
library(cowplot)
fLUE_PCWD <- fLUE_PCWD + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 5))
fLUE_PDSI <- fLUE_PDSI + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 5))
box_pdsi <- box_pdsi + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 5))
box_pcwd <- box_pcwd + theme(plot.margin = margin(t = 20, r = 0, b = 0, l = 5))

combined_fLUE <- plot_grid(
  fLUE_PCWD, box_pcwd ,fLUE_PDSI, box_pdsi, 
  nrow        = 2,
  ncol        = 2,
  align       = "vh",
  axis        = "l",
  labels      = c(
    "a) US-Ton Site ", "b)", "c)", "d)"
  ),
  # Push labels all the way to the left margin of each panel
  label_x = 0.01,   # 0 = extreme left edge, a little bit of padding
  hjust   = 0,      # left‑justify the label text
    # Tweak vertical positioning if you like
  label_y = 0.98,   # just below the top
  vjust   = 1,       # align text at its top
  rel_widths = c(2.5, 1),
  expand      = TRUE  
)

combined_fLUE

ggsave(
  filename = here::here("figures/combined_fLUE_US-Ton.pdf"),
  plot     = combined_fLUE,
  device   = cairo_pdf,    # better text rendering
  width    = 9,           # in inches
  height   = 7,            # in inches
  units    = "in",
  dpi      = 300,          # for raster elements (if any)
  bg       = "white"       # white background
)
```

