---
title: "Test_maps_and_stations"
author: "Patricia Helpap"
date: "2024-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r libraries, message=FALSE, echo=FALSE}
#load all necessary libraries
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(extRemes)
library(ggplot2)
library(cwd)
library(ncdf4)
library(readr)
library(reshape2)
library(ggpubr)
library(maps)
library(raster)
library(rnaturalearth)
library(sf)
```

## Global PCWD from one ensemble member calculated using *10mm* absolute threshold


```{r read data from netcdf, echo=FALSE}
##read in data - stored on Ubelix server
input_file <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850_AbsTrsh/PCWD_ANNMAX.nc"
nc_pwcd_1850 <- nc_open(input_file)
pcwd_annmax_1850 = ncvar_get(nc_pwcd_1850, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_1850, varid="lon")
lat = ncvar_get(nc_pwcd_1850, varid="lat")
time = ncvar_get(nc_pwcd_1850, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates <- reference_date + time

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1850)

```

```{r plot pcwd values from 1850, echo=FALSE}
# Extract the first year's data
first_year_data <- pcwd_annmax_1850[, , 1]

# Extract data for the first 30 years (assuming time dimension is the 3rd)
first_30_years_data <- pcwd_annmax_1850[,,1:30]
last_30_years_data <- pcwd_annmax_1850[,,131:160]
overall_avg_data10 <- pcwd_annmax_1850[,,1:160]

# Calculate the mean for each grid cell over the first 30 years
# Calculate the mean along the time dimension (3rd dimension) for each grid cell
mean_data1 <- apply(first_30_years_data, c(2, 1), mean, na.rm = TRUE)
mean_data2 <- apply(last_30_years_data, c(2, 1), mean, na.rm = TRUE)
diff_data <- mean_data2-mean_data1
overall_avg10 <- apply(overall_avg_data10, c(2, 1), mean, na.rm = TRUE)

```

Convert data to raster object for plotting
```{r prep data for plotting 1850, echo=FALSE}
# Convert first year's data into a raster object
r <- raster(t(first_year_data), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')  # Flip the raster to correct orientation

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations
```

```{r customise aestetics for plotting, echo=FALSE}
# Define custom color bins
custom_bins <- c(0, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, 1200, 2000, Inf)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Updated custom color palette with two additional colors
custom_colors <- c("#004f5a",  # New color: a darker turquoise at the beginning
                   "#006d75", "#1a9e96", "#55c1ab", "#85c68b", "#d1c54b", "#e8b249", 
                   "#e3842a", "#d45321", "#ba2e1f", "#9e2d2a", "#803c3d", "#63363f", 
                   "#9c276e",  # New color: a deeper shade in the pink tones
                   "#d238a5")

# Define the numeric labels for the breaks
break_labels <- c(as.character(scales::label_number(accuracy = 1)(custom_bins[-length(custom_bins)])), "Ocean")
```

```{r plot showing 1850, echo=FALSE, fig.show='hide'}
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "PCWD ANNMAX", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "PCWD ANNMAX for First Year", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "PCWD ANNMAX", reverse = TRUE))  # Adjust legend for better presentation
```


```{r prep data for first 30 years, echo=FALSE}
# Convert to raster
r_mean <- raster(mean_data1, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r_mean) <- "+proj=longlat +datum=WGS84"
r_mean <- flip(r_mean, direction='y')

r_masked_mean <- mask(r_mean, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df_mean <- as.data.frame(r_masked_mean, xy = TRUE)
colnames(r_df_mean) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

```

#### First 30 year mean
```{r plot showing 1850-1880, echo=FALSE}
ggplot() +
  geom_raster(data = r_df_mean, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "PCWD ANNMAX", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df_mean$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "Mean PCWD ANNMAX 1850-1880", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "PCWD ANNMAX", reverse = TRUE))  # Adjust legend for better presentation

```

```{r prep data for last 30 years}
# Convert to raster
r_mean <- raster(mean_data2, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r_mean) <- "+proj=longlat +datum=WGS84"
r_mean <- flip(r_mean, direction='y')

r_masked_mean <- mask(r_mean, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df_mean <- as.data.frame(r_masked_mean, xy = TRUE)
colnames(r_df_mean) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

```

#### Last 30 year mean
```{r plot showing 1979-2009}
ggplot() +
  geom_raster(data = r_df_mean, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "PCWD ANNMAX", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df_mean$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "Mean PCWD ANNMAX 1979-2009", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "PCWD ANNMAX", reverse = TRUE))  # Adjust legend for better presentation

```


```{r prep data for difference in periods}
# Convert to raster
r_diff <- raster(diff_data, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r_diff) <- "+proj=longlat +datum=WGS84"
r_diff <- flip(r_diff, direction='y')

r_masked_diff <- mask(r_diff, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df_diff <- as.data.frame(r_masked_diff, xy = TRUE)
colnames(r_df_diff) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

```

#### Difference between last 30 years and first 30 years
```{r plot showing differences}
# Plot using ggplot2
custom_bins <- c(-Inf,-100, -50, -20, -10, -1, 0, 1, 10, 20, 50, 100,Inf)  # Define breaks based on data range
# Create a 10-color scale from blue to white to red
custom_colors <- colorRampPalette(c("blue", "white", "red"))(13)

ggplot() +
  geom_raster(data = r_df_diff, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "PCWD ANNMAX", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df_diff$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "[1979-2009] - [1850-1880]", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "PCWD ANNMAX", reverse = TRUE))  # Adjust legend for better presentation


```

## Global PCWD from one ensemble member calculated using *30mm* absolute threshold

```{r read data from netcdf 2, echo=FALSE}
##read in data
input_file <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850_AbsTrsh_30/PCWD_ANNMAX.nc"
nc_pwcd_1850 <- nc_open(input_file)
pcwd_annmax_1850 = ncvar_get(nc_pwcd_1850, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_1850, varid="lon")
lat = ncvar_get(nc_pwcd_1850, varid="lat")
time = ncvar_get(nc_pwcd_1850, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates <- reference_date + time

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1850)

```

```{r plot pcwd values from 1850 2, echo=FALSE}

# Extract the first year's data
first_year_data <- pcwd_annmax_1850[, , 1]

# Extract data for the first 30 years (assuming time dimension is the 3rd)
first_30_years_data <- pcwd_annmax_1850[,,1:30]
last_30_years_data <- pcwd_annmax_1850[,,131:160]
overall_avg_data30 <- pcwd_annmax_1850[,,1:160]

# Calculate the mean for each grid cell over the first 30 years
# Calculate the mean along the time dimension (3rd dimension) for each grid cell
mean_data1 <- apply(first_30_years_data, c(2, 1), mean, na.rm = TRUE)
mean_data2 <- apply(last_30_years_data, c(2, 1), mean, na.rm = TRUE)
diff_data <- mean_data2 - mean_data1
overall_avg30 <- apply(overall_avg_data30, c(2, 1), mean, na.rm = TRUE)
overall_avg_diff <- overall_avg10 - overall_avg30
```

```{r prep data for plotting 1850 2, echo=FALSE}
# Convert first year's data into a raster object
r <- raster(t(first_year_data), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')  # Flip the raster to correct orientation

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations
```

```{r customise aestetics for plotting 2, echo=FALSE}
# Define custom color bins
custom_bins <- c(0, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, 1200, 2000, Inf)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Updated custom color palette with two additional colors
custom_colors <- c("#004f5a",  # New color: a darker turquoise at the beginning
                   "#006d75", "#1a9e96", "#55c1ab", "#85c68b", "#d1c54b", "#e8b249", 
                   "#e3842a", "#d45321", "#ba2e1f", "#9e2d2a", "#803c3d", "#63363f", 
                   "#9c276e",  # New color: a deeper shade in the pink tones
                   "#d238a5")

# Define the numeric labels for the breaks
break_labels <- c(as.character(scales::label_number(accuracy = 1)(custom_bins[-length(custom_bins)])), "Ocean")
```

```{r prep data for first 30 years 2, echo=FALSE}
# Convert to raster
r_mean <- raster(mean_data1, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r_mean) <- "+proj=longlat +datum=WGS84"
r_mean <- flip(r_mean, direction='y')

r_masked_mean <- mask(r_mean, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df_mean <- as.data.frame(r_masked_mean, xy = TRUE)
colnames(r_df_mean) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

```

#### First 30 year mean
```{r plot showing 1850-1880 2}
ggplot() +
  geom_raster(data = r_df_mean, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "PCWD ANNMAX", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df_mean$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "Mean PCWD ANNMAX 1850-1880", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "PCWD ANNMAX", reverse = TRUE))  # Adjust legend for better presentation

```

```{r prep data for last 30 years 2}
# Convert to raster
r_mean <- raster(mean_data2, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r_mean) <- "+proj=longlat +datum=WGS84"
r_mean <- flip(r_mean, direction='y')

r_masked_mean <- mask(r_mean, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df_mean <- as.data.frame(r_masked_mean, xy = TRUE)
colnames(r_df_mean) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

```

#### Last 30 year mean
```{r plot showing 1979-2009 2}
ggplot() +
  geom_raster(data = r_df_mean, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "PCWD ANNMAX", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df_mean$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "Mean PCWD ANNMAX 1979-2009", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "PCWD ANNMAX", reverse = TRUE))  # Adjust legend for better presentation

```

```{r prep data for difference in periods 2}
# Convert to raster
r_diff <- raster(diff_data, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r_diff) <- "+proj=longlat +datum=WGS84"
r_diff <- flip(r_diff, direction='y')

r_masked_diff <- mask(r_diff, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df_diff <- as.data.frame(r_masked_diff, xy = TRUE)
colnames(r_df_diff) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

```


#### Difference between last 30 years and first 30 years
```{r plot showing differences 2}
# Plot using ggplot2
custom_bins <- c(-Inf,-100, -50, -20, -10, -1, 0, 1, 10, 20, 50,100, Inf)  # Define breaks based on data range
# Create a 10-color scale from blue to white to red
custom_colors <- colorRampPalette(c("blue", "white", "red"))(13)

ggplot() +
  geom_raster(data = r_df_diff, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "PCWD ANNMAX", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df_diff$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(title = "[1979-2009] - [1850-1880]", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "PCWD ANNMAX", reverse = TRUE))  # Adjust legend for better presentation


```


## Comparison of methods
### Showing spatial difference in using *10mm* vs *30mm* on full period mean values
- full period mean *10mm* - full period mean *30mm*:

```{r prep data for difference in means between methods}
# Convert to raster
r_avg_diff <- raster(overall_avg_diff, xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r_avg_diff) <- "+proj=longlat +datum=WGS84"
r_avg_diff <- flip(r_avg_diff, direction='y')

r_masked_avg_diff <- mask(r_avg_diff, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df_avg_diff <- as.data.frame(r_masked_avg_diff, xy = TRUE)
colnames(r_df_avg_diff) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

```

```{r plot showing difference in means}
# Plot using ggplot2
custom_bins <- c(-Inf,-100, -50, -20, -10, -1, 0, 1, 10, 20, 50,100, Inf)  # Define breaks based on data range
custom_colors <- colorRampPalette(c("blue", "white", "red"))(13)

ggplot() +
  geom_raster(data = r_df_avg_diff, aes(x = x, y = y, fill = cut(layer, breaks = custom_bins)), na.rm = TRUE) +
  scale_fill_manual(name = "PCWD ANNMAX", values = custom_colors, na.value = "lightgrey", 
                    labels = c(levels(cut(r_df_avg_diff$layer, breaks = custom_bins)),"Ocean")) +  # Label ocean
  labs(x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.5) +
  coord_sf(xlim = range(lon), ylim = range(lat), expand = FALSE) +
  theme(legend.title = element_text(size = 12),  # Adjust legend title size
        legend.text = element_text(size = 10)) +  # Adjust legend text size
  guides(fill = guide_legend(title = "PCWD ANNMAX", reverse = TRUE))  # Adjust legend for better presentation


```

## Comparison of different methods for selected sites (see Patricia Gribi BSc Thesis)
m001 1850 sets, PCWD calculated
- with relative threshold (1850)
- with absolute threshold of 10mm (AT)
- with absolute threshold of 30mm (AT_30)

```{r read in set data}

##read in data from m001 1850 set
input_file <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850/PCWD_ANNMAX.nc"
nc_pwcd_1850 <- nc_open(input_file)
pcwd_annmax_1850 = ncvar_get(nc_pwcd_1850, varid="pcwd_annmax")
lon = ncvar_get(nc_pwcd_1850, varid="lon")
lat = ncvar_get(nc_pwcd_1850, varid="lat")
time_1850 = ncvar_get(nc_pwcd_1850, varid="time")
# Convert to actual dates (days since 2001-01-01)
reference_date <- as.Date("2001-01-01")
time_dates_1850 <- reference_date + time_1850

# # Print the resulting dates
# print(time_dates)

nc_close(nc_pwcd_1850)
melt_pcwd_annmax_1850 <- melt(pcwd_annmax_1850, varnames = c("lat", "lon", "time"), value.name = "pcwd_annmax")


##read in data from m001 1850 set -- ABSOLUTE THRESHOLD CWD 10mm
input_file <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850_AbsTrsh/PCWD_ANNMAX.nc"
nc_pwcd_1850_AT <- nc_open(input_file)
pcwd_annmax_1850_AT = ncvar_get(nc_pwcd_1850_AT, varid="pcwd_annmax")

nc_close(nc_pwcd_1850_AT)
melt_pcwd_annmax_1850_AT <- melt(pcwd_annmax_1850_AT, varnames = c("lat", "lon", "time"), value.name = "pcwd_annmax")


##read in data from m001 1850 set -- ABSOLUTE THRESHOLD CWD 10mm
input_file <- "/storage/research/giub_geco/data_2/scratch/phelpap/ModESim/m001_tidy/04_result_1850_AbsTrsh_30/PCWD_ANNMAX.nc"
nc_pwcd_1850_AT_30 <- nc_open(input_file)
pcwd_annmax_1850_AT_30 = ncvar_get(nc_pwcd_1850_AT_30, varid="pcwd_annmax")

nc_close(nc_pwcd_1850_AT_30)
melt_pcwd_annmax_1850_AT_30 <- melt(pcwd_annmax_1850_AT_30, varnames = c("lat", "lon", "time"), value.name = "pcwd_annmax")

```

```{r build dataframes for sites, echo=FALSE}
#used stations used in Patricia Gribis Bachelor thesis 
####################Russian site

time_AT <- c(1850:2009)
# Find the index of the closest latitude to 71.2
lat_index <- which.min(abs(as.numeric(lat) - 71.2))
lon_index <- which.min(abs(as.numeric(lon) - 111.25))
Russia_1850 <- pcwd_annmax_1850[lon_index,lat_index,1:160]
Russia_1850_AT <- pcwd_annmax_1850_AT[lon_index,lat_index,1:160]
Russia_1850_AT_30 <- pcwd_annmax_1850_AT_30[lon_index,lat_index,1:160]

# Prepend NAs to the shorter list
Russia_AT_df <- data.frame(time_AT, Russia_1850_AT, Russia_1850_AT_30, Russia_1850)



####################Algeria site
lat_index <- which.min(abs(as.numeric(lat) - 28.7))
lon_index <- which.min(abs(as.numeric(lon) - 3.75))
Algeria_1850<- pcwd_annmax_1850[lon_index,lat_index,1:160]
Algeria_1850_AT<- pcwd_annmax_1850_AT[lon_index,lat_index,1:160]
Algeria_1850_AT_30<- pcwd_annmax_1850_AT_30[lon_index,lat_index,1:160]

Algeria_AT_df <- data.frame(time_AT, Algeria_1850_AT, Algeria_1850_AT_30, Algeria_1850)


####################Antarctica site
lat_index <- which.min(abs(as.numeric(lat) - -81.5))
lon_index <- which.min(abs(as.numeric(lon) - 55))
Antarctica_1850<- pcwd_annmax_1850[lon_index,lat_index,1:160]
Antarctica_1850_AT<- pcwd_annmax_1850_AT[lon_index,lat_index,1:160]
Antarctica_1850_AT_30<- pcwd_annmax_1850_AT_30[lon_index,lat_index,1:160]

# Prepend NAs to the shorter list
Antarctica_AT_df <- data.frame(time_AT, Antarctica_1850_AT, Antarctica_1850_AT_30, Antarctica_1850)


####################Czechia site
lat_index <- which.min(abs(as.numeric(lat) - 50.4))
lon_index <- which.min(abs(as.numeric(lon) - 13.75))
Czechia_1850<- pcwd_annmax_1850[lon_index,lat_index,1:160]
Czechia_1850_AT<- pcwd_annmax_1850_AT[lon_index,lat_index,1:160]
Czechia_1850_AT_30<- pcwd_annmax_1850_AT_30[lon_index,lat_index,1:160]

Czechia_AT_df <- data.frame(time_AT, Czechia_1850_AT, Czechia_1850_AT_30, Czechia_1850)


####################Brazil site
lat_index <- which.min(abs(as.numeric(lat) - -7))
lon_index <- which.min(abs(as.numeric(lon) - -52.5))
Brazil_1850<- pcwd_annmax_1850[lon_index,lat_index,1:160]
Brazil_1850_AT<- pcwd_annmax_1850_AT[lon_index,lat_index,1:160]
Brazil_1850_AT_30<- pcwd_annmax_1850_AT_30[lon_index,lat_index,1:160]

Brazil_AT_df <- data.frame(time_AT, Brazil_1850_AT, Brazil_1850_AT_30, Brazil_1850)

```

```{r showing individual sites RT vs AT m001 for explicit sites}
Russia <- ggplot() +
  geom_line(data = Russia_AT_df, aes(x = time_AT, y = Russia_1850_AT, color = "Russia 1850_AT")) +
  geom_line(data = Russia_AT_df, aes(x = time_AT, y = Russia_1850, color = "Russia 1850")) +
  geom_line(data = Russia_AT_df, aes(x = time_AT, y = Russia_1850_AT_30, color = "Russia 1850_AT_30")) +
  theme_classic() +
  ylim(0, max(Russia_AT_df$Russia_1850)) +
  xlim(min(Russia_AT_df$time), max(Russia_AT_df$time)) +
  labs(
    x = "Year", 
    y = "PCWD(mm)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Russia 1850_AT" = "red", "Russia 1850" = "darkred","Russia 1850_AT_30" = "coral","Algeria 1850_AT" = "blue", "Algeria 1850" = "darkblue", "Antarctica 1850_AT" = "darkslategray4", "Antarctica 1850" = "darkslategray", "Brazil 1850_AT" = "green", "Brazil 1850" = "darkgreen", "Czechia 1850_AT" = "yellow", "Czechia 1850" = "gold"))

Brazil <- ggplot() +
  geom_line(data = Brazil_AT_df, aes(x = time_AT, y = Brazil_1850_AT, color = "Brazil 1850_AT")) +
  geom_line(data = Brazil_AT_df, aes(x = time_AT, y = Brazil_1850, color = "Brazil 1850")) +
    geom_line(data = Brazil_AT_df, aes(x = time_AT, y = Brazil_1850_AT_30, color = "Brazil 1850_AT_30")) +
  theme_classic() +
  ylim(0, max(Brazil_AT_df$Brazil_1850)) +
  xlim(min(Brazil_AT_df$time), max(Brazil_AT_df$time)) +
  labs(
    x = "Year", 
    y = "PCWD (mm)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Russia 1850_AT" = "red", "Russia 1850" = "darkred","Algeria 1850_AT" = "blue", "Algeria 1850" = "darkblue", "Antarctica 1850_AT" = "darkslategray4", "Antarctica 1850" = "darkslategray", "Brazil 1850_AT_30" = "darkkhaki","Brazil 1850_AT" = "green", "Brazil 1850" = "darkgreen", "Czechia 1850_AT" = "yellow", "Czechia 1850" = "gold"))

Algeria <- ggplot() +
  geom_line(data = Algeria_AT_df, aes(x = time_AT, y = Algeria_1850_AT, color = "Algeria 1850_AT")) +
  geom_line(data = Algeria_AT_df, aes(x = time_AT, y = Algeria_1850, color = "Algeria 1850")) +
    geom_line(data = Algeria_AT_df, aes(x = time_AT, y = Algeria_1850_AT_30, color = "Algeria 1850_AT_30")) +
  theme_classic() +
  ylim(0, max(Algeria_AT_df$Algeria_1850_AT)) +
  xlim(min(Algeria_AT_df$time), max(Algeria_AT_df$time)) +
  labs(
    x = "Year", 
    y = "PCWD(mm)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Russia 1850_AT" = "red", "Russia 1850" = "darkred","Algeria 1850_AT" = "deepskyblue","Algeria 1850_AT_30" = "darkslategray", "Algeria 1850" = "deepskyblue4", "Antarctica 1850_AT" = "darkslategray4", "Antarctica 1850" = "darkslategray", "Brazil 1850_AT" = "green", "Brazil 1850" = "darkgreen", "Czechia 1850_AT" = "yellow", "Czechia 1850" = "gold"))

Czechia <- ggplot() +
  geom_line(data = Czechia_AT_df, aes(x = time_AT, y = Czechia_1850_AT, color = "Czechia 1850_AT")) +
  geom_line(data = Czechia_AT_df, aes(x = time_AT, y = Czechia_1850, color = "Czechia 1850")) +
    geom_line(data = Czechia_AT_df, aes(x = time_AT, y = Czechia_1850_AT_30, color = "Czechia 1850_AT_30")) +
  theme_classic() +
  ylim(0, max(Czechia_AT_df$Czechia_1850)) +
  xlim(min(Czechia_AT_df$time), max(Czechia_AT_df$time)) +
  labs(
    x = "Year", 
    y = "PCWD (mm)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Russia 1850_AT" = "red", "Russia 1850" = "darkred","Algeria 1850_AT" = "blue", "Algeria 1850" = "darkblue", "Antarctica 1850_AT" = "darkslategray4", "Antarctica 1850" = "darkslategray", "Brazil 1850_AT" = "green", "Brazil 1850" = "darkgreen", "Czechia 1850_AT" = "yellow", "Czechia 1850" = "gold", "Czechia 1850_AT_30" = "darkgoldenrod4"))

Antarctica <- ggplot() +
  geom_line(data = Antarctica_AT_df, aes(x = time_AT, y = Antarctica_1850_AT, color = "Antarctica 1850_AT")) +
  geom_line(data = Antarctica_AT_df, aes(x = time_AT, y = Antarctica_1850, color = "Antarctica 1850")) +
    geom_line(data = Antarctica_AT_df, aes(x = time_AT, y = Antarctica_1850_AT_30, color = "Antarctica 1850_AT_30")) +
  theme_classic() +
  ylim(0, max(Antarctica_AT_df$Antarctica_1850)) +
  xlim(min(Antarctica_AT_df$time), max(Antarctica_AT_df$time)) +
  labs(
    x = "Year", 
    y = "PCWD (mm)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Russia 1850_AT" = "red", "Russia 1850" = "darkred","Algeria 1850_AT" = "deepskyblue", "Algeria 1850" = "deepskyblue4", "Antarctica 1850_AT" = "darkslategray4", "Antarctica 1850_AT" = "burlywood","Antarctica 1850" = "darkslategray", "Brazil 1850_AT" = "green", "Brazil 1850" = "darkgreen", "Czechia 1850_AT" = "yellow", "Czechia 1850" = "gold"))

ggarrange(Russia,Brazil , Algeria, Czechia, Antarctica,
                    ncol = 1, nrow = 5, common.legend = FALSE)

```

Focus in on Algeria
```{r focus on Algeria}
ggplot() +
    geom_line(data = Algeria_AT_df, aes(x = time_AT, y = Algeria_1850_AT_30, color = "Algeria 1850_AT_30")) +
    geom_line(data = Algeria_AT_df, aes(x = time_AT, y = Algeria_1850_AT, color = "Algeria 1850_AT")) +
  theme_classic() +
  ylim(0, max(Algeria_AT_df$Algeria_1850_AT_30)) +
  xlim(min(Algeria_AT_df$time), max(Algeria_AT_df$time)) +
  labs(
    x = "Year", 
    y = "PCWD(mm)",
    color = "Legend"
  ) +
  scale_color_manual(values = c("Algeria 1850_AT" = "deepskyblue","Algeria 1850_AT_30" = "darkslategray", "Algeria 1850" = "deepskyblue4"))
```

```{r}
#To save html markdown files: 
#rmarkdown::render("vignettes/Spatial_maps_m001.Rmd")

```

