---
title: "Analysis of volcanic forcing on PCWD"
author: "Patricia Helpap"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/Bern/Master thesis/R/lastmill_droughts")
knitr::opts_knit$set(root.dir = "~/Bern/Master thesis/R/lastmill_droughts")


```

```{r libraries, message=FALSE, echo=FALSE}
library(readr)
library(dplyr)
library(here)
library(lubridate)
library(patchwork)
library(ggplot2)
library(cwd)
library(ncdf4)
library(reshape2)
library(ggpubr)
library(maps)
library(raster)
library(rnaturalearth)
library(sf)
library(sp)
library(RColorBrewer)
library(tidyverse)
library(abind)
library(patchwork)
library(gridExtra)
library(ggsci)
```

## Analysis of volcanic forcing impact on hydroclimatic (PCWD, P, PET) variables after large volcanic eruptions

Read in extracted annmax pcwd, annual total precip and total pet data for 1420-1450, 1783, 1784, 1815 and 1816

```{r}
##get dimensions from one ensemble member
nc_file <- "data/volc_impact_aggregates/collected_p_pet/m019/tot_precip_tot_p_pet.nc"
nc_precip <- nc_open(nc_file)
precip <- ncvar_get(nc_precip, varid = "tot_precip")
lat <- ncvar_get(nc_precip, varid = "lat")
lon <- ncvar_get(nc_precip, varid = "lon")
time <- ncvar_get(nc_precip,varid = "time")
nc_close(nc_precip)

nc_file <- "data/volc_impact_aggregates/collected_p_pet/m019/tot_pet_tot_p_pet.nc"
nc_pet <- nc_open(nc_file)
pet <- ncvar_get(nc_pet, varid = "tot_pet")
nc_close(nc_pet)

nc_file <-"data/volc_impact_aggregates/collected_p_pet/m019/max_deficit_tot_p_pet.nc"
nc_pcwd <- nc_open(nc_file)
pcwd <- ncvar_get(nc_pcwd, varid = "max_deficit")
nc_close(nc_pcwd)

reference_date <- as.Date("2001-01-01")
time_dates <- reference_date + time
```


Create composites of ensemble for each variable
PRECIP
```{r}
#### total precipitation: 
# Define the base path
path <- "data/volc_impact_aggregates/collected_p_pet/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)

# Extract ensemble member identifiers from folder names
ensemble_members <- basename(folders)

# Identify the target files
target_files <- file.path(folders, "tot_precip_tot_p_pet.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included

# Filter ensemble members to match target files
ensemble_members <- ensemble_members[file.exists(target_files)]

# Initialize a list to store data for each ensemble member
data_by_ensemble_precip <- list()

# Extract dimensions from the first file to initialize the array size for consistency
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "tot_precip"))[1]
dim_y <- dim(ncvar_get(nc, "tot_precip"))[2]
dim_z <- dim(ncvar_get(nc, "tot_precip"))[3]
nc_close(nc)

# Convert time variable to actual dates (assuming "days since 2001-01-01")
reference_date <- as.Date("2001-01-01")
time_dates <- reference_date + time

# Find indices corresponding to years >= 1968
years <- as.numeric(format(time_dates, "%Y"))

# Loop through all the target files and filter data
for (i in seq_along(target_files)) {
  file <- target_files[i]
  member <- ensemble_members[i]
  
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "tot_precip")  # Adjust "pcwd_annmax" to your variable name

  data_by_ensemble_precip[[member]] <- data
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the data list
print(names(data_by_ensemble_precip))  # Check stored ensemble members
print(dim(data_by_ensemble_precip[[1]]))  # Check dimensions of the data for the first ensemble member



```

PET
```{r}
#### total precipitation: 
# Define the base path
path <- "data/volc_impact_aggregates/collected_p_pet/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)

# Extract ensemble member identifiers from folder names
ensemble_members <- basename(folders)

# Identify the target files
target_files <- file.path(folders, "tot_pet_tot_p_pet.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included

# Filter ensemble members to match target files
ensemble_members <- ensemble_members[file.exists(target_files)]

# Initialize a list to store data for each ensemble member
data_by_ensemble_pet <- list()

# Extract dimensions from the first file to initialize the array size for consistency
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "tot_pet"))[1]
dim_y <- dim(ncvar_get(nc, "tot_pet"))[2]
dim_z <- dim(ncvar_get(nc, "tot_pet"))[3]
nc_close(nc)

# Convert time variable to actual dates (assuming "days since 2001-01-01")
reference_date <- as.Date("2001-01-01")
time_dates <- reference_date + time

# Find indices corresponding to years >= 1968
years <- as.numeric(format(time_dates, "%Y"))

# Loop through all the target files and filter data
for (i in seq_along(target_files)) {
  file <- target_files[i]
  member <- ensemble_members[i]
  
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "tot_pet")  # Adjust "pcwd_annmax" to your variable name

  data_by_ensemble_pet[[member]] <- data
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the data list
print(names(data_by_ensemble_pet))  # Check stored ensemble members
print(dim(data_by_ensemble_pet[[1]]))  # Check dimensions of the data for the first ensemble member

```
PCWD
```{r}
#### total precipitation: 
# Define the base path
path <- "data/volc_impact_aggregates/collected_p_pet/"

# List all the directories within the base path
folders <- list.files(path, full.names = TRUE)

# Extract ensemble member identifiers from folder names
ensemble_members <- basename(folders)

# Identify the target files
target_files <- file.path(folders, "max_deficit_tot_p_pet.nc")
target_files <- target_files[file.exists(target_files)]  # Ensure only existing files are included

# Filter ensemble members to match target files
ensemble_members <- ensemble_members[file.exists(target_files)]

# Initialize a list to store data for each ensemble member
data_by_ensemble_pcwd <- list()

# Extract dimensions from the first file to initialize the array size for consistency
nc <- nc_open(target_files[1])
dim_x <- dim(ncvar_get(nc, "max_deficit"))[1]
dim_y <- dim(ncvar_get(nc, "max_deficit"))[2]
dim_z <- dim(ncvar_get(nc, "max_deficit"))[3]
nc_close(nc)

# Convert time variable to actual dates (assuming "days since 2001-01-01")
reference_date <- as.Date("2001-01-01")
time_dates <- reference_date + time

# Find indices corresponding to years >= 1968
years <- as.numeric(format(time_dates, "%Y"))

# Loop through all the target files and filter data
for (i in seq_along(target_files)) {
  file <- target_files[i]
  member <- ensemble_members[i]
  
  # Open the NetCDF file
  nc <- nc_open(file)
  
  # Extract the data variable
  data <- ncvar_get(nc, "max_deficit")  # Adjust "pcwd_annmax" to your variable name

  data_by_ensemble_pcwd[[member]] <- data
  
  # Close the file
  nc_close(nc)
}

# Output the structure of the data list
print(names(data_by_ensemble_pcwd))  # Check stored ensemble members
print(dim(data_by_ensemble_pcwd[[1]]))  # Check dimensions of the data for the first ensemble member

```
Calculate reference (1420-1450) mean for all variables
```{r}
#### precip
ensemble_array_precip <- simplify2array(data_by_ensemble_precip)  # Shape: [192, 96, 42, 20]
# Compute the ensemble mean across the 3rd dimension (time)
ensemble_ref_precip <- apply(ensemble_array_precip[,,1:30,], c(1, 2, 4), mean, na.rm = TRUE)  # Shape: [192, 96, 20]
dim(ensemble_ref_precip)

#ensemble means for reference, 1783, 1784, 1815, 1816
# Compute the ensemble mean across the 3rd dimension (time)
mean_precip_ref <- apply(ensemble_ref_precip, c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]
dim(mean_precip_ref)
mean_precip_1783 <- apply(ensemble_array_precip[,,31,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]
dim(mean_precip_1783)
mean_precip_1784 <- apply(ensemble_array_precip[,,32,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]

mean_precip_1815 <- apply(ensemble_array_precip[,,33,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]

mean_precip_1816 <- apply(ensemble_array_precip[,,34,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]


#### pet
ensemble_array_pet <- simplify2array(data_by_ensemble_pet)  # Shape: [192, 96, 42, 20]
# Compute the ensemble mean across the 3rd dimension (time)
ensemble_ref_pet <- apply(ensemble_array_pet[,,1:30,], c(1, 2, 4), mean, na.rm = TRUE)  # Shape: [192, 96, 20]
dim(ensemble_ref_pet)

#ensemble means for reference, 1783, 1784, 1815, 1816
# Compute the ensemble mean across the 3rd dimension (time)
mean_pet_ref <- apply(ensemble_ref_pet, c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]
dim(mean_pet_ref)
mean_pet_1783 <- apply(ensemble_array_pet[,,31,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]
dim(mean_pet_1783)
mean_pet_1784 <- apply(ensemble_array_pet[,,32,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]

mean_pet_1815 <- apply(ensemble_array_pet[,,33,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]

mean_pet_1816 <- apply(ensemble_array_pet[,,34,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]


#### annmax pcwd
ensemble_array_pcwd <- simplify2array(data_by_ensemble_pcwd)  # Shape: [192, 96, 42, 20]
# Compute the ensemble mean across the 3rd dimension (time)
ensemble_ref_pcwd <- apply(ensemble_array_pcwd[,,1:30,], c(1, 2, 4), mean, na.rm = TRUE)  # Shape: [192, 96, 20]
dim(ensemble_ref_pcwd)

#ensemble means for reference, 1783, 1784, 1815, 1816
# Compute the ensemble mean across the 3rd dimension (time)
mean_pcwd_ref <- apply(ensemble_ref_pcwd, c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]
dim(mean_pcwd_ref)
mean_pcwd_1783 <- apply(ensemble_array_pcwd[,,31,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]
dim(mean_pcwd_1783)
mean_pcwd_1784 <- apply(ensemble_array_pcwd[,,32,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]

mean_pcwd_1815 <- apply(ensemble_array_pcwd[,,33,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]

mean_pcwd_1816 <- apply(ensemble_array_pcwd[,,34,], c(1, 2), mean, na.rm = TRUE)  # Shape: [192, 96]


```

Maps for each year - ensemble mean

```{r}
################  Reference PCWD ###############
r <- raster(t(mean_pcwd_ref), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

custom_bins <- c(0, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, 1200, 2000, Inf)
custom_colors <- c("#004f5a",  # New color: a darker turquoise at the beginning
                   "#006d75", "#1a9e96", "#55c1ab", "#85c68b", "#d1c54b", "#e8b249",
                   "#e3842a", "#d45321", "#ba2e1f", "#9e2d2a", "#803c3d", "#63363f",
                   "#9c276e",  # New color: a deeper shade in the pink tones
                   "#d238a5")

# 1) build a vector of human–readable labels
#    for the finite intervals we do "a–b", 
#    and then for the last one ">2000"
finite_bins <- custom_bins[!is.infinite(custom_bins)]
lbls_finite <- paste0(finite_bins[-length(finite_bins)], "–", finite_bins[-1])
lbls <- c(lbls_finite, ">2000")

# 2) turn your raster values into a factor with exactly these levels
r_df$pcwd_cat <- cut(
  r_df$layer, 
  breaks = custom_bins, 
  labels = lbls, 
  include.lowest = TRUE, 
  right = FALSE
)

# Define bounding box for Africa
#bbox <- c(xmin = -30, xmax = 60, ymin = -35, ymax = 30)  # Approximate bounding box for Oceania


# Define the ggplot2 object with the gradient scale
ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = pcwd_cat), na.rm = TRUE) +
  scale_fill_manual(name = "mean PCWD (mm)", values = c("Ocean" = "grey50", setNames(custom_colors, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  labs(title = "Reference (1420-1450) PCWD", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PCWD [mm]", reverse = TRUE))  # Adjust legend for better presentation


################ 1783 PCWD ############
r <- raster(t(mean_pcwd_1783), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# 2) turn your raster values into a factor with exactly these levels
r_df$pcwd_cat <- cut(
  r_df$layer, 
  breaks = custom_bins, 
  labels = lbls, 
  include.lowest = TRUE, 
  right = FALSE
)

ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = pcwd_cat), na.rm = TRUE) +
  scale_fill_manual(name = "mean PCWD (mm)", values = c("Ocean" = "grey50", setNames(custom_colors, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  labs(title = "1783 (volc year) PCWD", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PCWD [mm]", reverse = TRUE))  # Adjust legend for better presentation


################ 1784 PCWD ############
r <- raster(t(mean_pcwd_1784), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# 2) turn your raster values into a factor with exactly these levels
r_df$pcwd_cat <- cut(
  r_df$layer, 
  breaks = custom_bins, 
  labels = lbls, 
  include.lowest = TRUE, 
  right = FALSE
)

ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = pcwd_cat), na.rm = TRUE) +
  scale_fill_manual(name = "mean PCWD (mm)", values = c("Ocean" = "grey50", setNames(custom_colors, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  labs(title = "1784 (volc year) PCWD", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PCWD [mm]", reverse = TRUE))  # Adjust legend for better presentation


################ 1815 PCWD ############
r <- raster(t(mean_pcwd_1815), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# 2) turn your raster values into a factor with exactly these levels
r_df$pcwd_cat <- cut(
  r_df$layer, 
  breaks = custom_bins, 
  labels = lbls, 
  include.lowest = TRUE, 
  right = FALSE
)

ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = pcwd_cat), na.rm = TRUE) +
  scale_fill_manual(name = "mean PCWD (mm)", values = c("Ocean" = "grey50", setNames(custom_colors, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  labs(title = "1815 (volc year) PCWD", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PCWD [mm]", reverse = TRUE))  # Adjust legend for better presentation


################ 1816 PCWD ############
r <- raster(t(mean_pcwd_1816), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# 2) turn your raster values into a factor with exactly these levels
r_df$pcwd_cat <- cut(
  r_df$layer, 
  breaks = custom_bins, 
  labels = lbls, 
  include.lowest = TRUE, 
  right = FALSE
)

ggplot() +
  geom_raster(data = r_df, aes(x = x, y = y, fill = pcwd_cat), na.rm = TRUE) +
  scale_fill_manual(name = "mean PCWD (mm)", values = c("Ocean" = "grey50", setNames(custom_colors, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  labs(title = "1816 (volc year) PCWD", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PCWD [mm]", reverse = TRUE))  # Adjust legend for better presentation


```

PCWD percent reduction comparing 1783 to reference for each gridcell

```{r}
# 1) Compute percent reduction
# new: percent change = (1783 - ref) / ref * 100
pcwd_pct <- (mean_pcwd_1783 - mean_pcwd_ref) / mean_pcwd_ref * 100

r <- raster(t(pcwd_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1783)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in PCWD: 1783 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))


###1784: 
# 1) Compute percent reduction
# new: percent change = (1784 - ref) / ref * 100
pcwd_pct <- (mean_pcwd_1784 - mean_pcwd_ref) / mean_pcwd_ref * 100

r <- raster(t(pcwd_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1784)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in PCWD: 1784 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))


# 1) Compute percent reduction
# new: percent change = (1815 - ref) / ref * 100
pcwd_pct <- (mean_pcwd_1815 - mean_pcwd_ref) / mean_pcwd_ref * 100

r <- raster(t(pcwd_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1815)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in PCWD: 1815 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))



# 1) Compute percent reduction
# new: percent change = (1816 - ref) / ref * 100
pcwd_pct <- (mean_pcwd_1816 - mean_pcwd_ref) / mean_pcwd_ref * 100

r <- raster(t(pcwd_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1816)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in PCWD: 1816 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))


```



PET plots
```{r}
################  Reference PCWD ###############
r <- raster(t(mean_pet_ref), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(-40,-20,0, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, 1200, 2000, 3000)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Updated custom color palette with two additional colors
 custom_colors <- c( "#041A40",  "#084594", "#2171B5", "#4292C6", "#6BAED6","#9ECAE1", "#C6DBEF",  "#A1D99B", "#74C476", "#41AB5D", "#FFFFB2", "#FECC5C", "#FD8D3C", "#FC4E2A",  "#E31A1C",  "#BD0026",  "#800026")

# simple interval labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],
  "–",
  custom_bins[-1]
)
# length(lbls) == 16


r_df$pet_cat <- cut(
  r_df$layer,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 3) draw the plot, manually specifying levels, labels and colours
ggplot() +
  geom_raster(
    data = r_df, 
    aes(x = x, y = y, fill = pet_cat), 
    na.rm = TRUE
  ) +
  # note we put "Ocean" *first* in the breaks, so it shows at the top
  scale_fill_manual(
    name      = "total PET (mm/yr)",
    values    = c("Ocean" = "grey50", setNames(custom_colors, lbls)), 
    breaks    = c("Ocean", lbls),
    labels    = c("Ocean", lbls),
    na.value  = "grey50"
  ) +
  labs(title = "Reference (1420-1450) PET", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PET [mm/yr]", reverse = TRUE))  # Adjust legend for better presentation


################  1783 PCWD ###############
r <- raster(t(mean_pet_1783), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(-40,-20,0, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, 1200, 2000, 3000)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Updated custom color palette with two additional colors
 custom_colors <- c( "#041A40",   "#084594",  "#2171B5", "#4292C6",  "#6BAED6",  "#9ECAE1", "#C6DBEF", "#A1D99B", "#74C476","#41AB5D","#FFFFB2", "#FECC5C", "#FD8D3C","#FC4E2A", "#E31A1C",  "#BD0026", "#800026")

r_df$pet_cat <- cut(
  r_df$layer,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 3) draw the plot, manually specifying levels, labels and colours
ggplot() +
  geom_raster(
    data = r_df, 
    aes(x = x, y = y, fill = pet_cat), 
    na.rm = TRUE
  ) +
  # note we put "Ocean" *first* in the breaks, so it shows at the top
  scale_fill_manual(
    name      = "total PET (mm/yr)",
    values    = c("Ocean" = "grey50", setNames(custom_colors, lbls)), 
    breaks    = c("Ocean", lbls),
    labels    = c("Ocean", lbls),
    na.value  = "grey50"
  ) +
  labs(title = "1783 (volcanic year) PET", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PET [mm/yr]", reverse = TRUE))  # Adjust legend for better presentation


################  1784 PCWD ###############
r <- raster(t(mean_pet_1784), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(-40,-20,0, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, 1200, 2000, 3000)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Updated custom color palette with two additional colors
 custom_colors <- c( "#041A40",   "#084594",  "#2171B5", "#4292C6",  "#6BAED6",  "#9ECAE1", "#C6DBEF", "#A1D99B", "#74C476","#41AB5D","#FFFFB2", "#FECC5C", "#FD8D3C","#FC4E2A", "#E31A1C",  "#BD0026", "#800026")

r_df$pet_cat <- cut(
  r_df$layer,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 3) draw the plot, manually specifying levels, labels and colours
ggplot() +
  geom_raster(
    data = r_df, 
    aes(x = x, y = y, fill = pet_cat), 
    na.rm = TRUE
  ) +
  # note we put "Ocean" *first* in the breaks, so it shows at the top
  scale_fill_manual(
    name      = "total PET (mm/yr)",
    values    = c("Ocean" = "grey50", setNames(custom_colors, lbls)), 
    breaks    = c("Ocean", lbls),
    labels    = c("Ocean", lbls),
    na.value  = "grey50"
  ) +
  labs(title = "1784 (volcanic year) PET", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PET [mm/yr]", reverse = TRUE))  # Adjust legend for better presentation

################  1815 PCWD ###############
r <- raster(t(mean_pet_1815), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(-40,-20,0, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, 1200, 2000, 3000)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Updated custom color palette with two additional colors
 custom_colors <- c( "#041A40",   "#084594",  "#2171B5", "#4292C6",  "#6BAED6",  "#9ECAE1", "#C6DBEF", "#A1D99B", "#74C476","#41AB5D","#FFFFB2", "#FECC5C", "#FD8D3C","#FC4E2A", "#E31A1C",  "#BD0026", "#800026")

r_df$pet_cat <- cut(
  r_df$layer,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 3) draw the plot, manually specifying levels, labels and colours
ggplot() +
  geom_raster(
    data = r_df, 
    aes(x = x, y = y, fill = pet_cat), 
    na.rm = TRUE
  ) +
  # note we put "Ocean" *first* in the breaks, so it shows at the top
  scale_fill_manual(
    name      = "total PET (mm/yr)",
    values    = c("Ocean" = "grey50", setNames(custom_colors, lbls)), 
    breaks    = c("Ocean", lbls),
    labels    = c("Ocean", lbls),
    na.value  = "grey50"
  ) +
  labs(title = "1815 (volcanic year) PET", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PET [mm/yr]", reverse = TRUE))  # Adjust legend for better presentation


################  1816 PCWD ###############
r <- raster(t(mean_pet_1816), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(-40,-20,0, 20, 40, 60, 80, 100, 150, 200, 300, 500, 700, 900, 1200, 2000, 3000)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Updated custom color palette with two additional colors
 custom_colors <- c( "#041A40",   "#084594",  "#2171B5", "#4292C6",  "#6BAED6",  "#9ECAE1", "#C6DBEF", "#A1D99B", "#74C476","#41AB5D","#FFFFB2", "#FECC5C", "#FD8D3C","#FC4E2A", "#E31A1C",  "#BD0026", "#800026")

r_df$pet_cat <- cut(
  r_df$layer,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 3) draw the plot, manually specifying levels, labels and colours
ggplot() +
  geom_raster(
    data = r_df, 
    aes(x = x, y = y, fill = pet_cat), 
    na.rm = TRUE
  ) +
  # note we put "Ocean" *first* in the breaks, so it shows at the top
  scale_fill_manual(
    name      = "total PET (mm/yr)",
    values    = c("Ocean" = "grey50", setNames(custom_colors, lbls)), 
    breaks    = c("Ocean", lbls),
    labels    = c("Ocean", lbls),
    na.value  = "grey50"
  ) +
  labs(title = "1816 (volcanic year) PET", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "PET [mm/yr]", reverse = TRUE))  # Adjust legend for better presentation

```


PET percent reduction comparing 1783 to reference for each gridcell

```{r}
# 1) Compute percent reduction
# new: percent change = (1783 - ref) / ref * 100
pet_pct <- (mean_pet_1783 - mean_pet_ref) / mean_pet_ref * 100

r <- raster(t(pet_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1783)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in PET: 1783 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))


###1784: 
# 1) Compute percent reduction
# new: percent change = (1784 - ref) / ref * 100
pet_pct <- (mean_pet_1784 - mean_pet_ref) / mean_pet_ref * 100

r <- raster(t(pet_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1784)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in PET: 1784 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))


# 1) Compute percent reduction
# new: percent change = (1815 - ref) / ref * 100
pet_pct <- (mean_pet_1815 - mean_pet_ref) / mean_pet_ref * 100

r <- raster(t(pet_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1815)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in PET: 1815 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))



# 1) Compute percent reduction
# new: percent change = (1816 - ref) / ref * 100
pet_pct <- (mean_pet_1816 - mean_pet_ref) / mean_pet_ref * 100

r <- raster(t(pet_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1816)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in PET: 1816 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))
```



PRECIP

```{r}
################  Reference Precip ###############
r <- raster(t(mean_precip_ref), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(0, 30, 50, 75, 100, 200, 300, 500, 700, 1000, 2000, 3000, Inf)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Base 9-colour Brewer ramp
base_pal <- brewer.pal(9, "GnBu")

# Interpolate to 15 colours
custom_colors <- colorRampPalette(base_pal)(13)
 
# simple interval labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],
  "–",
  custom_bins[-1]
)
# length(lbls) == 16


r_df$pet_cat <- cut(
  r_df$layer,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)



# Define bounding box for Africa
#bbox <- c(xmin = -30, xmax = 60, ymin = -35, ymax = 30)  # Approximate bounding box for Oceania

# 3) draw the plot, manually specifying levels, labels and colours
ggplot() +
  geom_raster(
    data = r_df, 
    aes(x = x, y = y, fill = pet_cat), 
    na.rm = TRUE
  ) +
  # note we put "Ocean" *first* in the breaks, so it shows at the top
  scale_fill_manual(
    name      = "total Precip (mm/yr)",
    values    = c("Ocean" = "grey50", setNames(custom_colors, lbls)), 
    breaks    = c("Ocean", lbls),
    labels    = c("Ocean", lbls),
    na.value  = "grey50"
  ) +
  labs(title = "Reference (1420-1450) Precipitation", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "P [mm/yr]", reverse = TRUE))  # Adjust legend for better presentation


################  1783 Precip ###############
r <- raster(t(mean_precip_1783), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(0, 30, 50, 75, 100, 200, 300, 500, 700, 1000, 2000, 3000, Inf)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Base 9-colour Brewer ramp
base_pal <- brewer.pal(9, "GnBu")

# Interpolate to 15 colours
custom_colors <- colorRampPalette(base_pal)(13)
 
# simple interval labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],
  "–",
  custom_bins[-1]
)
# length(lbls) == 16


r_df$pet_cat <- cut(
  r_df$layer,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)



# Define bounding box for Africa
#bbox <- c(xmin = -30, xmax = 60, ymin = -35, ymax = 30)  # Approximate bounding box for Oceania

# 3) draw the plot, manually specifying levels, labels and colours
ggplot() +
  geom_raster(
    data = r_df, 
    aes(x = x, y = y, fill = pet_cat), 
    na.rm = TRUE
  ) +
  # note we put "Ocean" *first* in the breaks, so it shows at the top
  scale_fill_manual(
    name      = "total Precip (mm/yr)",
    values    = c("Ocean" = "grey50", setNames(custom_colors, lbls)), 
    breaks    = c("Ocean", lbls),
    labels    = c("Ocean", lbls),
    na.value  = "grey50"
  ) +
  labs(title = "1783 (volcanic year) Precipitation", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "P [mm/yr]", reverse = TRUE))  # Adjust legend for better presentation


################  1784 Precip ###############
r <- raster(t(mean_precip_1784), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(0, 30, 50, 75, 100, 200, 300, 500, 700, 1000, 2000, 3000, Inf)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Base 9-colour Brewer ramp
base_pal <- brewer.pal(9, "GnBu")

# Interpolate to 15 colours
custom_colors <- colorRampPalette(base_pal)(13)
 
# simple interval labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],
  "–",
  custom_bins[-1]
)
# length(lbls) == 16


r_df$pet_cat <- cut(
  r_df$layer,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)



# Define bounding box for Africa
#bbox <- c(xmin = -30, xmax = 60, ymin = -35, ymax = 30)  # Approximate bounding box for Oceania

# 3) draw the plot, manually specifying levels, labels and colours
ggplot() +
  geom_raster(
    data = r_df, 
    aes(x = x, y = y, fill = pet_cat), 
    na.rm = TRUE
  ) +
  # note we put "Ocean" *first* in the breaks, so it shows at the top
  scale_fill_manual(
    name      = "total Precip (mm/yr)",
    values    = c("Ocean" = "grey50", setNames(custom_colors, lbls)), 
    breaks    = c("Ocean", lbls),
    labels    = c("Ocean", lbls),
    na.value  = "grey50"
  ) +
  labs(title = "1784 (volcanic year) Precipitation", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "P [mm/yr]", reverse = TRUE))  # Adjust legend for better presentation


################  1815 Precip ###############
r <- raster(t(mean_precip_1815), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(0, 30, 50, 75, 100, 200, 300, 500, 700, 1000, 2000, 3000, Inf)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Base 9-colour Brewer ramp
base_pal <- brewer.pal(9, "GnBu")

# Interpolate to 15 colours
custom_colors <- colorRampPalette(base_pal)(13)
 
# simple interval labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],
  "–",
  custom_bins[-1]
)
# length(lbls) == 16


r_df$pet_cat <- cut(
  r_df$layer,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)



# Define bounding box for Africa
#bbox <- c(xmin = -30, xmax = 60, ymin = -35, ymax = 30)  # Approximate bounding box for Oceania

# 3) draw the plot, manually specifying levels, labels and colours
ggplot() +
  geom_raster(
    data = r_df, 
    aes(x = x, y = y, fill = pet_cat), 
    na.rm = TRUE
  ) +
  # note we put "Ocean" *first* in the breaks, so it shows at the top
  scale_fill_manual(
    name      = "total Precip (mm/yr)",
    values    = c("Ocean" = "grey50", setNames(custom_colors, lbls)), 
    breaks    = c("Ocean", lbls),
    labels    = c("Ocean", lbls),
    na.value  = "grey50"
  ) +
  labs(title = "1815 (volcanic year) Precipitation", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "P [mm/yr]", reverse = TRUE))  # Adjust legend for better presentation


################  1816 Precip ###############
r <- raster(t(mean_precip_1816), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "layer")  # Rename columns to match ggplot expectations

# Define custom color bins
custom_bins <- c(0, 30, 50, 75, 100, 200, 300, 500, 700, 1000, 2000, 3000, Inf)  # Define breaks based on data range
# Define a custom 13-color palette from deep turquoise to bright pink with earthy tones in between
# Base 9-colour Brewer ramp
base_pal <- brewer.pal(9, "GnBu")

# Interpolate to 15 colours
custom_colors <- colorRampPalette(base_pal)(13)
 
# simple interval labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],
  "–",
  custom_bins[-1]
)
# length(lbls) == 16


r_df$pet_cat <- cut(
  r_df$layer,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)



# Define bounding box for Africa
#bbox <- c(xmin = -30, xmax = 60, ymin = -35, ymax = 30)  # Approximate bounding box for Oceania

# 3) draw the plot, manually specifying levels, labels and colours
ggplot() +
  geom_raster(
    data = r_df, 
    aes(x = x, y = y, fill = pet_cat), 
    na.rm = TRUE
  ) +
  # note we put "Ocean" *first* in the breaks, so it shows at the top
  scale_fill_manual(
    name      = "total Precip (mm/yr)",
    values    = c("Ocean" = "grey50", setNames(custom_colors, lbls)), 
    breaks    = c("Ocean", lbls),
    labels    = c("Ocean", lbls),
    na.value  = "grey50"
  ) +
  labs(title = "1816 (volcanic year) Precipitation", x = "Longitude", y = "Latitude") +
  theme_classic() +
  geom_sf(data = land, fill = NA, color = "black", lwd = 0.3) +
  coord_sf(xlim =	c(-180, 178.125), ylim =	c(-60, 90), expand = FALSE) + #leave out Antarctica
  theme(legend.position = "right", legend.title = element_text(size=15), legend.text = element_text(size=15), 
        axis.text = element_text(size=14), axis.title = element_text(size=14)) +
  guides(fill = guide_legend(title = "P [mm/yr]", reverse = TRUE))  # Adjust legend for better presentation

```
precip percent reduction comparing 1783 to reference for each gridcell

```{r}
# 1) Compute percent reduction
# new: percent change = (1783 - ref) / ref * 100
precip_pct <- (mean_precip_1783 - mean_precip_ref) / mean_precip_ref * 100

r <- raster(t(precip_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1783)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in Precipitation: 1783 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% Δ P", reverse = TRUE))


###1784: 
# 1) Compute percent reduction
# new: percent change = (1784 - ref) / ref * 100
precip_pct <- (mean_precip_1784 - mean_precip_ref) / mean_precip_ref * 100

r <- raster(t(precip_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1784)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in Precipitation: 1784 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))


# 1) Compute percent reduction
# new: percent change = (1815 - ref) / ref * 100
precip_pct <- (mean_precip_1815 - mean_precip_ref) / mean_precip_ref * 100

r <- raster(t(precip_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1815)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in Precipitation: 1815 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))



# 1) Compute percent reduction
# new: percent change = (1816 - ref) / ref * 100
precip_pct <- (mean_precip_1816 - mean_precip_ref) / mean_precip_ref * 100

r <- raster(t(precip_pct), xmn=min(lon), xmx=max(lon), ymn=min(lat), ymx=max(lat))
crs(r) <- "+proj=longlat +datum=WGS84"  # Set CRS for raster
r <- flip(r, direction='y')

# Retrieve land polygons and set the same CRS as the raster
land <- ne_countries(scale = "medium", returnclass = "sf")
land <- st_transform(land, crs = st_crs(r))  # Ensure CRS alignment

# Mask the raster with land polygons to remove ocean values
r_masked <- mask(r, as(land, "Spatial"))

# Convert the masked raster to a data frame for ggplot2
r_df <- as.data.frame(r_masked, xy = TRUE)
colnames(r_df) <- c("x", "y", "reduction")  # Rename columns to match ggplot expectations

# 5) set up bins & palette for % change including -10 and 10
custom_bins <- c(-100, -80, -60, -40, -20, -10, 0, 10, 20, 40, 60, 80, 100)

# we now have length(custom_bins)-1 = 12 intervals
n_intervals <- length(custom_bins) - 1    # 12

# build a smooth 12‐step RdBu diverging palette (blue for drops, red for increases)
library(RColorBrewer)
base_pal <- brewer.pal(11, "RdBu")       # RdBu has 11 base colours
custom_colors <- colorRampPalette(base_pal)(n_intervals)
custom_colors_rev <- rev(custom_colors)  # so negatives (drops) are blue

# 6) rebuild your labels
lbls <- paste0(
  custom_bins[-length(custom_bins)],   # from
  "–",
  custom_bins[-1]                      # to
)

r_df$cat <- cut(
  r_df$reduction,
  breaks         = custom_bins,
  labels         = lbls,
  include.lowest = TRUE,
  right          = FALSE
)

# 7) plot
library(ggplot2)
ggplot() +
  geom_raster(data = r_df, aes(x=x, y=y, fill=cat), na.rm = TRUE) +
  scale_fill_manual(name = "% change (ref → 1816)", values = c("Ocean" = "grey50", setNames(custom_colors_rev, lbls)),  
                    labels = c("Ocean", lbls), breaks = c("Ocean", lbls)) +  # Label ocean
  geom_sf(data = land, fill = NA, color = "black", size = 0.3) +
  coord_sf(xlim = c(-180,180), ylim = c(-60,90), expand = FALSE) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text  = element_text(size=12),
    axis.title = element_text(size=14)
  ) +
  labs(
    title = "Percent reduction in Precipitation: 1816 vs. Reference",
    x     = "Longitude",
    y     = "Latitude"
  )+
  guides(fill = guide_legend(title = "% change", reverse = TRUE))
```
